### 2016-01-11 papers/clusters.tex
start the paper

% This file is part of the Platypus project.
% Copyright 2016 the authors.

\documentclass[12pt, letterpaper, preprint]{aastex}

\begin{document}

\title{Identifying stellar clusters in chemical abundance space}
\author{DWH, AC, MKN, HWR, others}

\begin{abstract}
Hello World.
\end{abstract}

\end{document}

### 2016-01-11 papers/clusters.tex
now with abstract

@@
 
+\newcommand{\acronym}[1]{{\small{#1}}}
+\newcommand{\project}[1]{\textsl{#1}}
+\newcommand{\apogee}{\acronym{\project{APOGEE}}}
+\newcommand{\thecannon}{\project{The~Cannon}}
+
 \begin{document}
 
-\title{Identifying stellar clusters in chemical abundance space}
-\author{DWH, AC, MKN, HWR, others}
+\title{Identifying stellar clusters in chemical-abundance space}
+\author{DWH, MKN, AC, HWR, others}
 
 \begin{abstract}
-Hello World.
+% Context:
+There is a dusty, old, and un-realized idea (dubbed ``chemical tagging'')
+that every star ought to carry---%
+in its detailed chemical abundances---%
+an informative signature of its molecular-cloud birth location and time.
+This idea is pretty, but has never yielded scientific fruit.
+While astronomers have become more jaded, we have quietly been making
+spectroscopic measurements of chemical abundances far more precise with \thecannon\ 
+(now delivering 0.03-dex-level precision).
+% Aims:
+The chemical-tagging hypothesis is tested by looking at clusters in abundance space
+and confirming that they are clustered in phase space.
+% Methods:
+We identify (by eye) overdensities of stars observed by the \apogee\ spectroscopic survey
+in a 15-dimensional chemical-abundance space delivered by \thecannon,
+and plot their locations in phase-space.
+We use \emph{only} abundance-space information (no positional information) to identify stellar groups.
+We don't know full 6-d phase-space positions;
+we only have precise angular positions and radial velocities, and imprecise distances.
+% Results:
+We find that clusters in abundance space are indeed clusters in position space.
+This confirms the chemical-tagging hypothesis and
+verifies the precision of abundances delivered by \thecannon.
+This is the first-ever project to identify stellar clusters by blind search in abundance space.
 \end{abstract}
 
+Hello World!!
+
 \end{document}

### 2016-01-12 code/fof.py
starting k-means project

@@ -5,7 +5,6 @@ Copyright 2016 David W. Hogg (NYU).
 ## bugs:
 - does not create `dir`
 - needs a concept of a metric in the space (or a rescaling of the data)
-- does nothing
 """
 import numpy as np 
 import matplotlib.pyplot as plt

### 2016-01-12 code/kmeans.py

@@ -1,0 +1,52 @@
+"""
+This project is part of the Platypus project.
+Copyright 2016 David W. Hogg (NYU).
+
+## bugs:
+- does not create `dir`
+- needs a concept of a metric in the space (or a rescaling of the data)
+- does nothing
+"""
+import numpy as np 
+import sklearn.cluster as cl
+import corner
+import pickle as cp
+
+def pickle_to_file(fn, stuff):
+    fd = open(fn, "wb")
+    cp.dump(stuff, fd)
+    print("writing", fn)
+    fd.close()
+
+if __name__ == "__main__":
+    dir = "kmeans_figs"
+    suffix = "png"
+    
+    print("reading data")
+    filein2 = 'data/play_cnalmgnaosvmnni.txt' # ouch, ascii
+    t,g,feh,alpha,c,n,o,na,mg,al,s,v,mn,ni = np.loadtxt(filein2, usecols = (0,1,2,3,4,5,6,7,8,9,10,11,12,13), unpack =1, dtype = float) 
+    data = np.vstack((feh,c,n,o,na,mg,al,s,v,mn,ni) ).T
+    data = data[feh > -1.5] # according to MKN, but I don't believe her
+    N, D = data.shape
+    print(data, data.shape)
+
+    if False:
+        print("plotting corner")
+        figure = corner.corner(data, color="0.5")
+        fn = "{}/data.{}".format(dir, suffix)
+        figure.savefig(fn)
+        print(fn)
+
+    print("running k-means")
+    km = cl.KMeans()
+    labels = km.fit_predict(data)
+    centers = km.cluster_centers_.copy()
+    print(labels)
+    print(centers)
+
+    if False:
+        print("overplotting corner")
+        corner.corner(centers, color="r", fig=figure)
+        fn = "{}/centers.{}".format(dir, suffix)
+        figure.savefig(fn)
+        print(fn)

### 2016-01-12 code/kmeans.py
works, except for plotting

@@ -25,28 +25,32 @@ if __name__ == "__main__":
     print("reading data")
     filein2 = 'data/play_cnalmgnaosvmnni.txt' # ouch, ascii
     t,g,feh,alpha,c,n,o,na,mg,al,s,v,mn,ni = np.loadtxt(filein2, usecols = (0,1,2,3,4,5,6,7,8,9,10,11,12,13), unpack =1, dtype = float) 
-    data = np.vstack((feh,c,n,o,na,mg,al,s,v,mn,ni) ).T
+    data = np.vstack(
+             ( feh,         c,           n,           o,           na,
+               mg,          al,          s,           v,           mn,          ni) ).T
+    labels = ["Fe",        "C",         "N",         "O",         "Na",
+              "Mg",        "Al",        "S",         "V",         "Mn",        "Ni"]
+    ranges = [(-1.5, 0.5), (-0.6, 0.6), (-0.6, 0.6), (-0.3, 0.5), (-1.5, 1.0),
+              (-0.2, 0.6), (-1.5, 0.7), (-0.2, 0.6), (-1.5, 0.8), (-1.5, 1.0), (-1.5, 0.6)]
     data = data[feh > -1.5] # according to MKN, but I don't believe her
     N, D = data.shape
-    print(data, data.shape)
-
-    if False:
-        print("plotting corner")
-        figure = corner.corner(data, color="0.5")
-        fn = "{}/data.{}".format(dir, suffix)
-        figure.savefig(fn)
-        print(fn)
+    print(data.shape)
 
     print("running k-means")
-    km = cl.KMeans()
-    labels = km.fit_predict(data)
+    km = cl.KMeans(n_clusters=16, random_state=42, n_init=1)
+    clusters = km.fit_predict(data)
     centers = km.cluster_centers_.copy()
     print(labels)
     print(centers)
+    print(centers.shape)
 
-    if False:
-        print("overplotting corner")
-        corner.corner(centers, color="r", fig=figure)
+    if True:
+        print("plotting corner")
+        # corner.corner() order matters, apparently
+        figure = corner.corner(data, range=ranges, labels=labels, color="b",
+                               bins=128, plot_datapoints=True, plot_density=True, plot_contours=True)
+        corner.corner(centers, range=ranges, color="r", fig=figure,
+                      plot_datapoints=True, plot_density=False, plot_contours=False)
         fn = "{}/centers.{}".format(dir, suffix)
         figure.savefig(fn)
         print(fn)

### 2016-01-12 code/kmeans.py
lots more work

@@ -30,27 +30,32 @@ if __name__ == "__main__":
                mg,          al,          s,           v,           mn,          ni) ).T
     labels = ["Fe",        "C",         "N",         "O",         "Na",
               "Mg",        "Al",        "S",         "V",         "Mn",        "Ni"]
-    ranges = [(-1.5, 0.5), (-0.6, 0.6), (-0.6, 0.6), (-0.3, 0.5), (-1.5, 1.0),
-              (-0.2, 0.6), (-1.5, 0.7), (-0.2, 0.6), (-1.5, 0.8), (-1.5, 1.0), (-1.5, 0.6)]
+    ranges = [(-1.5, 0.6), (-0.5, 0.6), (-0.7, 0.7), (-0.2, 0.5), (-2.0, 1.0),
+              (-0.2, 0.6), (-1.7, 0.8), (-0.2, 0.7), (-2.0, 0.9), (-2.0, 1.0), (-2.0, 0.7)]
     data = data[feh > -1.5] # according to MKN, but I don't believe her
     N, D = data.shape
     print(data.shape)
 
     print("running k-means")
-    km = cl.KMeans(n_clusters=16, random_state=42, n_init=1)
+    K = 256
+    km = cl.KMeans(n_clusters=K, random_state=42, n_init=32)
     clusters = km.fit_predict(data)
     centers = km.cluster_centers_.copy()
-    print(labels)
-    print(centers)
     print(centers.shape)
 
-    if True:
-        print("plotting corner")
+    for k in range(-1,K):
+        print("plotting corner", k)
         # corner.corner() order matters, apparently
-        figure = corner.corner(data, range=ranges, labels=labels, color="b",
-                               bins=128, plot_datapoints=True, plot_density=True, plot_contours=True)
-        corner.corner(centers, range=ranges, color="r", fig=figure,
-                      plot_datapoints=True, plot_density=False, plot_contours=False)
-        fn = "{}/centers.{}".format(dir, suffix)
-        figure.savefig(fn)
-        print(fn)
+        if k < 0:
+            subdata = data
+            fn = "{}/data.{}".format(dir, suffix)
+        else:
+            subdata = data[clusters == k]
+            fn = "{}/cluster_{:04d}.{}".format(dir, k, suffix)
+        print(subdata.shape)
+        N, D = subdata.shape
+        if N > D:
+            figure = corner.corner(subdata, range=ranges, labels=labels, color="k",
[…]

### 2016-01-12 code/kmeans.py
save to pickle, don't plot

@@ -36,13 +36,20 @@ if __name__ == "__main__":
     N, D = data.shape
     print(data.shape)
 
-    print("running k-means")
-    K = 256
-    km = cl.KMeans(n_clusters=K, random_state=42, n_init=32)
-    clusters = km.fit_predict(data)
-    centers = km.cluster_centers_.copy()
-    print(centers.shape)
+    for K in [128, 256, 512]:
+        print("running k-means")
+        K = 512
+        km = cl.KMeans(n_clusters=K, random_state=42, n_init=32)
+        clusters = km.fit_predict(data)
+        centers = km.cluster_centers_.copy()
+        print(centers.shape)
 
+        print("writing pickle")
+        pfn = "kmeans_{:04d}.pkl".format(K)
+        pickle_to_file(pfn, (data, clusters, centers))
+        print(pfn)
+
+if False:
     for k in range(-1,K):
         print("plotting corner", k)
         # corner.corner() order matters, apparently

### 2016-01-12 code/kmeans.py
write pickles not plots

@@ -36,9 +36,8 @@ if __name__ == "__main__":
     N, D = data.shape
     print(data.shape)
 
-    for K in [128, 256, 512]:
-        print("running k-means")
-        K = 512
+    for K in 2 ** np.arange(3, 10):
+        print("running k-means at", K)
         km = cl.KMeans(n_clusters=K, random_state=42, n_init=32)
         clusters = km.fit_predict(data)
         centers = km.cluster_centers_.copy()

### 2016-01-12 code/kmeans.py
horrible abuse of try / except

@@ -3,14 +3,13 @@ This project is part of the Platypus project.
 Copyright 2016 David W. Hogg (NYU).
 
 ## bugs:
-- does not create `dir`
 - needs a concept of a metric in the space (or a rescaling of the data)
-- does nothing
 """
 import numpy as np 
 import sklearn.cluster as cl
 import corner
 import pickle as cp
+import os
 
 def pickle_to_file(fn, stuff):
     fd = open(fn, "wb")
@@ -18,40 +17,84 @@ def pickle_to_file(fn, stuff):
     print("writing", fn)
     fd.close()
 
+def read_pickle_file(fn):
+    fd = open(fn, "rb")
+    stuff = cp.load(fd)
+    fd.close()
+    return stuff
+
 if __name__ == "__main__":
-    dir = "kmeans_figs"
+    dir = "./kmeans_figs"
+    if not os.path.exists(dir):
+        os.mkdir(dir)
     suffix = "png"
+    data = None
+    metadata = None
     
-    print("reading data")
-    filein2 = 'data/play_cnalmgnaosvmnni.txt' # ouch, ascii
-    t,g,feh,alpha,c,n,o,na,mg,al,s,v,mn,ni = np.loadtxt(filein2, usecols = (0,1,2,3,4,5,6,7,8,9,10,11,12,13), unpack =1, dtype = float) 
-    data = np.vstack(
-             ( feh,         c,           n,           o,           na,
-               mg,          al,          s,           v,           mn,          ni) ).T
-    labels = ["Fe",        "C",         "N",         "O",         "Na",
-              "Mg",        "Al",        "S",         "V",         "Mn",        "Ni"]
-    ranges = [(-1.5, 0.6), (-0.5, 0.6), (-0.7, 0.7), (-0.2, 0.5), (-2.0, 1.0),
-              (-0.2, 0.6), (-1.7, 0.8), (-0.2, 0.7), (-2.0, 0.9), (-2.0, 1.0), (-2.0, 0.7)]
-    data = data[feh > -1.5] # according to MKN, but I don't believe her
-    N, D = data.shape
-    print(data.shape)
-
     for K in 2 ** np.arange(3, 10):
-        print("running k-means at", K)
-        km = cl.KMeans(n_clusters=K, random_state=42, n_init=32)
-        clusters = km.fit_predict(data)
-        centers = km.cluster_centers_.copy()
-        print(centers.shape)
-
-        print("writing pickle")
[…]

### 2016-01-12 papers/clusters.tex
outline

@@
 
-Hello World!!
+\section{Introduction}
+
+DWH: Chemical tagging yada yada (CITE).
+
+DWH: Open clusters look one-metal (CITE); globulars look peculiar (CITE); but no-one has ever gone the other way!
+
+DWH: Fast intro to \apogee.
+
+DWH: Fast intro to \thecannon.
+
+DWH: BASH the \apogee\ discontiguous field positioning.
+
+\section{Data and method}
+
+MKN: Details about \apogee\ data.
+
+AC: Details about \thecannon\ and implementation.
+
+DWH: Figure showing the full data set; discuss structure.
+
+DWH: Clustering methodology.
+
+\section{Results and discussion}
+
+DWH: For the first time, we see small-scale informative features in abundance space.
+
+DWH: When we identify the small-scale features in abundance space, they are informative in position space.
+
+DWH: How do we show that we aren't just getting clusters because \apogee\ has plate-to-plate calibration variations?
+
+DWH: Chemical tagging will work.
+
+\acknowledgements
+DWH: It is a pleasure to thank...
+
+DWH: github repo and hash.
+
+DWH: Be sure to cite software a la the new ApJ guidelines.
 

### 2016-01-12 code/kmeans.py
create data and metadata blocks

@@ -5,11 +5,12 @@ Copyright 2016 David W. Hogg (NYU).
 ## bugs:
 - needs a concept of a metric in the space (or a rescaling of the data)
 """
+import os
+import pickle as cp
 import numpy as np 
 import sklearn.cluster as cl
-import corner
-import pickle as cp
-import os
+from astropy.io import fits
+from corner import corner
 
 def pickle_to_file(fn, stuff):
     fd = open(fn, "wb")
@@ -32,7 +33,7 @@ if __name__ == "__main__":
     metadata = None
     
     for K in 2 ** np.arange(3, 10):
-        pfn = "kmeans_{:04d}.pkl".format(K)
+        pfn = "data/kmeans_{:04d}.pkl".format(K)
         try:
             print("attempting to read pickle", pfn)
             data, metadata, clusters, centers = read_pickle_file(pfn)
@@ -49,6 +50,45 @@ if __name__ == "__main__":
 
             if (data is None) or (metadata is None):
                 print("reading data")
+                dfn = "data/results-unregularized-matched.fits.gz"
+                hdulist = fits.open(dfn)
+                hdu = hdulist[1]
+                cols = hdu.columns
+                table = hdu.data
+                okay = ((table.field("TEFF_ASPCAP") > 3500.) *
+                        (table.field("TEFF_ASPCAP") < 5500.) *
+                        (table.field("LOGG_ASPCAP") > 0.) *
+                        (table.field("LOGG_ASPCAP") < 3.5))
+                table = table[okay]
+                metadata_labels = ["RA", "DEC", "TEFF_ASPCAP", "LOGG_ASPCAP"]
+                metadata = np.vstack((table.field(label) for label in metadata_labels)).T
+                okay = np.all(np.isfinite(metadata), axis=1)
+                table = table[okay]
+                metadata = metadata[okay]
+                data_labels = ["AL_H", "CA_H", "C_H", "FE_H", "K_H", "MG_H", "MN_H", "NA_H",
+                               "NI_H", "N_H", "O_H", "SI_H", "S_H", "TI_H", "V_H"]
+                data = np.vstack((table.field(label) for label in data_labels)).T
+                okay = np.all(np.isfinite(data), axis=1)
+                table = table[okay]
+                metadata = metadata[okay]
+                data = data[okay]
+                print(dfn, metadata.shape, data.shape)
+
+                print("plotting metadata")
+                cfn = "{}/metadata.{}".format(dir, suffix)
[…]

### 2016-01-13 code/kmeans.py
ranges, scales, plotting'

@@ -3,7 +3,8 @@ This project is part of the Platypus project.
 Copyright 2016 David W. Hogg (NYU).
 
 ## bugs:
-- needs a concept of a metric in the space (or a rescaling of the data)
+- need to set ranges on corner
+- needs better labels on corner
 """
 import os
 import pickle as cp
@@ -31,12 +32,48 @@ if __name__ == "__main__":
     suffix = "png"
     data = None
     metadata = None
-    
+
+    scale_dict = {"FE_H": 0.0191707168068,
+                  "AL_H": 0.0549037045265, # all from AC
+                  "CA_H": 0.0426365845422,
+                  "C_H": 0.0405909985963,
+                  "K_H": 0.0680897262727,
+                  "MG_H": 0.0324021951804,
+                  "MN_H": 0.0410348634747,
+                  "NA_H": 0.111044350016,
+                  "NI_H": 0.03438986215,
+                  "N_H": 0.0440559383568,
+                  "O_H": 0.037015877736,
+                  "SI_H": 0.0407894206516,
+                  "S_H": 0.0543424861906,
+                  "TI_H": 0.0718311106542,
+                  "V_H": 0.146438163035, }
+
+    range_dict = {"FE_H": (-1.9, 0.6),
+                  "AL_H - FE_H": (-0.6, 0.7),
+                  "CA_H - FE_H": (-0.4, 0.4),
+                  "C_H - FE_H": (-0.6, 0.7),
+                  "K_H - FE_H": (-0.9, 0.5),
+                  "MG_H - FE_H": (-0.25, 0.5),
+                  "MN_H - FE_H": (-0.4, 0.5),
+                  "NA_H - FE_H": (-1.4, 1.2),
+                  "NI_H - FE_H": (-0.4, 0.3),
+                  "N_H - FE_H": (-0.6, 0.9),
+                  "O_H - FE_H": (-0.25, 0.5),
+                  "SI_H - FE_H": (-0.25, 0.6),
+                  "S_H - FE_H": (-0.4, 0.9),
+                  "TI_H - FE_H": (-0.6, 0.5),
+                  "V_H - FE_H": (-0.9, 0.8),
+                  "RA": (0., 360.),
+                  "DEC": (-35., 90.),
+                  "TEFF_ASPCAP": (3500., 5500.),
+                  "LOGG_ASPCAP": (0., 3.9), }
+
     for K in 2 ** np.arange(3, 10):
         pfn = "data/kmeans_{:04d}.pkl".format(K)
[…]

### 2016-01-13 code/kmeans.py
switch to masking paradigm

@@ -3,7 +3,6 @@ This project is part of the Platypus project.
 Copyright 2016 David W. Hogg (NYU).
 
 ## bugs:
-- need to set ranges on corner
 - needs better labels on corner
 """
 import os
@@ -26,6 +25,7 @@ def read_pickle_file(fn):
     return stuff
 
 if __name__ == "__main__":
+    dfn = "./data/results-unregularized-matched.fits.gz"
     dir = "./kmeans_figs"
     if not os.path.exists(dir):
         os.mkdir(dir)
@@ -33,8 +33,8 @@ if __name__ == "__main__":
     data = None
     metadata = None
 
-    scale_dict = {"FE_H": 0.0191707168068,
-                  "AL_H": 0.0549037045265, # all from AC
+    scale_dict = {"FE_H": 0.0191707168068, # all from AC
+                  "AL_H": 0.0549037045265,
                   "CA_H": 0.0426365845422,
                   "C_H": 0.0405909985963,
                   "K_H": 0.0680897262727,
@@ -69,14 +69,35 @@ if __name__ == "__main__":
                   "TEFF_ASPCAP": (3500., 5500.),
                   "LOGG_ASPCAP": (0., 3.9), }
 
+    label_dict = {"FE_H": "[Fe/H] (dex)",
+                  "AL_H - FE_H": "[Al/Fe] (dex)",
+                  "CA_H - FE_H": "[Ca/Fe] (dex)",
+                  "C_H - FE_H":  "[C/Fe] (dex)",
+                  "K_H - FE_H":  "[K/Fe] (dex)",
+                  "MG_H - FE_H": "[Mg/Fe] (dex)",
+                  "MN_H - FE_H": "[Mn/Fe] (dex)",
+                  "NA_H - FE_H": "[Na/Fe] (dex)",
+                  "NI_H - FE_H": "[Ni/Fe] (dex)",
+                  "N_H - FE_H":  "[N/Fe] (dex)",
+                  "O_H - FE_H":  "[O/Fe] (dex)",
+                  "SI_H - FE_H": "[Si/Fe] (dex)",
+                  "S_H - FE_H":  "[S/Fe] (dex)",
+                  "TI_H - FE_H": "[Ti/Fe] (dex)",
+                  "V_H - FE_H":  "[V/Fe] (dex)",
+                  "RA":          "RA (J2000 deg)",
+                  "DEC":         "Dec (J2000 deg)",
[…]

### 2016-01-13 code/kmeans.py
typo

@@ -24,6 +24,36 @@ def read_pickle_file(fn):
     fd.close()
     return stuff
 
+def get_data(fn, labels):
+    print("get_data(): reading table")
+    hdulist = fits.open(fn)
+    hdu = hdulist[1]
+    cols = hdu.columns
+    table = hdu.data
+    mask = ((table.field("TEFF_ASPCAP") > 3500.) *
+            (table.field("TEFF_ASPCAP") < 5500.) *
+            (table.field("LOGG_ASPCAP") > 0.) *
+            (table.field("LOGG_ASPCAP") < 3.9)) # MKN recommendation
+    metadata_labels = ["RA", "DEC", "TEFF_ASPCAP", "LOGG_ASPCAP"]
+    metadata = np.vstack((table.field(label) for label in metadata_labels)).T
+    mask *= np.all(np.isfinite(metadata), axis=1)
+    data = np.vstack((table.field(label) for label in labels)).T
+    mask *= np.all(np.isfinite(data), axis=1)
+    data = data[mask]
+    print("get_data()", dfn, data.shape)
+    return data, mask
+
+def get_metadata(fn, labels, mask):
+    print("get_metadata(): reading table")
+    hdulist = fits.open(fn)
+    hdu = hdulist[1]
+    cols = hdu.columns
+    table = hdu.data
+    metadata = np.vstack((table.field(label) for label in labels)).T
+    metadata = metadata[mask]
+    print("get_metadata()", dfn, metadata.shape)
+    return metadata
+
 if __name__ == "__main__":
     dfn = "./data/results-unregularized-matched.fits.gz"
     dir = "./kmeans_figs"
@@ -31,7 +61,6 @@ if __name__ == "__main__":
         os.mkdir(dir)
     suffix = "png"
     data = None
-    metadata = None
 
     scale_dict = {"FE_H": 0.0191707168068, # all from AC
                   "AL_H": 0.0549037045265,
@@ -89,15 +118,20 @@ if __name__ == "__main__":
                   "TEFF_ASPCAP": "ASPCAP $T_\mathrm{eff}$ (K)",
                   "LOGG_ASPCAP": "ASPCAP log$g$ (dex)", }
 
-    for K in 2 ** np.arange(3, 10):
+    data_labels = ["FE_H",
+                   "AL_H", "CA_H", "C_H", "K_H",  "MG_H", "MN_H", "NA_H",
[…]

### 2016-01-13 papers/clusters.tex
cp in from Platypus
title and abstract improvements

@@
 
-\title{Identifying stellar clusters in chemical-abundance space}
-\author{DWH, MKN, AC, HWR, others}
+\title{Chemical tagging:
+       Identification of stellar structures purely by chemical-abundance similarity}
+\author{DWH, MKN, AC, HWR, DFM, others}
 
@@
 % Context:
-There is a dusty, old, and un-realized idea (dubbed ``chemical tagging'')
-that every star ought to carry---%
+The chemical-tagging hypothesis is that every star ought to carry---%
 in its detailed chemical abundances---%
 an informative signature of its molecular-cloud birth location and time.
-This idea is pretty, but has never yielded scientific fruit.
-While astronomers have become more jaded, we have quietly been making
-spectroscopic measurements of chemical abundances far more precise with \thecannon\ 
-(now delivering 0.03-dex-level precision).
+This idea has not previously yielded scientific successes, probably because of the
+noise in chemical-abundance measurements.
+However, we have succeeded in improving vastly spectroscopic measurements with \thecannon\ 
+(now delivering individual abundances with precisions around 0.04-dex).
 % Aims:
-The chemical-tagging hypothesis is tested by looking at clusters in abundance space
+We test the chemical-tagging hypothesis by looking at clusters in abundance space
 and confirming that they are clustered in phase space.
 % Methods:
-We identify (by eye) overdensities of stars observed by the \apogee\ spectroscopic survey
+We identify (by the k-means algorithm) overdensities of stars observed by the \apogee\ spectroscopic survey
 in a 15-dimensional chemical-abundance space delivered by \thecannon,
-and plot their locations in phase-space.
+and plot the associated stars in phase space.
 We use \emph{only} abundance-space information (no positional information) to identify stellar groups.
-We don't know full 6-d phase-space positions;
-we only have precise angular positions and radial velocities, and imprecise distances.
 % Results:
-We find that clusters in abundance space are indeed clusters in position space.
-This confirms the chemical-tagging hypothesis and
-verifies the precision of abundances delivered by \thecannon.
-This is the first-ever project to identify stellar clusters by blind search in abundance space.
+We find that clusters in abundance space are indeed clusters in phase space.
+We recover known clusters and stellar streams.
+This confirms the chemical-tagging hypothesis and verifies the precision of abundances delivered by \thecannon.
+This is the first-ever project to identify phase-space structures by blind search in abundance space;
+the prospects for future data sets are very good.
 \end{abstract}

### 2016-01-13 papers/clusters.tex
keywords

@@
 
-\title{Chemical tagging:
-       Identification of stellar structures purely by chemical-abundance similarity}
-\author{DWH, MKN, AC, HWR, DFM, others}
+\title{Chemical tagging: \\
+       Identification of stellar phase-space structures \\
+       purely by chemical-abundance similarity}
+\author{David~W.~Hogg\altaffilmark{1,2,3,4},
+        Andrew~R.~Casey\altaffilmark{5},
+        M.~Ness\altaffilmark{4},
+        Hans-Walter~Rix\altaffilmark{4}, \&
+        Daniel~Foreman-Mackey\altaffilmark{6}}
+\altaffiltext{1}{SCDA}
+\altaffiltext{2}{Center for Cosmology and Particle Physics, Department of Phyics,
+  New York University, 4 Washington Pl., room 424, New York, NY, 10003, USA}
+\altaffiltext{3}{Center for Data Science, New York University, 726 Broadway, 7th Floor, New York, NY 10003, USA}
+\altaffiltext{4}{Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, D-69117 Heidelberg, Germany}
+\altaffiltext{5}{IoA}
+\altaffiltext{6}{UW}
+\email{david.hogg@nyu.edu}
 
@@
 \end{abstract}
+
+\keywords{
+  Galaxy: abundances
+  ---
+  Galaxy: stellar content
+  ---
+  Galaxy: structure
+  ---
+  globular clusters: general
+  ---
+  open clusters and associations: general
+  ---
+  stars: abundances
+}
 

### 2016-01-13 code/kmeans.py
overhaul plotting; drop corner

@@ -10,7 +10,7 @@ import pickle as cp
 import numpy as np 
 import sklearn.cluster as cl
 from astropy.io import fits
-from corner import corner
+import pylab as plt
 
 def pickle_to_file(fn, stuff):
     fd = open(fn, "wb")
@@ -54,6 +54,39 @@ def get_metadata(fn, labels, mask):
     print("get_metadata()", dfn, metadata.shape)
     return metadata
 
+def plot_one_cluster(data, labels, mask, name, dir):
+    print("plot_one_cluster(): plotting", name)
+    suffix = "png"
+    for ly, lx, plotname in [
+        ("O_H - FE_H", "NA_H - FE_H", "ONa"),
+        ("C_H - FE_H", "N_H - FE_H", "CN"),
+        ("MG_H - FE_H", "AL_H - FE_H", "MgAl"),
+        ("S_H - FE_H", "AL_H - FE_H", "SAl"),
+        ("DEC", "RA", "sky"),
+        ("DEC", "VHELIO_AVG", "decv"),
+        ("LOGG_ASPCAP", "TEFF_ASPCAP", "HR"),
+        ]:
+        y = np.where(np.array(labels) == ly)[0][0]
+        x = np.where(np.array(labels) == lx)[0][0]
+        plt.figure(figsize=(4,4))
+        plt.subplots_adjust(left=0.2, right=0.96, bottom=0.2, top=0.96)
+        plt.clf()
+        kwargs = {"marker": ".", "ls": "none"}
+        if mask is None:
+            plt.plot(data[:,x], data[:,y], c="k", ms=1.0, alpha=0.20, **kwargs)
+        else:        
+            plt.plot(data[:,x], data[:,y], c="0.75", ms=1.0, alpha=0.20, **kwargs)
+            plt.plot(data[mask,x], data[mask,y], c="k", ms=5.0, alpha=0.5, **kwargs)
+        plt.ylim(range_dict[ly])
+        plt.xlim(range_dict[lx])
+        plt.ylabel(label_dict[ly])
+        plt.xlabel(label_dict[lx])
+        [l.set_rotation(45) for l in plt.gca().get_xticklabels()]
+        [l.set_rotation(45) for l in plt.gca().get_yticklabels()]
+        fn = "{}/{}_{}.{}".format(dir, name, plotname, suffix)
+        plt.savefig(fn)
+        print("plot_one_cluster(): plotting", fn)
+
 if __name__ == "__main__":
[…]

### 2016-01-13 code/kmeans.py
AC transpose

@@ -58,9 +58,9 @@ def plot_one_cluster(data, labels, mask, name, dir):
     print("plot_one_cluster(): plotting", name)
     suffix = "png"
     for ly, lx, plotname in [
-        ("O_H - FE_H", "NA_H - FE_H", "ONa"),
+        ("Na_H - FE_H", "O_H - FE_H", "NaO"),
         ("C_H - FE_H", "N_H - FE_H", "CN"),
-        ("MG_H - FE_H", "AL_H - FE_H", "MgAl"),
+        ("AL_H - FE_H", "MG_H - FE_H", "AlMg"),
         ("S_H - FE_H", "AL_H - FE_H", "SAl"),
         ("DEC", "RA", "sky"),
         ("DEC", "VHELIO_AVG", "decv"),

### 2016-01-13 code/kmeans.py
knobs to 11

@@ -157,7 +157,7 @@ if __name__ == "__main__":
                    "AL_H", "CA_H", "C_H", "K_H",  "MG_H", "MN_H", "NA_H",
                    "NI_H", "N_H",  "O_H", "SI_H", "S_H",  "TI_H", "V_H"]
 
-    for K in 2 ** np.arange(3, 9):
+    for K in 2 ** np.arange(3, 11):
         pfn = "data/kmeans_{:04d}.pkl".format(K)
         try:
             print("attempting to read pickle", pfn)

### 2016-01-13 papers/clusters.tex
outline, text

@@
 
-\section{Data and method}
+DWH: Think of this as a hypothesis test.
+
+\section{Sample and abundances}
 
@@
 
-DWH: Clustering methodology.
+\section{Method and results}
 
-\section{Results and discussion}
+Although only hints of small-scale clustering in the abundance space
+are visible in \figurename~DWH, exploration of the data indicates that
+known clusters do appear in the abundance space as over-densities.
+This encourages us to look for over-densities automatically in the
+abundance space and see if anything found that way would be over-dense
+in phase space.
+The simplest method for clustering points in $D$-dimensional space is
+the \emph{k-means} algorithm (DWH CITE).
+This algorithm is fast and performs well in practice in problems of
+this nature.
+
+DWH: Scaling of the data
+DWH: number of clusters
+DWH: restarts for local minima
+
+DWH: Reminder that we don't need the absolute best clustering to
+demonstrate the hypothesis.
+
+\section{Discussion}
 

### 2016-01-13 papers/clusters.tex
notes
outline

@@
 DWH: restarts for local minima
+DWH: Citation out to numpy and sklearn.
 
@@
 demonstrate the hypothesis.
+
+DWH: Visualize densest clusters.  Some of them are known clusters.
+Show examples.
+
+DWH: Do we have streams or other structures?
+
+DWH: What have we demonstrated?
 
@@
 
+DWH: grants and grant numbers.
+
+DWH: SDSS-III and SDSS-IV.
+
+DWH: NASA ADS
+
 DWH: github repo and hash.

### 2016-01-14 papers/from_mkn_1.tex
mkn text, to be merged

% This file is part of \tc\   project.
% Copyright 2015 the authors.

% style notes
% - \,percent not \%

\documentclass[12pt, preprint]{aastex}
\usepackage{bm, graphicx, subfigure, amsmath, morefloats}

% words
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\thecannon}{\project{The~Cannon}} 
\newcommand{\tc}{\project{The~Cannon}} 
\newcommand{\apogee}{\project{\textsc{apogee}}}
\newcommand{\apokasc}{\project{\textsc{apokasc}}}
\newcommand{\aspcap}{\project{\textsc{aspcap}}}
\newcommand{\corot}{\project{Corot}}
\newcommand{\kepler}{\project{Kepler}}
\newcommand{\gaia}{\project{Gaia}}
\newcommand{\gaiaeso}{\project{Gaia--\textsc{eso}}}
\newcommand{\galah}{\project{\textsc{galah}}}
\newcommand{\most}{\project{\textsc{most}}}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\documentname}{\textsl{Article}}

\newcommand{\teff}{\mbox{$\rm T_{eff}$}}
\newcommand{\kms}{\mbox{$\rm kms^{-1}$}}
\newcommand{\feh}{\mbox{$\rm [Fe/H]$}}
\newcommand{\xfe}{\mbox{$\rm [X/Fe]$}}
\newcommand{\alphafe}{\mbox{$\rm [\alpha/Fe]$}}
\newcommand{\mh}{\mbox{$\rm [M/H]$}}
\newcommand{\logg}{\mbox{$\rm \log g$}}
\newcommand{\noise}{\sigma_{n\lambda}}
\newcommand{\scatter}{s_{\lambda}}
\newcommand{\pix}{\mathrm{pix}}
\newcommand{\rfn}{\mathrm{ref}}
\newcommand{\rgc}{\mbox{$\rm R_{GC}$}}
\newcommand{\rgal}{\mbox{$\rm R_{GAL}$}}
\newcommand{\vgal}{\mbox{$\rm V_{GAL}$}}

% math and symbol macros
\newcommand{\set}[1]{\bm{#1}}
\newcommand{\starlabel}{\ell}
\newcommand{\starlabelvec}{\set{\starlabel}}
\newcommand{\mean}[1]{\overline{#1}}
\newcommand{\given}{\,|\,}

% math
\newcommand{\numax}{$\nu_{\max}$}
\newcommand{\deltanu}{$\Delta\nu$}
\bibliographystyle{apj}

\begin{document}

\title{The Signatures of formation of the Galaxy are in the chemical composition of its stars }

\section{Introduction}\label{sec:Intro}

Freeman and Bland-Hawthorn cite Weinberg (1997) in the opening of their 2002 review on `The New Galaxy: Signatures of Formation', with the assertion that while much progress has been made since 1977, the  theory galaxy formation remains a major outstanding problem. However, we are now in a new domain in stellar astronomy; one where it is possible to test the principles of galactic archeology and chemical tagging, precisely as laid out by Freeman and Bland-Hawthorn. Galactic archeology describes the approach of exploring the formation history of the Galaxy via examining the detailed chemical abundances of its stars. Within this realm lies the prospect of chemical tagging, which describes associating now spatially incoherent stellar groups which have built up the Milky Way, via their identical chemical compositions. These prospects rely on the Milky Way's bulge, disk and halo being comprised of numerous small satellite systems, which have built up our galaxy over time and which have formed from the same molecular gas and so are chemically indistinguishable \citep[see][and references therein]{DaSilva2015}. The systems that have formed the Galaxy are expected to no longer be spatially coherent, having been dispersed throughout the galaxy via processes of mixing and radial migration \citep[e.g.][]{Roskar et al., 2009, Quillen2015}. Therefore, the chemical phase space of the stars is the tool we have to reconstruct the building blocks of our galaxy at this snapshot in time.  By exploring the multi chemical abundance information of Milky Way stars (for many elements,  $>$ 15), the hope is that stellar siblings can be identified and that the dimensionality of this space can serve as fingerprints of the stars.

The prospects to test the principle that the idea that the Milky Way has been built up via multiple small satellite systems that are chemically distinct can now be put to the test with the multitude of data available from high resolution spectroscopic surveys. These surveys are systematically mapping the stellar content of the MIlky Way and its chemical and dynamical phase space. The detailed characterisation of the Milky Way's chemical phase space are the founding principles and goals of the major spectroscopic surveys of the current era, including GALAH \citep{Freeman2013, da Silva 2015}, Gaia-ESO \citep{Gilmore 2012} and APOGEE \citep{Majewsk2015}. These surveys are measuring high resolution (R $=$ 20,000-50,000), high signal to noise spectra (SNR $>$ 100) for hundreds of thousands of stars across a very wide spatial extent of the Milky Way's disk, bulge and halo. With this multitude of large stellar surveys underway, stars are now exceptionally powerful tools that can be used to reconstruct the galaxy's formation and evolution history.

A major limitation is now not the data quantify or quality, but the current precision on abundance measurements \citep[e.g.][]{Ting2015, Martel2015} and the traditional methodologies employed, which do not exploit the full information content nor dimensionality of the data. Achieving higher precision and exploiting the entire information content of the data is critical if we are to reconstruct the chemical components of the Galaxy and for chemical tagging to succeed. Attempts to date to identify the building blocks of the Milky Way via chemical tagging have relied on current abundance precisions and explored different approaches. These have not successfully identified components of the galaxy's assembly that had distinct formation sites (e.g. Ting et al., 2015, Mitschang. et al. 2013, Blanco-Cuaresma2015). However, they have highlighted  the need for higher precision and new approaches for exploiting the data $<$things that are good things that are bad$>$.  Blanco-Cuaresma et al., 2015 demonstrate that there is a large degree of overlap when exploring the chemical space of known open clusters which challenges chemically tagging and separating them, but also note that there is room for improvement if more elements were included in their analysis and models improved. Ting et al., 2015 presented an algorithm to detect groups of chemically homogeneous stars which they reported as challenging and with a low signal to noise in the APOGEE dataset. Mitschang et al., 2013 propose that  chemical tagging may be able to identify coeval groups of stars but these are not necessarily from a distinct origin of formation.

The by-hand study of solar twins, for a handful of stars, has shown however that the chemical information in some elements offers powerful signatures for chemical tagging (Melendez2014, Jofre2015) and PCA approach of Ting et al., 2014 shows there are many independent dimensions of information contained in the multi abundance phase space being delivered by the large spectroscopic surveys (e.g. Ting et al., 2014 report a dimensionality of 7 for GALAH survey). These approaches are important to establish which elements should be targeted in surveys.

We successfully achieve chemical tagging and we do so because we are working with a higher precision that previously and exploit the full information in the data....\\

With \tc\ we achieve both higher precision and exploit the full information content of the data.....cite Casey et al., 2016 Ness et al., 2016 in prep = do you want me to write things about this..\\

We do this with the APOGEE survey which covers a much wider spatial extent but GALAH will be much more powerful as they will report 29 elemental abundances which carry a larger quantity of nucleosynthetic information... \\

What else should I write ? \\

\end{document}

@ARTICLE{DeSilva2015,
   author = {{De Silva}, G.~M. and {Freeman}, K.~C. and {Bland-Hawthorn}, J. and 
	{Martell}, S. and {de Boer}, E.~W. and {Asplund}, M. and {Keller}, S. and 
	{Sharma}, S. and {Zucker}, D.~B. and {Zwitter}, T. and {Anguiano}, B. and 
	{Bacigalupo}, C. and {Bayliss}, D. and {Beavis}, M.~A. and {Bergemann}, M. and 
	{Campbell}, S. and {Cannon}, R. and {Carollo}, D. and {Casagrande}, L. and 
	{Casey}, A.~R. and {Da Costa}, G. and {D'Orazi}, V. and {Dotter}, A. and 
	{Duong}, L. and {Heger}, A. and {Ireland}, M.~J. and {Kafle}, P.~R. and 
	{Kos}, J. and {Lattanzio}, J. and {Lewis}, G.~F. and {Lin}, J. and 
	{Lind}, K. and {Munari}, U. and {Nataf}, D.~M. and {O'Toole}, S. and 
	{Parker}, Q. and {Reid}, W. and {Schlesinger}, K.~J. and {Sheinis}, A. and 
	{Simpson}, J.~D. and {Stello}, D. and {Ting}, Y.-S. and {Traven}, G. and 
	{Watson}, F. and {Wittenmyer}, R. and {Yong}, D. and {{\v Z}erjal}, M.
	},
    title = "{The GALAH survey: scientific motivation}",
  journal = {\mnras},
archivePrefix = "arXiv",
   eprint = {1502.04767},
 keywords = {Galaxy: stellar content},
     year = 2015,
    month = may,
   volume = 449,
    pages = {2604-2617},
      doi = {10.1093/mnras/stv327},
   adsurl = {http://adsabs.harvard.edu/abs/2015MNRAS.449.2604D},
  adsnote = {Provided by the SAO/NASA Astrophysics Data System}
}

@ARTICLE{Freeman2002,
   author = {{Freeman}, K. and {Bland-Hawthorn}, J.},
    title = "{The New Galaxy: Signatures of Its Formation}",
  journal = {\araa},
   eprint = {astro-ph/0208106},
 keywords = {osmology, local group, stellar populations, stellar kinematics},
     year = 2002,
   volume = 40,
    pages = {487-537},
      doi = {10.1146/annurev.astro.40.060401.093840},
   adsurl = {http://adsabs.harvard.edu/abs/2002ARA%26A..40..487F},
  adsnote = {Provided by the SAO/NASA Astrophysics Data System}
}

@ARTICLE{Hawkins2015,
   author = {{Hawkins}, K. and {Jofr{\'e}}, P. and {Masseron}, T. and {Gilmore}, G.
	},
    title = "{Using chemical tagging to redefine the interface of the Galactic disc and halo}",
  journal = {\mnras},
 keywords = {Galaxy: abundances, Galaxy: halo, Galaxy: structure},
     year = 2015,
    month = oct,
   volume = 453,
    pages = {758-774},
      doi = {10.1093/mnras/stv1586},
   adsurl = {http://adsabs.harvard.edu/abs/2015MNRAS.453..758H},
  adsnote = {Provided by the SAO/NASA Astrophysics Data System}
}

@ARTICLE{Ting2015,
   author = {{Ting}, Y.-S. and {Conroy}, C. and {Rix}, H.-W.},
    title = "{APOGEE chemical tagging constraint on the maximum star cluster mass in the $\alpha$-enhanced Galactic disk}",
  journal = {ArXiv e-prints},
archivePrefix = "arXiv",
   eprint = {1507.07563},
 keywords = {Astrophysics - Astrophysics of Galaxies},
     year = 2015,
    month = jul,
   adsurl = {http://adsabs.harvard.edu/abs/2015arXiv150707563T},
  adsnote = {Provided by the SAO/NASA Astrophysics Data System}
}

   author = {BC2015},
    title = "{Testing the chemical tagging technique with open clusters}",
  journal = {\aap},
archivePrefix = "arXiv",
   eprint = {1503.02082},
 primaryClass = "astro-ph.SR",
 keywords = {stars: abundances, techniques: spectroscopic, Galaxy: abundances},
     year = 2015,
    month = may,
   volume = 577,
      eid = {A47},
    pages = {A47},
      doi = {10.1051/0004-6361/201425232},
   adsurl = {http://adsabs.harvard.edu/abs/2015A%26A...577A..47B},
  adsnote = {Provided by the SAO/NASA Astrophysics Data System}
}

   author = {{M2014,
  journal = {\mnras},
archivePrefix = "arXiv",
   eprint = {1312.1759},
 keywords = {techniques: miscellaneous, stars: abundances, Galaxy: disc, Galaxy: evolution, open clusters and associations: general},
     year = 2014,
    month = mar,
   volume = 438,
    pages = {2753-2764},
      doi = {10.1093/mnras/stt2320},
   adsurl = {http://adsabs.harvard.edu/abs/2014MNRAS.438.2753M},
  adsnote = {Provided by the SAO/NASA Astrophysics Data System}
}

@ARTICLE{Melendez2014,
   author = {{Mel{\'e}ndez}, J. and {Ram{\'{\i}}rez}, I. and {Karakas}, A.~I. and 
	{Yong}, D. and {Monroe}, T.~R. and {Bedell}, M. and {Bergemann}, M. and 
	{Asplund}, M. and {Tucci Maia}, M. and {Bean}, J. and {do Nascimento}, Jr., J.-D. and 
	{Bazot}, M. and {Alves-Brito}, A. and {Freitas}, F.~C. and {Castro}, M.
	},
    title = "{18 Sco: A Solar Twin Rich in Refractory and Neutron-capture Elements. Implications for Chemical Tagging}",
  journal = {\apj},
archivePrefix = "arXiv",
   eprint = {1406.5244},
 primaryClass = "astro-ph.SR",
 keywords = {stars: abundances, stars: AGB and post-AGB, stars: fundamental parameters, Sun: abundances},
     year = 2014,
    month = aug,
   volume = 791,
      eid = {14},
    pages = {14},
      doi = {10.1088/0004-637X/791/1/14},
   adsurl = {http://adsabs.harvard.edu/abs/2014ApJ...791...14M},
  adsnote = {Provided by the SAO/NASA Astrophysics Data System}
}

@ARTICLE{Quillen2015,
   author = {{Quillen}, A.~C. and {Anguiano}, B. and {De Silva}, G. and {Freeman}, K. and 
	{Zucker}, D.~B. and {Minchev}, I. and {Bland-Hawthorn}, J.},
    title = "{The parent populations of six groups identified from chemical tagging in the solar neighbourhood}",
  journal = {\mnras},
archivePrefix = "arXiv",
   eprint = {1503.02887},
 keywords = {Galaxy: abundances, Galaxy: evolution, Galaxy: kinematics and dynamics, open clusters and associations: general},
     year = 2015,
    month = jul,
   volume = 450,
    pages = {2354-2366},
      doi = {10.1093/mnras/stv806},
   adsurl = {http://adsabs.harvard.edu/abs/2015MNRAS.450.2354Q},
  adsnote = {Provided by the SAO/NASA Astrophysics Data System}
}

\section*{Acknowledgments}

Don't acknowledge Ken as we would like him to be a ref.

\end{document}

### 2016-01-14 code/kmeans.py
tweaks; notes

@@ -58,7 +58,7 @@ def plot_one_cluster(data, labels, mask, name, dir):
     print("plot_one_cluster(): plotting", name)
     suffix = "png"
     for ly, lx, plotname in [
-        ("Na_H - FE_H", "O_H - FE_H", "NaO"),
+        ("NA_H - FE_H", "O_H - FE_H", "NaO"),
         ("C_H - FE_H", "N_H - FE_H", "CN"),
         ("AL_H - FE_H", "MG_H - FE_H", "AlMg"),
         ("S_H - FE_H", "AL_H - FE_H", "SAl"),
@@ -94,6 +94,7 @@ if __name__ == "__main__":
         os.mkdir(dir)
     suffix = "png"
     data = None
+    plotdata = None
 
     scale_dict = {"FE_H": 0.0191707168068, # all from AC
                   "AL_H": 0.0549037045265,
@@ -157,7 +158,7 @@ if __name__ == "__main__":
                    "AL_H", "CA_H", "C_H", "K_H",  "MG_H", "MN_H", "NA_H",
                    "NI_H", "N_H",  "O_H", "SI_H", "S_H",  "TI_H", "V_H"]
 
-    for K in 2 ** np.arange(3, 11):
+    for K in 2 ** np.arange(8, 10):
         pfn = "data/kmeans_{:04d}.pkl".format(K)
         try:
             print("attempting to read pickle", pfn)
@@ -207,22 +208,23 @@ if __name__ == "__main__":
         print(K, "logdet range:", np.min(logdets), np.median(logdets), np.max(logdets))
         print(K, "density range", np.min(densities), np.median(densities), np.max(densities))
 
-    # make plotting data
-    plotdata = data.copy()
-    plotdata_labels = data_labels.copy()
-    for d in range(1, D):
-        plotdata[:,d] = data[:,d] - data[:,0]
-        plotdata_labels[d] = data_labels[d] + " - " + data_labels[0]
-    metadata_labels = ["RA", "DEC", "VHELIO_AVG", "TEFF_ASPCAP", "LOGG_ASPCAP"]
-    metadata = get_metadata(dfn, metadata_labels, mask)
-    plotdata = np.hstack((plotdata, metadata))
-    plotdata_labels = plotdata_labels + metadata_labels
+        # make plotting data
+        if plotdata is None:
+            plotdata = data.copy()
+            plotdata_labels = data_labels.copy()
+            for d in range(1, D):
+                plotdata[:,d] = data[:,d] - data[:,0]
[…]

### 2016-01-14 code/kmeans.py
output tweaks

@@ -58,34 +58,39 @@ def plot_one_cluster(data, labels, mask, name, dir):
     print("plot_one_cluster(): plotting", name)
     suffix = "png"
     for ly, lx, plotname in [
+        ("MG_H - FE_H", "FE_H", "MgFe"),
         ("NA_H - FE_H", "O_H - FE_H", "NaO"),
         ("C_H - FE_H", "N_H - FE_H", "CN"),
         ("AL_H - FE_H", "MG_H - FE_H", "AlMg"),
         ("S_H - FE_H", "AL_H - FE_H", "SAl"),
+        ("MG_H - FE_H", "K_H - FE_H", "MgK"),
         ("DEC", "RA", "sky"),
         ("DEC", "VHELIO_AVG", "decv"),
         ("LOGG_ASPCAP", "TEFF_ASPCAP", "HR"),
         ]:
-        y = np.where(np.array(labels) == ly)[0][0]
-        x = np.where(np.array(labels) == lx)[0][0]
-        plt.figure(figsize=(4,4))
-        plt.subplots_adjust(left=0.2, right=0.96, bottom=0.2, top=0.96)
-        plt.clf()
-        kwargs = {"marker": ".", "ls": "none"}
-        if mask is None:
-            plt.plot(data[:,x], data[:,y], c="k", ms=1.0, alpha=0.20, **kwargs)
-        else:        
-            plt.plot(data[:,x], data[:,y], c="0.75", ms=1.0, alpha=0.20, **kwargs)
-            plt.plot(data[mask,x], data[mask,y], c="k", ms=5.0, alpha=0.5, **kwargs)
-        plt.ylim(range_dict[ly])
-        plt.xlim(range_dict[lx])
-        plt.ylabel(label_dict[ly])
-        plt.xlabel(label_dict[lx])
-        [l.set_rotation(45) for l in plt.gca().get_xticklabels()]
-        [l.set_rotation(45) for l in plt.gca().get_yticklabels()]
         fn = "{}/{}_{}.{}".format(dir, name, plotname, suffix)
-        plt.savefig(fn)
-        print("plot_one_cluster(): plotting", fn)
+        if os.path.exists(fn):
+            print("plot_one_cluster(): {} exists already".format(fn))
+        else:
+            y = np.where(np.array(labels) == ly)[0][0]
+            x = np.where(np.array(labels) == lx)[0][0]
+            fig = plt.figure(figsize=(4,4))
[…]

### 2016-01-14 papers/clusters.tex
introduction from MKN

@@
 % Copyright 2016 the authors.
+
+% to-do
+% - finish writing MKN's introduction
+% - write data and cannon section
+% - write method and results section
+% - write discussion section
+% - put in references
+% - cite codes!
+% - put in acknowledgement
+% - forward to Freeman, others?
+% - submit to journal and arXiv
 
@@
 \newcommand{\thecannon}{\project{The~Cannon}}
+\newcommand{\foreign}[1]{\textsl{#1}}
+\newcommand{\eg}{\foreign{e.\,g.}}
+\newcommand{\etal}{\foreign{et\,al.}}
 
@@
 
-DWH: Chemical tagging yada yada (CITE).
+\section{Introduction}\label{sec:Intro}
 
-DWH: Open clusters look one-metal (CITE); globulars look peculiar (CITE); but no-one has ever gone the other way!
+Freeman and Bland-Hawthorn cite Weinberg (1997) in the opening of
+their 2002 review on `The New Galaxy: Signatures of Formation', with
+the assertion that while much progress has been made since 1977, the
+theory galaxy formation remains a major outstanding problem.
+However, we are now in a new domain in stellar astronomy; one where it
+is possible to test the principles of galactic archeology and chemical
+tagging, precisely as laid out by Freeman and Bland-Hawthorn.
+Galactic archeology describes the approach of exploring the formation
+history of the Galaxy via examining the detailed chemical abundances
+of its stars.
+Within this realm lies the prospect of chemical tagging, which
+describes associating now spatially incoherent stellar groups which
+have built up the Milky Way, via their identical chemical
+compositions.
+These prospects rely on the Milky Way's bulge, disk and halo being
+comprised of numerous small satellite systems, which have built up our
+galaxy over time and which have formed from the same molecular gas and
+so are chemically indistinguishable \citep[see][and references
+  therein]{DaSilva2015}.
+The systems that have formed the Galaxy are expected to no longer be
+spatially coherent, having been dispersed throughout the galaxy via
+processes of mixing and radial migration \citep[\eg,][]{Roskar2009, Quillen2015}.
+Therefore, the chemical phase space of the stars is the tool we have
+to reconstruct the building blocks of our galaxy at this snapshot in
+time.
+By exploring the multi chemical abundance information of Milky Way
+stars (for many elements, $>15$), the hope is that stellar siblings
+can be identified and that the dimensionality of this space can serve
+as fingerprints of the stars.
+
+The prospects to test the principle that the idea that the Milky Way
+has been built up via multiple small satellite systems that are
+chemically distinct can now be put to the test with the multitude of
+data available from high resolution spectroscopic surveys.
+These surveys are systematically mapping the stellar content of the
+Milky Way and its chemical and dynamical phase space.
+The detailed characterisation of the Milky Way's chemical phase space
+are the founding principles and goals of the major spectroscopic
+surveys of the current era, including GALAH \citep{Freeman2013, da
+  Silva 2015}, Gaia-ESO \citep{Gilmore 2012} and APOGEE
+\citep{Majewsk2015}.
+These surveys are measuring high resolution (R $=$ 20,000-50,000),
+high signal to noise spectra (SNR $>$ 100) for hundreds of thousands
+of stars across a very wide spatial extent of the Milky Way's disk,
+bulge and halo.
+With this multitude of large stellar surveys underway, stars are now
+exceptionally powerful tools that can be used to reconstruct the
+galaxy's formation and evolution history.
+
+A major limitation is now not the data quantify or quality, but the
+current precision on abundance measurements \citep[\eg,][]{Ting2015,
+  Martel2015} and the traditional methodologies employed, which do not
+exploit the full information content nor dimensionality of the
+data.
+Achieving higher precision and exploiting the entire information
+content of the data is critical if we are to reconstruct the chemical
+components of the Galaxy and for chemical tagging to succeed.
+Attempts to date to identify the building blocks of the Milky Way via
+chemical tagging have relied on current abundance precisions and
+explored different approaches.
+These have not successfully identified components of the galaxy's
+assembly that had distinct formation sites (\eg, Ting et al., 2015, Mitschang. et al. 2013,
+Blanco-Cuaresma2015).
+However, they have highlighted the need for higher precision and new
+approaches for exploiting the data $<$things that are good things that
+are bad$>$.
+Blanco-Cuaresma et al., 2015 demonstrate that there is a large degree
+of overlap when exploring the chemical space of known open clusters
+which challenges chemically tagging and separating them, but also note
+that there is room for improvement if more elements were included in
+their analysis and models improved.
+Ting et al., 2015 presented an algorithm to detect groups of
+chemically homogeneous stars which they reported as challenging and
+with a low signal to noise in the APOGEE dataset.
+Mitschang et al., 2013 propose that chemical tagging may be able to
+identify coeval groups of stars but these are not necessarily from a
+distinct origin of formation.
+
+The by-hand study of solar twins, for a handful of stars, has shown
+however that the chemical information in some elements offers powerful
+signatures for chemical tagging (Melendez2014, Jofre2015) and PCA
+approach of Ting et al., 2014 shows there are many independent
+dimensions of information contained in the multi abundance phase space
+being delivered by the large spectroscopic surveys (e.g. Ting et al.,
+2014 report a dimensionality of 7 for GALAH survey).
+These approaches are important to establish which elements should be
+targeted in surveys.
+
+DWH:  We successfully achieve chemical tagging and we do so because we are
+working with a higher precision that previously and exploit the full
+information in the data....
+
+DWH: With \thecannon\ we achieve both higher precision and exploit the full
+information content of the data.....cite Casey et al., 2016 Ness et
+al., 2016 in prep = do you want me to write things about this...
+
+DWH: We do this with the APOGEE survey which covers a much wider
+spatial extent but GALAH will be much more powerful as they will
+report 29 elemental abundances which carry a larger quantity of
+nucleosynthetic information...
+
+DWH: Open clusters look one-metal (CITE); globulars look peculiar
+(CITE); but no-one has ever gone the other way!...
 

### 2016-01-14 code/kmeans.py
lb plot

@@ -65,6 +65,7 @@ def plot_one_cluster(data, labels, mask, name, dir):
         ("S_H - FE_H", "AL_H - FE_H", "SAl"),
         ("MG_H - FE_H", "K_H - FE_H", "MgK"),
         ("DEC", "RA", "sky"),
+        ("GLAT", "GLON", "gal"),
         ("DEC", "VHELIO_AVG", "decv"),
         ("LOGG_ASPCAP", "TEFF_ASPCAP", "HR"),
         ]:
@@ -134,6 +135,8 @@ if __name__ == "__main__":
                   "V_H - FE_H": (-0.9, 0.8),
                   "RA": (360., 0.),
                   "DEC": (-35., 90.),
+                  "GLON": (60., 0.),
+                  "GLAT": (-30., 30.),
                   "VHELIO_AVG": (-275., 275.),
                   "TEFF_ASPCAP": (5500., 3500.),
                   "LOGG_ASPCAP": (3.9, 0.0), }
@@ -155,6 +158,8 @@ if __name__ == "__main__":
                   "V_H - FE_H":  "[V/Fe] (dex)",
                   "RA":          "RA (J2000 deg)",
                   "DEC":         "Dec (J2000 deg)",
+                  "GLON":        "Galactic l (deg)",
+                  "GLAT":        "Galactic b (deg)",
                   "VHELIO_AVG":  "heliocentric RV (km/s)",
                   "TEFF_ASPCAP": "ASPCAP Teff (K)",
                   "LOGG_ASPCAP": "ASPCAP log g (dex)", }
@@ -220,7 +225,7 @@ if __name__ == "__main__":
             for d in range(1, D):
                 plotdata[:,d] = data[:,d] - data[:,0]
                 plotdata_labels[d] = data_labels[d] + " - " + data_labels[0]
-            metadata_labels = ["RA", "DEC", "VHELIO_AVG", "TEFF_ASPCAP", "LOGG_ASPCAP"]
+            metadata_labels = ["RA", "DEC", "GLON", "GLAT", "VHELIO_AVG", "TEFF_ASPCAP", "LOGG_ASPCAP"]
             metadata = get_metadata(dfn, metadata_labels, mask)
             plotdata = np.hstack((plotdata, metadata))
             plotdata_labels = plotdata_labels + metadata_labels

### 2016-01-14 papers/clusters.tex
butchering ness

--- 2016-01-13 papers/clusters.tex
@@
 \newcommand{\project}[1]{\textsl{#1}}
-\newcommand{\apogee}{\acronym{\project{APOGEE}}}
+\newcommand{\apogee}{\project{\acronym{APOGEE}}}
+\newcommand{\gaiaeso}{\project{Gaia--\acronym{ESO}}}
+\newcommand{\galah}{\project{\acronym{GALAH}}}
 \newcommand{\thecannon}{\project{The~Cannon}}
+\newcommand{\foreign}[1]{\textsl{#1}}
+\newcommand{\eg}{\foreign{e.\,g.}}
+\newcommand{\etal}{\foreign{et\,al.}}
 
@@
 We find that clusters in abundance space are indeed clusters in phase space.
-We recover known clusters and stellar streams.
+We recover some known clusters and find other interesting structures.
 This confirms the chemical-tagging hypothesis and verifies the precision of abundances delivered by \thecannon.
@@
 
-\section{Introduction}
+\section{Introduction}\label{sec:intro}
 
-DWH: Chemical tagging yada yada (CITE).
+If every star in the Milky Way was born in a molecular cloud that has
+unique chemical-abundance characteristics (as a function of time), and
+if every star preserves its surface abundances (or significant aspects
+of them) over its lifetime, then stars with common birthplaces and
+time ought to be identifiable in their surface abundances.
+That is, it ought to be possible to use stellar abundances to link up
+co-eval stars---even if at the present day they are no longer close to
+one another in phase space---and infer a detailed star-formation and
+dynamical history for the Galaxy.
+This idea has been dubbed ``chemical tagging''.
+It was proposed long ago \citep{Freeman2002} and is one of the
+principal motivations for a number of surveys, most notably
+\apogee\ \citep{apogee}, \gaiaeso\ \citep{gaiaeso} and \galah\ \citep{galah13, galah15}.
+In order to measure chemical tags and infer the Galaxy's dynamical
+properties and formation history, these surveys are each measuring high resolution,
+high signal-to-noise spectra for hundreds of thousands of stars across
+the Galaxy's disk, bulge and halo.
 
-DWH: Open clusters look one-metal (CITE); globulars look peculiar (CITE); but no-one has ever gone the other way!
+Chemical tagging holds the promise of revealing not just the
+star-formation history of the Galaxy, but also the accretion history
+(as things that fall in are expected to be chemically distinct from
+those that form in the parent body; \citealt{something}) and
+stellar-orbit diffusion process like radial mixing and radial
+migration (for example, \citealt{Roskar2009, Quillen2015}).  After
+stars are born---or after a star cluster is accreted and
+disrupted---associations or groups will disperse, through either
+coherent orbital evolution or else incoherent diffusion, but the
+chemical-abundance tags can in principle make the former members
+identifiable as co-eval stellar siblings.
+
+Although undeniably promising---and although it has motivated the
+launch of expensive, large-scale spectroscopic surveys---the
+chemical-tagging idea has yet to yield scientific fruit.
+Part of the reason is that the level of chemical specificity required
+is very high:
+If there are thousands of (relevant) molecular clouds forming stars in
+the recent history of the Milky Way, it will take many abundances,
+each measured accurately, to make a sufficiently unique fingerprint
+that stellar sibling identification will be reliable.
+
+Stellar spectroscopic surveys now have the resolution, signal-to-noise,
+wavelength coverage, and sample sizes to deliver chemical tags.
+The issue at present is measurement precision (and accuracy).
+The current precision on abundance measurements is not high enough
+(for example, \citealt{Ting2015, Martel2015}).
+That said, it is clear that the \emph{data} are precise enough;
+there is enough signal-to-noise at the relevant locations in spectrum
+space to deliver high-precision tags.
+Our view is that the traditional methods for making abundance
+measurements just aren't mature enough:
+There are issues with the physical models, and the data-analysis
+methods do not make use of all of the information in the data sets.
+Improving the models and exploiting the entire information
+content in the data is critical if we are going to deliver useful
+chemical tags.
+
+One note on accuracy and precision:
+
+In principle, the
+
+These have not successfully identified components of the galaxy's
+assembly that had distinct formation sites (\eg, Ting et al., 2015, Mitschang. et al. 2013,
+Blanco-Cuaresma2015).
+
+Blanco-Cuaresma et al., 2015 demonstrate that there is a large degree
+of overlap when exploring the chemical space of known open clusters
+which challenges chemically tagging and separating them, but also note
+that there is room for improvement if more elements were included in
+their analysis and models improved.
+Ting et al., 2015 presented an algorithm to detect groups of
+chemically homogeneous stars which they reported as challenging and
+with a low signal to noise in the APOGEE dataset.
+Mitschang et al., 2013 propose that chemical tagging may be able to
+identify coeval groups of stars but these are not necessarily from a
+distinct origin of formation.
+
+The by-hand study of solar twins, for a handful of stars, has shown
+however that the chemical information in some elements offers powerful
+signatures for chemical tagging (Melendez2014, Jofre2015) and PCA
+approach of Ting et al., 2014 shows there are many independent
+dimensions of information contained in the multi abundance phase space
+being delivered by the large spectroscopic surveys (e.g. Ting et al.,
+2014 report a dimensionality of 7 for GALAH survey).
+These approaches are important to establish which elements should be
+targeted in surveys.
+
+DWH:  We successfully achieve chemical tagging and we do so because we are
+working with a higher precision that previously and exploit the full
+information in the data....
+
+DWH: With \thecannon\ we achieve both higher precision and exploit the full
+information content of the data.....cite Casey et al., 2016 Ness et
+al., 2016 in prep = do you want me to write things about this...
+
+DWH: We do this with the APOGEE survey which covers a much wider
+spatial extent but GALAH will be much more powerful as they will
+report 29 elemental abundances which carry a larger quantity of
+nucleosynthetic information...
+
+DWH: Open clusters look one-metal (CITE); globulars look peculiar
+(CITE); but no-one has ever gone the other way!...
 

### 2016-01-14 papers/clusters.tex
intro text zeroth draft

@@
 % to-do
-% - finish writing MKN's introduction
 % - write data and cannon section
@@
 
+% text shit
 \newcommand{\acronym}[1]{{\small{#1}}}
@@
 \newcommand{\etal}{\foreign{et\,al.}}
+\newcommand{\documentname}{\textsl{Article}}
+
+% math shit
+\newcommand{\teff}{T_{\mathrm{eff}}}
+\newcommand{\logg}{\log g}
 
@@
 wavelength coverage, and sample sizes to deliver chemical tags.
-The issue at present is measurement precision (and accuracy).
+There are, however, two big issues.
+The first is that the physical assumptions behind the idea may require
+refinement:
+There may be chemical-abundance overlaps among open clusters
+\citep{blancocuaresma}; coeval groups of stars may have similar tags
+but different birth places \citep{mitschang}, and the
+chemical-abundance space might be low in dimensionality \citep{what}.
+On the other hand, precise studies of stellar twins \citep{melendez, jofre}
+and open clusters \citep{bovy} indicates that chemical abundances ought to be
+very precisely similar for co-eval stars.
+
+The second issue for chemical tagging---and the one we address
+here---is measurement precision (and accuracy).
 The current precision on abundance measurements is not high enough
@@
 
-One note on accuracy and precision:
+Our contribution in this space has been to develop
+\thecannon\ \citep{cannon, ages}, which is a data-driven model for
+stellar spectra.
+This model can deliver stellar parameters and chemical abundances to
+stars, making use of every pixel of every stellar spectrum (that is,
+all the information in the data) but making no use at all of physical
+models of stars.
+It relies only on there being training data---\emph{some} stars for
+which parameters and abundances are known and believed.
+In companion papers \citep{casey16, ness16} we show that
+\thecannon\ can deliver 15 to 19 abundances for stars in the
+\apogee\ survey, at precisions higher than even the training data on
+which the model is trained.
+We say more about this below.
+We will show here that \thecannon\ improves chemical abundance measurements
+to the point that \emph{chemical tagging is now possible.}
+We have thereby effectively saved the collective asses of all the
+collaborators in all of the above-referenced large spectroscopic
+surveys.
 
-In principle, the
+One note on \emph{accuracy} and \emph{precision:}
+In principle, the problem of chemical tagging does not require
+absolute accuracy for chemical-abundance measurements.
+We only require that we can precisely see that two stars are similar
+in their abundances; even if we are wrong about the absolute values of
+those abundances.
+This point might make it seem like we don't care that our models are
+wrong, so long as they are \emph{consistent.}
+However, this is a bit misleading:
+For chemical tagging to succeed, we need stars with different
+atmospheric parameters $(\teff, \logg)$ but the same chemical
+abundances to be assigned the same position in chemical-abundance
+space.
+That doesn't require overall accuracy, but it requires that the models
+have the right dependencies on atmospheric parameters such that the
+wrongness in abundance space is consistent across the H-R diagram.
+That is, we need a substantial amount of accuracy to achieve our goals.
 
-These have not successfully identified components of the galaxy's
-assembly that had distinct formation sites (\eg, Ting et al., 2015, Mitschang. et al. 2013,
-Blanco-Cuaresma2015).
+In what follows, we are going to use \apogee\ \acronym{DR12} data, in
+which we can measure 15 element abundances (C, N, O, Na, Mg, Al, Si,
+S, K, Ca, Ti, V, Mn, Fe, Ni).
+The \apogee\ data set covers a huge radial extent and---because the
+data are taken in the infrared---explores the thin disk.
+However, it has the disadvantage that it's spatial coverage is swiss
+cheese, which makes it hard to see, within the data set, linear or
+extended stellar structures.
+In many ways, \galah\ will deliver improvements, both because it will
+have more abundances (possibly 29) and contiguous sky coverage.
 
-Blanco-Cuaresma et al., 2015 demonstrate that there is a large degree
-of overlap when exploring the chemical space of known open clusters
-which challenges chemically tagging and separating them, but also note
-that there is room for improvement if more elements were included in
-their analysis and models improved.
-Ting et al., 2015 presented an algorithm to detect groups of
-chemically homogeneous stars which they reported as challenging and
-with a low signal to noise in the APOGEE dataset.
-Mitschang et al., 2013 propose that chemical tagging may be able to
-identify coeval groups of stars but these are not necessarily from a
-distinct origin of formation.
-
-The by-hand study of solar twins, for a handful of stars, has shown
-however that the chemical information in some elements offers powerful
-signatures for chemical tagging (Melendez2014, Jofre2015) and PCA
-approach of Ting et al., 2014 shows there are many independent
-dimensions of information contained in the multi abundance phase space
-being delivered by the large spectroscopic surveys (e.g. Ting et al.,
-2014 report a dimensionality of 7 for GALAH survey).
-These approaches are important to establish which elements should be
-targeted in surveys.
-
-DWH:  We successfully achieve chemical tagging and we do so because we are
-working with a higher precision that previously and exploit the full
-information in the data....
-
-DWH: With \thecannon\ we achieve both higher precision and exploit the full
-information content of the data.....cite Casey et al., 2016 Ness et
-al., 2016 in prep = do you want me to write things about this...
-
-DWH: We do this with the APOGEE survey which covers a much wider
-spatial extent but GALAH will be much more powerful as they will
-report 29 elemental abundances which carry a larger quantity of
-nucleosynthetic information...
-
-DWH: Open clusters look one-metal (CITE); globulars look peculiar
-(CITE); but no-one has ever gone the other way!...
-
-DWH: Fast intro to \apogee.
-
-DWH: Fast intro to \thecannon.
-
-DWH: BASH the \apogee\ discontiguous field positioning.
-
-DWH: Think of this as a hypothesis test.
+One final note:
+We think of this \documentname\ as performing a hypothesis test.
+We know that the stellar members of open and globular clusters---stars
+that are identified by being close in phase space---contain highly
+informative abundance information that identifies them also in
+chemical-abundance space.
+Does this work the other way around?
+Are groups of stars that are identified as being close in abundance
+space---collections of stars with near-identical chemical tags---also
+found to be close in phase space?
+If the answer to this question is ``yes'' then chemical tagging is
+\emph{on.}
 

### 2016-01-14 papers/clusters.tex
figure

@@
 \newcommand{\logg}{\log g}
+
+% figure shit
+\newcommand{\insanefigure}[1]{%
+\includegraphics[width=0.33\textwidth]{./figs/#1_MgFe.png}%
+\includegraphics[width=0.33\textwidth]{./figs/#1_NaO.png}%
+\includegraphics[width=0.33\textwidth]{./figs/#1_CN.png}\\
+\includegraphics[width=0.33\textwidth]{./figs/#1_AlMg.png}%
+\includegraphics[width=0.33\textwidth]{./figs/#1_SAl.png}%
+\includegraphics[width=0.33\textwidth]{./figs/#1_MgK.png}\\
+\includegraphics[width=0.33\textwidth]{./figs/#1_HR.png}%
+\includegraphics[width=0.33\textwidth]{./figs/#1_sky.png}%
+\includegraphics[width=0.33\textwidth]{./figs/#1_decv.png}}
 
@@
 
+\begin{figure}[!p]
+\insanefigure{all}
+\caption{Whatevsinator.\label{fig:all}}
+\end{figure}
+
+\clearpage{}
 \section{Method and results}
@@
 Although only hints of small-scale clustering in the abundance space
-are visible in \figurename~DWH, exploration of the data indicates that
+are visible in \figurename~\ref{fig:all}, exploration of the data indicates that
 known clusters do appear in the abundance space as over-densities.

### 2016-01-14 papers/clusters.tex
structure tweaks

@@
 % - put in acknowledgement
+% - fix all remaining occurrences of ARC, MKN, HWR, or DWH
 % - forward to Freeman, others?
@@
 \newcommand{\apogee}{\project{\acronym{APOGEE}}}
+\newcommand{\aspcap}{\project{\acronym{ASPCAP}}}
 \newcommand{\gaiaeso}{\project{Gaia--\acronym{ESO}}}
@@
 \includegraphics[width=0.33\textwidth]{./figs/#1_MgK.png}\\
-\includegraphics[width=0.33\textwidth]{./figs/#1_HR.png}%
 \includegraphics[width=0.33\textwidth]{./figs/#1_sky.png}%
-\includegraphics[width=0.33\textwidth]{./figs/#1_decv.png}}
+\includegraphics[width=0.33\textwidth]{./figs/#1_decv.png}%
+\includegraphics[width=0.33\textwidth]{./figs/#1_HR.png}}
 
-\begin{document}
+\begin{document}\sloppy\sloppypar
 
@@
 
+\clearpage
 \section{Introduction}\label{sec:intro}
@@
 
+\clearpage
 \section{Sample and abundances}
 
-MKN: Details about \apogee\ data.
+The details about how we select, reduce, and analyze the \apogee\ data
+are given in full detail in the companion papers \citep{casey16, ness16}.
+We summarize briefly here.
+For the spectroscopic data, we use the \apogee\ \acronym{DR12}
+(\cite{dr12}) spectroscopic data reductions for red-giant stars
+($\logg < 3.9$) that are not flagged as bad in the
+\apogee\ \aspcap\ (\cite{aspcap}) pipeline reductions.
+There are XXXXX stars that make our $\logg$ and flag cuts.
 
-AC: Details about \thecannon\ and implementation.
+The \apogee\ pipelines deliver stacked (co-added),
+pseudo-continuum-normalized spectra, but we do not use these; we
+perform our own continuum normalization on individual-exposure data
+and co-add it ourselves.
+This is required, because the performance of \thecannon\ depends
+critically on the methods used for spectral normalization.  It does
+not require accurate continuum estimation, but it requires that the
+normalization vector be obtained by a \emph{linear operation on the
+  data}.
+The linear operation we use is a smooth-function fit to pixels
+identified as continuum in our previous work \citep{thecannon}.
 
-DWH: Figure showing the full data set; discuss structure.
+For the training data labels (used to train \thecannon; see below), we
+use the \apogee\ \acronym{DR12} \aspcap\ atmospheric parameters and
+15-element abundances, but only for the subset of XXXXX red-giant
+stars that have spectral signal-to-noise ratio measurements greater
+than 200.
+This is a small fraction of the total sample, but these high
+signal-to-noise stars are observed to obtain consistent, low-scatter,
+sensible abundance-space measurements.
+
+DWH!!  How does \thecannon\ work here?
+
+Six two-dimensional projections of the abundance data, plus some other
+data quantities, are shown in \figurename~\ref{fig:all}.
 
@@
 
-\clearpage{}
+\clearpage
 \section{Method and results}
@@
 
+\clearpage
 \section{Discussion}
@@
 
+\clearpage
+\begin{thebibliography}{24}\raggedright
+\bibitem[Ness \etal(2016)]{ness16}
+whatev
+\end{thebibliography}
+
 \end{document}

### 2016-01-15 code/kmeans.py
plotting tweaks

@@ -61,7 +61,7 @@ def plot_one_cluster(data, labels, mask, name, dir):
         ("MG_H - FE_H", "FE_H", "MgFe"),
         ("NA_H - FE_H", "O_H - FE_H", "NaO"),
         ("C_H - FE_H", "N_H - FE_H", "CN"),
-        ("AL_H - FE_H", "MG_H - FE_H", "AlMg"),
+        ("MG_H - FE_H", "AL_H - FE_H", "AlMg"), # wrong name for file!!
         ("S_H - FE_H", "AL_H - FE_H", "SAl"),
         ("MG_H - FE_H", "K_H - FE_H", "MgK"),
         ("DEC", "RA", "sky"),
@@ -80,9 +80,9 @@ def plot_one_cluster(data, labels, mask, name, dir):
             plt.clf()
             kwargs = {"marker": ".", "ls": "none"}
             if mask is None:
-                plt.plot(data[:,x], data[:,y], c="k", ms=1.0, alpha=0.20, **kwargs)
+                plt.plot(data[:,x], data[:,y], c="k", ms=0.5, alpha=0.20, **kwargs)
             else:        
-                plt.plot(data[:,x], data[:,y], c="0.75", ms=1.0, alpha=0.20, **kwargs)
+                plt.plot(data[:,x], data[:,y], c="0.75", ms=0.5, alpha=0.20, **kwargs)
                 plt.plot(data[mask,x], data[mask,y], c="k", ms=5.0, alpha=0.5, **kwargs)
             plt.ylim(range_dict[ly])
             plt.xlim(range_dict[lx])
@@ -238,3 +238,5 @@ if __name__ == "__main__":
             plotcount += 1
             if plotcount == 32:
                 break
+
+    plot_one_cluster(plotdata, plotdata_labels, None, "all", dir)

### 2016-01-15 papers/clusters.tex
figures
figure tweaks
k-means

@@
 \includegraphics[width=0.33\textwidth]{./figs/#1_MgFe.png}%
+\includegraphics[width=0.33\textwidth]{./figs/#1_MgK.png}%
+\includegraphics[width=0.33\textwidth]{./figs/#1_AlMg.png}\\
 \includegraphics[width=0.33\textwidth]{./figs/#1_NaO.png}%
-\includegraphics[width=0.33\textwidth]{./figs/#1_CN.png}\\
-\includegraphics[width=0.33\textwidth]{./figs/#1_AlMg.png}%
-\includegraphics[width=0.33\textwidth]{./figs/#1_SAl.png}%
-\includegraphics[width=0.33\textwidth]{./figs/#1_MgK.png}\\
+\includegraphics[width=0.33\textwidth]{./figs/#1_CN.png}%
+\includegraphics[width=0.33\textwidth]{./figs/#1_SAl.png}\\
 \includegraphics[width=0.33\textwidth]{./figs/#1_sky.png}%
@@
         M.~Ness\altaffilmark{4},
-        Hans-Walter~Rix\altaffilmark{4}, \&
-        Daniel~Foreman-Mackey\altaffilmark{6}}
+        Hans-Walter~Rix\altaffilmark{4},
+    and~Daniel~Foreman-Mackey\altaffilmark{6}}
 \altaffiltext{1}{SCDA}
@@
 This confirms the chemical-tagging hypothesis and verifies the precision of abundances delivered by \thecannon.
-This is the first-ever project to identify phase-space structures by blind search in abundance space;
+This is the first-ever project to identify phase-space structures by blind search purely in abundance space;
 the prospects for future data sets are very good.
@@
 the \emph{k-means} algorithm (DWH CITE).
+Briefly, the k-means algorithm is to find the points closest (in
+$D$-space) to each of $K$ centers, and then take the mean of those
+points to update the locations of the centers; iterate to convergence.
+The output is the locations of the $K$ centers and the assignments of
+all points to centers.
 This algorithm is fast and performs well in practice in problems of
@@
 
-DWH: Scaling of the data
-DWH: number of clusters
-DWH: restarts for local minima
+The k-means algorithm has disadvantages, of course.
+One is that $K$ must be chosen by hand (or heuristically at best).
+Here we are only demonstrating a concept; we don't need to have the
+best possible clustering.
+For this reason we simply choose $K=128, 256, 512$ and look at all the
+results.
+The algorithm also only performs local optimization.
+At each $K$ we perform 32 restarts with different initializations, and
+preserve the best clustering (best according to the k-means score).
+
+Another issue with k-means is that it effectively uses metric
+distances in the $D$-space; it presumes Euclidean isotropy.
+We choose here to work in the hydrogen-normalized abundance space, the
+space of [C/H], [N/H], [O/H], [Na/H], [Mg/H], [Al/H], [Si/H], [S/H],
+[K/H], [Ca/H], [Ti/H], [V/H], [Mn/H], [Fe/H], [Ni/H].
+But in addition to this, we re-scale these by approximate measurement
+precisions obtained by \thecannon\ before running k-means.
+This makes the space close to isotropic in measurment uncertainty or
+observational precision.
+
+When the k-means results are returned, we construct the $D\times D$ empirical variance tensor for the DWH THIS IS WRONG!
+
 DWH: Citation out to numpy and sklearn.
@@
 
+\begin{figure}[!p]
+\insanefigure{cluster_0256_0253}
+\caption{Whatevsinator.\label{fig:infallcluster}}
+\end{figure}
+
+\begin{figure}[!p]
+\insanefigure{cluster_0256_0034}
+\caption{Whatevsinator.\label{fig:Pal5}}
+\end{figure}
+
+\begin{figure}[!p]
+\insanefigure{cluster_0256_0177}
+\caption{Whatevsinator.\label{fig:Sgr}}
+\end{figure}
+
+\begin{figure}[!p]
+\insanefigure{cluster_0256_0010}
+\caption{Whatevsinator.\label{fig:halo}}
+\end{figure}
+
+\begin{figure}[!p]
+\insanefigure{cluster_0256_0141}
+\caption{Whatevsinator.\label{fig:disk}}
+\end{figure}
+
 \clearpage
@@
 
-DWH: For the first time, we see small-scale informative features in abundance space.
+DWH: For the first time/H], [we see small-scale informative features in abundance space.
 
-DWH: When we identify the small-scale features in abundance space, they are informative in position space.
+DWH: When we identify the small-scale features in abundance space,they are informative in position space.
 

### 2016-01-15 code/kmeans.py
tweaks

@@ -65,9 +65,10 @@ def plot_one_cluster(data, labels, mask, name, dir):
         ("S_H - FE_H", "AL_H - FE_H", "SAl"),
         ("MG_H - FE_H", "K_H - FE_H", "MgK"),
         ("DEC", "RA", "sky"),
-        ("GLAT", "GLON", "gal"),
+        # ("GLAT", "GLON", "gal"),
         ("DEC", "VHELIO_AVG", "decv"),
         ("LOGG_ASPCAP", "TEFF_ASPCAP", "HR"),
+        # ("DEC", "AL_H - FE_H", "DecAl")
         ]:
         fn = "{}/{}_{}.{}".format(dir, name, plotname, suffix)
         if os.path.exists(fn):
@@ -168,7 +169,7 @@ if __name__ == "__main__":
                    "AL_H", "CA_H", "C_H", "K_H",  "MG_H", "MN_H", "NA_H",
                    "NI_H", "N_H",  "O_H", "SI_H", "S_H",  "TI_H", "V_H"]
 
-    for K in 2 ** np.arange(7, 10):
+    for K in [256, 512]:
         pfn = "data/kmeans_{:04d}.pkl".format(K)
         try:
             print("attempting to read pickle", pfn)
@@ -206,7 +207,7 @@ if __name__ == "__main__":
         for k in range(K):
             I = (clusters == k)
             sizes[k] = np.sum(I)
-            subdata = data[I]
+            subdata = data[I,:] - (np.mean(data[I], axis=0))[None, :]
             if sizes[k] > (D + 1):
                 variance = np.sum(subdata[:,:,None] * subdata[:,None,:], axis=0)
                 s, logdets[k] = np.linalg.slogdet(variance)
@@ -227,6 +228,7 @@ if __name__ == "__main__":
                 plotdata_labels[d] = data_labels[d] + " - " + data_labels[0]
             metadata_labels = ["RA", "DEC", "GLON", "GLAT", "VHELIO_AVG", "TEFF_ASPCAP", "LOGG_ASPCAP"]
             metadata = get_metadata(dfn, metadata_labels, mask)
+            fields =   get_metadata(dfn, ["FIELD", ], mask)
             plotdata = np.hstack((plotdata, metadata))
             plotdata_labels = plotdata_labels + metadata_labels
 
@@ -234,6 +236,7 @@ if __name__ == "__main__":
         plotcount = 0
         for k in (np.argsort(densities))[::-1]:
[…]

### 2016-01-15 papers/clusters.tex
results

@@
 precisions obtained by \thecannon\ before running k-means.
-This makes the space close to isotropic in measurment uncertainty or
+This makes the space close to isotropic in measurement uncertainty or
 observational precision.
-
-When the k-means results are returned, we construct the $D\times D$ empirical variance tensor for the DWH THIS IS WRONG!
-
-DWH: Citation out to numpy and sklearn.
+Importantly we use \emph{only} abundance-space information, and no
+positional or velocity information (nor $\teff$ nor $\logg$ nor any
+targeting or observational meta-data) as input to the clustering
+algorithm.
 
@@
 
-DWH: Visualize densest clusters.  Some of them are known clusters.
-Show examples.
+When the k-means results are returned, the $D\times D$ empirical
+variance tensor for the members of each cluster can be constructed,
+and an effective density can be computed as the number of points in
+the cluster divided by the square-root of the determinant of the
+tensor.
+This density was used to rank abundance-space overdensities (from highest
+to lowest) for visual inspection.
+We chose a few interesting cases from many and show them in
+\figurename~\ref{fig:M13}, \ref{fig:Pal5}, \ref{fig:Sgr},
+\ref{fig:halo}, and \ref{fig:disk}.
+The first three of these are known clusters M13, Pal5, and Sagittarius (MKN, ARC, help!).
+The fourth is a halo structure with high velocity dispersion and possibly accreted.
+The fifth is a thin-disk star-formation feature.
+We will discuss these more below.
 
-DWH: Do we have streams or other structures?
-
-DWH: What have we demonstrated?
+All of the code for the clustering and visualization is built on numpy
+(DWH CITE), scikit-learn (DWH CITE), and matplotlib (DWH CITE).
 
@@
 
+DWH: What have we demonstrated?
+
 DWH: For the first time/H], [we see small-scale informative features in abundance space.

### 2016-01-15 papers/from_mkn_1.tex
introduction from MKN
references, acks, tweaks

@@
 
-@ARTICLE{DeSilva2015,
-   author = {{De Silva}, G.~M. and {Freeman}, K.~C. and {Bland-Hawthorn}, J. and 
-	{Martell}, S. and {de Boer}, E.~W. and {Asplund}, M. and {Keller}, S. and 
-	{Sharma}, S. and {Zucker}, D.~B. and {Zwitter}, T. and {Anguiano}, B. and 
-	{Bacigalupo}, C. and {Bayliss}, D. and {Beavis}, M.~A. and {Bergemann}, M. and 
-	{Campbell}, S. and {Cannon}, R. and {Carollo}, D. and {Casagrande}, L. and 
-	{Casey}, A.~R. and {Da Costa}, G. and {D'Orazi}, V. and {Dotter}, A. and 
-	{Duong}, L. and {Heger}, A. and {Ireland}, M.~J. and {Kafle}, P.~R. and 
-	{Kos}, J. and {Lattanzio}, J. and {Lewis}, G.~F. and {Lin}, J. and 
-	{Lind}, K. and {Munari}, U. and {Nataf}, D.~M. and {O'Toole}, S. and 
-	{Parker}, Q. and {Reid}, W. and {Schlesinger}, K.~J. and {Sheinis}, A. and 
-	{Simpson}, J.~D. and {Stello}, D. and {Ting}, Y.-S. and {Traven}, G. and 
-	{Watson}, F. and {Wittenmyer}, R. and {Yong}, D. and {{\v Z}erjal}, M.
-	},
-    title = "{The GALAH survey: scientific motivation}",
-  journal = {\mnras},
-archivePrefix = "arXiv",
-   eprint = {1502.04767},
- keywords = {Galaxy: stellar content},
-     year = 2015,
-    month = may,
-   volume = 449,
-    pages = {2604-2617},
-      doi = {10.1093/mnras/stv327},
-   adsurl = {http://adsabs.harvard.edu/abs/2015MNRAS.449.2604D},
-  adsnote = {Provided by the SAO/NASA Astrophysics Data System}
-}
-
-@ARTICLE{Freeman2002,
-   author = {{Freeman}, K. and {Bland-Hawthorn}, J.},
-    title = "{The New Galaxy: Signatures of Its Formation}",
-  journal = {\araa},
-   eprint = {astro-ph/0208106},
- keywords = {osmology, local group, stellar populations, stellar kinematics},
-     year = 2002,
-   volume = 40,
-    pages = {487-537},
-      doi = {10.1146/annurev.astro.40.060401.093840},
-   adsurl = {http://adsabs.harvard.edu/abs/2002ARA%26A..40..487F},
-  adsnote = {Provided by the SAO/NASA Astrophysics Data System}
-}
-
-@ARTICLE{Hawkins2015,
-   author = {{Hawkins}, K. and {Jofr{\'e}}, P. and {Masseron}, T. and {Gilmore}, G.
-	},
-    title = "{Using chemical tagging to redefine the interface of the Galactic disc and halo}",
-  journal = {\mnras},
- keywords = {Galaxy: abundances, Galaxy: halo, Galaxy: structure},
-     year = 2015,
-    month = oct,
-   volume = 453,
-    pages = {758-774},
-      doi = {10.1093/mnras/stv1586},
-   adsurl = {http://adsabs.harvard.edu/abs/2015MNRAS.453..758H},
-  adsnote = {Provided by the SAO/NASA Astrophysics Data System}
-}
-
 @ARTICLE{Ting2015,
@@
 
-   author = {BC2015},
-    title = "{Testing the chemical tagging technique with open clusters}",
-  journal = {\aap},
-archivePrefix = "arXiv",
-   eprint = {1503.02082},
- primaryClass = "astro-ph.SR",
- keywords = {stars: abundances, techniques: spectroscopic, Galaxy: abundances},
-     year = 2015,
-    month = may,
-   volume = 577,
-      eid = {A47},
-    pages = {A47},
-      doi = {10.1051/0004-6361/201425232},
-   adsurl = {http://adsabs.harvard.edu/abs/2015A%26A...577A..47B},
-  adsnote = {Provided by the SAO/NASA Astrophysics Data System}
-}
-
-   author = {{M2014,
-  journal = {\mnras},
-archivePrefix = "arXiv",
-   eprint = {1312.1759},
- keywords = {techniques: miscellaneous, stars: abundances, Galaxy: disc, Galaxy: evolution, open clusters and associations: general},
-     year = 2014,
-    month = mar,
-   volume = 438,
-    pages = {2753-2764},
-      doi = {10.1093/mnras/stt2320},
-   adsurl = {http://adsabs.harvard.edu/abs/2014MNRAS.438.2753M},
-  adsnote = {Provided by the SAO/NASA Astrophysics Data System}
-}
-
-@ARTICLE{Melendez2014,
-   author = {{Mel{\'e}ndez}, J. and {Ram{\'{\i}}rez}, I. and {Karakas}, A.~I. and 
-	{Yong}, D. and {Monroe}, T.~R. and {Bedell}, M. and {Bergemann}, M. and 
-	{Asplund}, M. and {Tucci Maia}, M. and {Bean}, J. and {do Nascimento}, Jr., J.-D. and 
-	{Bazot}, M. and {Alves-Brito}, A. and {Freitas}, F.~C. and {Castro}, M.
-	},
-    title = "{18 Sco: A Solar Twin Rich in Refractory and Neutron-capture Elements. Implications for Chemical Tagging}",
-  journal = {\apj},
-archivePrefix = "arXiv",
-   eprint = {1406.5244},
- primaryClass = "astro-ph.SR",
- keywords = {stars: abundances, stars: AGB and post-AGB, stars: fundamental parameters, Sun: abundances},
-     year = 2014,
-    month = aug,
-   volume = 791,
-      eid = {14},
-    pages = {14},
-      doi = {10.1088/0004-637X/791/1/14},
-   adsurl = {http://adsabs.harvard.edu/abs/2014ApJ...791...14M},
-  adsnote = {Provided by the SAO/NASA Astrophysics Data System}
-}
-
-@ARTICLE{Quillen2015,
-   author = {{Quillen}, A.~C. and {Anguiano}, B. and {De Silva}, G. and {Freeman}, K. and 
-	{Zucker}, D.~B. and {Minchev}, I. and {Bland-Hawthorn}, J.},
-    title = "{The parent populations of six groups identified from chemical tagging in the solar neighbourhood}",
-  journal = {\mnras},
-archivePrefix = "arXiv",
-   eprint = {1503.02887},
- keywords = {Galaxy: abundances, Galaxy: evolution, Galaxy: kinematics and dynamics, open clusters and associations: general},
-     year = 2015,
-    month = jul,
-   volume = 450,
-    pages = {2354-2366},
-      doi = {10.1093/mnras/stv806},
-   adsurl = {http://adsabs.harvard.edu/abs/2015MNRAS.450.2354Q},
-  adsnote = {Provided by the SAO/NASA Astrophysics Data System}
-}
-
-\section*{Acknowledgments}
-
-Don't acknowledge Ken as we would like him to be a ref.
-
 \end{document}

### 2016-01-15 papers/clusters.tex

@@
 % - cite codes!
-% - put in acknowledgement
-% - fix all remaining occurrences of ARC, MKN, HWR, or DWH
+% - fix all remaining occurrences of ARC, MKN, HWR, DWH, or CITE
 % - forward to Freeman, others?
@@
 \newcommand{\project}[1]{\textsl{#1}}
+\newcommand{\sdssiii}{\project{\acronym{SDSS-III}}}
 \newcommand{\apogee}{\project{\acronym{APOGEE}}}
@@
 This idea has been dubbed ``chemical tagging''.
-It was proposed long ago \citep{Freeman2002} and is one of the
+It was proposed long ago \citep{freeman} and is one of the
 principal motivations for a number of surveys, most notably
-\apogee\ \citep{apogee}, \gaiaeso\ \citep{gaiaeso} and \galah\ \citep{galah13, galah15}.
+\apogee\ \citep{apogee}, \gaiaeso\ \citep{gaiaeso} and \galah\ \citep{galah}.
 In order to measure chemical tags and infer the Galaxy's dynamical
@@
 stellar-orbit diffusion process like radial mixing and radial
-migration (for example, \citealt{Roskar2009, Quillen2015}).  After
+migration (for example, \citealt{Roskar2009, quillen}).  After
 stars are born---or after a star cluster is accreted and
@@
 The current precision on abundance measurements is not high enough
-(for example, \citealt{Ting2015, Martel2015}).
+(for example, \citealt{Ting2015, martel}).
 That said, it is clear that the \emph{data} are precise enough;
@@
 Our contribution in this space has been to develop
-\thecannon\ \citep{cannon, ages}, which is a data-driven model for
+\thecannon\ \citep{thecannon, ages}, which is a data-driven model for
 stellar spectra.
@@
 
-In what follows, we are going to use \apogee\ \acronym{DR12} data, in
+In what follows, we are going to use \apogee\ \acronym{DR12} data \citep{dr12}, in
 which we can measure 15 element abundances (C, N, O, Na, Mg, Al, Si,
@@
 \apogee\ \aspcap\ (\cite{aspcap}) pipeline reductions.
-There are XXXXX stars that make our $\logg$ and flag cuts.
+There are 98462 [DWH CHECK] stars that make our $\logg$ and flag cuts.
 
@@
 use the \apogee\ \acronym{DR12} \aspcap\ atmospheric parameters and
-15-element abundances, but only for the subset of XXXXX red-giant
+15-element abundances, but only for the subset of ARC:XXXXX red-giant
 stars that have spectral signal-to-noise ratio measurements greater
@@
 The simplest method for clustering points in $D$-dimensional space is
-the \emph{k-means} algorithm (DWH CITE).
+the \emph{k-means} algorithm (see CITE for a pedagogical introduction).
 Briefly, the k-means algorithm is to find the points closest (in
@@
 best possible clustering.
-For this reason we simply choose $K=128, 256, 512$ and look at all the
+For this reason we simply choose $K=128$, $K=256$, and $K=512$ and look at all the
 results.
@@
 
-All of the code for the clustering and visualization is built on numpy
-(DWH CITE), scikit-learn (DWH CITE), and matplotlib (DWH CITE).
+All of the code for the data manipulation, clustering, and
+visualization is built on numpy (CITE), scikit-learn (CITE),
+and matplotlib (CITE).
 
@@
 \insanefigure{cluster_0256_0253}
-\caption{Whatevsinator.\label{fig:infallcluster}}
+\caption{Whatevsinator.\label{fig:M13}}
 \end{figure}
@@
 \acknowledgements
-DWH: It is a pleasure to thank...
+DWH was partially supported by
+  the NSF (grants IIS-1124794, AST-1517237),
+  NASA (grant NNX12AI50G), and
+  the Moore-Sloan Data Science Environment at NYU.
+This research made use of the NASA \project{Astrophysics Data System}.
 
-DWH: grants and grant numbers.
+This project made use of \sdssiii\ data.
+Funding for \sdssiii\ has been provided by the Alfred P. Sloan
+Foundation, the Participating Institutions, the National Science
+Foundation, and the U.S. Department of Energy Office of Science. The
+\sdssiii\ web site is http://www.sdss3.org/.
 
-DWH: SDSS-III and SDSS-IV.
-
-DWH: NASA ADS
-
-DWH: github repo and hash.
-
-DWH: Be sure to cite software a la the new ApJ guidelines.
+\sdssiii\ is managed by the Astrophysical Research Consortium for the
+Participating Institutions of the \sdssiii\ Collaboration including the
+University of Arizona, the Brazilian Participation Group, Brookhaven
+National Laboratory, Carnegie Mellon University, University of
+Florida, the French Participation Group, the German Participation
+Group, Harvard University, the Instituto de Astrofisica de Canarias,
+the Michigan State/Notre Dame/JINA Participation Group, Johns Hopkins
+University, Lawrence Berkeley National Laboratory, Max Planck
+Institute for Astrophysics, Max Planck Institute for Extraterrestrial
+Physics, New Mexico State University, New York University, Ohio State
+University, Pennsylvania State University, University of Portsmouth,
+Princeton University, the Spanish Participation Group, University of
+Tokyo, University of Utah, Vanderbilt University, University of
+Virginia, University of Washington, and Yale University.
 
@@
 \begin{thebibliography}{24}\raggedright
-\bibitem[Ness \etal(2016)]{ness16}
-whatev
+\bibitem[Alam \etal(2015)]{dr12}
+Alam,~S., Albareti,~F.~D., Allende~Prieto,~C., \etal, 2015, \apjs, 219, 12
+\bibitem[Blanco-Cuaresma \etal(2015)]{blancocuaresma}
+Blanco-Cuaresma,~S., Soubiran,~C., Heiter,~U., \etal, 2015, \aap, 577, A47
+\bibitem[Bovy(2015)]{bovy}
+Bovy,~J., 2015, arXiv:1510.06745
+\bibitem[Casey \etal(2016)]{casey16}
+Casey,~A.~R., Hogg,~D.~W., Ness,~M., Rix,~H.-W., 2016, in preparation
+\bibitem[De~Silva \etal(2015)]{galah}
+De~Silva,~G.~M., Freeman,~K.~C., Bland-Hawthorn,~J., \etal, 2015, \mnras, 449, 2604
+\bibitem[Freeman \& Bland-Hawthorn(2002)]{freeman}
+Freeman,~K., \& Bland-Hawthorn,~J., 2002, \araa, 40, 487
+\bibitem[Garc{\'{\i}}a~P{\'e}rez \etal(2015)]{aspcap}
+Garc{\'{\i}}a~P{\'e}rez,~A.~E., Allende~Prieto,~C., Holtzman,~J.~A., \etal, 2015, arXiv:1510.07635
+\bibitem[Majewski \etal(2015)]{apogee}
+Majewski,~S.~R., Schiavon,~R.~P., Frinchaboy,~P.~M., \etal, 2015, arXiv:1509.05420
+\bibitem[Melendez \etal(2014)]{melendez}
+Mel{\'e}ndez,~J., Ram{\'{\i}}rez,~I., Karakas,~A.~I., \etal, 2014, \apj, 791, 14
+\bibitem[Mitschang \etal(2014)]{mitschang}
+Mitschang,~A.~W., De~Silva,~G., Zucker,~D.~B., \etal, 2014, \mnras, 438, 2753
+\bibitem[Ness \etal(2015)]{thecannon}
+Ness,~M., Hogg,~D.~W., Rix,~H.-W., Ho,~A.~Y.~Q., Zasowski,~G., 2015, \apj, 808, 16
+\bibitem[Ness \etal(2016a)]{ages}
+Ness,~M., Hogg,~D.~W., Rix,~H.-W., \etal, 2016a, \apj, in press (arXiv:1511.08204)
+\bibitem[Ness \etal(2016b)]{ness16}
+Ness,~M., Hogg,~D.~W., Casey,~A.~R., Rix,~H.-W., 2016b, in preparation
+\bibitem[Quillen \etal(2015)]{quillen}
+Quillen,~A.~C., Anguiano,~B., De~Silva,~G., \etal, 2015, \mnras, 450, 2354
+\bibitem[Ting \etal(2016)]{ting}
+Ting,~Y.-S., Conroy,~C., \& Rix,~H.-W., 2016, \apj, in press (arXiv:1507.07563)
 \end{thebibliography}

### 2016-01-15 papers/clusters.tex
zeroth draft DONE

@@
 \newcommand{\documentname}{\textsl{Article}}
+\newcommand{\sectionname}{Section}
 
@@
 \includegraphics[width=0.33\textwidth]{./figs/#1_HR.png}}
+
+% results shit
+\newcommand{\totalnumber}{98462} %% DWH CHECK
 
@@
 noise in chemical-abundance measurements.
-However, we have succeeded in improving vastly spectroscopic measurements with \thecannon\ 
+However, we have succeeded in substantially improving spectroscopic measurements with \thecannon\ 
 (now delivering individual abundances with precisions around 0.04-dex).
@@
 The current precision on abundance measurements is not high enough
-(for example, \citealt{Ting2015, martel}).
+(for example, \citealt{martel, ting}).
 That said, it is clear that the \emph{data} are precise enough;
@@
 
-\clearpage
 \section{Sample and abundances}
@@
 \apogee\ \aspcap\ (\cite{aspcap}) pipeline reductions.
-There are 98462 [DWH CHECK] stars that make our $\logg$ and flag cuts.
+There are \totalnumber\ stars that make our $\logg$ and flag cuts.
 
@@
 
-DWH!!  How does \thecannon\ work here?
+As described in the original paper \citep{thecannon} and the companion
+papers \citep{casey16, ness16}, \thecannon\ is operated in two stages.
+In the first stage---the traning step---the training set of stars with
+measured spectra and known labels (known parameters and abundances) is
+used to establish a likelihood function for a set of coefficients that
+parameterize a smooth, simple prediction of spectral pixels given
+labels.
+\thecannon\ finds the set of coefficients that optimizes the
+training-step likelihood function.
+In the implementation used in this \documentname, the spectral
+prediction is quadratic in labels (including all cross-terms to second
+order), and there are 17 labels ($\teff$, $\logg$, and 15 abundances
+referenced to H) and 8500 spectral pixels, so there are on the order
+of 1.4 million coefficients set by a convex optimization in the
+training step.
+
+In the second stage---the test step---each unlabeled star is used to
+establish a single-star likelihood function for labels, holding the
+coefficients fixed.
+\thecannon\ finds the labels that minimize the single-star likelihood
+function for each test-set star.
+This optimization is not convex, but it is embarrassingly parallel.
+For the results shown here, the training step takes hours and the test
+step takes less than an hour on a small research cluster in Cambridge.
 
@@
 \insanefigure{all}
-\caption{Whatevsinator.\label{fig:all}}
+\caption{The full sample of \totalnumber\ stars used in this study.
+  The top six panels show six two-dimensional projections of the
+  empirical abundance space distribution.
+  The clustering algorithm (described in
+  \sectionname~\ref{sec:method}) performs clustering in a full
+  15-dimensional abundance space of elements referenced to H (not Fe),
+  but plots are shown here referenced to Fe for familiarity reasons.
+  The bottom three panels show stellar meta-data not used in the
+  clustering but used to test whether the abundance-space clusters
+  are interesting spatially.\label{fig:all}}
 \end{figure}
@@
 \clearpage
-\section{Method and results}
+\section{Method and results}\label{sec:method}
 
@@
 known clusters do appear in the abundance space as over-densities.
+(In general, the collapse of 15 dimensions down to 2-d projections
+will hide, smooth, or dilute any structure; there is no guarantee that
+a highly featured distribution in 15-d will show features in
+\emph{any} 2-d projection, let alone axis-aligned, human-selected projections.)
 This encourages us to look for over-densities automatically in the
@@
 The simplest method for clustering points in $D$-dimensional space is
-the \emph{k-means} algorithm (see CITE for a pedagogical introduction).
+the \emph{k-means} algorithm (see \citealt{bishop} for a pedagogical introduction).
 Briefly, the k-means algorithm is to find the points closest (in
@@
 
-DWH: Reminder that we don't need the absolute best clustering to
-demonstrate the hypothesis.
+DWH: EXPLICITLY GIVE THE ABUNDANCE-SPACE SCALES USED.
+
+One note on clustering algorithm decisions is warranted:
+Because this project is just a demonstration of the chemical-tagging
+concept, there is no requirement that we choose the best clustering
+algorithm or the method that reveals the best structures.
+Indeed, it is a feature of this work that even the simplest, most
+generic, least tuned clustering algorithm returns interesting
+structures (as shown below).
+The prospects for Milky Way science only improve as the clustering
+algorithm improves.
 
@@
 
+DWH \& HWR: How do we show that we aren't just getting clusters because \apogee\ has plate-to-plate calibration variations?
+
 All of the code for the data manipulation, clustering, and
-visualization is built on numpy (CITE), scikit-learn (CITE),
-and matplotlib (CITE).
+visualization is built on \project{numpy} \citep{numpy},
+\project{scikit-learn} \citep{sklearn}, and \project{matplotlib}
+\citep{matplotlib}.
+All of the code written specifically for this project is available in
+two open-source code repositories at
+\url{https://github.com/andycasey/AnniesLasso} and
+\url{https://github.com/davidwhogg/Platypus}.
 
@@
 \insanefigure{cluster_0256_0253}
-\caption{Whatevsinator.\label{fig:M13}}
+\caption{Same as \figurename~\ref{fig:all} except that the entire
+  sample has been made gray and the members of one of the most dense
+  $K=256$ abundance-space clusters has been made black (and prominent).
+  This cluster, which was identified only in abundance space (six
+  projections of which are the top six panels in this \figurename),
+  turns out to be dominated by the halo globular cluster
+  M13.\label{fig:M13}}
 \end{figure}
-
 \begin{figure}[!p]
 \insanefigure{cluster_0256_0034}
-\caption{Whatevsinator.\label{fig:Pal5}}
+\caption{Same as \figurename~\ref{fig:M13} but for another dense
+  abundance-space cluster.
+  This one turns out to be dominated by the disrupting globular
+  cluster Palomar 5.\label{fig:Pal5}}
 \end{figure}
-
 \begin{figure}[!p]
 \insanefigure{cluster_0256_0177}
-\caption{Whatevsinator.\label{fig:Sgr}}
+\caption{Same as \figurename~\ref{fig:M13} but for another dense
+  abundance-space cluster.
+  This one turns out to be dominated by the Sagittarius star
+  cloud.\label{fig:Sgr}}
 \end{figure}
-
 \begin{figure}[!p]
 \insanefigure{cluster_0256_0010}
-\caption{Whatevsinator.\label{fig:halo}}
+\caption{Same as \figurename~\ref{fig:M13} but for another dense
+  abundance-space cluster.
+  This one turns out to be dominated by a high-velocity-dispersion
+  structure in the Galaxy halo.\label{fig:halo}}
 \end{figure}
-
 \begin{figure}[!p]
 \insanefigure{cluster_0256_0141}
-\caption{Whatevsinator.\label{fig:disk}}
+\caption{Same as \figurename~\ref{fig:M13} but for another dense
+  abundance-space cluster.
+  This one turns out to be dominated by a very thin young stellar
+  structure in the Galaxy disk.\label{fig:disk}}
 \end{figure}
@@
 
-DWH: What have we demonstrated?
+We have demonstrated for the first time that, in a large-scale
+spectroscopic survey of stars, stars that appear similar in
+chemical-abundance properties are often associated physically in
+clusters or other structures.
+The reverse is well established---that is, that stars that are
+associated physically in clusters are similar in chemical-abundance
+properties (with a recent model-free test by \citealt{bovy}).
+MKN: WHAT ELSE TO CITE HERE?
+However, it was not known whether abundance-space information would be
+precise enough, or \emph{measurable} precisely enough to find this
+structure in surveys dominated by a large mix of stars from different
+origins and ages (and there was good reason to be pessimistic;
+\citealt{ting}).
+We have demonstrated with a straightforward (and in no way optimal)
+clustering in chemical-abundance space that it is easy to find co-eval
+stellar structures.
 
-DWH: For the first time/H], [we see small-scale informative features in abundance space.
+What made this possible is substantial improvements in precision
+delivered by \thecannon.
+We don't have an absolute measure of the precision of the measurements,
+but from looking at open and globular clusters, it appears to be on the
+order of or better than 0.04~dex for the median element in the list of 15.
+These improvements are discussed in detail in the companion papers
+\citep{casey16, ness16} about the abundance measurements.
+We believe these improvements come from a combination of factors, not
+limited to (a)~improvements in the determination of pseudo-continuum,
+(b)~the use of more spectral range than just unblended element windows
+\citep{aspcap}, and (c)~accurate spectral predictions
+(\thecannon\ delivers accurate predictions because it is fit to
+observed data).
+Furthermore, because abundance space is high in dimensionality, even
+small improvements in abundance measurement get taken to a significant
+power when thinking about information gains.
 
-DWH: When we identify the small-scale features in abundance space,they are informative in position space.
+There is no sense at all in which \figurename~\ref{fig:M13},
+\ref{fig:Pal5}, \ref{fig:Sgr}, \ref{fig:halo}, and \ref{fig:disk} show
+a representative or complete set of features found in abundance space.
+These features were hand-chosen to be obviously interesting and
+interpetable.
+There are many other things to be found in this data set!
 
-DWH: How do we show that we aren't just getting clusters because \apogee\ has plate-to-plate calibration variations?
+On the particulars of these features, a few comments:
+DWH: Comments about Sag and M13: They appear in multiple clusters.
+MKN: Comments about the young disk feature and what it probably is?
 
-DWH: Chemical tagging will work.
+The \apogee\ project shows great promise for these studies, and with
+the appearance of the companion papers, we will release the
+chemical-abundance measurements we used here for further study.
+The future is even brighter, however: \apogee\ is expanding to 19
+elements, more stars, and all-sky angular coverage with \apogee2, and
+\galah\ is nearly ready to release chemical abundances on a larger set
+of 29 elements.
+The \apogee\ elements employed in this project include alpha
+(Mg, O, Si, Ca, Ti), light proton-capture (C, N, S), odd-Z (Na, Al, K)
+and iron peak (Fe, Ni, V, Mn) elements.
+\galah\ will deliver element abundances from all the major
+nucleosynthetic processes, including light proton capture elements,
+alpha, odd-Z, iron-peak, light and heavy slow neutron capture, and
+rapid neutron capture.
+Chemical tagging capabilities are expected to grow as the
+\emph{product} of the measured nucleosynthetic pathways.
 

### 2016-01-16 code/kmeans.py
HWR-inspired plotting tweaks

@@ -65,8 +65,9 @@ def plot_one_cluster(data, labels, mask, name, dir):
         ("S_H - FE_H", "AL_H - FE_H", "SAl"),
         ("MG_H - FE_H", "K_H - FE_H", "MgK"),
         ("DEC", "RA", "sky"),
-        # ("GLAT", "GLON", "gal"),
+        ("GLAT", "GLON", "gal"),
         ("DEC", "VHELIO_AVG", "decv"),
+        ("VHELIO_AVG", "GLON", "vlon"),
         ("LOGG_ASPCAP", "TEFF_ASPCAP", "HR"),
         # ("DEC", "AL_H - FE_H", "DecAl")
         ]:
@@ -76,8 +77,12 @@ def plot_one_cluster(data, labels, mask, name, dir):
         else:
             y = np.where(np.array(labels) == ly)[0][0]
             x = np.where(np.array(labels) == lx)[0][0]
-            fig = plt.figure(figsize=(4,4))
-            plt.subplots_adjust(left=0.2, right=0.96, bottom=0.2, top=0.96)
+            if plotname == "sky" or plotname == "gal" or plotname == "vlon":
+                fig = plt.figure(figsize=(8,4))
+                plt.subplots_adjust(left=0.1, right=0.98, bottom=0.2, top=0.96)
+            else:
+                fig = plt.figure(figsize=(4,4))
+                plt.subplots_adjust(left=0.2, right=0.96, bottom=0.2, top=0.96)
             plt.clf()
             kwargs = {"marker": ".", "ls": "none"}
             if mask is None:
@@ -134,10 +139,10 @@ if __name__ == "__main__":
                   "S_H - FE_H": (-0.4, 0.9),
                   "TI_H - FE_H": (-0.6, 0.5),
                   "V_H - FE_H": (-0.9, 0.8),
-                  "RA": (360., 0.),
-                  "DEC": (-35., 90.),
-                  "GLON": (60., 0.),
-                  "GLAT": (-30., 30.),
+                  "RA":   (360., 0.),
+                  "DEC":  (-60., 90.),
+                  "GLON": (360., 0.),
+                  "GLAT": (-65., 85.),
                   "VHELIO_AVG": (-275., 275.),
                   "TEFF_ASPCAP": (5500., 3500.),
                   "LOGG_ASPCAP": (3.9, 0.0), }
@@ -226,9 +231,10 @@ if __name__ == "__main__":
             for d in range(1, D):
[…]

### 2016-01-17 papers/clusters.tex
import from Rix

@@
 
-\title{Chemical tagging: \\
+\title{Chemical tagging can work: \\
        Identification of stellar phase-space structures \\
@@
 
-If every star in the Milky Way was born in a molecular cloud that has
-unique chemical-abundance characteristics (as a function of time), and
-if every star preserves its surface abundances (or significant aspects
-of them) over its lifetime, then stars with common birthplaces and
-time ought to be identifiable in their surface abundances.
-That is, it ought to be possible to use stellar abundances to link up
-co-eval stars---even if at the present day they are no longer close to
-one another in phase space---and infer a detailed star-formation and
-dynamical history for the Galaxy.
-This idea has been dubbed ``chemical tagging''.
-It was proposed long ago \citep{freeman} and is one of the
+Ensembles of stars in the Milky Way are born in molecular clouds,
+which are presumed to be near-homogeneous in their chemical element composition.
+However, most stars are born in unbound associations, or are born in star clusters that disperse rapidly: they will eventually end up in very different parts of phase space, on differing orbits.
+Yet, if every star preserved its photospheric element abundances over its lifetime (at least for most elements), then stars of common birth origin ought to be identifiable through their detailed photospheric abundances, long after any spatial or phase-space proximity has vanished.
+
+This idea, dubbed ``chemical tagging'', was proposed long ago \citep{freeman} and is one of the
 principal motivations for a number of surveys, most notably
 \apogee\ \citep{apogee}, \gaiaeso\ \citep{gaiaeso} and \galah\ \citep{galah}.
-In order to measure chemical tags and infer the Galaxy's dynamical
-properties and formation history, these surveys are each measuring high resolution,
+In order to determine the precise abundance labels for chemical tagging,  these surveys are each measuring high resolution,
 high signal-to-noise spectra for hundreds of thousands of stars across
@@
 disrupted---associations or groups will disperse, through either
-coherent orbital evolution or else incoherent diffusion, but the
-chemical-abundance tags can in principle make the former members
-identifiable as co-eval stellar siblings.
+internal two-body processes or external tidal forces.
 
-Although undeniably promising---and although it has motivated the
-launch of expensive, large-scale spectroscopic surveys---the
-chemical-tagging idea has yet to yield scientific fruit.
-Part of the reason is that the level of chemical specificity required
+Although undeniably promising---and motivating the 
+launch of costly large-scale spectroscopic surveys---
+chemical-tagging as a search technique has yet to be proven in practice: identifying stars of common birth origin {\it purely} on the basis of their near-identical abundances patterns, without any consideration of position or velocity.
+
+Part of the reason is that the level of abundance specificity required
 is very high:
 If there are thousands of (relevant) molecular clouds forming stars in
-the recent history of the Milky Way, it will take many abundances,
-each measured accurately, to make a sufficiently unique fingerprint
-that stellar sibling identification will be reliable.
+the recent history of the Milky Way, clumps of stars can only stand out in a high-dimensional abundance space: therefore accurate, or at least precise, measurements of many different abundances are needed for stellar sibling to have sufficiently unique fingerprints.
+Stellar spectroscopic surveys now have the resolution, signal-to-noise,
+wavelength coverage, and sample sizes to deliver many different chemical tags for each star.
 
-Stellar spectroscopic surveys now have the resolution, signal-to-noise,
-wavelength coverage, and sample sizes to deliver chemical tags.
 There are, however, two big issues.
@@
 here---is measurement precision (and accuracy).
-The current precision on abundance measurements is not high enough
-(for example, \citealt{martel, ting}).
-That said, it is clear that the \emph{data} are precise enough;
-there is enough signal-to-noise at the relevant locations in spectrum
-space to deliver high-precision tags.
-Our view is that the traditional methods for making abundance
-measurements just aren't mature enough:
-There are issues with the physical models, and the data-analysis
-methods do not make use of all of the information in the data sets.
+The current precision on abundance measurements in the published survey catalogs is not high enough (for example, \citealt{martel, ting}).
+However, our recent work \citep{thecannon} suggests that \emph{data} are precise enough: there is enough signal-to-noise at the relevant 
+locations in spectrum space to deliver high-precision tags.
+The existence of very large data sets, homogeneous in data/spectrum space, suggest that one can use data-driven approaches to determining stellar abundances, which can considerably outperform traditional methods.
+These traditional approaches are based on ab-initio physical models
+that have shortcomings that become apparent in this age of high-quality spectra; and the data-analysis methods do not make use of all of the information in the data sets.
 Improving the models and exploiting the entire information
@@
 
-Our contribution in this space has been to develop
+Our specific contribution in this space has been to develop
 \thecannon\ \citep{thecannon, ages}, which is a data-driven model for
@@
 In what follows, we are going to use \apogee\ \acronym{DR12} data \citep{dr12}, in
-which we can measure 15 element abundances (C, N, O, Na, Mg, Al, Si,
-S, K, Ca, Ti, V, Mn, Fe, Ni).
+which we can re-derive 15 element abundances (C, N, O, Na, Mg, Al, Si,
+S, K, Ca, Ti, V, Mn, Fe, Ni), using \thecannon\ \citep{thecannon}.
 The \apogee\ data set covers a huge radial extent and---because the
@@
 One final note:
-We think of this \documentname\ as performing a hypothesis test.
+we think of this \documentname\ as performing a proof of concept.
 We know that the stellar members of open and globular clusters---stars
@@
 Does this work the other way around?
-Are groups of stars that are identified as being close in abundance
-space---collections of stars with near-identical chemical tags---also
-found to be close in phase space?
-If the answer to this question is ``yes'' then chemical tagging is
+I.e.
+can chemical tagging identify small subsets of stars, among a vastly greater background sample, that have a common birth origin.
+If we find stars purely by their clustering in abundance space and subsequently show that they are part of a still spatially coherent cluster, then chemical tagging is
 \emph{on.}
 
-\section{Sample and abundances}
+\section{Data: \apogee\  giants with [X/H] re-derived from \thecannon }
 
-The details about how we select, reduce, and analyze the \apogee\ data
-are given in full detail in the companion papers \citep{casey16, ness16}.
-We summarize briefly here.
-For the spectroscopic data, we use the \apogee\ \acronym{DR12}
-(\cite{dr12}) spectroscopic data reductions for red-giant stars
-($\logg < 3.9$) that are not flagged as bad in the
-\apogee\ \aspcap\ (\cite{aspcap}) pipeline reductions.
-There are \totalnumber\ stars that make our $\logg$ and flag cuts.
+Our analysis draws on the spectra of \totalnumber\ giant stars ($\logg < 3.9$) from \apogee\ \acronym{DR12} (\cite{dr12}), alll those not flagged as bad in the \apogee\ \aspcap\ (\cite{aspcap}) pipeline reductions.
+We re-analyze these
+\emph{spectra} using \thecannon\ , because this deliver stellar abundance labels of higher precision, especially for stars of SNR$\le 150$.
 
-The \apogee\ pipelines deliver stacked (co-added),
-pseudo-continuum-normalized spectra, but we do not use these; we
-perform our own continuum normalization on individual-exposure data
-and co-add it ourselves.
-This is required, because the performance of \thecannon\ depends
-critically on the methods used for spectral normalization.  It does
-not require accurate continuum estimation, but it requires that the
-normalization vector be obtained by a \emph{linear operation on the
-  data}.
-The linear operation we use is a smooth-function fit to pixels
-identified as continuum in our previous work \citep{thecannon}.
-
-For the training data labels (used to train \thecannon; see below), we
-use the \apogee\ \acronym{DR12} \aspcap\ atmospheric parameters and
-15-element abundances, but only for the subset of ARC:XXXXX red-giant
-stars that have spectral signal-to-noise ratio measurements greater
-than 200.
-This is a small fraction of the total sample, but these high
-signal-to-noise stars are observed to obtain consistent, low-scatter,
+The details about how we select, reduce, and analyze the \apogee\ data are given in full detail in the companion papers \citep{casey16, ness16}, and we only summarize briefly here.
+We re-derived 17 labels ($\teff$, $\logg$, and 15 abundances
+referenced to H) form the \apogee\ \acronym{DR12} spectra. 
+In the \thecannon 's  training step we trained on the
+ARC:XXXXX red-giant stars that have spectral SNR$\ge 200$.
+For this small fraction of the sample with the highest SNR objects the labels provided by \aspcap provide consistent, low-scatter,
 sensible abundance-space measurements.
-
-As described in the original paper \citep{thecannon} and the companion
-papers \citep{casey16, ness16}, \thecannon\ is operated in two stages.
-In the first stage---the traning step---the training set of stars with
-measured spectra and known labels (known parameters and abundances) is
-used to establish a likelihood function for a set of coefficients that
-parameterize a smooth, simple prediction of spectral pixels given
-labels.
-\thecannon\ finds the set of coefficients that optimizes the
-training-step likelihood function.
-In the implementation used in this \documentname, the spectral
-prediction is quadratic in labels (including all cross-terms to second
-order), and there are 17 labels ($\teff$, $\logg$, and 15 abundances
-referenced to H) and 8500 spectral pixels, so there are on the order
-of 1.4 million coefficients set by a convex optimization in the
-training step.
-
-In the second stage---the test step---each unlabeled star is used to
-establish a single-star likelihood function for labels, holding the
+The training step fixes a spectral model that predics the normalized spectrum as a quadratic function of all labels.
+In the second stage---the test step---each unlabeled star is used to establish a single-star likelihood function for its labels, holding the model
 coefficients fixed.
@@
 function for each test-set star.
-This optimization is not convex, but it is embarrassingly parallel.
-For the results shown here, the training step takes hours and the test
+This optimization is not convex, but it is embarrassingly parallel,
+and therefore fast. The test
 step takes less than an hour on a small research cluster in Cambridge.
+
+% We start by performing our own continuum normalization on individual-exposure spectra and co-add them ourselves.
+% This is required, because the performance of \thecannon\ depends
+% critically on the methods used for spectral normalization.  It does
+% not require accurate continuum estimation, but it requires that the
+% normalization vector be obtained by a \emph{linear operation on the
+%   data}. The linear operation we use is a smooth-function fit to pixels
+% identified as continuum in our previous work \citep{thecannon}.
+
+% For the training data labels (used to train \thecannon; see below), we
+% use the \apogee\ \acronym{DR12} \aspcap\ atmospheric parameters and
+% 15-element abundances, but only for the subset of ARC:XXXXX red-giant
+% stars that have spectral signal-to-noise ratio measurements greater
+% than 200.
+% This is a small fraction of the total sample, but these highest
+% signal-to-noise stars are observed to obtain consistent, low-scatter,
+% sensible abundance-space measurements.
+
+% As described in the original paper \citep{thecannon} and the companion
+% papers \citep{casey16, ness16}, \thecannon\ is operated in two stages.
+% In the first stage---the training step---the training set of stars with
+% measured spectra and known labels (known parameters and abundances) is
+% used to establish a likelihood function for a set of coefficients that
+% parameterize a smooth, simple prediction of spectral pixels given
+% labels.
+% \thecannon\ finds the set of coefficients that optimizes the
+% training-step likelihood function.
+% In the implementation used in this \documentname, the spectral
+% prediction is quadratic in labels (including all cross-terms to second
+% order), and there are 17 labels ($\teff$, $\logg$, and 15 abundances
+% referenced to H) and 8500 spectral pixels, so there are on the order
+% of 1.4 million coefficients set by a convex optimization in the
+% training step.
+
+% In the second stage---the test step---each unlabeled star is used to
+% establish a single-star likelihood function for labels, holding the
+% coefficients fixed.
+% \thecannon\ finds the labels that minimize the single-star likelihood
+% function for each test-set star.
+% This optimization is not convex, but it is embarrassingly parallel.
+% For the results shown here, the training step takes hours and the test
+% step takes less than an hour on a small research cluster in Cambridge.
 

### 2016-01-17 papers/clusters.tex
small tweaks of HWR stuff

@@
 % to-do
-% - write data and cannon section
-% - write method and results section
-% - write discussion section
+% - write discussion section w MKN input
+% - expand kmeans discussion
 % - put in references
@@
 % - submit to journal and arXiv
+
+% style notes
+% - never use \cite{}, always \citep{} or \citealt{}.
 
@@
 Ensembles of stars in the Milky Way are born in molecular clouds,
-which are presumed to be near-homogeneous in their chemical element composition.
-However, most stars are born in unbound associations, or are born in star clusters that disperse rapidly: they will eventually end up in very different parts of phase space, on differing orbits.
-Yet, if every star preserved its photospheric element abundances over its lifetime (at least for most elements), then stars of common birth origin ought to be identifiable through their detailed photospheric abundances, long after any spatial or phase-space proximity has vanished.
+which are presumed to be near-homogeneous in their chemical element
+composition.
+However, most stars are born in unbound associations, or are born in
+star clusters that disperse rapidly; they will eventually end up in
+very different parts of phase space, on differing orbits.
+Yet, if every star preserved its photospheric element abundances over
+its lifetime (at least for most elements), then stars of common birth
+origin ought to be identifiable through their detailed photospheric
+abundances, long after any spatial or phase-space proximity has
+vanished.
 
-This idea, dubbed ``chemical tagging'', was proposed long ago \citep{freeman} and is one of the
-principal motivations for a number of surveys, most notably
-\apogee\ \citep{apogee}, \gaiaeso\ \citep{gaiaeso} and \galah\ \citep{galah}.
-In order to determine the precise abundance labels for chemical tagging,  these surveys are each measuring high resolution,
-high signal-to-noise spectra for hundreds of thousands of stars across
-the Galaxy's disk, bulge and halo.
+This idea, dubbed ``chemical tagging'', was proposed long ago
+\citep{freeman} and is one of the principal motivations for a number
+of surveys, most notably \apogee\ \citep{apogee},
+\gaiaeso\ \citep{gaiaeso} and \galah\ \citep{galah}.
+In order to determine the precise abundance labels for chemical
+tagging, these surveys are each measuring high resolution, high
+signal-to-noise spectra for hundreds of thousands of stars across the
+Galaxy's disk, bulge and halo.
 
@@
 Although undeniably promising---and motivating the 
-launch of costly large-scale spectroscopic surveys---
-chemical-tagging as a search technique has yet to be proven in practice: identifying stars of common birth origin {\it purely} on the basis of their near-identical abundances patterns, without any consideration of position or velocity.
+launch of costly large-scale spectroscopic surveys---%
+chemical-tagging as a search technique has yet to be proven in
+practice: identifying stars of common birth origin {\it purely} on the
+basis of their near-identical abundances patterns, without any
+consideration of position or velocity.
 
@@
 If there are thousands of (relevant) molecular clouds forming stars in
-the recent history of the Milky Way, clumps of stars can only stand out in a high-dimensional abundance space: therefore accurate, or at least precise, measurements of many different abundances are needed for stellar sibling to have sufficiently unique fingerprints.
+the recent history of the Milky Way, clumps of stars can only stand
+out in a high-dimensional abundance space: therefore accurate, or at
+least precise, measurements of many different abundances are needed
+for stellar sibling to have sufficiently unique fingerprints.
 Stellar spectroscopic surveys now have the resolution, signal-to-noise,
@@
 
-\section{Data: \apogee\  giants with [X/H] re-derived from \thecannon }
+\section{Data: \apogee\ giants with abundances from \thecannon}\label{sec:data}
 
-Our analysis draws on the spectra of \totalnumber\ giant stars ($\logg < 3.9$) from \apogee\ \acronym{DR12} (\cite{dr12}), alll those not flagged as bad in the \apogee\ \aspcap\ (\cite{aspcap}) pipeline reductions.
-We re-analyze these
-\emph{spectra} using \thecannon\ , because this deliver stellar abundance labels of higher precision, especially for stars of SNR$\le 150$.
-
-The details about how we select, reduce, and analyze the \apogee\ data are given in full detail in the companion papers \citep{casey16, ness16}, and we only summarize briefly here.
+Our analysis draws on the spectra of \totalnumber\ giant stars ($\logg
+< 3.9$) from \apogee\ \acronym{DR12} \citep{dr12}, alll those not
+flagged as bad in the \apogee\ \aspcap\ \citep{aspcap} pipeline
+reductions.
+We re-analyze these \emph{spectra} using \thecannon, because this
+deliver stellar abundance labels of higher precision, especially for
+stars of SNR$\le 150$.
+The details about how we select, reduce, and analyze the \apogee\ data
+are given in full detail in the companion papers \citep{casey16,
+  ness16}, and we only summarize briefly here.
 We re-derived 17 labels ($\teff$, $\logg$, and 15 abundances
 referenced to H) form the \apogee\ \acronym{DR12} spectra. 
-In the \thecannon 's  training step we trained on the
-ARC:XXXXX red-giant stars that have spectral SNR$\ge 200$.
-For this small fraction of the sample with the highest SNR objects the labels provided by \aspcap provide consistent, low-scatter,
-sensible abundance-space measurements.
-The training step fixes a spectral model that predics the normalized spectrum as a quadratic function of all labels.
-In the second stage---the test step---each unlabeled star is used to establish a single-star likelihood function for its labels, holding the model
-coefficients fixed.
+For the training step of \thecannon, 
+ARC:XXXXX red-giant stars with spectral SNR$\ge 200$ were used.
+For this small fraction of the sample with the highest SNR objects the
+labels provided by \aspcap provide consistent, relatively low-scatter, sensible
+abundance-space measurements.
+The training step fixes a spectral model
+that predics the normalized spectrum as a quadratic function of all
+labels. In the second stage---the test step---each unlabeled star is
+used to establish a single-star likelihood function for its labels,
+holding the model coefficients fixed.
 \thecannon\ finds the labels that minimize the single-star likelihood
@@
 step takes less than an hour on a small research cluster in Cambridge.
+
+% HWR commented this (below) out
 
@@
 \clearpage
-\section{Method and results}\label{sec:method}
+\section{Clustering method and results}\label{sec:method}
 

### 2016-01-17 papers/clusters.tex
figure improvements
thanks

@@
 % figure shit
+\newlength{\figwidth}\setlength{\figwidth}{0.3\textwidth}
 \newcommand{\insanefigure}[1]{%
-\includegraphics[width=0.33\textwidth]{./figs/#1_MgFe.png}%
-\includegraphics[width=0.33\textwidth]{./figs/#1_MgK.png}%
-\includegraphics[width=0.33\textwidth]{./figs/#1_AlMg.png}\\
-\includegraphics[width=0.33\textwidth]{./figs/#1_NaO.png}%
-\includegraphics[width=0.33\textwidth]{./figs/#1_CN.png}%
-\includegraphics[width=0.33\textwidth]{./figs/#1_SAl.png}\\
-\includegraphics[width=0.33\textwidth]{./figs/#1_sky.png}%
-\includegraphics[width=0.33\textwidth]{./figs/#1_decv.png}%
-\includegraphics[width=0.33\textwidth]{./figs/#1_HR.png}}
+\includegraphics[width=\figwidth]{./figs/#1_MgFe.png}%
+\includegraphics[width=\figwidth]{./figs/#1_MgK.png}%
+\includegraphics[width=\figwidth]{./figs/#1_AlMg.png}\\
+\includegraphics[width=\figwidth]{./figs/#1_NaO.png}%
+\includegraphics[width=\figwidth]{./figs/#1_CN.png}%
+\includegraphics[width=\figwidth]{./figs/#1_SAl.png}\\
+\includegraphics[width=2\figwidth]{./figs/#1_gal.png}%
+\includegraphics[width=\figwidth]{./figs/#1_HR.png}\\
+\includegraphics[width=2\figwidth]{./figs/#1_vlon.png}}
 
@@
 \acknowledgements
+It is a pleasure to thank
+  Jeremy Magland (SCDA), and
+  Eddie Schlafly (LBNL)
+for valuable discussions.
 DWH was partially supported by

### 2016-01-17 papers/clusters.tex
references

@@
 (as things that fall in are expected to be chemically distinct from
-those that form in the parent body; \citealt{something}) and
+those that form in the parent body; for example, \citealt{font}) and
 stellar-orbit diffusion process like radial mixing and radial
-migration (for example, \citealt{Roskar2009, quillen}).  After
+migration (for example, \citealt{roskar, quillen}).  After
 stars are born---or after a star cluster is accreted and
@@
 but different birth places \citep{mitschang}, and the
-chemical-abundance space might be low in dimensionality \citep{what}.
+chemical-abundance space might be low in dimensionality.
 On the other hand, precise studies of stellar twins \citep{melendez, jofre}
@@
 here---is measurement precision (and accuracy).
-The current precision on abundance measurements in the published survey catalogs is not high enough (for example, \citealt{martel, ting}).
-However, our recent work \citep{thecannon} suggests that \emph{data} are precise enough: there is enough signal-to-noise at the relevant 
+The current precision on abundance measurements in the published
+survey catalogs is not high enough (for example, \citealt{martell,
+  ting}).
+However, our recent work \citep{thecannon} suggests that \emph{data}
+are precise enough: there is enough signal-to-noise at the relevant
 locations in spectrum space to deliver high-precision tags.
-The existence of very large data sets, homogeneous in data/spectrum space, suggest that one can use data-driven approaches to determining stellar abundances, which can considerably outperform traditional methods.
-These traditional approaches are based on ab-initio physical models
-that have shortcomings that become apparent in this age of high-quality spectra; and the data-analysis methods do not make use of all of the information in the data sets.
+The existence of very large data sets, homogeneous in data/spectrum
+space, suggest that one can use data-driven approaches to determining
+stellar abundances, which can considerably outperform traditional
+methods.
+These traditional approaches are based on \foreign{ab-initio} physical models
+that have shortcomings that become apparent in this age of
+high-quality spectra, and the data-analysis methods do not make use of
+all of the information in the data sets.
 Improving the models and exploiting the entire information
@@
 For this small fraction of the sample with the highest SNR objects the
-labels provided by \aspcap provide consistent, relatively low-scatter, sensible
+labels provided by \aspcap\ provide consistent, relatively low-scatter, sensible
 abundance-space measurements.
@@
 It is a pleasure to thank
-  Jeremy Magland (SCDA), and
-  Eddie Schlafly (LBNL)
+  Jeremy Magland (\acronym{SCDA}), and
+  Eddie Schlafly (\acronym{LBNL})
 for valuable discussions.
-DWH was partially supported by
-  the NSF (grants IIS-1124794, AST-1517237),
-  NASA (grant NNX12AI50G), and
-  the Moore-Sloan Data Science Environment at NYU.
-This research made use of the NASA \project{Astrophysics Data System}.
+This project was funded in part by
+  the \acronym{NSF} (grants \acronym{IIS-1124794}, \acronym{AST-1517237}),
+  \acronym{NASA} (grant \acronym{NNX12AI50G}),
+  the Moore-Sloan Data Science Environment at \acronym{NYU}, and
+  the European Research Council under the
+  European Union's Seventh Framework Programme (FP~7)
+  \acronym{ERC} Grant Agreement n.~\acronym{[321035]}).
+This research made use of the \acronym{NASA} \project{Astrophysics Data System}.
 
@@
 Foundation, the Participating Institutions, the National Science
-Foundation, and the U.S. Department of Energy Office of Science. The
+Foundation, and the \acronym{U.S.} Department of Energy Office of Science. The
 \sdssiii\ web site is http://www.sdss3.org/.
@@
 Group, Harvard University, the Instituto de Astrofisica de Canarias,
-the Michigan State/Notre Dame/JINA Participation Group, Johns Hopkins
+the Michigan State/Notre Dame/\acronym{JINA} Participation Group, Johns Hopkins
 University, Lawrence Berkeley National Laboratory, Max Planck
@@
 Alam,~S., Albareti,~F.~D., Allende~Prieto,~C., \etal, 2015, \apjs, 219, 12
+\bibitem[Bishop(2006)]{bishop}
+Bishop,~C.~M., 2006, \textit{Pattern Recognition and Machine Learning}, Springer, New York
 \bibitem[Blanco-Cuaresma \etal(2015)]{blancocuaresma}
@@
 De~Silva,~G.~M., Freeman,~K.~C., Bland-Hawthorn,~J., \etal, 2015, \mnras, 449, 2604
+\bibitem[Font \etal(2006)]{font}
+Font,~A.~S., Johnston,~K.~V., Bullock,~J.~S., Robertson,~B.~E., 2006, \apj, 638, 585 
 \bibitem[Freeman \& Bland-Hawthorn(2002)]{freeman}
-Freeman,~K., \& Bland-Hawthorn,~J., 2002, \araa, 40, 487
+Freeman,~K., Bland-Hawthorn,~J., 2002, \araa, 40, 487
 \bibitem[Garc{\'{\i}}a~P{\'e}rez \etal(2015)]{aspcap}
 Garc{\'{\i}}a~P{\'e}rez,~A.~E., Allende~Prieto,~C., Holtzman,~J.~A., \etal, 2015, arXiv:1510.07635
+\bibitem[Hunter(2007)]{matplotlib}
+Hunter,~J.~D., 2007, Computing in Science \& Engineering, 9, 90
+\bibitem[Jofre(2015)]{jofre}
+Jofre,~P., 2015, arXiv:1512.05216
 \bibitem[Majewski \etal(2015)]{apogee}
 Majewski,~S.~R., Schiavon,~R.~P., Frinchaboy,~P.~M., \etal, 2015, arXiv:1509.05420
+\bibitem[Martell(2015)]{martell}
+Martell,~S.~L., 2015, arXiv:1507.00079
 \bibitem[Melendez \etal(2014)]{melendez}
@@
 Ness,~M., Hogg,~D.~W., Casey,~A.~R., Rix,~H.-W., 2016b, in preparation
+\bibitem[Pedregosa \etal(2011)]{sklearn}
+Pedregosa,~F., Varoquaux,~G., Gramfort,~A., \etal, 2011, Journal of Machine Learning Research, 12, 2825
 \bibitem[Quillen \etal(2015)]{quillen}
 Quillen,~A.~C., Anguiano,~B., De~Silva,~G., \etal, 2015, \mnras, 450, 2354
+\bibitem[Ro{\v s}kar \etal(2008)]{roskar}
+Ro{\v s}kar,~R., Debattista,~V.~P., Quinn,~T.~R., Stinson,~G.~S., Wadsley,~J., 2008, \apjl, 684, L79
 \bibitem[Ting \etal(2016)]{ting}
-Ting,~Y.-S., Conroy,~C., \& Rix,~H.-W., 2016, \apj, in press (arXiv:1507.07563)
+Ting,~Y.-S., Conroy,~C., Rix,~H.-W., 2016, \apj, in press (arXiv:1507.07563)
+\bibitem[van~der~Walt \etal(2011)]{numpy}
+van~der~Walt,~S., Colbert,~S.~C., Varoquaux,~G., 2011, Computing in Science \& Engineering, 13, 22
 \end{thebibliography}
-
 \end{document}

### 2016-01-17 papers/clusters.tex
affils

@@
     and~Daniel~Foreman-Mackey\altaffilmark{6}}
-\altaffiltext{1}{SCDA}
+\altaffiltext{1}{Simons Center for Data Analysis, 160 Fifth Avenue, 7th Floor, New York, NY 10010, USA}
 \altaffiltext{2}{Center for Cosmology and Particle Physics, Department of Phyics,
@@
 \altaffiltext{4}{Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, D-69117 Heidelberg, Germany}
-\altaffiltext{5}{IoA}
-\altaffiltext{6}{UW}
+\altaffiltext{5}{Institute of Astronomy, University of Cambridge, Madingley Road, Cambridge CB3~0HA, UK}
+\altaffiltext{6}{Department of Astronomy, University of Washington, Box 351580, Seattle, WA 98195-1580}
 \email{david.hogg@nyu.edu}

### 2016-01-17 papers/clusters.tex
pronoun trouble

@@
 stellar spectra.
-This model can deliver stellar parameters and chemical abundances to
+This model can deliver stellar parameters and chemical abundances for
 stars, making use of every pixel of every stellar spectrum (that is,

### 2016-01-18 papers/clusters.tex
expand Ness' nickname

@@
         Andrew~R.~Casey\altaffilmark{5},
-        M.~Ness\altaffilmark{4},
+        Melissa~K.~Ness\altaffilmark{4},
         Hans-Walter~Rix\altaffilmark{4},

### 2016-01-18 papers/clusters.tex
Add Gaia-ESO reference
Add grant number for ARC
It's not just the noise, it's the fact that a lot of abundances then become 'upper limits' or unreported

@@
 This idea has not previously yielded scientific successes, probably because of the
-noise in chemical-abundance measurements.
+noise and incompleteness in chemical-abundance measurements.
 However, we have succeeded in substantially improving spectroscopic measurements with \thecannon\ 
-(now delivering individual abundances with precisions around 0.04-dex).
+(now delivering individual abundances with precisions around 0.04~dex).
 % Aims:
@@
   European Union's Seventh Framework Programme (FP~7)
-  \acronym{ERC} Grant Agreement n.~\acronym{[321035]}).
+  \acronym{ERC} Grant Agreement n.~\acronym{[320360, 321035]}).
 This research made use of the \acronym{NASA} \project{Astrophysics Data System}.
@@
 Garc{\'{\i}}a~P{\'e}rez,~A.~E., Allende~Prieto,~C., Holtzman,~J.~A., \etal, 2015, arXiv:1510.07635
+\bibitem[Gilmore et al.(2012)]{gaiaeso} Gilmore, G., Randich, 
+S., Asplund, M., et al.\ 2012, The Messenger, 147, 25 
 \bibitem[Hunter(2007)]{matplotlib}

### 2016-01-18 papers/clusters.tex
Update Jofre reference
Minor wording in intro

@@
 
-Part of the reason is that the level of abundance specificity required
-is very high:
+Part of the reason that chemical tagging still remains unrealized
+is because the level of abundance specificity required
+is very high.
 If there are thousands of (relevant) molecular clouds forming stars in
@@
 Hunter,~J.~D., 2007, Computing in Science \& Engineering, 9, 90
-\bibitem[Jofre(2015)]{jofre}
-Jofre,~P., 2015, arXiv:1512.05216
+\bibitem[Jofr{\'e} \etal(2015)]{jofre} Jofr{\'e}, P., 
+M{\"a}dler, T., Gilmore, G., et al.\ 2015, \mnras, 453, 1428 
 \bibitem[Majewski \etal(2015)]{apogee}

### 2016-01-18 papers/clusters.tex
I definitely think that this is something that we would forget to remove later on. Therefore, I'm saving my collective ass by removing it now. Sorry @davidwhogg

@@
 to the point that \emph{chemical tagging is now possible.}
-We have thereby effectively saved the collective asses of all the
-collaborators in all of the above-referenced large spectroscopic
-surveys.
 

### 2016-01-18 papers/clusters.tex
Added introductory paragraph about how knowingly wrong ab-initio models/analysis techniques are

@@
 That is, we need a substantial amount of accuracy to achieve our goals.
+
+It is important to realize that analysis techniques based on \foreign{ab-initio}
+physical models are inaccurate, yet they too strive to improve precision
+in the face of knowingly inaccurate models.  Incomplete atomic data, 
+simplifications of photospheric structure, presumptions about convective
+motions, and inconsistencies resulting from assumptions of the local thermal
+equilibrium, all contribute to produce knowingly inaccurate abundances.
+\thecannon\ stands out because it demonstrably improves the precision on
+chemical abundances, whereas the accuracy of those labels is limited only
+by the training set employed.
 

### 2016-01-18 papers/clusters.tex
Hogg: never write while hungry

@@
 data are taken in the infrared---explores the thin disk.
-However, it has the disadvantage that it's spatial coverage is swiss
-cheese, which makes it hard to see, within the data set, linear or
+However, it has the disadvantage that it's spatial coverage is incomplete
+(i.e., limited pointings produce a \foreign{`swiss cheese'}-like structure),
+which makes it hard to see, within the data set, linear or
 extended stellar structures.

### 2016-01-18 papers/clusters.tex
Define 'on'

@@
 informative abundance information that identifies them also in
-chemical-abundance space.
-Does this work the other way around?
-I.e.
-can chemical tagging identify small subsets of stars, among a vastly greater background sample, that have a common birth origin.
-If we find stars purely by their clustering in abundance space and subsequently show that they are part of a still spatially coherent cluster, then chemical tagging is
-\emph{on.}
+chemical-abundance space.  Does this work the other way around? 
+Can chemical tagging identify small subsets of stars, among a vastly
+greater background sample, that have a common birth origin? If we 
+find stars purely by their clustering in abundance space and subsequently
+show that they are part of a still spatially coherent cluster, then
+we will have resolved all practical outstanding issues plaguing 
+chemical tagging, thereby enabling tantalizing potential to truly
+unravel the formation of the Milky Way.
 

### 2016-01-18 papers/clusters.tex
Speed details and size of training set

@@
 For the training step of \thecannon, 
-ARC:XXXXX red-giant stars with spectral SNR$\ge 200$ were used.
+12,681 red-giant stars with spectral SNR$\ge 200$ were used.
 For this small fraction of the sample with the highest SNR objects the
@@
 The training step fixes a spectral model
-that predics the normalized spectrum as a quadratic function of all
-labels. In the second stage---the test step---each unlabeled star is
+that predicts the normalized spectrum as a quadratic function of all
+labels. In the second stage---the test step---each unlabelled star is
 used to establish a single-star likelihood function for its labels,
@@
 and therefore fast. The test
-step takes less than an hour on a small research cluster in Cambridge.
+step for 150,000 spectra takes less than 30~minutes on a small 
+research cluster in Cambridge.
 

### 2016-01-18 papers/clusters.tex
Minor changes, incl. a question

@@
 % use the \apogee\ \acronym{DR12} \aspcap\ atmospheric parameters and
-% 15-element abundances, but only for the subset of ARC:XXXXX red-giant
+% 15-element abundances, but only for the subset of 12,681 red-giant
 % stars that have spectral signal-to-noise ratio measurements greater
@@
 % For the results shown here, the training step takes hours and the test
-% step takes less than an hour on a small research cluster in Cambridge.
+%step for 150,000 spectra takes less than 30~minutes on a small 
+%research cluster in Cambridge.
 
@@
 
+% ARC asks: how did you chose the initializations?
+
 Another issue with k-means is that it effectively uses metric
@@
 Indeed, it is a feature of this work that even the simplest, most
-generic, least tuned clustering algorithm returns interesting
+generic, least-tuned clustering algorithm returns interesting
 structures (as shown below).

### 2016-01-18 papers/clusters.tex
Some intro text for #6 (can be expanded,.. to infinity really)

@@
 Six two-dimensional projections of the abundance data, plus some other
-data quantities, are shown in \figurename~\ref{fig:all}.
+data quantities, are shown in \figurename~\ref{fig:all}.  Throughout
+this work we show two-dimensional abundance projections for only a few
+elements. The entire sequence of 15-dimensional combinations is too
+immense to visualize. The projections shown are ones which
+have been extensively used in abundance studies of globular clusters
+and satellite systems. These are typically light element abundances,
+and our projections include C--N, Na--O, and Mg--Al. Globular clusters
+famously demonstrate correlations in these elements, thereby making
+them suitable projections for us to show in verifying and examining
+any substructure identified by k-means (or any other algorithm).
+
+% ARC asks DWH: Why aren't we showing The Cannon Teff/logg in these
+% 				figures? Even if they aren't used for clustering..
 

### 2016-01-18 papers/clusters.tex
A question for Hogg, and minor char changes

@@
 In order to determine the precise abundance labels for chemical
-tagging, these surveys are each measuring high resolution, high
-signal-to-noise spectra for hundreds of thousands of stars across the
+tagging, these surveys are each measuring high-resolution, high
+signal-to-noise (S/N) spectra for hundreds of thousands of stars across the
 Galaxy's disk, bulge and halo.
@@
 Although undeniably promising---and motivating the 
-launch of costly large-scale spectroscopic surveys---%
+launch of costly large-scale spectroscopic surveys---
 chemical-tagging as a search technique has yet to be proven in
@@
 wavelength coverage, and sample sizes to deliver many different chemical tags for each star.
+
+% ARC to DWH:   "stand out in a high-dimensional abundance space" -->
+%		 	    They'll only stand out in a high-dimensional abundance
+% 				space, or a low-dimensional abundance space where those
+% 				dimensions are all very orthogonal (e.g., different
+% 				nucleosynthetic pathways). Should we add anything to that
+%				effect?
 

### 2016-01-18 papers/clusters.tex
Minor intro changes

@@
 have the right dependencies on atmospheric parameters such that the
-wrongness in abundance space is consistent across the H-R diagram.
+wrongness in abundance space is consistent across the Hertzsprung-Russell diagram.
 That is, we need a substantial amount of accuracy to achieve our goals.
@@
 simplifications of photospheric structure, presumptions about convective
-motions, and inconsistencies resulting from assumptions of the local thermal
+motion, and inconsistencies resulting from assumptions of the local thermal
 equilibrium, all contribute to produce knowingly inaccurate abundances.
@@
 we will have resolved all practical outstanding issues plaguing 
-chemical tagging, thereby enabling tantalizing potential to truly
-unravel the formation of the Milky Way.
+chemical tagging, thereby coming closer to the tantalizing potential 
+in truly unravelling the formation of the Milky Way.
 

### 2016-01-18 papers/clusters.tex
SNR-> S/N to match in-text description. Code details at end of ack

@@
 deliver stellar abundance labels of higher precision, especially for
-stars of SNR$\le 150$.
+stars of $S/N \le 150$.
 The details about how we select, reduce, and analyze the \apogee\ data
@@
 For the training step of \thecannon, 
-12,681 red-giant stars with spectral SNR$\ge 200$ were used.
-For this small fraction of the sample with the highest SNR objects the
+12,681 red-giant stars with spectral $S/N \ge 200$ were used.
+For this small fraction of the sample with the highest $S/N$ objects the
 labels provided by \aspcap\ provide consistent, relatively low-scatter, sensible
@@
 
-All of the code for the data manipulation, clustering, and
-visualization is built on \project{numpy} \citep{numpy},
-\project{scikit-learn} \citep{sklearn}, and \project{matplotlib}
-\citep{matplotlib}.
-All of the code written specifically for this project is available in
-two open-source code repositories at
-\url{https://github.com/andycasey/AnniesLasso} and
-\url{https://github.com/davidwhogg/Platypus}.
-
 \begin{figure}[!p]
@@
 
+All of the code for the data manipulation, clustering, and
+visualization is built on \project{numpy} \citep{numpy},
+\project{scikit-learn} \citep{sklearn}, and \project{matplotlib}
+\citep{matplotlib}.
+All of the code written specifically for this project is available in
+two open-source code repositories at
+\url{https://github.com/andycasey/AnniesLasso} and
+\url{https://github.com/davidwhogg/Platypus}.
+
 \clearpage

### 2016-01-18 papers/clusters.tex
Added some stubs for me to fill in with the discussion

@@
 
+ARC: Which known clusters were detected? Which comparably-sized clusters
+	were not detected? Why not? (Why not M 15?)
+
+ARC: How do the abundances that we find in these known clusters compare
+	with the literature?
+
+ARC: Expand on why these particular abundance projections, with
+	comparisons to the literature.
+
+ARC: We should note that the anti-correlations known to exist in GCs
+	actually make this a *HARDER* task, since the abundances are
+	somewhat arbitrarily structured (rather than normally concentrated)
+	in a multi-dimensional space.
+
+ARC: What can be said about the M 53 stars? Since we find GCs which are
+	mostly spatially concentrated (e.g., very few sitting elsewhere),
+	and then we find M53 "co-cluster" stars all over the halo, does this
+	directly speak for evidence of accretion by proto-stellar clusters
+	in the high-redshift analogues of the Milky Way?
+
+ARC: In addition to the increased precision, it is very important to
+	mention the fact that we report an "abundance" for *everything*.
+	Gaps in abundance data from classical spectroscopy are rampant,
+	and pose a serious problem even if you want to perform Real World(tm)
+	tests of chemical tagging. \thecannon\ has solved this entire
+	subset of astronomy.
+
 There is no sense at all in which \figurename~\ref{fig:M13},

### 2016-01-18 papers/clusters.tex
s/n always math mode

@@
 tagging, these surveys are each measuring high-resolution, high
-signal-to-noise (S/N) spectra for hundreds of thousands of stars across the
+signal-to-noise ($S/N$) spectra for hundreds of thousands of stars across the
 Galaxy's disk, bulge and halo.

### 2016-01-18 code/kmeans.py
plotting improvements, I hope

@@ -84,12 +84,19 @@ def plot_one_cluster(data, labels, mask, name, dir):
                 fig = plt.figure(figsize=(4,4))
                 plt.subplots_adjust(left=0.2, right=0.96, bottom=0.2, top=0.96)
             plt.clf()
-            kwargs = {"marker": ".", "ls": "none"}
+            kwargs = {"ls": "none", "marker": "."}
             if mask is None:
-                plt.plot(data[:,x], data[:,y], c="k", ms=0.5, alpha=0.20, **kwargs)
+                plt.plot(data[:,x], data[:,y], c="k", mec="none", ms=0.5, alpha=0.20, **kwargs)
             else:        
-                plt.plot(data[:,x], data[:,y], c="0.75", ms=0.5, alpha=0.20, **kwargs)
-                plt.plot(data[mask,x], data[mask,y], c="k", ms=5.0, alpha=0.5, **kwargs)
+                plt.plot(data[:,x], data[:,y], c="0.75", mec="none", ms=0.5, alpha=0.20, **kwargs)
+                angle = 0.
+                thisb = 0.025 # magic
+                cc = plt.get_cmap('brg')
+                for i in np.arange(len(data))[mask]:
+                    plt.plot([data[i,x], ], [data[i,y], ], c=cc(thisb), alpha=0.75,
+                             marker=(3, 1, angle), ms=7.0, ls="none")
+                    angle = (angle + 55.) % 360. # magic
+                    thisb = (thisb + 0.05) % 1.0 # magic
             plt.ylim(range_dict[ly])
             plt.xlim(range_dict[lx])
             plt.ylabel(label_dict[ly])
@@ -97,7 +104,27 @@ def plot_one_cluster(data, labels, mask, name, dir):
             [l.set_rotation(45) for l in plt.gca().get_xticklabels()]
             [l.set_rotation(45) for l in plt.gca().get_yticklabels()]
             plt.savefig(fn)
-            print("plot_one_cluster(): plotting", fn)
+            print("plot_one_cluster(): saving", fn)
+
+def plot_cluster_stats(sizes, densities, dir):
+    K = len(sizes)
+    print("plot_one_cluster(): plotting", K)
+    suffix = "png"
+    fn = "{}/clusters_{:04d}.{}".format(dir, K, suffix)
+    if os.path.exists(fn):
+        print("plot_one_cluster(): {} exists already".format(fn))
+        return None
+    assert len(densities) == K
[…]

### 2016-01-18 papers/clusters.tex
figure changes

@@
 \caption{Same as \figurename~\ref{fig:all} except that the entire
-  sample has been made gray and the members of one of the most dense
-  $K=256$ abundance-space clusters has been made black (and prominent).
+  sample has been made gray and the members of the most dense
+  $K=256$ abundance-space clusters have been rendered as unique, prominent triangles.
   This cluster, which was identified only in abundance space (six

### 2016-01-18 code/kmeans.py
spacing

@@ -86,7 +86,7 @@ def plot_one_cluster(data, labels, mask, name, dir):
             plt.clf()
             kwargs = {"ls": "none", "marker": "."}
             if mask is None:
-                plt.plot(data[:,x], data[:,y], c="k", mec="none", ms=0.5, alpha=0.20, **kwargs)
+                plt.plot(data[:,x], data[:,y], c="k",    mec="none", ms=0.5, alpha=0.20, **kwargs)
             else:        
                 plt.plot(data[:,x], data[:,y], c="0.75", mec="none", ms=0.5, alpha=0.20, **kwargs)
                 angle = 0.

### 2016-01-18 papers/clusters.tex
notes to self
notes to self, figure

@@
 % - write discussion section w MKN input
-% - expand kmeans discussion
-% - put in references
-% - cite codes!
 % - fix all remaining occurrences of ARC, MKN, HWR, DWH, or CITE
-% - forward to Freeman, others?
 % - submit to journal and arXiv
+% - forward to Freeman, Bovy, others?
 
@@
 % results shit
-\newcommand{\totalnumber}{98462} %% DWH CHECK
+\newcommand{\totalnumber}{98462}
 
@@
 
-% ARC asks DWH: Why aren't we showing The Cannon Teff/logg in these
-% 				figures? Even if they aren't used for clustering..
+ARC asks DWH: Why aren't we showing The Cannon Teff/logg in these
+				figures? Even if they aren't used for clustering..
 
@@
 [K/H], [Ca/H], [Ti/H], [V/H], [Mn/H], [Fe/H], [Ni/H].
-But in addition to this, we re-scale these by approximate measurement
+But in addition to this, we re-scale these by approximate typical measurement
 precisions obtained by \thecannon\ before running k-means.
@@
 to lowest) for visual inspection.
-We chose a few interesting cases from many and show them in
+In \figurename~\ref{fig:densities} we show the distribution of cluster membership
+and density for the $K=256$ run of k-means.
+
+We chose a few interesting cases from the high-density clusters in the $K=256$ run and show them in
 \figurename~\ref{fig:M13}, \ref{fig:Pal5}, \ref{fig:Sgr},
 \ref{fig:halo}, and \ref{fig:disk}.
-The first three of these are known clusters M13, Pal5, and Sagittarius (MKN, ARC, help!).
-The fourth is a halo structure with high velocity dispersion and possibly accreted.
+The first three of these are dominated by stars in the known clusters M13, Pal5, and Sagittarius.
+The fourth is a halo structure with high velocitay dispersion and possibly accreted.
 The fifth is a thin-disk star-formation feature.
@@
 
-DWH \& HWR: How do we show that we aren't just getting clusters because \apogee\ has plate-to-plate calibration variations?
+DWH: The clusters shown in \figurename~\ref{fig:M13}, \ref{fig:Pal5},
+and \ref{fig:Sgr} might not look dense in abundance space, but they
+are very dense in the 15-dimensional space.  Indeed,
+\figurename~\ref{fig:M13} is the densest cluster found in the 15-d
+space \emph{by far} (see \figurename~\ref{fig:densities}).
 
+DWH: The figures show results only from the $K=256$ run, why?
+
+DWH: Even if $K$ was chosen very carefully, it would still break up
+real clusters into many clusters, why?  Abundance spreads!
+
+DWH: The algorithm sees many different clusters; the figures show only
+a few.  Do we see all clusters we \emph{should} see?  That question is
+beyond the scope of this \documentname.
+
+DWH: Each cluster figure and the disk figure shows stars that are part
+of the structure and some that are not.  Every cluster will have
+background contamination, why?
+
+\begin{figure}[!p]
+\includegraphics[width=2\figurewidth]{./figs/clusters_0256.pdf}
+\caption{The distribution of membership and density for the 256
+  abundance-space clusters returned by the k-means algorithm at
+  $K=256$.  The densest cluster---which is shown in
+  \figurename~\ref{fig:M13}---is far more dense than the second
+  densest.\label{fig:densities}}
+\end{figure}
 \begin{figure}[!p]

### 2016-01-18 papers/clusters.tex
typo
caption
kmeans text expansion badge unlocked

@@
 If there are thousands of (relevant) molecular clouds forming stars in
-the recent history of the Milky Way, clumps of stars can only stand
-out in a high-dimensional abundance space: therefore accurate, or at
-least precise, measurements of many different abundances are needed
+the recent history of the Milky Way, clumps of stars can only be
+identified in abundance space if abundance space is high in dimensionality.
+(In principle, it needs to be high in dimensionality both in terms
+of the number of measured abundances \emph{and} in terms of the number of
+neucleosynthetic pathways or the dimensionality of the true abundance
+space.)
+Therefore accurate---or at
+least precise---measurements of many different abundances are needed
 for stellar sibling to have sufficiently unique fingerprints.
-Stellar spectroscopic surveys now have the resolution, signal-to-noise,
-wavelength coverage, and sample sizes to deliver many different chemical tags for each star.
-
-% ARC to DWH:   "stand out in a high-dimensional abundance space" -->
-%		 	    They'll only stand out in a high-dimensional abundance
-% 				space, or a low-dimensional abundance space where those
-% 				dimensions are all very orthogonal (e.g., different
-% 				nucleosynthetic pathways). Should we add anything to that
-%				effect?
+Stellar spectroscopic surveys now have the resolution,
+signal-to-noise, wavelength coverage, and sample sizes to deliver many
+different chemical tags for each star.
 
@@
 
-ARC asks DWH: Why aren't we showing The Cannon Teff/logg in these
-				figures? Even if they aren't used for clustering..
-
 \begin{figure}[!p]
@@
 preserve the best clustering (best according to the k-means score).
-
-% ARC asks: how did you chose the initializations?
+Each of the 32 initializations was performed with the
+\project{scikit-learn} standard k-means initialization procedure
+\citep{sklearn}.
 
@@
 algorithm.
-
-DWH: EXPLICITLY GIVE THE ABUNDANCE-SPACE SCALES USED.
 
@@
 and density for the $K=256$ run of k-means.
+The densest clusters tend to have fewer members, presumably because
+the bulk of the population is not extremely strongly clustered, or not
+in ways to which k-means is sensitive.
 
-We chose a few interesting cases from the high-density clusters in the $K=256$ run and show them in
+Importantly, the k-means algorithm assigns \emph{every} star to at
+least one cluster.
+For this reason, there is no sense in which each ``cluster'' returned
+by k-means is in any sense an overdensity in abundance space.
+In what follows, we only consider high-density clusters---clusters
+that are much more dense than average; these ought to represent true
+over-densities in abundance space.
+
+We chose a few interesting cases from the high-density clusters
+in the $K=256$ run and show them in
 \figurename~\ref{fig:M13}, \ref{fig:Pal5}, \ref{fig:Sgr},
 \ref{fig:halo}, and \ref{fig:disk}.
-The first three of these are dominated by stars in the known clusters M13, Pal5, and Sagittarius.
-The fourth is a halo structure with high velocitay dispersion and possibly accreted.
+The first three of these are dominated by stars in the known clusters
+M13, Pal5, and Sagittarius.
+The fourth is a halo structure with high velocity dispersion and
+possibly accreted.
 The fifth is a thin-disk star-formation feature.
-We will discuss these more below.
+We will discuss the astrophysical implications of these results in
+\sectionname~\ref{sec:discussion}.
 
-DWH: The clusters shown in \figurename~\ref{fig:M13}, \ref{fig:Pal5},
-and \ref{fig:Sgr} might not look dense in abundance space, but they
+The abundance-space clusters shown in the \figurename s
+might not look dense in abundance space, but they
 are very dense in the 15-dimensional space.  Indeed,
@@
 space \emph{by far} (see \figurename~\ref{fig:densities}).
+The challenge of this work is to find structure in the high
+dimensionality space that is not visible in any 2-d projection.
+Although we have by no means any model of the 15-d space, we do know
+that the structures shown in the figures are very high in 15-d
+density.
 
-DWH: The figures show results only from the $K=256$ run, why?
+We have only shown results from the $K=256$ run of k-means.
+This run was chosen because its densest abundance-space clusters map
+well onto known stellar clusters.
+At $K=128$ the densest abundance-space clusters tended to combine
+multiple known stellar clusters into single, large abundance-space
+groupings.
+At $K=512$ k-means tended to split even mono-abundance groups into
+smaller sub-groups.
+This is a reminder that k-means has been chosen just for simplicity
+and reliability here; it is by no means matched to the discovery of
+stellar structures.
+An important follow-up project is to build a model of abundance space
+that captures the features expected from stellar populations.
+Even if we were able to tune k-means in some sense ``perfectly'', it
+would still break up many stellar clusters, since they often show
+multiple stellar populations.
+It is also the case that each of the abundance-space clusters we
+\emph{do} show in the \figurename s includes both star-cluster members
+and some background contamination, and also is missing some true
+star-cluster members.
 
-DWH: Even if $K$ was chosen very carefully, it would still break up
-real clusters into many clusters, why?  Abundance spreads!
-
-DWH: The algorithm sees many different clusters; the figures show only
-a few.  Do we see all clusters we \emph{should} see?  That question is
-beyond the scope of this \documentname.
-
-DWH: Each cluster figure and the disk figure shows stars that are part
-of the structure and some that are not.  Every cluster will have
-background contamination, why?
+Although we show only three stellar clusters in
+\figurename~\ref{fig:M13}, \ref{fig:Pal5}, and \ref{fig:Sgr}, many
+other clusters are visible among our densest abundance-space k-means
+clusters.
+These include M15, M92, and M107, among others.
+We haven't asked yet whether we detect---as abundance-space
+overdensities---all of the stellar clusters we expect to find.
+This \documentname\ is simply a demonstration of the chemical-tagging
+concept; an investigation of whether we can construct complete
+catalogs of stellar clusters is beyond the scope.
 
 \begin{figure}[!p]
-\includegraphics[width=2\figurewidth]{./figs/clusters_0256.pdf}
+\includegraphics[width=2\figwidth]{./figs/clusters_0256.png}
 \caption{The distribution of membership and density for the 256
   abundance-space clusters returned by the k-means algorithm at
-  $K=256$.  The densest cluster---which is shown in
-  \figurename~\ref{fig:M13}---is far more dense than the second
+  $K=256$.  The densest cluster (which is displayed in more detail in 
+  \figurename~\ref{fig:M13}) is far more dense than the second
   densest.\label{fig:densities}}

### 2016-01-18 papers/clusters.tex
reference tweaks

@@
 Garc{\'{\i}}a~P{\'e}rez,~A.~E., Allende~Prieto,~C., Holtzman,~J.~A., \etal, 2015, arXiv:1510.07635
-\bibitem[Gilmore et al.(2012)]{gaiaeso} Gilmore, G., Randich, 
-S., Asplund, M., et al.\ 2012, The Messenger, 147, 25 
+\bibitem[Gilmore \etal(2012)]{gaiaeso}
+Gilmore,~G., Randich,~S., Asplund,~M., \etal, 2012, The Messenger, 147, 25 
 \bibitem[Hunter(2007)]{matplotlib}
 Hunter,~J.~D., 2007, Computing in Science \& Engineering, 9, 90
-\bibitem[Jofr{\'e} \etal(2015)]{jofre} Jofr{\'e}, P., 
-M{\"a}dler, T., Gilmore, G., et al.\ 2015, \mnras, 453, 1428 
+\bibitem[Jofr{\'e} \etal(2015)]{jofre}
+Jofr{\'e},~P., M{\"a}dler,~T., Gilmore,~G., \etal, 2015, \mnras, 453, 1428 
 \bibitem[Majewski \etal(2015)]{apogee}

### 2016-01-19 papers/clusters.tex
MKN comments

@@
 % - never use \cite{}, always \citep{} or \citealt{}.
+% - never use et al., always use \etal
 
@@
 The simplest method for clustering points in $D$-dimensional space is
-the \emph{k-means} algorithm (see \citealt{bishop} for a pedagogical introduction).
+the \emph{k-means} algorithm (see \citealt{bishop} for a pedagogical
+introduction and references to the original literature).
 Briefly, the k-means algorithm is to find the points closest (in
@@
 This algorithm is fast and performs well in practice in problems of
-this nature.
+this nature; also we are not the first to use the k-means algorithm in
+abundance space \citep{gratton}.
 
@@
 \clearpage
-\section{Discussion}
+\section{Discussion}\label{sec:discussion}
 
@@
 associated physically in clusters are similar in chemical-abundance
-properties (with a recent model-free test by \citealt{bovy}).
-MKN: WHAT ELSE TO CITE HERE?
-However, it was not known whether abundance-space information would be
-precise enough, or \emph{measurable} precisely enough to find this
+properties.
+That is, although they often show chemical diversity, that chemical
+diversity is small, and follows well-defined trends or a subspace in
+abundance space (for example, \citealt{gratton, meszaros, bovy},
+and references within those).
+However, it was not known---prior to this work---whether abundance-space information would be
+informative enough, or \emph{measurable} precisely enough to find this
 structure in surveys dominated by a large mix of stars from different
@@
 
-ARC: Which known clusters were detected? Which comparably-sized clusters
-	were not detected? Why not? (Why not M 15?)
+ARC: In addition to the increased precision, it is very important to
+	mention the fact that we report an "abundance" for *everything*.
+	Gaps in abundance data from classical spectroscopy are rampant,
+	and pose a serious problem even if you want to perform Real World(tm)
+	tests of chemical tagging. \thecannon\ has solved this entire
+	subset of astronomy.  DWH say: I agree; moved this point earlier.
+
+%ARC: Which known clusters were detected? Which comparably-sized clusters
+%	were not detected? Why not? (Why not M 15?)  DWH say: M15 WAS detected, as the paper now notes above; I say drop this issue entirely.
 
@@
 	mostly spatially concentrated (e.g., very few sitting elsewhere),
-	and then we find M53 "co-cluster" stars all over the halo, does this
+	and then we find M53 ``co-cluster'' stars all over the halo, does this
 	directly speak for evidence of accretion by proto-stellar clusters
-	in the high-redshift analogues of the Milky Way?
+	in the high-redshift analogues of the Milky Way?  DWH say: If you use the name ``M53'' then define it or say where you see it.
 
-ARC: In addition to the increased precision, it is very important to
-	mention the fact that we report an "abundance" for *everything*.
-	Gaps in abundance data from classical spectroscopy are rampant,
-	and pose a serious problem even if you want to perform Real World(tm)
-	tests of chemical tagging. \thecannon\ has solved this entire
-	subset of astronomy.
+The abundance-space group of stars in the disk structure shown in
+\figurename~\ref{fig:disk} may be associated with a very thin
+Milky-Way bar component reported previously \citep{wegg}, which is
+found to exist predominantly toward the end of the bar (at $l$
+$\approx$ 30$^\circ$).
+The stars in this structure are metal-rich and have a low
+[C/N] ratio and are therefore likely young (following the [C/N]--mass
+relationship; \citealt{martig}).
+It is possible that this structure has very low scale height because
+it is very young and hasn't experienced dynamical heating or radial
+migration.
 
+It is worth reiterating here what is written above:
 There is no sense at all in which \figurename~\ref{fig:M13},
@@
 interpetable.
-There are many other things to be found in this data set!
-
-On the particulars of these features, a few comments:
-DWH: Comments about Sag and M13: They appear in multiple clusters.
-MKN: Comments about the young disk feature and what it probably is?
+There are many other things to be found in this data set, and much
+more effective models to be built of abundance-space structure.
+The only strong conclusion of this investigation is that chemical
+tagging \emph{can} work; the abundance measurements with \thecannon\ are
+precise enough and the chemical-abundance space is informative enough.
 
@@
 Gilmore,~G., Randich,~S., Asplund,~M., \etal, 2012, The Messenger, 147, 25 
+\bibitem[Gratton \etal(2012)]{gratton}
+Gratton,~R.~G., Carretta,~E., Bragaglia,~A., 2012, \aapr, 20, 50
 \bibitem[Hunter(2007)]{matplotlib}
@@
 Martell,~S.~L., 2015, arXiv:1507.00079
+\bibitem[Martig \etal(2016)]{martig}
+Martig,~M., Fouesneau,~M., Rix,~H.-W., \etal, 2016, \mnras, 456, 3655
 \bibitem[Melendez \etal(2014)]{melendez}
 Mel{\'e}ndez,~J., Ram{\'{\i}}rez,~I., Karakas,~A.~I., \etal, 2014, \apj, 791, 14
+\bibitem[M{\'e}sz{\'a}ros \etal(2015)]{meszaros}
+M{\'e}sz{\'a}ros,~S., Martell,~S.~L., Shetrone,~M., \etal, 2015, \aj, 149, 153
 \bibitem[Mitschang \etal(2014)]{mitschang}
@@
 van~der~Walt,~S., Colbert,~S.~C., Varoquaux,~G., 2011, Computing in Science \& Engineering, 13, 22
+\bibitem[Wegg \etal(2015)]{wegg}
+Wegg,~C., Gerhard,~O., Portail,~M., 2015, \mnras, 450, 4050
 \end{thebibliography}

### 2016-01-19 papers/clusters.tex
figuretweak
figure placement
Extremely important change...

@@
         Hans-Walter~Rix\altaffilmark{4},
-    and~Daniel~Foreman-Mackey\altaffilmark{6}}
+    and~Daniel~Foreman-Mackey\altaffilmark{6,7}}
 \altaffiltext{1}{Simons Center for Data Analysis, 160 Fifth Avenue, 7th Floor, New York, NY 10010, USA}
@@
 \altaffiltext{5}{Institute of Astronomy, University of Cambridge, Madingley Road, Cambridge CB3~0HA, UK}
-\altaffiltext{6}{Department of Astronomy, University of Washington, Box 351580, Seattle, WA 98195-1580}
+\altaffiltext{6}{Sagan Fellow}
+\altaffiltext{7}{Department of Astronomy, University of Washington, Box 351580, Seattle, WA 98195-1580}
 \email{david.hogg@nyu.edu}
@@
 
-\begin{figure}[!p]
-\includegraphics[width=2\figwidth]{./figs/clusters_0256.png}
+\begin{figure}[!bp]
+\includegraphics[width=1.5\figwidth]{./figs/clusters_0256.png}
 \caption{The distribution of membership and density for the 256

### 2016-01-19 code/kmeans.py
figure tweak; move to prepare for merge

@@ -115,8 +115,8 @@ def plot_cluster_stats(sizes, densities, dir):
         print("plot_one_cluster(): {} exists already".format(fn))
         return None
     assert len(densities) == K
-    fig = plt.figure(figsize=(4,4))
-    plt.subplots_adjust(left=0.2, right=0.96, bottom=0.2, top=0.96)
+    fig = plt.figure(figsize=(6,6))
+    plt.subplots_adjust(left=0.15, right=0.97, bottom=0.15, top=0.97)
     plt.clf()
     kwargs = {"marker": ".", "ls": "none"}
     plt.plot(sizes, densities, c="k", ms=5.0, alpha=0.5, **kwargs)

### 2016-01-19 papers/clusters.tex
etal
MKN name

@@
 \newcommand{\foreign}[1]{\textsl{#1}}
-\newcommand{\eg}{\foreign{e.\,g.}}
-\newcommand{\etal}{\foreign{et\,al.}}
+\newcommand{\etal}{\foreign{et~al.}}
 \newcommand{\documentname}{\textsl{Article}}
@@
         Andrew~R.~Casey\altaffilmark{5},
-        Melissa~K.~Ness\altaffilmark{4},
+        Melissa~Ness\altaffilmark{4}, % typeset as per MKN-DWH email conversation 2016-01-18
         Hans-Walter~Rix\altaffilmark{4},

### 2016-01-19 schmapers/clusters.tex
stuff for MKN

% This file is part of the Platypus project.
% Copyright 2016 the authors.

\documentclass[12pt, letterpaper, preprint]{aastex}

\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\apogee}{\acronym{\project{APOGEE}}}
\newcommand{\thecannon}{\project{The~Cannon}}

% figure shit
\newlength{\figwidth}\setlength{\figwidth}{0.3\textwidth}
\newcommand{\insanefigure}[1]{%
\includegraphics[width=\figwidth]{./figs/#1_MgFe.png}%
\includegraphics[width=\figwidth]{./figs/#1_MgK.png}%
\includegraphics[width=\figwidth]{./figs/#1_AlMg.png}\\
\includegraphics[width=\figwidth]{./figs/#1_NaO.png}%
\includegraphics[width=\figwidth]{./figs/#1_CN.png}%
\includegraphics[width=\figwidth]{./figs/#1_SAl.png}\\
\includegraphics[width=2\figwidth]{./figs/#1_gal.png}%
\includegraphics[width=\figwidth]{./figs/#1_HR.png}\\
\includegraphics[width=2\figwidth]{./figs/#1_vlon.png}}

\begin{document}

\title{Identifying stellar clusters in chemical-abundance space}
\author{DWH, MKN, AC, HWR, DFM, others}

\begin{abstract}
% Context:
There is a dusty, old, and un-realized idea (dubbed ``chemical tagging'')
that every star ought to carry---%
in its detailed chemical abundances---%
an informative signature of its molecular-cloud birth location and time.
This idea is pretty, but has never yielded scientific fruit.
While astronomers have become more jaded, we have quietly been making
spectroscopic measurements of chemical abundances far more precise with \thecannon\ 
(now delivering 0.03-dex-level precision).
% Aims:
The chemical-tagging hypothesis is tested by looking at clusters in abundance space
and confirming that they are clustered in phase space.
% Methods:
We identify (by eye) overdensities of stars observed by the \apogee\ spectroscopic survey
in a 15-dimensional chemical-abundance space delivered by \thecannon,
and plot their locations in phase-space.
We use \emph{only} abundance-space information (no positional information) to identify stellar groups.
We don't know full 6-d phase-space positions;
we only have precise angular positions and radial velocities, and imprecise distances.
% Results:
We find that clusters in abundance space are indeed clusters in position space.
This confirms the chemical-tagging hypothesis and
verifies the precision of abundances delivered by \thecannon.
This is the first-ever project to identify stellar clusters by blind search in abundance space.
\end{abstract}

\section{Introduction}

DWH: Chemical tagging yada yada (CITE).

DWH: Open clusters look one-metal (CITE); globulars look peculiar (CITE); but no-one has ever gone the other way!

DWH: Fast intro to \apogee.

DWH: Fast intro to \thecannon.

DWH: BASH the \apogee\ discontiguous field positioning.

\section{Data and method}

MKN: Details about \apogee\ data.

AC: Details about \thecannon\ and implementation.

DWH: Figure showing the full data set; discuss structure.

DWH: Clustering methodology.

\section{Results and discussion}

DWH: For the first time, we see small-scale informative features in abundance space.

DWH: When we identify the small-scale features in abundance space, they are informative in position space.

DWH: How do we show that we aren't just getting clusters because \apogee\ has plate-to-plate calibration variations?

DWH: Chemical tagging will work.

\acknowledgements
DWH: It is a pleasure to thank...

DWH: github repo and hash.

DWH: Be sure to cite software a la the new ApJ guidelines.

\begin{figure}[!p]
\insanefigure{noncluster_0256_0080}
\caption{The median-density $K=256$ cluster.\label{fig:median}}
\end{figure}
\begin{figure}[!p]
\insanefigure{noncluster_0256_0196}
\caption{The lowest density $K=256$ cluster.\label{fig:lowest}}
\end{figure}

\end{document}

### 2016-01-19 code/kmeans.py

@@ -252,7 +252,7 @@ if __name__ == "__main__":
         print(K, "density range", np.min(densities), np.median(densities), np.max(densities))
         plot_cluster_stats(sizes, densities, dir)
 
-    # make plotting data
+        # make plotting data
         if plotdata is None:
             plotdata = data.copy()
             plotdata_labels = data_labels.copy()
@@ -266,7 +266,7 @@ if __name__ == "__main__":
             plotdata = np.hstack((plotdata, metadata))
             plotdata_labels = plotdata_labels + metadata_labels
 
-        # plot clusters from the last K run
+        # plot clusters from this K
         plotcount = 0
         for k in (np.argsort(densities))[::-1]:
             clustername = "cluster_{:04d}_{:04d}".format(K, k)
@@ -275,6 +275,14 @@ if __name__ == "__main__":
             plotcount += 1
             if plotcount == 32:
                 break
+
+        # plot some non-clusters for Ness
+        kmed = (np.argsort(densities))[K/2]
+        klow = (np.argsort(densities))[0]
+        for k in (kmed, klow):
+            clustername = "noncluster_{:04d}_{:04d}".format(K, k)
+            print(clustername, sizes[k], logdets[k], densities[k], fields[(clusters == k)].T)
+            plot_one_cluster(plotdata, plotdata_labels, (clusters==k), clustername, dir)
 
     # summary plots
     plot_one_cluster(plotdata, plotdata_labels, None, "all", dir)

### 2016-01-19 code/kmeans.py
plotting cluster context for HWR

@@ -1,9 +1,6 @@
 """
 This project is part of the Platypus project.
 Copyright 2016 David W. Hogg (NYU).
-
-## bugs:
-- needs better labels on corner
 """
 import os
 import pickle as cp
@@ -54,9 +51,8 @@ def get_metadata(fn, labels, mask):
     print("get_metadata()", dfn, metadata.shape)
     return metadata
 
-def plot_one_cluster(data, labels, mask, name, dir):
+def plot_one_cluster(data, labels, mask, name, dir, suffix="png"):
     print("plot_one_cluster(): plotting", name)
-    suffix = "png"
     for ly, lx, plotname in [
         ("MG_H - FE_H", "FE_H", "MgFe"),
         ("NA_H - FE_H", "O_H - FE_H", "NaO"),
@@ -104,15 +100,35 @@ def plot_one_cluster(data, labels, mask, name, dir):
             [l.set_rotation(45) for l in plt.gca().get_xticklabels()]
             [l.set_rotation(45) for l in plt.gca().get_yticklabels()]
             plt.savefig(fn)
-            print("plot_one_cluster(): saving", fn)
+            print("plot_one_cluster(): wrote", fn)
 
-def plot_cluster_stats(sizes, densities, dir):
+def plot_cluster_context(sizes, densities, k, name, dir, suffix="png"):
+    print("plot_cluster_context(): plotting", name)
+    fn = "{}/{}_context.{}".format(dir, name, suffix)
+    if os.path.exists(fn):
+        print("plot_cluster_context(): {} exists already".format(fn))
+        return
+    fig = plt.figure(figsize=(4,4))
+    plt.subplots_adjust(left=0.2, right=0.96, bottom=0.2, top=0.96)
+    plt.clf()
+    if k is None:
+        plt.savefig(fn)
+        print("plot_cluster_context(): wrote", fn)
+        return
+    plt.plot(sizes, densities, marker=".", c="k", ms=5.0, alpha=0.5, ls="none")
+    plt.plot(sizes[k], densities[k], marker="o", c="k", mfc="none", ms=5.0, alpha=1., ls="none")
+    plt.ylabel("cluster density")
+    plt.xlabel("number of stars in cluster")
+    plt.loglog()
+    plt.savefig(fn)
+    print("plot_cluster_context(): wrote", fn)
+
+def plot_cluster_stats(sizes, densities, dir, suffix="png"):
     K = len(sizes)
-    print("plot_one_cluster(): plotting", K)
-    suffix = "png"
[…]

### 2016-01-19 papers/clusters.tex
Typos and washing Hogg's mouth with soap

@@
 
-% text shit
+% text
 \newcommand{\acronym}[1]{{\small{#1}}}
@@
 
-% math shit
+% math
 \newcommand{\teff}{T_{\mathrm{eff}}}
@@
 
-% figure shit
+% figure 
 \newlength{\figwidth}\setlength{\figwidth}{0.3\textwidth}
@@
 
-% results shit
+% results
 \newcommand{\totalnumber}{98462}
@@
         Andrew~R.~Casey\altaffilmark{5},
-        Melissa~Ness\altaffilmark{4}, % typeset as per MKN-DWH email conversation 2016-01-18
+        Melissa~Ness\altaffilmark{4},
         Hans-Walter~Rix\altaffilmark{4},
@@
 \altaffiltext{1}{Simons Center for Data Analysis, 160 Fifth Avenue, 7th Floor, New York, NY 10010, USA}
-\altaffiltext{2}{Center for Cosmology and Particle Physics, Department of Phyics,
+\altaffiltext{2}{Center for Cosmology and Particle Physics, Department of Physics,
   New York University, 4 Washington Pl., room 424, New York, NY, 10003, USA}

### 2016-01-19 papers/clusters.tex
Spectroscopists see <0.05 dex and think 'differential analysis', let's not let them think that for here

@@
 However, we have succeeded in substantially improving spectroscopic measurements with \thecannon\ 
-(now delivering individual abundances with precisions around 0.04~dex).
+(now delivering $\geq$15 individual abundances for $\sim10^5$ stars with precisions around 0.04~dex).
 % Aims:

### 2016-01-19 papers/clusters.tex
Typos, question for Hogg, and signal-to-noise --> /N$ post-definition

@@
 of the number of measured abundances \emph{and} in terms of the number of
-neucleosynthetic pathways or the dimensionality of the true abundance
+nucleosynthetic pathways or the dimensionality of the true abundance
 space.)
@@
 Stellar spectroscopic surveys now have the resolution,
-signal-to-noise, wavelength coverage, and sample sizes to deliver many
+$S/N$, wavelength coverage, and sample sizes to deliver many
 different chemical tags for each star.
@@
 However, our recent work \citep{thecannon} suggests that \emph{data}
-are precise enough: there is enough signal-to-noise at the relevant
+are precise enough: there is enough $S/N$ at the relevant
 locations in spectrum space to deliver high-precision tags.
@@
 motion, and inconsistencies resulting from assumptions of the local thermal
-equilibrium, all contribute to produce knowingly inaccurate abundances.
+equilibrium all contribute to produce knowingly inaccurate abundances.
 \thecannon\ stands out because it demonstrably improves the precision on
@@
 
+% ARC asks: "not flagged as bad" --> what exactly do we mean here? ASPCAPFLAG = 0?
+
 Our analysis draws on the spectra of \totalnumber\ giant stars ($\logg
-< 3.9$) from \apogee\ \acronym{DR12} \citep{dr12}, alll those not
+< 3.9$) from \apogee\ \acronym{DR12} \citep{dr12}, all those not
 flagged as bad in the \apogee\ \aspcap\ \citep{aspcap} pipeline

### 2016-01-19 papers/clusters.tex
Minor edits to intro

@@
 known clusters do appear in the abundance space as over-densities.
-(In general, the collapse of 15 dimensions down to 2-d projections
+In general, the collapse of 15 dimensions down to 2-d projections
 will hide, smooth, or dilute any structure; there is no guarantee that
 a highly featured distribution in 15-d will show features in
-\emph{any} 2-d projection, let alone axis-aligned, human-selected projections.)
+\emph{any} 2-d projection, let alone axis-aligned, human-selected projections.
 This encourages us to look for over-densities automatically in the
@@
 
+% ARC TO DO: Make claim of the fourth being M53.
 We chose a few interesting cases from the high-density clusters
@@
 The challenge of this work is to find structure in the high
-dimensionality space that is not visible in any 2-d projection.
+dimensionality space that is not obviously visible in any 2-d projection.
 Although we have by no means any model of the 15-d space, we do know
@@
 would still break up many stellar clusters, since they often show
-multiple stellar populations.
+multiple stellar populations and non-linear correlations between elemental abundances.
 It is also the case that each of the abundance-space clusters we
@@
 We haven't asked yet whether we detect---as abundance-space
-overdensities---all of the stellar clusters we expect to find.
+overdensities---all of the stellar clusters we expect to find. (Many
+clusters within the \apogee\ data set only contain a few plausible 
+members, thereby  complicating any inferences we wish to make about 
+detection completeness using simple k-means.) 
 This \documentname\ is simply a demonstration of the chemical-tagging
@@
 catalogs of stellar clusters is beyond the scope.
+
+% ARC to DWH: Are the colors of broken glass indicative of anything?
 

### 2016-01-19 papers/clusters.tex
Expanded stub about k-means w.r.t. chemical abundance anti-correlations

@@
 
-ARC: In addition to the increased precision, it is very important to
-	mention the fact that we report an "abundance" for *everything*.
-	Gaps in abundance data from classical spectroscopy are rampant,
-	and pose a serious problem even if you want to perform Real World(tm)
-	tests of chemical tagging. \thecannon\ has solved this entire
-	subset of astronomy.  DWH say: I agree; moved this point earlier.
-
-%ARC: Which known clusters were detected? Which comparably-sized clusters
-%	were not detected? Why not? (Why not M 15?)  DWH say: M15 WAS detected, as the paper now notes above; I say drop this issue entirely.
+The abundances we find for clusters identified in k-space are consistent
+with detailed studies in the literature.
 
@@
 
-ARC: Expand on why these particular abundance projections, with
-	comparisons to the literature.
+% Overall abundances (mean [fE/h])
+% individual abundances
+% anti-correlations
+% ARC: Expand on why these particular abundance projections, with
+%	comparisons to the literature.
 
-ARC: We should note that the anti-correlations known to exist in GCs
-	actually make this a *HARDER* task, since the abundances are
-	somewhat arbitrarily structured (rather than normally concentrated)
-	in a multi-dimensional space.
+There are a few points to highlight about the abundance anti-correlations
+found in globular clusters. First, the detection of these signatures gives
+confirmation to our analysis: The training set was not purposefully selected
+to contain members of any globular or open cluster. However amongst the test
+set, we recover globular cluster members and their anti-correlations 
+in chemical abundances. That is to
+say we recover peculiar chemical abundances that are not dominant in our
+training set, but are expected from previous studies. Secondly, the very
+presence of complex abundance correlations makes the detection of clusters
+(by k-means) substantially more difficult. Briefly, the k-means algorithm
+is most effective for near-normally distributed clusters in high-dimensional space.
+While globular clusters often show distributions of this kind in a few dimensions, they also demonstrate non-linear correlations which would be
+sub-optimally selected by k-means. For these reasons the fact that (a)~\thecannon\
+reports these abundance correlations and (b)~k-means--as a simple algorithm without information about expectations in abundance space--does not appear to be
+severely impacted by these correlations, provides high promise for more sophisticated approaches.
 

### 2016-01-19 papers/clusters.tex
HWR figure request

--- 2016-01-19 papers/clusters.tex
@@
 % to-do
-% - write discussion section w MKN input
 % - fix all remaining occurrences of ARC, MKN, HWR, DWH, or CITE
-% - submit to journal and arXiv
+% - submit to arXiv
 % - forward to Freeman, Bovy, others?
+% - submit to journal
 
@@
 
-% text
+% text shit
 \newcommand{\acronym}[1]{{\small{#1}}}
@@
 
-% math
+% math shit
 \newcommand{\teff}{T_{\mathrm{eff}}}
@@
 
-% figure 
+% figure shit
 \newlength{\figwidth}\setlength{\figwidth}{0.3\textwidth}
@@
 \includegraphics[width=\figwidth]{./figs/#1_HR.png}\\
-\includegraphics[width=2\figwidth]{./figs/#1_vlon.png}}
+\includegraphics[width=2\figwidth]{./figs/#1_vlon.png}%
+\includegraphics[width=\figwidth]{./figs/#1_context.png}}
 
-% results
+% results shit
 \newcommand{\totalnumber}{98462}
@@
         Andrew~R.~Casey\altaffilmark{5},
-        Melissa~Ness\altaffilmark{4},
+        Melissa~Ness\altaffilmark{4}, % typeset as per MKN-DWH email conversation 2016-01-18
         Hans-Walter~Rix\altaffilmark{4},
@@
 \altaffiltext{1}{Simons Center for Data Analysis, 160 Fifth Avenue, 7th Floor, New York, NY 10010, USA}
-\altaffiltext{2}{Center for Cosmology and Particle Physics, Department of Physics,
+\altaffiltext{2}{Center for Cosmology and Particle Physics, Department of Phyics,
   New York University, 4 Washington Pl., room 424, New York, NY, 10003, USA}
@@
 However, we have succeeded in substantially improving spectroscopic measurements with \thecannon\ 
-(now delivering $\geq$15 individual abundances for $\sim10^5$ stars with precisions around 0.04~dex).
+(now delivering individual abundances with precisions around 0.04~dex).
 % Aims:
@@
 of the number of measured abundances \emph{and} in terms of the number of
-nucleosynthetic pathways or the dimensionality of the true abundance
+neucleosynthetic pathways or the dimensionality of the true abundance
 space.)
@@
 Stellar spectroscopic surveys now have the resolution,
-$S/N$, wavelength coverage, and sample sizes to deliver many
+signal-to-noise, wavelength coverage, and sample sizes to deliver many
 different chemical tags for each star.
@@
 However, our recent work \citep{thecannon} suggests that \emph{data}
-are precise enough: there is enough $S/N$ at the relevant
+are precise enough: there is enough signal-to-noise at the relevant
 locations in spectrum space to deliver high-precision tags.
@@
 motion, and inconsistencies resulting from assumptions of the local thermal
-equilibrium all contribute to produce knowingly inaccurate abundances.
+equilibrium, all contribute to produce knowingly inaccurate abundances.
 \thecannon\ stands out because it demonstrably improves the precision on
@@
 
-% ARC asks: "not flagged as bad" --> what exactly do we mean here? ASPCAPFLAG = 0?
-
 Our analysis draws on the spectra of \totalnumber\ giant stars ($\logg
-< 3.9$) from \apogee\ \acronym{DR12} \citep{dr12}, all those not
+< 3.9$) from \apogee\ \acronym{DR12} \citep{dr12}, alll those not
 flagged as bad in the \apogee\ \aspcap\ \citep{aspcap} pipeline
@@
   sample has been made gray and the members of the most dense
-  $K=256$ abundance-space clusters have been rendered as unique, prominent triangles.
+  $K=256$ abundance-space cluster have been rendered as unique, prominent triangles.
+  The lower-right plot shows this cluster (circle) in context of the other clusters (dots).
   This cluster, which was identified only in abundance space (six

### 2016-01-19 papers/clusters.tex

--- 2016-01-19 papers/clusters.tex
@@
 \includegraphics[width=\figwidth]{./figs/#1_HR.png}\\
-\includegraphics[width=2\figwidth]{./figs/#1_vlon.png}}
+\includegraphics[width=2\figwidth]{./figs/#1_vlon.png}%
+\includegraphics[width=\figwidth]{./figs/#1_context.png}}
 

### 2016-01-19 papers/clusters.tex
Some zeroth text and notes on M13

--- 2016-01-19 papers/clusters.tex
@@
 
-The abundances we find for clusters identified in k-space are consistent
-with detailed studies in the literature.
+The abundances we find for clusters identified in k-space are in excellent
+agreement with detailed studies that exist in the literature. The mean
+metallicity for stars we identify in abundance space that are later attributable
+to M13 have a mean of [Fe/H]$ = -1.55$, in agreement with other works \citep{Kraft_1992,Cohen_Melendez,Johnson_Pilachowski} and the \citet{Harris} catalog.
+The detailed abundances are also in agreement: we find a correlation in 
+[C/Fe] and [N/Fe] abundances, and an intrinsic spreads in each element
+that reflects deep mixing along the red giant branch.
+
+% M13
+% ANDY IS EDITING AND COLLECTING CRAP HERE.
+We have a smaller spread in [O/Fe] than Johnson & Pilachowski (http://iopscience.iop.org/article/10.1088/2041-8205/754/2/L38/pdf)
+Theirs is -1 to +1
+
+The mean of [Na/Fe] and [O/Fe] is consistent with J & P.
+
+%\bibitem[Johnson 
+%\& Pilachowski(2012)]{2012ApJ...754L..38J} Johnson, C.~I., \& Pilachowski, C.~A.\ 2012, \apjl, 754, L38
+
+Our [N/Fe] and [C/Fe] overlaps perfectly with Smith et al (2005).
+Our scatter in [C/Fe] is slightly less: this is because those with the lowest [C/Fe] ratios are the brightest cluster members, which are not in this cluster (e.g., lowest Teff is ~4500) -- did APGOEE observe them?
+
+%\bibitem[Smith et al.(2005)]{2005AJ....129.1589S} Smith, G.~H., Briley, 
+%M.~M., \& Harbeck, D.\ 2005, \aj, 129, 1589
+
+C-depletion wrt Teff (due to mixing) also seen in
+
+ \bibitem[Briley et al.(2004)]{2004AJ....127.1579B} Briley, M.~M., Cohen, 
+J.~G., \& Stetson, P.~B.\ 2004, \aj, 127, 1579
+
+% Abundances (Mg, Al, Na, O, Fe, and probably other projections) perfectly consistent with 
+%\bibitem[Cohen 
+%\& Mel{\'e}ndez(2005)]{2005AJ....129..303C} Cohen, J.~G., \& Mel{\'e}ndez, J.\ 2005, \aj, 129, 303
+
+and
+
+%\bibitem[Kraft et al.(1992)]{1992AJ....104..645K} Kraft, R.~P., Sneden, C., 
+%Langer, G.~E., \& Prosser, C.~F.\ 1992, \aj, 104, 645 
+% except perhaps we have fewer low [O/Fe] objects than Kraft et al.
+
+% Palomar 5
+
+% Sagittarius "star cloud"?
+% Fig 6: high-velocity-dispersion structure in the Galactic halo.
 

### 2016-01-19 papers/clusters.tex
Added abundance discussion for M13

--- 2016-01-19 papers/clusters.tex
@@
 
-The abundances we find for clusters identified in k-space are consistent
-with detailed studies in the literature.
-
-ARC: How do the abundances that we find in these known clusters compare
-	with the literature?
-
-% Overall abundances (mean [fE/h])
-% individual abundances
-% anti-correlations
-% ARC: Expand on why these particular abundance projections, with
-%	comparisons to the literature.
-
 There are a few points to highlight about the abundance anti-correlations
@@
 severely impacted by these correlations, provides high promise for more sophisticated approaches.
+
+The abundances of known clusters that we identify by k-means are in
+excellent agreement with the literature. For the group shown in Figure \ref{fig:M13},
+which we later attribute to M13, we find a mean of [Fe/H]$ = -1.55$ (DWH CHECK).
+The mean and spread agree well with other works dedicated to studying this cluster
+\citep{Kraft_1992,Cohen_Melendez_2005,Johnson_Pilachowski_2012}, and compiled catalogs of
+globular cluster properties \citep{Harris}. The detailed abundances are also
+consistent: we find a correlation in [C/Fe] and [N/Fe] abundances \citep{Smith_2005}, and their
+projection in Figure \ref{fig:M13} show a spread that reflects deep mixing
+along the red giant branch \citep{Briley_2004}. We find that the light element
+abundances (most notably Mg--Al) for this cluster are anti-correlated, as
+expected from other studies \citep[e.g.,][]{Gratton}. However we find a
+smaller spread in [O/Fe] than reported by others who have looked exclusively
+at M13 \citep{Johson_Pilachowski_2012}.  As we discuss above, this is expected:
+k-means is by no means optimal for identifying arbitrary-shaped structures
+in high-dimensional space, and the first stars that would be `discarded' from
+this group would be those with the most extreme abundances: low [O/Fe] and
+high [Na/Fe].
+
+% Palomar 5
+
+% Sagittarius "star cloud"?
+% Fig 6: high-velocity-dispersion structure in the Galactic halo.
+
+ARC: How do the abundances that we find in these known clusters compare
+	with the literature?
+
+% Overall abundances (mean [fE/h])
+% individual abundances
+% anti-correlations
+% ARC: Expand on why these particular abundance projections, with
+%	comparisons to the literature.
 
@@
 Bovy,~J., 2015, arXiv:1510.06745
+ \bibitem[Briley \etal(2004)]{Briley_2004} Briley, M.~M., Cohen, 
+J.~G., \& Stetson, P.~B.\ 2004, \aj, 127, 1579 
 \bibitem[Casey \etal(2016)]{casey16}
 Casey,~A.~R., Hogg,~D.~W., Ness,~M., Rix,~H.-W., 2016, in preparation
+\bibitem[Cohen \& Mel{\'e}ndez(2005)]{Cohen_Melendez_2005} Cohen, J.~G., \& Mel{\'e}ndez, J.\ 2005, \aj, 129, 303 
 \bibitem[De~Silva \etal(2015)]{galah}
@@
 Jofr{\'e},~P., M{\"a}dler,~T., Gilmore,~G., \etal, 2015, \mnras, 453, 1428 
+\bibitem[Johnson \& Pilachowski(2012)]{Johnson_Pilachowski_2012} Johnson, C.~I., \& Pilachowski, C.~A.\ 2012, \apjl, 754, L38 
+\bibitem[Kraft \etal(1992)]{Kraft_1992} Kraft, R.~P., Sneden, C., Langer, G.~E., \& Prosser, C.~F.\ 1992, \aj, 104, 645 
 \bibitem[Majewski \etal(2015)]{apogee}
@@
 Ro{\v s}kar,~R., Debattista,~V.~P., Quinn,~T.~R., Stinson,~G.~S., Wadsley,~J., 2008, \apjl, 684, L79
+\bibitem[Smith \etal(2005)]{Smith_2005} Smith, G.~H., Briley, 
+M.~M., \& Harbeck, D.\ 2005, \aj, 129, 1589 
 \bibitem[Ting \etal(2016)]{ting}

### 2016-01-19 papers/clusters.tex
Suggested new abstract opener

@@
 % Context:
-The chemical-tagging hypothesis is that every star ought to carry---%
-in its detailed chemical abundances---%
-an informative signature of its molecular-cloud birth location and time.
+Chemical tagging promises to robustly group co-natal, spatially extended stars by their present-day  photospheric chemical abundances.
 This idea has not previously yielded scientific successes, probably because of the
 noise and incompleteness in chemical-abundance measurements.
-However, we have succeeded in substantially improving spectroscopic measurements with \thecannon\ 
-(now delivering $\geq$15 individual abundances for $\sim10^5$ stars with precisions around 0.04~dex).
+However, we have succeeded in substantially improving spectroscopic measurements with \thecannon: 
+delivering $\geq$15 individual abundances for $\sim10^5$ stars with precisions around 0.04~dex.
 % Aims:

### 2016-01-19 papers/clusters.tex
Added some zero-th order discussion about Palomar 5 abundances (it looks like we are cranking it)

@@
 abundances (most notably Mg--Al) for this cluster are anti-correlated, as
-expected from other studies \citep[e.g.,][]{Gratton}. However we find a
+expected from other studies \citep[e.g.,][]{gratton}. However we find a
 smaller spread in [O/Fe] than reported by others who have looked exclusively
-at M13 \citep{Johson_Pilachowski_2012}.  As we discuss above, this is expected:
+at M13 \citep{Johnson_Pilachowski_2012}.  As we discuss above, this is expected:
 k-means is by no means optimal for identifying arbitrary-shaped structures
@@
 % Palomar 5
+There are scant abundance measurements reported for Palomar 5, for which
+most stars in the cluster in Figure \ref{fig:Pal5} belong. We find a mean
+metallicity of [Fe/H]$ = -1.3$ (DWH CHECK), which lies between the only
+photometric \citep{Geisler_1997} and spectroscopic \citep{Smith_1985,Smith_2002} abundance
+measurements for this cluster.  The anti-correlation in C--N and Na--O abundance
+have been seen elsewhere \citep[][, respectively]{Smith_1985,Smith_2002}, although
+to our knowledge the Mg--Al anti-correlation has never been shown in this cluster.
+Indeed, this work contains the largest number of abundances for Palomar 5,
+permitting a thorough search of tidally disrupted members \citep[e.g.,][]{Kuzma_2015} 
+that have retained the unique cluster abundance signature .
 
@@
 
-ARC: How do the abundances that we find in these known clusters compare
-	with the literature?
-
-% Overall abundances (mean [fE/h])
-% individual abundances
-% anti-correlations
 % ARC: Expand on why these particular abundance projections, with
@@
 Garc{\'{\i}}a~P{\'e}rez,~A.~E., Allende~Prieto,~C., Holtzman,~J.~A., \etal, 2015, arXiv:1510.07635
+\bibitem[Geisler \etal(1997)]{Geisler_1997} Geisler, D., Claria, 
+J.~J., \& Minniti, D.\ 1997, \pasp, 109, 799 
 \bibitem[Gilmore \etal(2012)]{gaiaeso}
@@
 \bibitem[Kraft \etal(1992)]{Kraft_1992} Kraft, R.~P., Sneden, C., Langer, G.~E., \& Prosser, C.~F.\ 1992, \aj, 104, 645 
+\bibitem[Kuzma \etal(2015)]{Kuzma_2015} Kuzma, P.~B., Da Costa, 
+G.~S., Keller, S.~C., \& Maunder, E.\ 2015, \mnras, 446, 3297 
 \bibitem[Majewski \etal(2015)]{apogee}
@@
 Ro{\v s}kar,~R., Debattista,~V.~P., Quinn,~T.~R., Stinson,~G.~S., Wadsley,~J., 2008, \apjl, 684, L79
+\bibitem[Smith(1985)]{Smith_1985} Smith, G.~H.\ 1985, \apj, 298, 
+249 
+\bibitem[Smith \etal(2002)]{Smith_2002} Smith, G.~H., Sneden, C., 
+\& Kraft, R.~P.\ 2002, \aj, 123, 1502 
 \bibitem[Smith \etal(2005)]{Smith_2005} Smith, G.~H., Briley, 

### 2016-01-19 papers/clusters.tex
Typos

@@
 measurements for this cluster.  The anti-correlation in C--N and Na--O abundance
-have been seen elsewhere \citep[][, respectively]{Smith_1985,Smith_2002}, although
-to our knowledge the Mg--Al anti-correlation has never been shown in this cluster.
+have been seen elsewhere \citep[][respectively]{Smith_1985,Smith_2002}, although
+to our knowledge the Mg--Al anti-correlation has not previously been shown for this cluster.
 Indeed, this work contains the largest number of abundances for Palomar 5,
 permitting a thorough search of tidally disrupted members \citep[e.g.,][]{Kuzma_2015} 
-that have retained the unique cluster abundance signature .
+that have retained the unique cluster abundance signature.
 

### 2016-01-19 papers/clusters.tex
A word
Add Harris citation, clarify 'Sagittarius star cloud'

@@
 % Context:
-Chemical tagging promises to robustly group co-natal, spatially extended stars by their present-day  photospheric chemical abundances.
+Chemical tagging promises to robustly group co-natal, spatially extended stars by their present-day  photospheric abundances.
 This idea has not previously yielded scientific successes, probably because of the
@@
   abundance-space cluster.
-  This one turns out to be dominated by the Sagittarius star
-  cloud.\label{fig:Sgr}}
+  This one turns out to be dominated by the Sagittarius dwarf spheroidal galaxy.\label{fig:Sgr}}
 \end{figure}
@@
 \citep{Kraft_1992,Cohen_Melendez_2005,Johnson_Pilachowski_2012}, and compiled catalogs of
-globular cluster properties \citep{Harris}. The detailed abundances are also
+globular cluster properties \citep[][accessed 2016]{Harris_1996}. The detailed abundances are also
 consistent: we find a correlation in [C/Fe] and [N/Fe] abundances \citep{Smith_2005}, and their
@@
 
-% Palomar 5
 There are scant abundance measurements reported for Palomar 5, for which
@@
 
-% Sagittarius "star cloud"?
-% Fig 6: high-velocity-dispersion structure in the Galactic halo.
-
-% ARC: Expand on why these particular abundance projections, with
-%	comparisons to the literature.
+% Sagittarius
 
@@
 Gratton,~R.~G., Carretta,~E., Bragaglia,~A., 2012, \aapr, 20, 50
+\bibitem[Harris(1996)]{Harris_1996} Harris, W.~E.\ 1996, \aj, 112, 
+1487 
 \bibitem[Hunter(2007)]{matplotlib}

### 2016-01-19 code/kmeans.py
hwr plots, triangles mean something

@@ -71,6 +71,8 @@ def plot_one_cluster(data, labels, mask, name, dir, suffix="png"):
         if os.path.exists(fn):
             print("plot_one_cluster(): {} exists already".format(fn))
         else:
+            logg = np.where(np.array(labels) == "LOGG_ASPCAP")[0][0]
+            rv = np.where(np.array(labels) == "VHELIO_AVG")[0][0]
             y = np.where(np.array(labels) == ly)[0][0]
             x = np.where(np.array(labels) == lx)[0][0]
             if plotname == "sky" or plotname == "gal" or plotname == "vlon":
@@ -83,16 +85,15 @@ def plot_one_cluster(data, labels, mask, name, dir, suffix="png"):
             kwargs = {"ls": "none", "marker": "."}
             if mask is None:
                 plt.plot(data[:,x], data[:,y], c="k",    mec="none", ms=0.5, alpha=0.20, **kwargs)
-            else:        
+            else:
                 plt.plot(data[:,x], data[:,y], c="0.75", mec="none", ms=0.5, alpha=0.20, **kwargs)
-                angle = 0.
-                thisb = 0.025 # magic
-                cc = plt.get_cmap('brg')
-                for i in np.arange(len(data))[mask]:
-                    plt.plot([data[i,x], ], [data[i,y], ], c=cc(thisb), alpha=0.75,
-                             marker=(3, 1, angle), ms=7.0, ls="none")
-                    angle = (angle + 55.) % 360. # magic
-                    thisb = (thisb + 0.05) % 1.0 # magic
+                cc = plt.get_cmap('viridis')
+                M = len(data[mask])
+                angles = (120. / M) * (data[mask, logg].argsort().argsort())
+                colors = [cc(q) for q in (1. / M) * (data[mask, rv].argsort().argsort())]
+                for j, i in enumerate(np.arange(len(data))[mask]):
+                    plt.plot([data[i,x], ], [data[i,y], ], c=colors[j], alpha=0.75,
+                             marker=(3, 1, angles[j]), ms=7.0, ls="none")
             plt.ylim(range_dict[ly])
             plt.xlim(range_dict[lx])
             plt.ylabel(label_dict[ly])
@@ -101,6 +102,10 @@ def plot_one_cluster(data, labels, mask, name, dir, suffix="png"):
[…]

### 2016-01-19 papers/clusters.tex
figure caption

--- 2016-01-19 papers/clusters.tex
@@
 % Context:
-Chemical tagging promises to robustly group co-natal, spatially extended stars by their present-day  photospheric chemical abundances.
+The chemical-tagging hypothesis is that every star ought to carry---%
+in its detailed chemical abundances---%
+an informative signature of its molecular-cloud birth location and time.
 This idea has not previously yielded scientific successes, probably because of the
 noise and incompleteness in chemical-abundance measurements.
-However, we have succeeded in substantially improving spectroscopic measurements with \thecannon: 
-delivering $\geq$15 individual abundances for $\sim10^5$ stars with precisions around 0.04~dex.
+However, we have succeeded in substantially improving spectroscopic measurements with \thecannon\ 
+(now delivering $\geq$15 individual abundances for $\sim10^5$ stars with precisions around 0.04~dex).
 % Aims:
@@
   sample has been made gray and the members of the most dense
-  $K=256$ abundance-space cluster have been rendered as unique, prominent triangles.
+  $K=256$ abundance-space cluster have been rendered as unique, prominent triangles
+  (color is velocity rank, orientation is $\logg$ rank).
   The lower-right plot shows this cluster (circle) in context of the other clusters (dots).
@@
 
+The abundances we find for clusters identified in k-space are in excellent
+agreement with detailed studies that exist in the literature. The mean
+metallicity for stars we identify in abundance space that are later attributable
+to M13 have a mean of [Fe/H]$ = -1.55$, in agreement with other works \citep{Kraft_1992,Cohen_Melendez,Johnson_Pilachowski} and the \citet{Harris} catalog.
+The detailed abundances are also in agreement: we find a correlation in 
+[C/Fe] and [N/Fe] abundances, and an intrinsic spreads in each element
+that reflects deep mixing along the red giant branch.
+
+% M13
+% ANDY IS EDITING AND COLLECTING CRAP HERE.
+We have a smaller spread in [O/Fe] than Johnson & Pilachowski (http://iopscience.iop.org/article/10.1088/2041-8205/754/2/L38/pdf)
+Theirs is -1 to +1
+
+The mean of [Na/Fe] and [O/Fe] is consistent with J & P.
+
+%\bibitem[Johnson 
+%\& Pilachowski(2012)]{2012ApJ...754L..38J} Johnson, C.~I., \& Pilachowski, C.~A.\ 2012, \apjl, 754, L38
+
+Our [N/Fe] and [C/Fe] overlaps perfectly with Smith et al (2005).
+Our scatter in [C/Fe] is slightly less: this is because those with the lowest [C/Fe] ratios are the brightest cluster members, which are not in this cluster (e.g., lowest Teff is ~4500) -- did APGOEE observe them?
+
+%\bibitem[Smith et al.(2005)]{2005AJ....129.1589S} Smith, G.~H., Briley, 
+%M.~M., \& Harbeck, D.\ 2005, \aj, 129, 1589
+
+C-depletion wrt Teff (due to mixing) also seen in
+
+ \bibitem[Briley et al.(2004)]{2004AJ....127.1579B} Briley, M.~M., Cohen, 
+J.~G., \& Stetson, P.~B.\ 2004, \aj, 127, 1579
+
+% Abundances (Mg, Al, Na, O, Fe, and probably other projections) perfectly consistent with 
+%\bibitem[Cohen 
+%\& Mel{\'e}ndez(2005)]{2005AJ....129..303C} Cohen, J.~G., \& Mel{\'e}ndez, J.\ 2005, \aj, 129, 303
+
+and
+
+%\bibitem[Kraft et al.(1992)]{1992AJ....104..645K} Kraft, R.~P., Sneden, C., 
+%Langer, G.~E., \& Prosser, C.~F.\ 1992, \aj, 104, 645 
+% except perhaps we have fewer low [O/Fe] objects than Kraft et al.
+
+% Palomar 5
+
+% Sagittarius "star cloud"?
+% Fig 6: high-velocity-dispersion structure in the Galactic halo.
+
+ARC: How do the abundances that we find in these known clusters compare
+	with the literature?
+
+% Overall abundances (mean [fE/h])
+% individual abundances
+% anti-correlations
+% ARC: Expand on why these particular abundance projections, with
+%	comparisons to the literature.
+
 There are a few points to highlight about the abundance anti-correlations
@@
 severely impacted by these correlations, provides high promise for more sophisticated approaches.
-
-The abundances of known clusters that we identify by k-means are in
-excellent agreement with the literature. For the group shown in Figure \ref{fig:M13},
-which we later attribute to M13, we find a mean of [Fe/H]$ = -1.55$ (DWH CHECK).
-The mean and spread agree well with other works dedicated to studying this cluster
-\citep{Kraft_1992,Cohen_Melendez_2005,Johnson_Pilachowski_2012}, and compiled catalogs of
-globular cluster properties \citep{Harris}. The detailed abundances are also
-consistent: we find a correlation in [C/Fe] and [N/Fe] abundances \citep{Smith_2005}, and their
-projection in Figure \ref{fig:M13} show a spread that reflects deep mixing
-along the red giant branch \citep{Briley_2004}. We find that the light element
-abundances (most notably Mg--Al) for this cluster are anti-correlated, as
-expected from other studies \citep[e.g.,][]{gratton}. However we find a
-smaller spread in [O/Fe] than reported by others who have looked exclusively
-at M13 \citep{Johnson_Pilachowski_2012}.  As we discuss above, this is expected:
-k-means is by no means optimal for identifying arbitrary-shaped structures
-in high-dimensional space, and the first stars that would be `discarded' from
-this group would be those with the most extreme abundances: low [O/Fe] and
-high [Na/Fe].
-
-% Palomar 5
-There are scant abundance measurements reported for Palomar 5, for which
-most stars in the cluster in Figure \ref{fig:Pal5} belong. We find a mean
-metallicity of [Fe/H]$ = -1.3$ (DWH CHECK), which lies between the only
-photometric \citep{Geisler_1997} and spectroscopic \citep{Smith_1985,Smith_2002} abundance
-measurements for this cluster.  The anti-correlation in C--N and Na--O abundance
-have been seen elsewhere \citep[][, respectively]{Smith_1985,Smith_2002}, although
-to our knowledge the Mg--Al anti-correlation has never been shown in this cluster.
-Indeed, this work contains the largest number of abundances for Palomar 5,
-permitting a thorough search of tidally disrupted members \citep[e.g.,][]{Kuzma_2015} 
-that have retained the unique cluster abundance signature .
-
-% Sagittarius "star cloud"?
-% Fig 6: high-velocity-dispersion structure in the Galactic halo.
-
-% ARC: Expand on why these particular abundance projections, with
-%	comparisons to the literature.
 
@@
 Bovy,~J., 2015, arXiv:1510.06745
- \bibitem[Briley \etal(2004)]{Briley_2004} Briley, M.~M., Cohen, 
-J.~G., \& Stetson, P.~B.\ 2004, \aj, 127, 1579 
 \bibitem[Casey \etal(2016)]{casey16}
 Casey,~A.~R., Hogg,~D.~W., Ness,~M., Rix,~H.-W., 2016, in preparation
-\bibitem[Cohen \& Mel{\'e}ndez(2005)]{Cohen_Melendez_2005} Cohen, J.~G., \& Mel{\'e}ndez, J.\ 2005, \aj, 129, 303 
 \bibitem[De~Silva \etal(2015)]{galah}
@@
 Garc{\'{\i}}a~P{\'e}rez,~A.~E., Allende~Prieto,~C., Holtzman,~J.~A., \etal, 2015, arXiv:1510.07635
-\bibitem[Geisler \etal(1997)]{Geisler_1997} Geisler, D., Claria, 
-J.~J., \& Minniti, D.\ 1997, \pasp, 109, 799 
 \bibitem[Gilmore \etal(2012)]{gaiaeso}
@@
 Jofr{\'e},~P., M{\"a}dler,~T., Gilmore,~G., \etal, 2015, \mnras, 453, 1428 
-\bibitem[Johnson \& Pilachowski(2012)]{Johnson_Pilachowski_2012} Johnson, C.~I., \& Pilachowski, C.~A.\ 2012, \apjl, 754, L38 
-\bibitem[Kraft \etal(1992)]{Kraft_1992} Kraft, R.~P., Sneden, C., Langer, G.~E., \& Prosser, C.~F.\ 1992, \aj, 104, 645 
-\bibitem[Kuzma \etal(2015)]{Kuzma_2015} Kuzma, P.~B., Da Costa, 
-G.~S., Keller, S.~C., \& Maunder, E.\ 2015, \mnras, 446, 3297 
 \bibitem[Majewski \etal(2015)]{apogee}
@@
 Ro{\v s}kar,~R., Debattista,~V.~P., Quinn,~T.~R., Stinson,~G.~S., Wadsley,~J., 2008, \apjl, 684, L79
-\bibitem[Smith(1985)]{Smith_1985} Smith, G.~H.\ 1985, \apj, 298, 
-249 
-\bibitem[Smith \etal(2002)]{Smith_2002} Smith, G.~H., Sneden, C., 
-\& Kraft, R.~P.\ 2002, \aj, 123, 1502 
-\bibitem[Smith \etal(2005)]{Smith_2005} Smith, G.~H., Briley, 
-M.~M., \& Harbeck, D.\ 2005, \aj, 129, 1589
 \bibitem[Ting \etal(2016)]{ting}

### 2016-01-19 papers/clusters.tex
Added wild speculation about the high-velocity-dispersion cluster

--- 2016-01-19 papers/clusters.tex
@@
 
-ARC: What can be said about the M 53 stars? Since we find GCs which are
-	mostly spatially concentrated (e.g., very few sitting elsewhere),
-	and then we find M53 ``co-cluster'' stars all over the halo, does this
-	directly speak for evidence of accretion by proto-stellar clusters
-	in the high-redshift analogues of the Milky Way?  DWH say: If you use the name ``M53'' then define it or say where you see it.
+The high-velocity-dispersion structure seen in Figure \ref{halo} stands
+in stark contrast to the rest of the clusters we have identified. In
+addition to being highly clustered in 15-dimensions, it is visibly
+clustered in our 2-d projections. Indeed, it is more clustered in our
+abundance projections than the young stellar structure in the disk,
+and comparably so to the other known globular clusters that we have
+shown here. The disparity between this structure and others presented
+here is the lack of co-spatial stars. 12 stars within this cluster are
+marked as `M53' candidates by \apogee, but the remaining 145 stars
+are spread amongst fields throughout the halo.
+
+% The high-velocity-dispersion structure in the Galaxy halo.
+We freely admit that k-means is by no means optimal, and other clusters
+we identify are accompanied by a few stars that are not currently co-spatial.
+Those stars may be unassociated interlopers that have been misclassified,
+or they may be true cluster members that are now unbound. However in the
+case of the high-velocity-dispersion structure in Figure \ref{fig:halo},
+the situation is far more extreme. We identify a group of stars with
+highly similar abundances in 15 dimensions that are now spread throughout
+the Galaxy. Because some stars are candidates of the massive globular
+cluster M53, it provides tantalizing possibility that these stars may
+indeed be all members of the same co-natal gas cloud, implying that 
+these stars have been accreted onto the halo by a single globular
+cluster early in the Milky Way's formation. While speculative (and we
+make no consideration to whether this is indeed the true scenario),
+this example demonstrates the ultimate goal of chemical tagging.
 

### 2016-01-19 papers/clusters.tex
rix-based edits

--- 2016-01-19 papers/clusters.tex
@@
 \clearpage
-\section{Clustering method and results}\label{sec:method}
+\section{Identifying abundance-space clusters}\label{sec:method}
 
 Although only hints of small-scale clustering in the abundance space
-are visible in \figurename~\ref{fig:all}, exploration of the data indicates that
-known clusters do appear in the abundance space as over-densities.
+are visible in \figurename~\ref{fig:all}, exploration of the data by hand indicates that
+known clusters do appear in the high-dimensional abundance space as over-densities.
 In general, the collapse of 15 dimensions down to 2-d projections
@@
 introduction and references to the original literature).
-Briefly, the k-means algorithm is to find the points closest (in
-$D$-space) to each of $K$ centers, and then take the mean of those
-points to update the locations of the centers; iterate to convergence.
+
+Briefly, the k-means algorithm is the following:
+$\bullet$~Starting at some guess for $K$ $D$-space cluster centers ($K$
+locations in the $D$-dimensional space), assign each point in the
+space (each star in our case) to the closest of the $K$ centers.
+$\bullet$~Given that assignment of stars to centers, update each center (each of
+the $K$ $D$-space positions) by taking the mean of the locations of
+the stars assigned to that center.
+Iterate the assignment and mean steps to convergence.
 The output is the locations of the $K$ centers and the assignments of
@@
 
-The k-means algorithm has disadvantages, of course.
+The k-means algorithm has a number of limitations:
 One is that $K$ must be chosen by hand (or heuristically at best).
@@
 results.
-The algorithm also only performs local optimization.
+Also, the k-means algorithm only performs local optimization.
 At each $K$ we perform 32 restarts with different initializations, and
@@
 But in addition to this, we re-scale these by approximate typical measurement
-precisions obtained by \thecannon\ before running k-means.
+precisions obtained by \thecannon\ before running k-means.\footnote{The scalings were DWH DWH.}
 This makes the space close to isotropic in measurement uncertainty or
@@
 
-One note on clustering algorithm decisions is warranted:
-Because this project is just a demonstration of the chemical-tagging
-concept, there is no requirement that we choose the best clustering
+Nothing about this algorithm or our choices are particularly
+tuned or optimized; this is in no sense the
 algorithm or the method that reveals the best structures.
 Indeed, it is a feature of this work that even the simplest, most
-generic, least-tuned clustering algorithm returns interesting
+generic clustering algorithm returns interesting
 structures (as shown below).
@@
 When the k-means results are returned, the $D\times D$ empirical
-variance tensor for the members of each cluster can be constructed,
-and an effective density can be computed as the number of points in
-the cluster divided by the square-root of the determinant of the
-tensor.
+variance tensor for the members of each cluster can be constructed.
+From this, an effective density in the abundance space can be computed
+as the number of points in the cluster divided by the square-root of
+the determinant of the tensor.
 This density was used to rank abundance-space overdensities (from highest

### 2016-01-19 papers/clusters.tex
biblio tweaks

@@
 
-% ARC to DWH: Are the colors of broken glass indicative of anything?
-
 \begin{figure}[!bp]
@@
 which we later attribute to M13, we find a mean of [Fe/H]$ = -1.55$ (DWH CHECK).
-The mean and spread agree well with other works dedicated to studying this cluster
+The mean and spread agree well with other work on this cluster
 \citep{Kraft_1992,Cohen_Melendez_2005,Johnson_Pilachowski_2012}, and compiled catalogs of
@@
 abundances (most notably Mg--Al) for this cluster are anti-correlated, as
-expected from other studies \citep[e.g.,][]{gratton}. However we find a
+expected from other studies \citep[for example,][]{gratton}. However we find a
 smaller spread in [O/Fe] than reported by others who have looked exclusively
@@
 Indeed, this work contains the largest number of abundances for Palomar 5,
-permitting a thorough search of tidally disrupted members \citep[e.g.,][]{Kuzma_2015} 
+permitting a thorough search of tidally disrupted members \citep[for example,][]{Kuzma_2015} 
 that have retained the unique cluster abundance signature.
@@
 Bovy,~J., 2015, arXiv:1510.06745
- \bibitem[Briley \etal(2004)]{Briley_2004} Briley, M.~M., Cohen, 
-J.~G., \& Stetson, P.~B.\ 2004, \aj, 127, 1579 
-\bibitem[Casey \etal(2016)]{casey16}
+ \bibitem[Briley \etal(2004)]{Briley_2004}
+Briley, M.~M., Cohen, J.~G., Stetson, P.~B., 2004, \aj, 127, 1579 
+\bibitem[Casey \etal(2016)]{ycasey16}
 Casey,~A.~R., Hogg,~D.~W., Ness,~M., Rix,~H.-W., 2016, in preparation
-\bibitem[Cohen \& Mel{\'e}ndez(2005)]{Cohen_Melendez_2005} Cohen, J.~G., \& Mel{\'e}ndez, J.\ 2005, \aj, 129, 303 
+\bibitem[Cohen \& Mel{\'e}ndez(2005)]{Cohen_Melendez_2005}
+Cohen, J.~G.,  Mel{\'e}ndez, J., 2005, \aj, 129, 303 
 \bibitem[De~Silva \etal(2015)]{galah}
@@
 Garc{\'{\i}}a~P{\'e}rez,~A.~E., Allende~Prieto,~C., Holtzman,~J.~A., \etal, 2015, arXiv:1510.07635
-\bibitem[Geisler \etal(1997)]{Geisler_1997} Geisler, D., Claria, 
-J.~J., \& Minniti, D.\ 1997, \pasp, 109, 799 
+\bibitem[Geisler \etal(1997)]{Geisler_1997}
+Geisler, D., Claria, J.~J., Minniti, D., 1997, \pasp, 109, 799 
 \bibitem[Gilmore \etal(2012)]{gaiaeso}
@@
 Gratton,~R.~G., Carretta,~E., Bragaglia,~A., 2012, \aapr, 20, 50
-\bibitem[Harris(1996)]{Harris_1996} Harris, W.~E.\ 1996, \aj, 112, 
-1487 
+\bibitem[Harris(1996)]{Harris_1996}
+Harris,~W.~E., 1996, \aj, 112, 1487 
 \bibitem[Hunter(2007)]{matplotlib}
@@
 Jofr{\'e},~P., M{\"a}dler,~T., Gilmore,~G., \etal, 2015, \mnras, 453, 1428 
-\bibitem[Johnson \& Pilachowski(2012)]{Johnson_Pilachowski_2012} Johnson, C.~I., \& Pilachowski, C.~A.\ 2012, \apjl, 754, L38 
-\bibitem[Kraft \etal(1992)]{Kraft_1992} Kraft, R.~P., Sneden, C., Langer, G.~E., \& Prosser, C.~F.\ 1992, \aj, 104, 645 
-\bibitem[Kuzma \etal(2015)]{Kuzma_2015} Kuzma, P.~B., Da Costa, 
-G.~S., Keller, S.~C., \& Maunder, E.\ 2015, \mnras, 446, 3297 
+\bibitem[Johnson \& Pilachowski(2012)]{Johnson_Pilachowski_2012}
+Johnson, C.~I.,  Pilachowski, C.~A., 2012, \apjl, 754, L38 
+\bibitem[Kraft \etal(1992)]{Kraft_1992}
+Kraft, R.~P., Sneden, C., Langer, G.~E.,  Prosser, C.~F., 1992, \aj, 104, 645 
+\bibitem[Kuzma \etal(2015)]{Kuzma_2015}
+Kuzma, P.~B., Da Costa, G.~S., Keller, S.~C., Maunder, E., 2015, \mnras, 446, 3297 
 \bibitem[Majewski \etal(2015)]{apogee}
@@
 Ro{\v s}kar,~R., Debattista,~V.~P., Quinn,~T.~R., Stinson,~G.~S., Wadsley,~J., 2008, \apjl, 684, L79
-\bibitem[Smith(1985)]{Smith_1985} Smith, G.~H.\ 1985, \apj, 298, 
-249 
-\bibitem[Smith \etal(2002)]{Smith_2002} Smith, G.~H., Sneden, C., 
-\& Kraft, R.~P.\ 2002, \aj, 123, 1502 
-\bibitem[Smith \etal(2005)]{Smith_2005} Smith, G.~H., Briley, 
-M.~M., \& Harbeck, D.\ 2005, \aj, 129, 1589
+\bibitem[Smith(1985)]{Smith_1985}
+Smith, G.~H., 1985, \apj, 298, 249 
+\bibitem[Smith \etal(2002)]{Smith_2002}
+Smith, G.~H., Sneden, C., Kraft, R.~P., 2002, \aj, 123, 1502 
+\bibitem[Smith \etal(2005)]{Smith_2005}
+Smith, G.~H., Briley,~M.~M.,  Harbeck, D., 2005, \aj, 129, 1589
 \bibitem[Ting \etal(2016)]{ting}

### 2016-01-19 papers/clusters.tex
small tweaks

@@
 % to-do
-% - fix all remaining occurrences of ARC, MKN, HWR, DWH, or CITE
 % - submit to arXiv
@@
 
-% ARC asks: "not flagged as bad" --> what exactly do we mean here? ASPCAPFLAG = 0?
-
 Our analysis draws on the spectra of \totalnumber\ giant stars ($\logg
-< 3.9$) from \apogee\ \acronym{DR12} \citep{dr12}, all those not
-flagged as bad in the \apogee\ \aspcap\ \citep{aspcap} pipeline
+< 3.9$) from \apogee\ \acronym{DR12} \citep{dr12}, with no warning flags
+set in the \apogee\ \aspcap\ \citep{aspcap} pipeline
 reductions.
@@
 research cluster in Cambridge.
-
-% HWR commented this (below) out
-
-% We start by performing our own continuum normalization on individual-exposure spectra and co-add them ourselves.
-% This is required, because the performance of \thecannon\ depends
-% critically on the methods used for spectral normalization.  It does
-% not require accurate continuum estimation, but it requires that the
-% normalization vector be obtained by a \emph{linear operation on the
-%   data}. The linear operation we use is a smooth-function fit to pixels
-% identified as continuum in our previous work \citep{thecannon}.
-
-% For the training data labels (used to train \thecannon; see below), we
-% use the \apogee\ \acronym{DR12} \aspcap\ atmospheric parameters and
-% 15-element abundances, but only for the subset of 12,681 red-giant
-% stars that have spectral signal-to-noise ratio measurements greater
-% than 200.
-% This is a small fraction of the total sample, but these highest
-% signal-to-noise stars are observed to obtain consistent, low-scatter,
-% sensible abundance-space measurements.
-
-% As described in the original paper \citep{thecannon} and the companion
-% papers \citep{casey16, ness16}, \thecannon\ is operated in two stages.
-% In the first stage---the training step---the training set of stars with
-% measured spectra and known labels (known parameters and abundances) is
-% used to establish a likelihood function for a set of coefficients that
-% parameterize a smooth, simple prediction of spectral pixels given
-% labels.
-% \thecannon\ finds the set of coefficients that optimizes the
-% training-step likelihood function.
-% In the implementation used in this \documentname, the spectral
-% prediction is quadratic in labels (including all cross-terms to second
-% order), and there are 17 labels ($\teff$, $\logg$, and 15 abundances
-% referenced to H) and 8500 spectral pixels, so there are on the order
-% of 1.4 million coefficients set by a convex optimization in the
-% training step.
-
-% In the second stage---the test step---each unlabeled star is used to
-% establish a single-star likelihood function for labels, holding the
-% coefficients fixed.
-% \thecannon\ finds the labels that minimize the single-star likelihood
-% function for each test-set star.
-% This optimization is not convex, but it is embarrassingly parallel.
-% For the results shown here, the training step takes hours and the test
-%step for 150,000 spectra takes less than 30~minutes on a small 
-%research cluster in Cambridge.
 
@@
 Briefly, the k-means algorithm is the following:
-$\bullet$~Starting at some guess for $K$ $D$-space cluster centers ($K$
-locations in the $D$-dimensional space), assign each point in the
+$\bullet$~Start at some inital guess for the locations of $K$ $D$-space cluster centers ($K$
+locations in the $D$-dimensional space).
+$\bullet$~Assign each point in the
 space (each star in our case) to the closest of the $K$ centers.
-$\bullet$~Given that assignment of stars to centers, update each center (each of
+$\bullet$~Given this assignment of stars to centers, update each center (each of
 the $K$ $D$-space positions) by taking the mean of the locations of
 the stars assigned to that center.
-Iterate the assignment and mean steps to convergence.
-The output is the locations of the $K$ centers and the assignments of
-all points to centers.
+$\bullet$~Iterate these steps (assignment of points, followed by taking of means) to convergence.
+The output of this algorithm is the converged locations of the $K$ centers and the assignments of
+all points to those centers.
 This algorithm is fast and performs well in practice in problems of
@@
 But in addition to this, we re-scale these by approximate typical measurement
-precisions obtained by \thecannon\ before running k-means.\footnote{The scalings were DWH DWH.}
+precisions obtained by \thecannon\ before running k-means.\footnote{%
+  The scalings were
+  [C/H]/[0.041],
+  [N/H]/[0.044],
+  [O/H]/[0.037],
+  [Na/H]/[0.111],
+  [Mg/H]/[0.032],
+  [Al/H]/[0.055],
+  [Si/H]/[0.041],
+  [S/H]/[0.054],
+  [K/H]/[0.069],
+  [Ca/H]/[0.043],
+  [Ti/H]/[0.072],
+  [V/H]/[0.146],
+  [Mn/H]/[0.041],
+  [Fe/H]/[0.019], and
+  [Ni/H]/[0.034].}
 This makes the space close to isotropic in measurement uncertainty or
@@
 
-% ARC TO DO: Make claim of the fourth being M53.
 We chose a few interesting cases from the high-density clusters
@@
   abundance-space cluster.
-  This one turns out to be dominated by a high-velocity-dispersion
+  This one turns out to be dominated by a hitherto unrecognized high-velocity-dispersion
   structure in the Galaxy halo.\label{fig:halo}}
@@
   This one turns out to be dominated by a very thin young stellar
-  structure in the Galaxy disk.\label{fig:disk}}
+  structure in the Galaxy disk, probably one that has been discovered previously \citep{wegg}.\label{fig:disk}}
 \end{figure}
@@
 The abundances of known clusters that we identify by k-means are in
-excellent agreement with the literature. For the group shown in Figure \ref{fig:M13},
-which we later attribute to M13, we find a mean of [Fe/H]$ = -1.55$ (DWH CHECK).
+excellent agreement with the literature. For the group shown in \figurename~\ref{fig:M13},
+which we later attribute to M13, we find a mean of [Fe/H]$ = -1.55$.
 The mean and spread agree well with other work on this cluster
@@
 consistent: we find a correlation in [C/Fe] and [N/Fe] abundances \citep{Smith_2005}, and their
-projection in Figure \ref{fig:M13} show a spread that reflects deep mixing
+projection in \figurename~\ref{fig:M13} show a spread that reflects deep mixing
 along the red giant branch \citep{Briley_2004}. We find that the light element
@@
 There are scant abundance measurements reported for Palomar 5, for which
-most stars in the cluster in Figure \ref{fig:Pal5} belong. We find a mean
-metallicity of [Fe/H]$ = -1.3$ (DWH CHECK), which lies between the only
+most stars in the cluster in \figurename~\ref{fig:Pal5} belong. We find a mean
+metallicity of [Fe/H]$ = -1.3$, which lies between the only
 photometric \citep{Geisler_1997} and spectroscopic \citep{Smith_1985,Smith_2002} abundance
@@
 
-% Sagittarius
-
-The high-velocity-dispersion structure seen in Figure \ref{halo} stands
+The high-velocity-dispersion structure seen in \figurename~\ref{fig:halo} stands
 in stark contrast to the rest of the clusters we have identified. In
@@
 
-% The high-velocity-dispersion structure in the Galaxy halo.
 We freely admit that k-means is by no means optimal, and other clusters
@@
 or they may be true cluster members that are now unbound. However in the
-case of the high-velocity-dispersion structure in Figure \ref{fig:halo},
+case of the high-velocity-dispersion structure in \figurename~\ref{fig:halo},
 the situation is far more extreme. We identify a group of stars with
@@
 Briley, M.~M., Cohen, J.~G., Stetson, P.~B., 2004, \aj, 127, 1579 
-\bibitem[Casey \etal(2016)]{ycasey16}
+\bibitem[Casey \etal(2016)]{casey16}
 Casey,~A.~R., Hogg,~D.~W., Ness,~M., Rix,~H.-W., 2016, in preparation

### 2016-01-19 papers/clusters.tex
text tweaks, no footnote

@@
 % Copyright 2016 the authors.
-
-% to-do
-% - submit to arXiv
-% - forward to Freeman, Bovy, others?
-% - submit to journal
 
@@
 However, it has the disadvantage that it's spatial coverage is incomplete
-(i.e., limited pointings produce a \foreign{`swiss cheese'}-like structure),
+(that is, limited pointings produce a fractured sky map),
 which makes it hard to see, within the data set, linear or
@@
 But in addition to this, we re-scale these by approximate typical measurement
-precisions obtained by \thecannon\ before running k-means.\footnote{%
-  The scalings were
+precisions obtained by \thecannon\ before running k-means.
+These scalings were
   [C/H]/[0.041],
@@
   [Fe/H]/[0.019], and
-  [Ni/H]/[0.034].}
-This makes the space close to isotropic in measurement uncertainty or
+  [Ni/H]/[0.034].
+This scaling makes the space close to isotropic in measurement uncertainty or
 observational precision.

### 2016-01-19 papers/clusters.tex
abstract tweak
wording tweaks

@@
 % Context:
-Chemical tagging promises to robustly group co-natal, spatially extended stars by their present-day  photospheric abundances.
+Chemical tagging promises to use surface chemical abundance measurements to find sets of stars
+that were born together but which are now in spatially extended distributions.
 This idea has not previously yielded scientific successes, probably because of the
@@
 However, we have succeeded in substantially improving spectroscopic measurements with \thecannon: 
-delivering $\geq$15 individual abundances for $\sim10^5$ stars with precisions around 0.04~dex.
+delivering $\geq 15$ individual abundances for $\sim 10^5$ stars with precisions around 0.04~dex.
 % Aims:
@@
 stellar structures.
+Although this is very strong evidence that chemical tagging will work,
+we have not performed, in any sense, the canonical procedure of
+looking for features that are consistent with being delta functions in
+chemical-abundance space (convolved with a measurement noise
+distribution); we have looked only for over-densities, not delta
+functions.
 
-What made this possible is substantial improvements in precision
+What made our success here possible is substantial improvements in precision
 delivered by \thecannon.
@@
 
-We freely admit that k-means is by no means optimal, and other clusters
+The k-means algorithm employed here is by no means optimal, and other clusters
 we identify are accompanied by a few stars that are not currently co-spatial.
@@
 highly similar abundances in 15 dimensions that are now spread throughout
-the Galaxy. Because some stars are candidates of the massive globular
+the Galaxy. Because some stars are candidate members of the massive globular
 cluster M53, it provides tantalizing possibility that these stars may
@@
 these stars have been accreted onto the halo by a single globular
-cluster early in the Milky Way's formation. While speculative (and we
-make no consideration to whether this is indeed the true scenario),
-this example demonstrates the ultimate goal of chemical tagging.
+cluster early in the Milky Way's formation. While speculative,
+this example demonstrates the promise of chemical tagging.
 
@@
 found to exist predominantly toward the end of the bar (at $l$
-$\approx$ 30$^\circ$).
+$\approx 30$~deg).
 The stars in this structure are metal-rich and have a low
@@
 It is a pleasure to thank
-  Jeremy Magland (\acronym{SCDA}), and
-  Eddie Schlafly (\acronym{LBNL})
+  Jeremy Magland (\acronym{SCDA})
 for valuable discussions.
@@
 \bibitem[Ting \etal(2016)]{ting}
-Ting,~Y.-S., Conroy,~C., Rix,~H.-W., 2016, \apj, in press (arXiv:1507.07563)
+Ting,~Y.-S., Conroy,~C., Rix,~H.-W., 2016, \apj, 816, 10
 \bibitem[van~der~Walt \etal(2011)]{numpy}

### 2016-01-19 papers/clusters.tex
tweaks

@@
 noise and incompleteness in chemical-abundance measurements.
-However, we have succeeded in substantially improving spectroscopic measurements with \thecannon: 
-delivering $\geq 15$ individual abundances for $\sim 10^5$ stars with precisions around 0.04~dex.
+However, we have succeeded in substantially improving spectroscopic measurements with \thecannon,
+which has delivered $\geq 15$ individual abundances for $\sim 10^5$ stars with precisions around 0.04~dex.
 % Aims:
@@
 We find that clusters in abundance space are indeed clusters in phase space.
-We recover some known clusters and find other interesting structures.
+We recover some known phase-space clusters and find other interesting structures.
 This confirms the chemical-tagging hypothesis and verifies the precision of abundances delivered by \thecannon.

### 2016-01-19 code/kmeans.py
figure improvements

@@ -123,8 +123,8 @@ def plot_cluster_context(sizes, densities, k, name, dir, suffix="png"):
     plt.plot(sizes, densities, marker=".", c="k", ms=5.0, alpha=0.5, ls="none")
     plt.plot(sizes[k], densities[k], marker="o", c="k", mfc="none", ms=7.0, mew=2., alpha=1., ls="none")
     _clusterlims(sizes, densities)
-    plt.ylabel("cluster density")
-    plt.xlabel("number of stars in cluster")
+    plt.ylabel("cluster abundance-space density")
+    plt.xlabel("number in abundance-space cluster")
     plt.loglog()
     [l.set_rotation(45) for l in plt.gca().get_xticklabels()]
     [l.set_rotation(45) for l in plt.gca().get_yticklabels()]
@@ -143,10 +143,10 @@ def plot_cluster_stats(sizes, densities, dir, suffix="png"):
     plt.subplots_adjust(left=0.15, right=0.97, bottom=0.15, top=0.97)
     plt.clf()
     kwargs = {"marker": ".", "ls": "none"}
-    plt.plot(sizes, densities, c="k", ms=5.0, alpha=0.5, **kwargs)
+    plt.plot(sizes, densities, c="k", ms=7.5, alpha=0.5, **kwargs)
     _clusterlims(sizes, densities)
-    plt.ylabel("cluster density")
-    plt.xlabel("number of stars in cluster")
+    plt.ylabel("cluster abundance-space density")
+    plt.xlabel("number of stars in abundance-space cluster")
     plt.loglog()
     plt.savefig(fn)
     print("plot_cluster_stats(): wrote", fn)

### 2016-01-20 papers/clusters.tex
Abstract tweak (refrained from using chemical twice in the same sentence)

@@
 % Context:
-Chemical tagging promises to use surface chemical abundance measurements to find sets of stars
-that were born together but which are now in spatially extended distributions.
+Chemical tagging promises to use detailed abundance measurements to find sets of stars that were born together but are now in spatially extended distributions.
 This idea has not previously yielded scientific successes, probably because of the

### 2016-01-20 papers/clusters.tex
Typos, a word, and comma for large numbers

@@
 % results
-\newcommand{\totalnumber}{98462}
+\newcommand{\totalnumber}{98,462}
 
@@
 
-Part of the reason that chemical tagging still remains unrealized
+Part of the reason that chemical tagging remains unrealized
 is because the level of abundance specificity required
@@
 We re-derived 17 labels ($\teff$, $\logg$, and 15 abundances
-referenced to H) form the \apogee\ \acronym{DR12} spectra. 
+referenced to H) from the \apogee\ \acronym{DR12} spectra. 
 For the training step of \thecannon, 
 12,681 red-giant stars with spectral $S/N \ge 200$ were used.
-For this small fraction of the sample with the highest $S/N$ objects the
+For this small fraction of the sample with the highest $S/N$, the
 labels provided by \aspcap\ provide consistent, relatively low-scatter, sensible

### 2016-01-20 papers/clusters.tex
Subtle emphasis on the model size
Surely the [logarithmic] brackets around these scalings was unintended, right?? @davidwhogg

@@
 that predicts the normalized spectrum as a quadratic function of all
-labels. In the second stage---the test step---each unlabelled star is
+labels ($\sim 1.5\times10^6$ model coefficients). In the second stage---the test step---each unlabelled star is
 used to establish a single-star likelihood function for its labels,
@@
 These scalings were
-  [C/H]/[0.041],
-  [N/H]/[0.044],
-  [O/H]/[0.037],
-  [Na/H]/[0.111],
-  [Mg/H]/[0.032],
-  [Al/H]/[0.055],
-  [Si/H]/[0.041],
-  [S/H]/[0.054],
-  [K/H]/[0.069],
-  [Ca/H]/[0.043],
-  [Ti/H]/[0.072],
-  [V/H]/[0.146],
-  [Mn/H]/[0.041],
-  [Fe/H]/[0.019], and
-  [Ni/H]/[0.034].
+  [C/H]/0.041,
+  [N/H]/0.044,
+  [O/H]/0.037,
+  [Na/H]/0.111,
+  [Mg/H]/0.032,
+  [Al/H]/0.055,
+  [Si/H]/0.041,
+  [S/H]/0.054,
+  [K/H]/0.069,
+  [Ca/H]/0.043,
+  [Ti/H]/0.072,
+  [V/H]/0.146,
+  [Mn/H]/0.041,
+  [Fe/H]/0.019, and
+  [Ni/H]/0.034.
 This scaling makes the space close to isotropic in measurement uncertainty or

### 2016-01-20 papers/clusters.tex
Give fraction of M53 stars in text

@@
 highly similar abundances in 15 dimensions that are now spread throughout
-the Galaxy. Because some stars are candidate members of the massive globular
+the Galaxy. Because some stars ($<8\%$) are candidate members of the massive globular
 cluster M53, it provides tantalizing possibility that these stars may

### 2016-01-20 papers/clusters.tex
Anna Y. Q. Ho

@@
 least precise---measurements of many different abundances are needed
-for stellar sibling to have sufficiently unique fingerprints.
+for stellar siblings to have sufficiently unique fingerprints.
 Stellar spectroscopic surveys now have the resolution,
@@
 There may be chemical-abundance overlaps among open clusters
-\citep{blancocuaresma}; coeval groups of stars may have similar tags
+\citep{blancocuaresma}, coeval groups of stars may have similar tags
 but different birth places \citep{mitschang}, and the
@@
 On the other hand, precise studies of stellar twins \citep{melendez, jofre}
-and open clusters \citep{bovy} indicates that chemical abundances ought to be
-very precisely similar for co-eval stars.
+and open clusters \citep{bovy} indicate that chemical abundances ought to be
+very similar for co-eval stars.
 
@@
 models of stars.
-It relies only on there being training data---\emph{some} stars for
+It relies only on there being training data---\emph{some} reference stars for
 which parameters and abundances are known and believed.
@@
 \thecannon\ can deliver 15 to 19 abundances for stars in the
-\apogee\ survey, at precisions higher than even the training data on
+\apogee\ survey, at precisions higher than even the training (reference) data on
 which the model is trained.
@@
 chemical abundances, whereas the accuracy of those labels is limited only
-by the training set employed.
+by the training (reference) set employed.
 
@@
 data are taken in the infrared---explores the thin disk.
-However, it has the disadvantage that it's spatial coverage is incomplete
+However, it has the disadvantage that its spatial coverage is incomplete
 (that is, limited pointings produce a fractured sky map),
@@
 we will have resolved all practical outstanding issues plaguing 
-chemical tagging, thereby coming closer to the tantalizing potential 
-in truly unravelling the formation of the Milky Way.
+chemical tagging, thereby bringing us much closer to
+unravelling the formation of the Milky Way.
 
@@
   The bottom three panels show stellar meta-data not used in the
-  clustering but used to test whether the abundance-space clusters
-  are interesting spatially.\label{fig:all}}
+  clustering described below.\label{fig:all}}
 \end{figure}
@@
 Briefly, the k-means algorithm is the following:
-$\bullet$~Start at some inital guess for the locations of $K$ $D$-space cluster centers ($K$
+$\bullet$~Start at some initial guess for the locations of $K$ $D$-space cluster centers ($K$
 locations in the $D$-dimensional space).
@@
 least one cluster.
-For this reason, there is no sense in which each ``cluster'' returned
-by k-means is in any sense an overdensity in abundance space.
+For this reason, there is no sense in which \emph{every} ``cluster'' returned
+by k-means is a distinct overdensity in abundance space.
 In what follows, we only consider high-density clusters---clusters
@@
 This \documentname\ is simply a demonstration of the chemical-tagging
-concept; an investigation of whether we can construct complete
-catalogs of stellar clusters is beyond the scope.
+concept; a full investigation of whether we can construct complete
+catalogs of stellar clusters is beyond our present scope.
 
@@
 found in globular clusters. First, the detection of these signatures gives
-confirmation to our analysis: The training set was not purposefully selected
+confirmation to our analysis: The training (reference) set was not purposefully selected
 to contain members of any globular or open cluster. However amongst the test
@@
 It is a pleasure to thank
-  Jeremy Magland (\acronym{SCDA})
-for valuable discussions.
+  Anna~Y.~Q.~Ho (Caltech),
+  Jeremy~Magland (\acronym{SCDA}), and
+  the Blanton--Hogg group meeting
+for valuable comments and discussions.
 This project was funded in part by

### 2016-01-20 papers/clusters.tex
Hogg comments

@@
     and~Daniel~Foreman-Mackey\altaffilmark{6,7}}
-\altaffiltext{1}{Simons Center for Data Analysis, 160 Fifth Avenue, 7th Floor, New York, NY 10010, USA}
+\altaffiltext{1}{Simons Center for Data Analysis, 160 Fifth Avenue, 7th floor, New York, NY 10010, USA}
 \altaffiltext{2}{Center for Cosmology and Particle Physics, Department of Physics,
-  New York University, 4 Washington Pl., room 424, New York, NY, 10003, USA}
-\altaffiltext{3}{Center for Data Science, New York University, 726 Broadway, 7th Floor, New York, NY 10003, USA}
+  New York University, 4 Washington Pl., room 424, New York, NY 10003, USA}
+\altaffiltext{3}{Center for Data Science, New York University, 726 Broadway, 7th floor, New York, NY 10003, USA}
 \altaffiltext{4}{Max-Planck-Institut f\"ur Astronomie, K\"onigstuhl 17, D-69117 Heidelberg, Germany}
@@
 \altaffiltext{6}{Sagan Fellow}
-\altaffiltext{7}{Department of Astronomy, University of Washington, Box 351580, Seattle, WA 98195-1580}
+\altaffiltext{7}{Department of Astronomy, University of Washington, Box 351580, Seattle, WA 98195, USA}
 \email{david.hogg@nyu.edu}
@@
 % Context:
-Chemical tagging promises to use detailed abundance measurements to find sets of stars that were born together but are now in spatially extended distributions.
+Chemical tagging promises to use detailed abundance measurements to
+identify spatially separated stars that were in fact born together
+(in the same molecular cloud), long ago.
 This idea has not previously yielded scientific successes, probably because of the
 noise and incompleteness in chemical-abundance measurements.
-However, we have succeeded in substantially improving spectroscopic measurements with \thecannon,
-which has delivered $\geq 15$ individual abundances for $\sim 10^5$ stars with precisions around 0.04~dex.
+However, we have succeeded in substantially improving spectroscopic
+measurements with \thecannon, which has delivered $15$ individual
+abundances for $\sim 10^5$ stars observed as part of the
+\apogee\ spectroscopic survey, with precisions around 0.04~dex.
 % Aims:
@@
 % Methods:
-We identify (by the k-means algorithm) overdensities of stars observed by the \apogee\ spectroscopic survey
-in a 15-dimensional chemical-abundance space delivered by \thecannon,
+We identify (by the k-means algorithm) overdensities of stars
+in the 15-dimensional chemical-abundance space delivered by \thecannon,
 and plot the associated stars in phase space.
@@
 We recover some known phase-space clusters and find other interesting structures.
-This confirms the chemical-tagging hypothesis and verifies the precision of abundances delivered by \thecannon.
+This confirms the chemical-tagging hypothesis and verifies the precision of the abundance measurements delivered by \thecannon.
 This is the first-ever project to identify phase-space structures by blind search purely in abundance space;
@@
 star clusters that disperse rapidly; they will eventually end up in
-very different parts of phase space, on differing orbits.
+very different parts of phase space in the Galaxy.
 Yet, if every star preserved its photospheric element abundances over
@@
 origin ought to be identifiable through their detailed photospheric
-abundances, long after any spatial or phase-space proximity has
+abundances, long after any spatial proximity has
 vanished.
 
-This idea, dubbed ``chemical tagging'', was proposed long ago
-\citep{freeman} and is one of the principal motivations for a number
-of surveys, most notably \apogee\ \citep{apogee},
+This idea---dubbed ``chemical tagging''
+\citep{freeman}---is one of the principal motivations for a number
+of surveys, including \apogee\ \citep{apogee},
 \gaiaeso\ \citep{gaiaeso} and \galah\ \citep{galah}.
@@
 stars are born---or after a star cluster is accreted and
-disrupted---associations or groups will disperse, through either
-internal two-body processes or external tidal forces.
+disrupted---associations or groups will disperse, through
+two-body processes, interactions with resonances, or tides from the whole Galaxy.
 
@@
 The current precision on abundance measurements in the published
-survey catalogs is not high enough (for example, \citealt{martell,
-  ting}).
-However, our recent work \citep{thecannon} suggests that \emph{data}
-are precise enough: there is enough $S/N$ at the relevant
+survey catalogs is not high enough (for example, \citealt{martell, ting}).
+However, our recent work \citep{thecannon} suggests that the \emph{data}
+are precise enough: There is enough $S/N$ at the relevant
 locations in spectrum space to deliver high-precision tags.
-The existence of very large data sets, homogeneous in data/spectrum
-space, suggest that one can use data-driven approaches to determining
-stellar abundances, which can considerably outperform traditional
+The existence of very large data sets, homogeneous in spectrum
+space, suggests that data-driven approaches to the determination
+of stellar abundances might considerably outperform traditional
 methods.
@@
 absolute accuracy for chemical-abundance measurements.
-We only require that we can precisely see that two stars are similar
-in their abundances; even if we are wrong about the absolute values of
+It only requires that we can precisely see that two stars are similar
+in their abundances, even if we happen to be wrong about the absolute values of
 those abundances.
@@
 space.
-That doesn't require overall accuracy, but it requires that the models
+That doesn't require \emph{overall} accuracy, but it requires that the models
 have the right dependencies on atmospheric parameters such that the
@@
 in the face of knowingly inaccurate models.  Incomplete atomic data, 
-simplifications of photospheric structure, presumptions about convective
-motion, and inconsistencies resulting from assumptions of the local thermal
-equilibrium all contribute to produce knowingly inaccurate abundances.
+simplifications of photospheric structure, assumptions about convective
+motion, and inconsistencies resulting from positing local thermal
+equilibrium all contribute to produce inaccurate abundances.
 \thecannon\ stands out because it demonstrably improves the precision on
@@
 reductions.
-We re-analyze these \emph{spectra} using \thecannon, because this
+We re-analyze these \emph{spectra} using \thecannon, because it can
 deliver stellar abundance labels of higher precision, especially for
@@
 are given in full detail in the companion papers \citep{casey16,
-  ness16}, and we only summarize briefly here.
+  ness16}, and we only summarize briefly here:
 We re-derived 17 labels ($\teff$, $\logg$, and 15 abundances
@@
 used to establish a single-star likelihood function for its labels,
-holding the model coefficients fixed.
+holding the spectral model coefficients (the parameters of the model) fixed.
 \thecannon\ finds the labels that minimize the single-star likelihood
@@
 this work we show two-dimensional abundance projections for only a few
-elements. The entire sequence of 15-dimensional combinations is too
+elements. The entire sequence of 15-choose-2 combinations is too
 immense to visualize. The projections shown are ones which
@@
 them suitable projections for us to show in verifying and examining
-any substructure identified by k-means (or any other algorithm).
+any substructure identified by k-means (or any other clustering algorithm).
 
@@
 The abundance-space clusters shown in the \figurename s
-might not look dense in abundance space, but they
+might not look dense in 2-d projections of the abundance space, but they
 are very dense in the 15-dimensional space.  Indeed,
 \figurename~\ref{fig:M13} is the densest cluster found in the 15-d
-space \emph{by far} (see \figurename~\ref{fig:densities}).
+space \emph{by far}.
 The challenge of this work is to find structure in the high
@@
 This is a reminder that k-means has been chosen just for simplicity
-and reliability here; it is by no means matched to the discovery of
+here; it is by no means matched to the discovery of
 stellar structures.
 An important follow-up project is to build a model of abundance space
-that captures the features expected from stellar populations.
+that captures the features expected from stellar populations (and observational uncertainties).
 Even if we were able to tune k-means in some sense ``perfectly'', it
 would still break up many stellar clusters, since they often show
-multiple stellar populations and non-linear correlations between elemental abundances.
+multiple stellar populations and non-linear correlations between element abundances.
 It is also the case that each of the abundance-space clusters we
@@
 We have demonstrated for the first time that, in a large-scale
-spectroscopic survey of stars, stars that appear similar in
+spectroscopic survey, stars that appear similar in
 chemical-abundance properties are often associated physically in
@@
 (by k-means) substantially more difficult. Briefly, the k-means algorithm
-is most effective for near-normally distributed clusters in high-dimensional space.
-While globular clusters often show distributions of this kind in a few dimensions, they also demonstrate non-linear correlations which would be
+is most effective for near-normally distributed, isotropic clusters in high-dimensional space.
+While globular clusters often show distributions of this kind in a few dimensions,
+they also demonstrate non-linear correlations which would be
 sub-optimally selected by k-means. For these reasons the fact that (a)~\thecannon\
-reports these abundance correlations and (b)~k-means--as a simple algorithm without information about expectations in abundance space--does not appear to be
-severely impacted by these correlations, provides high promise for more sophisticated approaches.
+reports these abundance correlations and (b)~k-means (as a simple algorithm
+without information about expectations in abundance space) does not appear to be
+severely impacted by these correlations, gives us great hope for more sophisticated approaches.
 
@@
 excellent agreement with the literature. For the group shown in \figurename~\ref{fig:M13},
-which we later attribute to M13, we find a mean of [Fe/H]$ = -1.55$.
+which we attribute to M13, we find a mean of [Fe/H]$ = -1.55$.
 The mean and spread agree well with other work on this cluster
@@
 k-means is by no means optimal for identifying arbitrary-shaped structures
-in high-dimensional space, and the first stars that would be `discarded' from
-this group would be those with the most extreme abundances: low [O/Fe] and
+in high-dimensional space, and the first stars that would be assigned to another k-means
+cluster would be those with the most extreme abundances: low [O/Fe] and
 high [Na/Fe].
@@
 
-The high-velocity-dispersion structure seen in \figurename~\ref{fig:halo} stands
+The high velocity-dispersion structure seen in \figurename~\ref{fig:halo} stands
 in stark contrast to the rest of the clusters we have identified. In
@@
 the situation is far more extreme. We identify a group of stars with
-highly similar abundances in 15 dimensions that are now spread throughout
-the Galaxy. Because some stars ($<8\%$) are candidate members of the massive globular
+very similar abundances in 15 dimensions that are now spread throughout
+the Galaxy halo.
+Because some stars ($<8\%$) are candidate members of the massive globular
 cluster M53, it provides tantalizing possibility that these stars may
@@
 cluster early in the Milky Way's formation. While speculative,
-this example demonstrates the promise of chemical tagging.
+this idea demonstrates the promise of chemical tagging.
 
@@
 interpetable.
-There are many other things to be found in this data set, and much
+There are many other things to be found in this data set, and many
 more effective models to be built of abundance-space structure.

### 2016-01-21 code/kmeans.py
make plot_everything option

@@ -154,11 +154,13 @@ def plot_cluster_stats(sizes, densities, dir, suffix="png"):
 if __name__ == "__main__":
     dfn = "./data/results-unregularized-matched.fits.gz"
     dir = "./kmeans_figs"
-    if not os.path.exists(dir):
-        os.mkdir(dir)
     suffix = "png"
     data = None
     plotdata = None
+    # Ks = 2 ** np.arange(3, 10) # do everything
+    Ks = [256, ] # just do the winner
+    plot_everything = True # set to true only for MKN atlas
+    # plot_everything = False
 
     scale_dict = {"FE_H": 0.0191707168068, # all from AC
                   "AL_H": 0.0549037045265,
@@ -226,7 +228,12 @@ if __name__ == "__main__":
                    "AL_H", "CA_H", "C_H", "K_H",  "MG_H", "MN_H", "NA_H",
                    "NI_H", "N_H",  "O_H", "SI_H", "S_H",  "TI_H", "V_H"]
 
-    for K in [256,]: # just do the winner
+    # get ready...
+    if not os.path.exists(dir):
+        os.mkdir(dir)
+
+    # go!!
+    for K in Ks:
         pfn = "data/kmeans_{:04d}.pkl".format(K)
         try:
             print("attempting to read pickle", pfn)
@@ -299,7 +306,7 @@ if __name__ == "__main__":
             plot_cluster_context(sizes, densities, k, clustername, dir)
             plot_one_cluster(plotdata, plotdata_labels, (clusters==k), clustername, dir)
             plotcount += 1
-            if plotcount == 32:
+            if not plot_everything and plotcount >= 32:
                 break
 
         # plot some non-clusters for Ness

### 2016-01-21 code/kmeans.py
make an atlas for MKN

@@ -102,6 +102,7 @@ def plot_one_cluster(data, labels, mask, name, dir, suffix="png"):
             [l.set_rotation(45) for l in plt.gca().get_yticklabels()]
             plt.savefig(fn)
             print("plot_one_cluster(): wrote", fn)
+    plt.close("all")
 
 def _clusterlims(sizes, densities):
     plt.ylim(np.min(densities) / 5., np.max(densities) * 5.)

### 2016-01-21 code/kmeans.py
turn off the insanity

@@ -160,8 +160,7 @@ if __name__ == "__main__":
     plotdata = None
     # Ks = 2 ** np.arange(3, 10) # do everything
     Ks = [256, ] # just do the winner
-    plot_everything = True # set to true only for MKN atlas
-    # plot_everything = False
+    plot_everything = False # set to true only for MKN atlas
 
     scale_dict = {"FE_H": 0.0191707168068, # all from AC
                   "AL_H": 0.0549037045265,

### 2016-01-22 code/distances/distances.py
bringing in Andy's distances code for safe keeping

@@ -1,0 +1,222 @@
+# coding: utf-8
+
+""" Interpolate absolute magnitudes from the PARSEC isochrones """
+
+from __future__ import division, print_function
+
+__author__ = "Andy Casey <arc@ast.cam.ac.uk>"
+
+# Standard library
+from glob import glob
+
+# Third party
+import numpy as np
+import scipy.interpolate
+
+
+def load_parsec_isochrone(filename):
+
+    # Load the relevant information from the isochrone filename
+
+    """
+    # File generated by CMD 2.5 (http://stev.oapd.inaf.it/cmd) on Wed Jan  8 16:38:19 CET 2014
+    # PARSEC isochrones, release v1.1
+    # Basic reference: Bressan et al. (2012), MNRAS, 427, 127
+    # Warning: the TP-AGB phase is not included! TP-AGB tracks are in preparation by Marigo et al.
+    # Photometric system:  2MASS <i>JHK</i><sub>s</sub>
+    # BCs of Carbon stars derive from Loidl et al. (2001, A&A 342, 531)
+    # O-rich circumstellar dust ignored
+    # C-rich circumstellar dust ignored
+    # IMF: Chabrier (2001) lognormal 
+    # On RGB, assumed Reimers mass loss with efficiency eta=0.2
+    # Kind of output: isochrone tables
+    #   Isochrone  Z = 0.01210  Y = 0.27040 [M/H] = -0.089  eta_R = 0.200   Age =   1.0000e+09 yr
+    #      Z    log(age/yr) M_ini       M_act   logL/Lo logTe   logG    mbol    J       H       Ks  int_IMF stage
+        0.012100    9.0000  0.10000000  0.1000  -2.9005 3.5040  5.3005  12.021  10.125  9.580   9.342   1.57040203  0
+    """
+
+
+    # We require: Teff, Logg, J, H, K
+    # We will arrange it like this though: (J, H, K, Teff, logg) because
+    # that will make things easier later
+    isochrone_data = np.loadtxt(filename, usecols=(8, 9, 10, 5, 6))
+
+    # Parse the overall metallicity as well:
+    with open(filename, 'r') as fp:
+        # There are cleverer ways to do this
+        content = fp.readlines()
+
+    try:
+        feh = float(content[11].split()[10])
+    except:
+        feh = float(content[12].split()[10])
+
+    # Put this as a column to the data array
[…]

### 2016-01-23 code/kmeans.py
small bug in variance / density estimate; plot high-Z only

@@ -158,9 +158,10 @@ if __name__ == "__main__":
     suffix = "png"
     data = None
     plotdata = None
-    # Ks = 2 ** np.arange(3, 10) # do everything
+    Ks = 2 ** np.arange(8, 12) # do everything
     Ks = [256, ] # just do the winner
     plot_everything = False # set to true only for MKN atlas
+    only_high_Z = True
 
     scale_dict = {"FE_H": 0.0191707168068, # all from AC
                   "AL_H": 0.0549037045265,
@@ -268,18 +269,21 @@ if __name__ == "__main__":
         N, D = data.shape
         sizes = np.zeros(K).astype(int)
         logdets = np.zeros(K)
+        fehs = np.zeros(K)
         for k in range(K):
             I = (clusters == k)
             sizes[k] = np.sum(I)
             subdata = data[I,:] - (np.mean(data[I], axis=0))[None, :]
             if sizes[k] > (D + 1):
-                variance = np.sum(subdata[:,:,None] * subdata[:,None,:], axis=0)
+                variance = np.sum(subdata[:,:,None] * subdata[:,None,:], axis=0) / (sizes[k] - 1.)
                 s, logdets[k] = np.linalg.slogdet(variance)
                 assert s > 0
             else:
                 logdets[k] = -np.Inf
+            fehs[k] = np.mean(data[I, 0])
         densities = sizes * np.exp(-0.5 * logdets)
         print(K, "size range:", np.min(sizes), np.median(sizes), np.max(sizes))
+        print(K, "[Fe/H] range", np.min(fehs), np.median(fehs), np.max(fehs))
         print(K, "logdet range:", np.min(logdets), np.median(logdets), np.max(logdets))
         print(K, "density range", np.min(densities), np.median(densities), np.max(densities))
         plot_cluster_stats(sizes, densities, dir)
@@ -302,12 +306,13 @@ if __name__ == "__main__":
         plotcount = 0
         for k in (np.argsort(densities))[::-1]:
             clustername = "cluster_{:04d}_{:04d}".format(K, k)
-            print(clustername, sizes[k], logdets[k], densities[k], fields[(clusters == k)].T)
-            plot_cluster_context(sizes, densities, k, clustername, dir)
-            plot_one_cluster(plotdata, plotdata_labels, (clusters==k), clustername, dir)
-            plotcount += 1
-            if not plot_everything and plotcount >= 32:
-                break
[…]

### 2016-01-23 code/kmeans.py
tweaks

@@ -159,9 +159,9 @@ if __name__ == "__main__":
     data = None
     plotdata = None
     Ks = 2 ** np.arange(8, 12) # do everything
-    Ks = [256, ] # just do the winner
+    # Ks = [256, ] # just do the winner
     plot_everything = False # set to true only for MKN atlas
-    only_high_Z = True
+    only_high_Z = True # set to true to only plot things in the metallicity bulk
 
     scale_dict = {"FE_H": 0.0191707168068, # all from AC
                   "AL_H": 0.0549037045265,

### 2016-01-23 code/kmeans.py
noncluster no longer necessary

@@ -314,15 +314,6 @@ if __name__ == "__main__":
                 if (not plot_everything) and (plotcount >= 32):
                     break
 
-        # plot some non-clusters for Ness
-        kmed = (np.argsort(densities))[K/2]
-        klow = (np.argsort(densities))[0]
-        for k in (kmed, klow):
-            clustername = "noncluster_{:04d}_{:04d}".format(K, k)
-            print(clustername, sizes[k], logdets[k], densities[k], fields[(clusters == k)].T)
-            plot_cluster_context(sizes, densities, k, clustername, dir)
-            plot_one_cluster(plotdata, plotdata_labels, (clusters==k), clustername, dir)
-
     # summary plots
     plot_cluster_context(sizes, densities, None, "all", dir)
     plot_one_cluster(plotdata, plotdata_labels, None, "all", dir)

### 2016-01-23 code/kmeans.py
condition bug

@@ -306,8 +306,8 @@ if __name__ == "__main__":
         plotcount = 0
         for k in (np.argsort(densities))[::-1]:
             clustername = "cluster_{:04d}_{:04d}".format(K, k)
-            print(clustername, sizes[k], logdets[k], densities[k], fehs[k])
-            if (not only_high_Z) or (fehs[k] > 0.5 and fehs[k] < 0.3):
+            if (not only_high_Z) or (fehs[k] > -0.5 and fehs[k] < 0.3):
+                print(clustername, sizes[k], logdets[k], densities[k], fehs[k])
                 plot_cluster_context(sizes, densities, k, clustername, dir)
                 plot_one_cluster(plotdata, plotdata_labels, (clusters==k), clustername, dir)
                 plotcount += 1

### 2016-01-23 code/kmeans.py
refactor density plotting

@@ -105,24 +105,46 @@ def plot_one_cluster(data, labels, mask, name, dir, suffix="png"):
     plt.close("all")
 
 def _clusterlims(sizes, densities):
-    plt.ylim(np.min(densities) / 5., np.max(densities) * 5.)
+    plt.ylim(np.min(densities) / 5., np.max(densities[np.isfinite(densities)]) * 5.)
     plt.xlim(np.min(sizes) / 1.4, np.max(sizes) * 1.4)
 
-def plot_cluster_context(sizes, densities, k, name, dir, suffix="png"):
+def _clusterplot(xs, ys, k, ms=5.0):
+    finite = np.isfinite(ys)
+    infinite = np.logical_not(finite)
+    ybig = 2. * np.max(ys[finite])
+    thisys = np.clip(ys, 0., ybig)
+    plt.plot(xs[finite], thisys[finite], marker=".", c="k",  ms=ms, alpha=0.5, ls="none")
+    plt.plot(xs[infinite], thisys[infinite], marker="^", c="k", ms=0.5*ms, alpha=0.5, ls="none")
+    if k is not None:
+        plt.plot(xs[k], thisys[k], marker="o", c="k", mfc="none", ms=1.5*ms, mew=2., alpha=1., ls="none")
+
+def plot_cluster_context(sizes, densities, dir, name=None, k=None, suffix="png"):
+    """
+    so many conditionals!
+    """
     print("plot_cluster_context(): plotting", name)
-    fn = "{}/{}_context.{}".format(dir, name, suffix)
+    if name is None:
+        K = len(sizes)
+        fn = "{}/clusters_{:04d}.{}".format(dir, K, suffix)
+    else:
+        fn = "{}/{}_context.{}".format(dir, name, suffix)
     if os.path.exists(fn):
         print("plot_cluster_context(): {} exists already".format(fn))
         return
-    fig = plt.figure(figsize=(4,4))
-    plt.subplots_adjust(left=0.2, right=0.96, bottom=0.2, top=0.96)
-    plt.clf()
     if k is None:
+        fig = plt.figure(figsize=(6,6))
+        plt.subplots_adjust(left=0.15, right=0.97, bottom=0.15, top=0.97)
+        ms = 7.5
+    else:
+        fig = plt.figure(figsize=(4,4))
+        plt.subplots_adjust(left=0.2, right=0.96, bottom=0.2, top=0.96)
+        ms = 5.0
+    plt.clf()
[…]

### 2016-01-23 papers/high_Z_clusters.tex
examples at higher metallicity

--- 2016-01-20 papers/clusters.tex
@@
 % Copyright 2016 the authors.
-
-% style notes
-% - never use \cite{}, always \citep{} or \citealt{}.
-% - never use et al., always use \etal
 
@@
 \newcommand{\insanefigure}[1]{%
-\includegraphics[width=\figwidth]{./figs/#1_MgFe.png}%
-\includegraphics[width=\figwidth]{./figs/#1_MgK.png}%
-\includegraphics[width=\figwidth]{./figs/#1_AlMg.png}\\
-\includegraphics[width=\figwidth]{./figs/#1_NaO.png}%
-\includegraphics[width=\figwidth]{./figs/#1_CN.png}%
-\includegraphics[width=\figwidth]{./figs/#1_SAl.png}\\
-\includegraphics[width=2\figwidth]{./figs/#1_gal.png}%
-\includegraphics[width=\figwidth]{./figs/#1_HR.png}\\
-\includegraphics[width=2\figwidth]{./figs/#1_vlon.png}%
-\includegraphics[width=\figwidth]{./figs/#1_context.png}}
+\includegraphics[width=\figwidth]{../code/kmeans_figs/#1_MgFe.png}%
+\includegraphics[width=\figwidth]{../code/kmeans_figs/#1_MgK.png}%
+\includegraphics[width=\figwidth]{../code/kmeans_figs/#1_AlMg.png}\\
+\includegraphics[width=\figwidth]{../code/kmeans_figs/#1_NaO.png}%
+\includegraphics[width=\figwidth]{../code/kmeans_figs/#1_CN.png}%
+\includegraphics[width=\figwidth]{../code/kmeans_figs/#1_SAl.png}\\
+\includegraphics[width=2\figwidth]{../code/kmeans_figs/#1_gal.png}%
+\includegraphics[width=\figwidth]{../code/kmeans_figs/#1_HR.png}\\
+\includegraphics[width=2\figwidth]{../code/kmeans_figs/#1_vlon.png}%
+\includegraphics[width=\figwidth]{../code/kmeans_figs/#1_context.png}}
 
@@
 
-\title{Chemical tagging can work: \\
-       Identification of stellar phase-space structures \\
-       purely by chemical-abundance similarity}
+\title{High-metallicity clusters}
 \author{David~W.~Hogg\altaffilmark{1,2,3,4},
@@
 % Context:
-Chemical tagging promises to use detailed abundance measurements to
-identify spatially separated stars that were in fact born together
-(in the same molecular cloud), long ago.
-This idea has not previously yielded scientific successes, probably because of the
-noise and incompleteness in chemical-abundance measurements.
-However, we have succeeded in substantially improving spectroscopic
-measurements with \thecannon, which has delivered $15$ individual
-abundances for $\sim 10^5$ stars observed as part of the
-\apogee\ spectroscopic survey, with precisions around 0.04~dex.
-% Aims:
-We test the chemical-tagging hypothesis by looking at clusters in abundance space
-and confirming that they are clustered in phase space.
-% Methods:
-We identify (by the k-means algorithm) overdensities of stars
-in the 15-dimensional chemical-abundance space delivered by \thecannon,
-and plot the associated stars in phase space.
-We use \emph{only} abundance-space information (no positional information) to identify stellar groups.
-% Results:
-We find that clusters in abundance space are indeed clusters in phase space.
-We recover some known phase-space clusters and find other interesting structures.
-This confirms the chemical-tagging hypothesis and verifies the precision of the abundance measurements delivered by \thecannon.
-This is the first-ever project to identify phase-space structures by blind search purely in abundance space;
-the prospects for future data sets are very good.
+Just some high-metallicity examples
 \end{abstract}
 
-\keywords{
-  Galaxy: abundances
-  ---
-  Galaxy: stellar content
-  ---
-  Galaxy: structure
-  ---
-  globular clusters: general
-  ---
-  open clusters and associations: general
-  ---
-  stars: abundances
-}
-
-\clearpage
-\section{Introduction}\label{sec:intro}
-
-Ensembles of stars in the Milky Way are born in molecular clouds,
-which are presumed to be near-homogeneous in their chemical element
-composition.
-However, most stars are born in unbound associations, or are born in
-star clusters that disperse rapidly; they will eventually end up in
-very different parts of phase space in the Galaxy.
-Yet, if every star preserved its photospheric element abundances over
-its lifetime (at least for most elements), then stars of common birth
-origin ought to be identifiable through their detailed photospheric
-abundances, long after any spatial proximity has
-vanished.
-
-This idea---dubbed ``chemical tagging''
-\citep{freeman}---is one of the principal motivations for a number
-of surveys, including \apogee\ \citep{apogee},
-\gaiaeso\ \citep{gaiaeso} and \galah\ \citep{galah}.
-In order to determine the precise abundance labels for chemical
-tagging, these surveys are each measuring high-resolution, high
-signal-to-noise ($S/N$) spectra for hundreds of thousands of stars across the
-Galaxy's disk, bulge and halo.
-
-Chemical tagging holds the promise of revealing not just the
-star-formation history of the Galaxy, but also the accretion history
-(as things that fall in are expected to be chemically distinct from
-those that form in the parent body; for example, \citealt{font}) and
-stellar-orbit diffusion process like radial mixing and radial
-migration (for example, \citealt{roskar, quillen}).  After
-stars are born---or after a star cluster is accreted and
-disrupted---associations or groups will disperse, through
-two-body processes, interactions with resonances, or tides from the whole Galaxy.
-
-Although undeniably promising---and motivating the 
-launch of costly large-scale spectroscopic surveys---
-chemical-tagging as a search technique has yet to be proven in
-practice: identifying stars of common birth origin {\it purely} on the
-basis of their near-identical abundances patterns, without any
-consideration of position or velocity.
-
-Part of the reason that chemical tagging remains unrealized
-is because the level of abundance specificity required
-is very high.
-If there are thousands of (relevant) molecular clouds forming stars in
-the recent history of the Milky Way, clumps of stars can only be
-identified in abundance space if abundance space is high in dimensionality.
-(In principle, it needs to be high in dimensionality both in terms
-of the number of measured abundances \emph{and} in terms of the number of
-nucleosynthetic pathways or the dimensionality of the true abundance
-space.)
-Therefore accurate---or at
-least precise---measurements of many different abundances are needed
-for stellar siblings to have sufficiently unique fingerprints.
-Stellar spectroscopic surveys now have the resolution,
-$S/N$, wavelength coverage, and sample sizes to deliver many
-different chemical tags for each star.
-
-There are, however, two big issues.
-The first is that the physical assumptions behind the idea may require
-refinement:
-There may be chemical-abundance overlaps among open clusters
-\citep{blancocuaresma}, coeval groups of stars may have similar tags
-but different birth places \citep{mitschang}, and the
-chemical-abundance space might be low in dimensionality.
-On the other hand, precise studies of stellar twins \citep{melendez, jofre}
-and open clusters \citep{bovy} indicate that chemical abundances ought to be
-very similar for co-eval stars.
-
-The second issue for chemical tagging---and the one we address
-here---is measurement precision (and accuracy).
-The current precision on abundance measurements in the published
-survey catalogs is not high enough (for example, \citealt{martell, ting}).
-However, our recent work \citep{thecannon} suggests that the \emph{data}
-are precise enough: There is enough $S/N$ at the relevant
-locations in spectrum space to deliver high-precision tags.
-The existence of very large data sets, homogeneous in spectrum
-space, suggests that data-driven approaches to the determination
-of stellar abundances might considerably outperform traditional
-methods.
-These traditional approaches are based on \foreign{ab-initio} physical models
-that have shortcomings that become apparent in this age of
-high-quality spectra, and the data-analysis methods do not make use of
-all of the information in the data sets.
-Improving the models and exploiting the entire information
-content in the data is critical if we are going to deliver useful
-chemical tags.
-
-Our specific contribution in this space has been to develop
-\thecannon\ \citep{thecannon, ages}, which is a data-driven model for
-stellar spectra.
-This model can deliver stellar parameters and chemical abundances for
-stars, making use of every pixel of every stellar spectrum (that is,
-all the information in the data) but making no use at all of physical
-models of stars.
-It relies only on there being training data---\emph{some} reference stars for
-which parameters and abundances are known and believed.
-In companion papers \citep{casey16, ness16} we show that
-\thecannon\ can deliver 15 to 19 abundances for stars in the
-\apogee\ survey, at precisions higher than even the training (reference) data on
-which the model is trained.
-We say more about this below.
-We will show here that \thecannon\ improves chemical abundance measurements
-to the point that \emph{chemical tagging is now possible.}
-
-One note on \emph{accuracy} and \emph{precision:}
-In principle, the problem of chemical tagging does not require
-absolute accuracy for chemical-abundance measurements.
-It only requires that we can precisely see that two stars are similar
-in their abundances, even if we happen to be wrong about the absolute values of
-those abundances.
-This point might make it seem like we don't care that our models are
-wrong, so long as they are \emph{consistent.}
-However, this is a bit misleading:
-For chemical tagging to succeed, we need stars with different
-atmospheric parameters $(\teff, \logg)$ but the same chemical
-abundances to be assigned the same position in chemical-abundance
-space.
-That doesn't require \emph{overall} accuracy, but it requires that the models
-have the right dependencies on atmospheric parameters such that the
-wrongness in abundance space is consistent across the Hertzsprung-Russell diagram.
-That is, we need a substantial amount of accuracy to achieve our goals.
-
-It is important to realize that analysis techniques based on \foreign{ab-initio}
-physical models are inaccurate, yet they too strive to improve precision
-in the face of knowingly inaccurate models.  Incomplete atomic data, 
-simplifications of photospheric structure, assumptions about convective
-motion, and inconsistencies resulting from positing local thermal
-equilibrium all contribute to produce inaccurate abundances.
-\thecannon\ stands out because it demonstrably improves the precision on
-chemical abundances, whereas the accuracy of those labels is limited only
-by the training (reference) set employed.
-
-In what follows, we are going to use \apogee\ \acronym{DR12} data \citep{dr12}, in
-which we can re-derive 15 element abundances (C, N, O, Na, Mg, Al, Si,
-S, K, Ca, Ti, V, Mn, Fe, Ni), using \thecannon\ \citep{thecannon}.
-The \apogee\ data set covers a huge radial extent and---because the
-data are taken in the infrared---explores the thin disk.
-However, it has the disadvantage that its spatial coverage is incomplete
-(that is, limited pointings produce a fractured sky map),
-which makes it hard to see, within the data set, linear or
-extended stellar structures.
-In many ways, \galah\ will deliver improvements, both because it will
-have more abundances (possibly 29) and contiguous sky coverage.
-
-One final note:
-we think of this \documentname\ as performing a proof of concept.
-We know that the stellar members of open and globular clusters---stars
-that are identified by being close in phase space---contain highly
-informative abundance information that identifies them also in
-chemical-abundance space.  Does this work the other way around? 
-Can chemical tagging identify small subsets of stars, among a vastly
-greater background sample, that have a common birth origin? If we 
-find stars purely by their clustering in abundance space and subsequently
-show that they are part of a still spatially coherent cluster, then
-we will have resolved all practical outstanding issues plaguing 
-chemical tagging, thereby bringing us much closer to
-unravelling the formation of the Milky Way.
-
-\section{Data: \apogee\ giants with abundances from \thecannon}\label{sec:data}
-
-Our analysis draws on the spectra of \totalnumber\ giant stars ($\logg
-< 3.9$) from \apogee\ \acronym{DR12} \citep{dr12}, with no warning flags
-set in the \apogee\ \aspcap\ \citep{aspcap} pipeline
-reductions.
-We re-analyze these \emph{spectra} using \thecannon, because it can
-deliver stellar abundance labels of higher precision, especially for
-stars of $S/N \le 150$.
-The details about how we select, reduce, and analyze the \apogee\ data
-are given in full detail in the companion papers \citep{casey16,
-  ness16}, and we only summarize briefly here:
-We re-derived 17 labels ($\teff$, $\logg$, and 15 abundances
-referenced to H) from the \apogee\ \acronym{DR12} spectra. 
-For the training step of \thecannon, 
-12,681 red-giant stars with spectral $S/N \ge 200$ were used.
-For this small fraction of the sample with the highest $S/N$, the
-labels provided by \aspcap\ provide consistent, relatively low-scatter, sensible
-abundance-space measurements.
-The training step fixes a spectral model
-that predicts the normalized spectrum as a quadratic function of all
-labels ($\sim 1.5\times10^6$ model coefficients). In the second stage---the test step---each unlabelled star is
-used to establish a single-star likelihood function for its labels,
-holding the spectral model coefficients (the parameters of the model) fixed.
-\thecannon\ finds the labels that minimize the single-star likelihood
-function for each test-set star.
-This optimization is not convex, but it is embarrassingly parallel,
-and therefore fast. The test
-step for 150,000 spectra takes less than 30~minutes on a small 
-research cluster in Cambridge.
-
-Six two-dimensional projections of the abundance data, plus some other
-data quantities, are shown in \figurename~\ref{fig:all}.  Throughout
-this work we show two-dimensional abundance projections for only a few
-elements. The entire sequence of 15-choose-2 combinations is too
-immense to visualize. The projections shown are ones which
-have been extensively used in abundance studies of globular clusters
-and satellite systems. These are typically light element abundances,
-and our projections include C--N, Na--O, and Mg--Al. Globular clusters
-famously demonstrate correlations in these elements, thereby making
-them suitable projections for us to show in verifying and examining
-any substructure identified by k-means (or any other clustering algorithm).
-
 \begin{figure}[!p]
-\insanefigure{all}
-\caption{The full sample of \totalnumber\ stars used in this study.
-  The top six panels show six two-dimensional projections of the
-  empirical abundance space distribution.
-  The clustering algorithm (described in
-  \sectionname~\ref{sec:method}) performs clustering in a full
-  15-dimensional abundance space of elements referenced to H (not Fe),
-  but plots are shown here referenced to Fe for familiarity reasons.
-  The bottom three panels show stellar meta-data not used in the
-  clustering described below.\label{fig:all}}
+\insanefigure{cluster_1024_0068}
+\caption{$K=1024$ abundance-space cluster $68$.}
+\end{figure}
+\begin{figure}[!p]
+\insanefigure{cluster_1024_0116}
+\caption{$K=1024$ abundance-space cluster $116$.}
+\end{figure}
+\begin{figure}[!p]
+\insanefigure{cluster_1024_0318}
+\caption{$K=1024$ abundance-space cluster $318$.}
+\end{figure}
+\begin{figure}[!p]
+\insanefigure{cluster_1024_0595}
+\caption{$K=1024$ abundance-space cluster $595$.}
+\end{figure}
+\begin{figure}[!p]
+\insanefigure{cluster_1024_0595}
+\caption{$K=1024$ abundance-space cluster $595$.}
 \end{figure}
 
-\clearpage
-\section{Identifying abundance-space clusters}\label{sec:method}
-
-Although only hints of small-scale clustering in the abundance space
-are visible in \figurename~\ref{fig:all}, exploration of the data by hand indicates that
-known clusters do appear in the high-dimensional abundance space as over-densities.
-In general, the collapse of 15 dimensions down to 2-d projections
-will hide, smooth, or dilute any structure; there is no guarantee that
-a highly featured distribution in 15-d will show features in
-\emph{any} 2-d projection, let alone axis-aligned, human-selected projections.
-This encourages us to look for over-densities automatically in the
-abundance space and see if anything found that way would be over-dense
-in phase space.
-The simplest method for clustering points in $D$-dimensional space is
-the \emph{k-means} algorithm (see \citealt{bishop} for a pedagogical
-introduction and references to the original literature).
-
-Briefly, the k-means algorithm is the following:
-$\bullet$~Start at some initial guess for the locations of $K$ $D$-space cluster centers ($K$
-locations in the $D$-dimensional space).
-$\bullet$~Assign each point in the
-space (each star in our case) to the closest of the $K$ centers.
-$\bullet$~Given this assignment of stars to centers, update each center (each of
-the $K$ $D$-space positions) by taking the mean of the locations of
-the stars assigned to that center.
-$\bullet$~Iterate these steps (assignment of points, followed by taking of means) to convergence.
-The output of this algorithm is the converged locations of the $K$ centers and the assignments of
-all points to those centers.
-This algorithm is fast and performs well in practice in problems of
-this nature; also we are not the first to use the k-means algorithm in
-abundance space \citep{gratton}.
-
-The k-means algorithm has a number of limitations:
-One is that $K$ must be chosen by hand (or heuristically at best).
-Here we are only demonstrating a concept; we don't need to have the
-best possible clustering.
-For this reason we simply choose $K=128$, $K=256$, and $K=512$ and look at all the
-results.
-Also, the k-means algorithm only performs local optimization.
-At each $K$ we perform 32 restarts with different initializations, and
-preserve the best clustering (best according to the k-means score).
-Each of the 32 initializations was performed with the
-\project{scikit-learn} standard k-means initialization procedure
-\citep{sklearn}.
-
-Another issue with k-means is that it effectively uses metric
-distances in the $D$-space; it presumes Euclidean isotropy.
-We choose here to work in the hydrogen-normalized abundance space, the
-space of [C/H], [N/H], [O/H], [Na/H], [Mg/H], [Al/H], [Si/H], [S/H],
-[K/H], [Ca/H], [Ti/H], [V/H], [Mn/H], [Fe/H], [Ni/H].
-But in addition to this, we re-scale these by approximate typical measurement
-precisions obtained by \thecannon\ before running k-means.
-These scalings were
-  [C/H]/0.041,
-  [N/H]/0.044,
-  [O/H]/0.037,
-  [Na/H]/0.111,
-  [Mg/H]/0.032,
-  [Al/H]/0.055,
-  [Si/H]/0.041,
-  [S/H]/0.054,
-  [K/H]/0.069,
-  [Ca/H]/0.043,
-  [Ti/H]/0.072,
-  [V/H]/0.146,
-  [Mn/H]/0.041,
-  [Fe/H]/0.019, and
-  [Ni/H]/0.034.
-This scaling makes the space close to isotropic in measurement uncertainty or
-observational precision.
-Importantly we use \emph{only} abundance-space information, and no
-positional or velocity information (nor $\teff$ nor $\logg$ nor any
-targeting or observational meta-data) as input to the clustering
-algorithm.
-
-Nothing about this algorithm or our choices are particularly
-tuned or optimized; this is in no sense the
-algorithm or the method that reveals the best structures.
-Indeed, it is a feature of this work that even the simplest, most
-generic clustering algorithm returns interesting
-structures (as shown below).
-The prospects for Milky Way science only improve as the clustering
-algorithm improves.
-
-When the k-means results are returned, the $D\times D$ empirical
-variance tensor for the members of each cluster can be constructed.
-From this, an effective density in the abundance space can be computed
-as the number of points in the cluster divided by the square-root of
-the determinant of the tensor.
-This density was used to rank abundance-space overdensities (from highest
-to lowest) for visual inspection.
-In \figurename~\ref{fig:densities} we show the distribution of cluster membership
-and density for the $K=256$ run of k-means.
-The densest clusters tend to have fewer members, presumably because
-the bulk of the population is not extremely strongly clustered, or not
-in ways to which k-means is sensitive.
-
-Importantly, the k-means algorithm assigns \emph{every} star to at
-least one cluster.
-For this reason, there is no sense in which \emph{every} ``cluster'' returned
-by k-means is a distinct overdensity in abundance space.
-In what follows, we only consider high-density clusters---clusters
-that are much more dense than average; these ought to represent true
-over-densities in abundance space.
-
-We chose a few interesting cases from the high-density clusters
-in the $K=256$ run and show them in
-\figurename~\ref{fig:M13}, \ref{fig:Pal5}, \ref{fig:Sgr},
-\ref{fig:halo}, and \ref{fig:disk}.
-The first three of these are dominated by stars in the known clusters
-M13, Pal5, and Sagittarius.
-The fourth is a halo structure with high velocity dispersion and
-possibly accreted.
-The fifth is a thin-disk star-formation feature.
-We will discuss the astrophysical implications of these results in
-\sectionname~\ref{sec:discussion}.
-
-The abundance-space clusters shown in the \figurename s
-might not look dense in 2-d projections of the abundance space, but they
-are very dense in the 15-dimensional space.  Indeed,
-\figurename~\ref{fig:M13} is the densest cluster found in the 15-d
-space \emph{by far}.
-The challenge of this work is to find structure in the high
-dimensionality space that is not obviously visible in any 2-d projection.
-Although we have by no means any model of the 15-d space, we do know
-that the structures shown in the figures are very high in 15-d
-density.
-
-We have only shown results from the $K=256$ run of k-means.
-This run was chosen because its densest abundance-space clusters map
-well onto known stellar clusters.
-At $K=128$ the densest abundance-space clusters tended to combine
-multiple known stellar clusters into single, large abundance-space
-groupings.
-At $K=512$ k-means tended to split even mono-abundance groups into
-smaller sub-groups.
-This is a reminder that k-means has been chosen just for simplicity
-here; it is by no means matched to the discovery of
-stellar structures.
-An important follow-up project is to build a model of abundance space
-that captures the features expected from stellar populations (and observational uncertainties).
-Even if we were able to tune k-means in some sense ``perfectly'', it
-would still break up many stellar clusters, since they often show
-multiple stellar populations and non-linear correlations between element abundances.
-It is also the case that each of the abundance-space clusters we
-\emph{do} show in the \figurename s includes both star-cluster members
-and some background contamination, and also is missing some true
-star-cluster members.
-
-Although we show only three stellar clusters in
-\figurename~\ref{fig:M13}, \ref{fig:Pal5}, and \ref{fig:Sgr}, many
-other clusters are visible among our densest abundance-space k-means
-clusters.
-These include M15, M92, and M107, among others.
-We haven't asked yet whether we detect---as abundance-space
-overdensities---all of the stellar clusters we expect to find. (Many
-clusters within the \apogee\ data set only contain a few plausible 
-members, thereby  complicating any inferences we wish to make about 
-detection completeness using simple k-means.) 
-This \documentname\ is simply a demonstration of the chemical-tagging
-concept; a full investigation of whether we can construct complete
-catalogs of stellar clusters is beyond our present scope.
-
-\begin{figure}[!bp]
-\includegraphics[width=1.5\figwidth]{./figs/clusters_0256.png}
-\caption{The distribution of membership and density for the 256
-  abundance-space clusters returned by the k-means algorithm at
-  $K=256$.  The densest cluster (which is displayed in more detail in 
-  \figurename~\ref{fig:M13}) is far more dense than the second
-  densest.\label{fig:densities}}
-\end{figure}
-\begin{figure}[!p]
-\insanefigure{cluster_0256_0253}
-\caption{Same as \figurename~\ref{fig:all} except that the entire
-  sample has been made gray and the members of the most dense
-  $K=256$ abundance-space cluster have been rendered as unique, prominent triangles
-  (color is velocity rank, orientation is $\logg$ rank).
-  The lower-right plot shows this cluster (circle) in context of the other clusters (dots).
-  This cluster, which was identified only in abundance space (six
-  projections of which are the top six panels in this \figurename),
-  turns out to be dominated by the halo globular cluster
-  M13.\label{fig:M13}}
-\end{figure}
-\begin{figure}[!p]
-\insanefigure{cluster_0256_0034}
-\caption{Same as \figurename~\ref{fig:M13} but for another dense
-  abundance-space cluster.
-  This one turns out to be dominated by the disrupting globular
-  cluster Palomar 5.\label{fig:Pal5}}
-\end{figure}
-\begin{figure}[!p]
-\insanefigure{cluster_0256_0177}
-\caption{Same as \figurename~\ref{fig:M13} but for another dense
-  abundance-space cluster.
-  This one turns out to be dominated by the Sagittarius dwarf spheroidal galaxy.\label{fig:Sgr}}
-\end{figure}
-\begin{figure}[!p]
-\insanefigure{cluster_0256_0010}
-\caption{Same as \figurename~\ref{fig:M13} but for another dense
-  abundance-space cluster.
-  This one turns out to be dominated by a hitherto unrecognized high-velocity-dispersion
-  structure in the Galaxy halo.\label{fig:halo}}
-\end{figure}
-\begin{figure}[!p]
-\insanefigure{cluster_0256_0141}
-\caption{Same as \figurename~\ref{fig:M13} but for another dense
-  abundance-space cluster.
-  This one turns out to be dominated by a very thin young stellar
-  structure in the Galaxy disk, probably one that has been discovered previously \citep{wegg}.\label{fig:disk}}
-\end{figure}
-
-\clearpage
-\section{Discussion}\label{sec:discussion}
-
-We have demonstrated for the first time that, in a large-scale
-spectroscopic survey, stars that appear similar in
-chemical-abundance properties are often associated physically in
-clusters or other structures.
-The reverse is well established---that is, that stars that are
-associated physically in clusters are similar in chemical-abundance
-properties.
-That is, although they often show chemical diversity, that chemical
-diversity is small, and follows well-defined trends or a subspace in
-abundance space (for example, \citealt{gratton, meszaros, bovy},
-and references within those).
-However, it was not known---prior to this work---whether abundance-space information would be
-informative enough, or \emph{measurable} precisely enough to find this
-structure in surveys dominated by a large mix of stars from different
-origins and ages (and there was good reason to be pessimistic;
-\citealt{ting}).
-We have demonstrated with a straightforward (and in no way optimal)
-clustering in chemical-abundance space that it is easy to find co-eval
-stellar structures.
-Although this is very strong evidence that chemical tagging will work,
-we have not performed, in any sense, the canonical procedure of
-looking for features that are consistent with being delta functions in
-chemical-abundance space (convolved with a measurement noise
-distribution); we have looked only for over-densities, not delta
-functions.
-
-What made our success here possible is substantial improvements in precision
-delivered by \thecannon.
-We don't have an absolute measure of the precision of the measurements,
-but from looking at open and globular clusters, it appears to be on the
-order of or better than 0.04~dex for the median element in the list of 15.
-These improvements are discussed in detail in the companion papers
-\citep{casey16, ness16} about the abundance measurements.
-We believe these improvements come from a combination of factors, not
-limited to (a)~improvements in the determination of pseudo-continuum,
-(b)~the use of more spectral range than just unblended element windows
-\citep{aspcap}, and (c)~accurate spectral predictions
-(\thecannon\ delivers accurate predictions because it is fit to
-observed data). 
-Furthermore, because abundance space is high in dimensionality, even
-small improvements in abundance measurement get taken to a significant
-power when thinking about information gains.
-
-There are a few points to highlight about the abundance anti-correlations
-found in globular clusters. First, the detection of these signatures gives
-confirmation to our analysis: The training (reference) set was not purposefully selected
-to contain members of any globular or open cluster. However amongst the test
-set, we recover globular cluster members and their anti-correlations 
-in chemical abundances. That is to
-say we recover peculiar chemical abundances that are not dominant in our
-training set, but are expected from previous studies. Secondly, the very
-presence of complex abundance correlations makes the detection of clusters
-(by k-means) substantially more difficult. Briefly, the k-means algorithm
-is most effective for near-normally distributed, isotropic clusters in high-dimensional space.
-While globular clusters often show distributions of this kind in a few dimensions,
-they also demonstrate non-linear correlations which would be
-sub-optimally selected by k-means. For these reasons the fact that (a)~\thecannon\
-reports these abundance correlations and (b)~k-means (as a simple algorithm
-without information about expectations in abundance space) does not appear to be
-severely impacted by these correlations, gives us great hope for more sophisticated approaches.
-
-The abundances of known clusters that we identify by k-means are in
-excellent agreement with the literature. For the group shown in \figurename~\ref{fig:M13},
-which we attribute to M13, we find a mean of [Fe/H]$ = -1.55$.
-The mean and spread agree well with other work on this cluster
-\citep{Kraft_1992,Cohen_Melendez_2005,Johnson_Pilachowski_2012}, and compiled catalogs of
-globular cluster properties \citep[][accessed 2016]{Harris_1996}. The detailed abundances are also
-consistent: we find a correlation in [C/Fe] and [N/Fe] abundances \citep{Smith_2005}, and their
-projection in \figurename~\ref{fig:M13} show a spread that reflects deep mixing
-along the red giant branch \citep{Briley_2004}. We find that the light element
-abundances (most notably Mg--Al) for this cluster are anti-correlated, as
-expected from other studies \citep[for example,][]{gratton}. However we find a
-smaller spread in [O/Fe] than reported by others who have looked exclusively
-at M13 \citep{Johnson_Pilachowski_2012}.  As we discuss above, this is expected:
-k-means is by no means optimal for identifying arbitrary-shaped structures
-in high-dimensional space, and the first stars that would be assigned to another k-means
-cluster would be those with the most extreme abundances: low [O/Fe] and
-high [Na/Fe].
-
-There are scant abundance measurements reported for Palomar 5, for which
-most stars in the cluster in \figurename~\ref{fig:Pal5} belong. We find a mean
-metallicity of [Fe/H]$ = -1.3$, which lies between the only
-photometric \citep{Geisler_1997} and spectroscopic \citep{Smith_1985,Smith_2002} abundance
-measurements for this cluster.  The anti-correlation in C--N and Na--O abundance
-have been seen elsewhere \citep[][respectively]{Smith_1985,Smith_2002}, although
-to our knowledge the Mg--Al anti-correlation has not previously been shown for this cluster.
-Indeed, this work contains the largest number of abundances for Palomar 5,
-permitting a thorough search of tidally disrupted members \citep[for example,][]{Kuzma_2015} 
-that have retained the unique cluster abundance signature.
-
-The high velocity-dispersion structure seen in \figurename~\ref{fig:halo} stands
-in stark contrast to the rest of the clusters we have identified. In
-addition to being highly clustered in 15-dimensions, it is visibly
-clustered in our 2-d projections. Indeed, it is more clustered in our
-abundance projections than the young stellar structure in the disk,
-and comparably so to the other known globular clusters that we have
-shown here. The disparity between this structure and others presented
-here is the lack of co-spatial stars. 12 stars within this cluster are
-marked as `M53' candidates by \apogee, but the remaining 145 stars
-are spread amongst fields throughout the halo.
-
-The k-means algorithm employed here is by no means optimal, and other clusters
-we identify are accompanied by a few stars that are not currently co-spatial.
-Those stars may be unassociated interlopers that have been misclassified,
-or they may be true cluster members that are now unbound. However in the
-case of the high-velocity-dispersion structure in \figurename~\ref{fig:halo},
-the situation is far more extreme. We identify a group of stars with
-very similar abundances in 15 dimensions that are now spread throughout
-the Galaxy halo.
-Because some stars ($<8\%$) are candidate members of the massive globular
-cluster M53, it provides tantalizing possibility that these stars may
-indeed be all members of the same co-natal gas cloud, implying that 
-these stars have been accreted onto the halo by a single globular
-cluster early in the Milky Way's formation. While speculative,
-this idea demonstrates the promise of chemical tagging.
-
-The abundance-space group of stars in the disk structure shown in
-\figurename~\ref{fig:disk} may be associated with a very thin
-Milky-Way bar component reported previously \citep{wegg}, which is
-found to exist predominantly toward the end of the bar (at $l$
-$\approx 30$~deg).
-The stars in this structure are metal-rich and have a low
-[C/N] ratio and are therefore likely young (following the [C/N]--mass
-relationship; \citealt{martig}).
-It is possible that this structure has very low scale height because
-it is very young and hasn't experienced dynamical heating or radial
-migration.
-
-It is worth reiterating here what is written above:
-There is no sense at all in which \figurename~\ref{fig:M13},
-\ref{fig:Pal5}, \ref{fig:Sgr}, \ref{fig:halo}, and \ref{fig:disk} show
-a representative or complete set of features found in abundance space.
-These features were hand-chosen to be obviously interesting and
-interpetable.
-There are many other things to be found in this data set, and many
-more effective models to be built of abundance-space structure.
-The only strong conclusion of this investigation is that chemical
-tagging \emph{can} work; the abundance measurements with \thecannon\ are
-precise enough and the chemical-abundance space is informative enough.
-
-The \apogee\ project shows great promise for these studies, and with
-the appearance of the companion papers, we will release the
-chemical-abundance measurements we used here for further study.
-The future is even brighter, however: \apogee\ is expanding to 19
-elements, more stars, and all-sky angular coverage with \apogee2, and
-\galah\ is nearly ready to release chemical abundances on a larger set
-of 29 elements.
-The \apogee\ elements employed in this project include alpha
-(Mg, O, Si, Ca, Ti), light proton-capture (C, N, S), odd-Z (Na, Al, K)
-and iron peak (Fe, Ni, V, Mn) elements.
-\galah\ will deliver element abundances from all the major
-nucleosynthetic processes, including light proton capture elements,
-alpha, odd-Z, iron-peak, light and heavy slow neutron capture, and
-rapid neutron capture.
-Chemical tagging capabilities are expected to grow as the
-\emph{product} of the measured nucleosynthetic pathways.
-
-\acknowledgements
-It is a pleasure to thank
-  Anna~Y.~Q.~Ho (Caltech),
-  Jeremy~Magland (\acronym{SCDA}), and
-  the Blanton--Hogg group meeting
-for valuable comments and discussions.
-This project was funded in part by
-  the \acronym{NSF} (grants \acronym{IIS-1124794}, \acronym{AST-1517237}),
-  \acronym{NASA} (grant \acronym{NNX12AI50G}),
-  the Moore-Sloan Data Science Environment at \acronym{NYU}, and
-  the European Research Council under the
-  European Union's Seventh Framework Programme (FP~7)
-  \acronym{ERC} Grant Agreement n.~\acronym{[320360, 321035]}).
-This research made use of the \acronym{NASA} \project{Astrophysics Data System}.
-
-This project made use of \sdssiii\ data.
-Funding for \sdssiii\ has been provided by the Alfred P. Sloan
-Foundation, the Participating Institutions, the National Science
-Foundation, and the \acronym{U.S.} Department of Energy Office of Science. The
-\sdssiii\ web site is http://www.sdss3.org/.
-
-\sdssiii\ is managed by the Astrophysical Research Consortium for the
-Participating Institutions of the \sdssiii\ Collaboration including the
-University of Arizona, the Brazilian Participation Group, Brookhaven
-National Laboratory, Carnegie Mellon University, University of
-Florida, the French Participation Group, the German Participation
-Group, Harvard University, the Instituto de Astrofisica de Canarias,
-the Michigan State/Notre Dame/\acronym{JINA} Participation Group, Johns Hopkins
-University, Lawrence Berkeley National Laboratory, Max Planck
-Institute for Astrophysics, Max Planck Institute for Extraterrestrial
-Physics, New Mexico State University, New York University, Ohio State
-University, Pennsylvania State University, University of Portsmouth,
-Princeton University, the Spanish Participation Group, University of
-Tokyo, University of Utah, Vanderbilt University, University of
-Virginia, University of Washington, and Yale University.
-
-All of the code for the data manipulation, clustering, and
-visualization is built on \project{numpy} \citep{numpy},
-\project{scikit-learn} \citep{sklearn}, and \project{matplotlib}
-\citep{matplotlib}.
-All of the code written specifically for this project is available in
-two open-source code repositories at
-\url{https://github.com/andycasey/AnniesLasso} and
-\url{https://github.com/davidwhogg/Platypus}.
-
-\clearpage
-\begin{thebibliography}{24}\raggedright
-\bibitem[Alam \etal(2015)]{dr12}
-Alam,~S., Albareti,~F.~D., Allende~Prieto,~C., \etal, 2015, \apjs, 219, 12
-\bibitem[Bishop(2006)]{bishop}
-Bishop,~C.~M., 2006, \textit{Pattern Recognition and Machine Learning}, Springer, New York
-\bibitem[Blanco-Cuaresma \etal(2015)]{blancocuaresma}
-Blanco-Cuaresma,~S., Soubiran,~C., Heiter,~U., \etal, 2015, \aap, 577, A47
-\bibitem[Bovy(2015)]{bovy}
-Bovy,~J., 2015, arXiv:1510.06745
- \bibitem[Briley \etal(2004)]{Briley_2004}
-Briley, M.~M., Cohen, J.~G., Stetson, P.~B., 2004, \aj, 127, 1579 
-\bibitem[Casey \etal(2016)]{casey16}
-Casey,~A.~R., Hogg,~D.~W., Ness,~M., Rix,~H.-W., 2016, in preparation
-\bibitem[Cohen \& Mel{\'e}ndez(2005)]{Cohen_Melendez_2005}
-Cohen, J.~G.,  Mel{\'e}ndez, J., 2005, \aj, 129, 303 
-\bibitem[De~Silva \etal(2015)]{galah}
-De~Silva,~G.~M., Freeman,~K.~C., Bland-Hawthorn,~J., \etal, 2015, \mnras, 449, 2604
-\bibitem[Font \etal(2006)]{font}
-Font,~A.~S., Johnston,~K.~V., Bullock,~J.~S., Robertson,~B.~E., 2006, \apj, 638, 585 
-\bibitem[Freeman \& Bland-Hawthorn(2002)]{freeman}
-Freeman,~K., Bland-Hawthorn,~J., 2002, \araa, 40, 487
-\bibitem[Garc{\'{\i}}a~P{\'e}rez \etal(2015)]{aspcap}
-Garc{\'{\i}}a~P{\'e}rez,~A.~E., Allende~Prieto,~C., Holtzman,~J.~A., \etal, 2015, arXiv:1510.07635
-\bibitem[Geisler \etal(1997)]{Geisler_1997}
-Geisler, D., Claria, J.~J., Minniti, D., 1997, \pasp, 109, 799 
-\bibitem[Gilmore \etal(2012)]{gaiaeso}
-Gilmore,~G., Randich,~S., Asplund,~M., \etal, 2012, The Messenger, 147, 25 
-\bibitem[Gratton \etal(2012)]{gratton}
-Gratton,~R.~G., Carretta,~E., Bragaglia,~A., 2012, \aapr, 20, 50
-\bibitem[Harris(1996)]{Harris_1996}
-Harris,~W.~E., 1996, \aj, 112, 1487 
-\bibitem[Hunter(2007)]{matplotlib}
-Hunter,~J.~D., 2007, Computing in Science \& Engineering, 9, 90
-\bibitem[Jofr{\'e} \etal(2015)]{jofre}
-Jofr{\'e},~P., M{\"a}dler,~T., Gilmore,~G., \etal, 2015, \mnras, 453, 1428 
-\bibitem[Johnson \& Pilachowski(2012)]{Johnson_Pilachowski_2012}
-Johnson, C.~I.,  Pilachowski, C.~A., 2012, \apjl, 754, L38 
-\bibitem[Kraft \etal(1992)]{Kraft_1992}
-Kraft, R.~P., Sneden, C., Langer, G.~E.,  Prosser, C.~F., 1992, \aj, 104, 645 
-\bibitem[Kuzma \etal(2015)]{Kuzma_2015}
-Kuzma, P.~B., Da Costa, G.~S., Keller, S.~C., Maunder, E., 2015, \mnras, 446, 3297 
-\bibitem[Majewski \etal(2015)]{apogee}
-Majewski,~S.~R., Schiavon,~R.~P., Frinchaboy,~P.~M., \etal, 2015, arXiv:1509.05420
-\bibitem[Martell(2015)]{martell}
-Martell,~S.~L., 2015, arXiv:1507.00079
-\bibitem[Martig \etal(2016)]{martig}
-Martig,~M., Fouesneau,~M., Rix,~H.-W., \etal, 2016, \mnras, 456, 3655
-\bibitem[Melendez \etal(2014)]{melendez}
-Mel{\'e}ndez,~J., Ram{\'{\i}}rez,~I., Karakas,~A.~I., \etal, 2014, \apj, 791, 14
-\bibitem[M{\'e}sz{\'a}ros \etal(2015)]{meszaros}
-M{\'e}sz{\'a}ros,~S., Martell,~S.~L., Shetrone,~M., \etal, 2015, \aj, 149, 153
-\bibitem[Mitschang \etal(2014)]{mitschang}
-Mitschang,~A.~W., De~Silva,~G., Zucker,~D.~B., \etal, 2014, \mnras, 438, 2753
-\bibitem[Ness \etal(2015)]{thecannon}
-Ness,~M., Hogg,~D.~W., Rix,~H.-W., Ho,~A.~Y.~Q., Zasowski,~G., 2015, \apj, 808, 16
-\bibitem[Ness \etal(2016a)]{ages}
-Ness,~M., Hogg,~D.~W., Rix,~H.-W., \etal, 2016a, \apj, in press (arXiv:1511.08204)
-\bibitem[Ness \etal(2016b)]{ness16}
-Ness,~M., Hogg,~D.~W., Casey,~A.~R., Rix,~H.-W., 2016b, in preparation
-\bibitem[Pedregosa \etal(2011)]{sklearn}
-Pedregosa,~F., Varoquaux,~G., Gramfort,~A., \etal, 2011, Journal of Machine Learning Research, 12, 2825
-\bibitem[Quillen \etal(2015)]{quillen}
-Quillen,~A.~C., Anguiano,~B., De~Silva,~G., \etal, 2015, \mnras, 450, 2354
-\bibitem[Ro{\v s}kar \etal(2008)]{roskar}
-Ro{\v s}kar,~R., Debattista,~V.~P., Quinn,~T.~R., Stinson,~G.~S., Wadsley,~J., 2008, \apjl, 684, L79
-\bibitem[Smith(1985)]{Smith_1985}
-Smith, G.~H., 1985, \apj, 298, 249 
-\bibitem[Smith \etal(2002)]{Smith_2002}
-Smith, G.~H., Sneden, C., Kraft, R.~P., 2002, \aj, 123, 1502 
-\bibitem[Smith \etal(2005)]{Smith_2005}
-Smith, G.~H., Briley,~M.~M.,  Harbeck, D., 2005, \aj, 129, 1589
-\bibitem[Ting \etal(2016)]{ting}
-Ting,~Y.-S., Conroy,~C., Rix,~H.-W., 2016, \apj, 816, 10
-\bibitem[van~der~Walt \etal(2011)]{numpy}
-van~der~Walt,~S., Colbert,~S.~C., Varoquaux,~G., 2011, Computing in Science \& Engineering, 13, 22
-\bibitem[Wegg \etal(2015)]{wegg}
-Wegg,~C., Gerhard,~O., Portail,~M., 2015, \mnras, 450, 4050
-\end{thebibliography}
 \end{document}

### 2016-01-23 code/kmeans.py
plot more

@@ -313,7 +313,7 @@ if __name__ == "__main__":
                 plot_cluster_context(sizes, densities, dir, k=k, name=clustername)
                 plot_one_cluster(plotdata, plotdata_labels, (clusters==k), clustername, dir)
                 plotcount += 1
-                if (not plot_everything) and (plotcount >= 32):
+                if (not plot_everything) and (plotcount >= 64):
                     break
 
     # summary plots

### 2016-01-23 code/kmeans.py
looking at 2048

@@ -162,6 +162,7 @@ if __name__ == "__main__":
     plotdata = None
     Ks = 2 ** np.arange(8, 12) # do everything
     # Ks = [256, ] # just do the winner
+    Ks = [2048, ] # just do the craziest
     plot_everything = False # set to true only for MKN atlas
     only_high_Z = True # set to true to only plot things in the metallicity bulk
 
@@ -313,7 +314,7 @@ if __name__ == "__main__":
                 plot_cluster_context(sizes, densities, dir, k=k, name=clustername)
                 plot_one_cluster(plotdata, plotdata_labels, (clusters==k), clustername, dir)
                 plotcount += 1
-                if (not plot_everything) and (plotcount >= 64):
+                if (not plot_everything) and (plotcount >= 128):
                     break
 
     # summary plots

### 2016-01-25 code/kmeans.py
looking for an open cluster

@@ -161,8 +161,8 @@ if __name__ == "__main__":
     data = None
     plotdata = None
     Ks = 2 ** np.arange(8, 12) # do everything
-    # Ks = [256, ] # just do the winner
-    Ks = [2048, ] # just do the craziest
+    Ks = [256, ] # just do the winner
+    # Ks = [2048, ] # just do the craziest
     plot_everything = False # set to true only for MKN atlas
     only_high_Z = True # set to true to only plot things in the metallicity bulk
 
@@ -232,6 +232,36 @@ if __name__ == "__main__":
                    "AL_H", "CA_H", "C_H", "K_H",  "MG_H", "MN_H", "NA_H",
                    "NI_H", "N_H",  "O_H", "SI_H", "S_H",  "TI_H", "V_H"]
 
+    NGC6819_names = ["2M19403684+4015172",
+                     "2M19405601+4013395",
+                     "2M19405797+4008174",
+                     "2M19410203+4006280",
+                     "2M19410622+4010532",
+                     "2M19410858+4013299",
+                     "2M19410926+4014436",
+                     "2M19410994+4009056",
+                     "2M19411355+4012205",
+                     "2M19411476+4011008",
+                     "2M19411631+4005508",
+                     "2M19412147+4013573",
+                     "2M19412386+4021444",
+                     "2M19412953+4012210",
+                     "2M19413027+4015218",
+                     "2M19413444+4008462",
+                     "2M19415064+4016010",
+                     "2M19404803+4008085",
+                     "2M19404965+4014313",
+                     "2M19411319+4014567",
+                     "2M19411367+4003382",
+                     "2M19411705+4010517",
+                     "2M19412176+4012111",
+                     "2M19412222+4016442",
+                     "2M19412707+4012283",
+                     "2M19413031+4009005",
+                     "2M19413330+4012349",
+                     "2M19413439+4017482",
+                     "2M19414427+4005527"]
+
     # get ready...
     if not os.path.exists(dir):
         os.mkdir(dir)
@@ -298,12 +328,19 @@ if __name__ == "__main__":
             for d in range(1, D):
                 plotdata[:,d] = data[:,d] - data[:,0]
                 plotdata_labels[d] = data_labels[d] + " - " + data_labels[0]
[…]

### 2016-01-26 code/kmeans.py
visualize NGC 6819

@@ -338,8 +338,11 @@ if __name__ == "__main__":
 
         # find NGC 6819 members, because I care
         print(K, "looking for NGC 6819 members")
+        NGC6819_indices = []
         for name in NGC6819_names:
-            print(K, name, clusters[names == name])
+            NGC6819_indices += list(np.arange(len(data))[names == name])
+        NGC6819_indices = np.array(NGC6819_indices)
+        plot_one_cluster(plotdata, plotdata_labels, NGC6819_indices, "NGC_6819", dir)
         assert False
 
         # plot clusters from this K

### 2016-02-16 papers/gradients.tex
shell for gradients paper

--- 2016-01-20 papers/clusters.tex
@@
 
-\title{Chemical tagging can work: \\
-       Identification of stellar phase-space structures \\
-       purely by chemical-abundance similarity}
+\title{The fifteen-dimensional chemical-abundance gradient in the Milky Way disk}
 \author{David~W.~Hogg\altaffilmark{1,2,3,4},
+        Melissa~Ness\altaffilmark{4},
         Andrew~R.~Casey\altaffilmark{5},
-        Melissa~Ness\altaffilmark{4},
-        Hans-Walter~Rix\altaffilmark{4},
-    and~Daniel~Foreman-Mackey\altaffilmark{6,7}}
+        Hans-Walter~Rix\altaffilmark{4}}
 \altaffiltext{1}{Simons Center for Data Analysis, 160 Fifth Avenue, 7th floor, New York, NY 10010, USA}
@@
 \altaffiltext{5}{Institute of Astronomy, University of Cambridge, Madingley Road, Cambridge CB3~0HA, UK}
-\altaffiltext{6}{Sagan Fellow}
-\altaffiltext{7}{Department of Astronomy, University of Washington, Box 351580, Seattle, WA 98195, USA}
 \email{david.hogg@nyu.edu}
@@
 % Context:
-Chemical tagging promises to use detailed abundance measurements to
-identify spatially separated stars that were in fact born together
-(in the same molecular cloud), long ago.
-This idea has not previously yielded scientific successes, probably because of the
-noise and incompleteness in chemical-abundance measurements.
-However, we have succeeded in substantially improving spectroscopic
-measurements with \thecannon, which has delivered $15$ individual
-abundances for $\sim 10^5$ stars observed as part of the
-\apogee\ spectroscopic survey, with precisions around 0.04~dex.
+The \apogee\ infrared spectroscopic survey and \thecannon\ model of
+stellar spectra have together produced precise abundances for 15
+elements for $10^5$ red-giant stars throughout the Milky Way disk.
+These data make possible extremely precise measurements of chemical
+abundance differences between different components of the Milky Way.
 % Aims:
-We test the chemical-tagging hypothesis by looking at clusters in abundance space
-and confirming that they are clustered in phase space.
+We visualize and measure the 15-dimensional chemical-abundance distribution of
+red-giant stars as a function of galactocentric radius, near the plane
+of the disk, to determine the chemical-abundance gradients, but also
+test ideas of star formation, Galactic nucleosynthesis, and radial
+migration.
 % Methods:
-We identify (by the k-means algorithm) overdensities of stars
-in the 15-dimensional chemical-abundance space delivered by \thecannon,
-and plot the associated stars in phase space.
-We use \emph{only} abundance-space information (no positional information) to identify stellar groups.
+Red-giant distances are determined HOW?
+Chemical-abundance gradients---in mean abundance and abundance
+dispersion---are determined by linear regression.
 % Results:
-We find that clusters in abundance space are indeed clusters in phase space.
-We recover some known phase-space clusters and find other interesting structures.
-This confirms the chemical-tagging hypothesis and verifies the precision of the abundance measurements delivered by \thecannon.
-This is the first-ever project to identify phase-space structures by blind search purely in abundance space;
-the prospects for future data sets are very good.
+We find STUFF.
 \end{abstract}
@@
   ---
-  Galaxy: structure
+  stars: abundances
   ---
-  globular clusters: general
+  stars: formation
   ---
-  open clusters and associations: general
+  stars: kinematics and dynamics
   ---
-  stars: abundances
+  Galaxy: kinematics and dynamics
 }
@@
 
-Ensembles of stars in the Milky Way are born in molecular clouds,
-which are presumed to be near-homogeneous in their chemical element
-composition.
-However, most stars are born in unbound associations, or are born in
-star clusters that disperse rapidly; they will eventually end up in
-very different parts of phase space in the Galaxy.
-Yet, if every star preserved its photospheric element abundances over
-its lifetime (at least for most elements), then stars of common birth
-origin ought to be identifiable through their detailed photospheric
-abundances, long after any spatial proximity has
-vanished.
+foo.
 
-This idea---dubbed ``chemical tagging''
-\citep{freeman}---is one of the principal motivations for a number
-of surveys, including \apogee\ \citep{apogee},
-\gaiaeso\ \citep{gaiaeso} and \galah\ \citep{galah}.
-In order to determine the precise abundance labels for chemical
-tagging, these surveys are each measuring high-resolution, high
-signal-to-noise ($S/N$) spectra for hundreds of thousands of stars across the
-Galaxy's disk, bulge and halo.
-
-Chemical tagging holds the promise of revealing not just the
-star-formation history of the Galaxy, but also the accretion history
-(as things that fall in are expected to be chemically distinct from
-those that form in the parent body; for example, \citealt{font}) and
-stellar-orbit diffusion process like radial mixing and radial
-migration (for example, \citealt{roskar, quillen}).  After
-stars are born---or after a star cluster is accreted and
-disrupted---associations or groups will disperse, through
-two-body processes, interactions with resonances, or tides from the whole Galaxy.
-
-Although undeniably promising---and motivating the 
-launch of costly large-scale spectroscopic surveys---
-chemical-tagging as a search technique has yet to be proven in
-practice: identifying stars of common birth origin {\it purely} on the
-basis of their near-identical abundances patterns, without any
-consideration of position or velocity.
-
-Part of the reason that chemical tagging remains unrealized
-is because the level of abundance specificity required
-is very high.
-If there are thousands of (relevant) molecular clouds forming stars in
-the recent history of the Milky Way, clumps of stars can only be
-identified in abundance space if abundance space is high in dimensionality.
-(In principle, it needs to be high in dimensionality both in terms
-of the number of measured abundances \emph{and} in terms of the number of
-nucleosynthetic pathways or the dimensionality of the true abundance
-space.)
-Therefore accurate---or at
-least precise---measurements of many different abundances are needed
-for stellar siblings to have sufficiently unique fingerprints.
-Stellar spectroscopic surveys now have the resolution,
-$S/N$, wavelength coverage, and sample sizes to deliver many
-different chemical tags for each star.
-
-There are, however, two big issues.
-The first is that the physical assumptions behind the idea may require
-refinement:
-There may be chemical-abundance overlaps among open clusters
-\citep{blancocuaresma}, coeval groups of stars may have similar tags
-but different birth places \citep{mitschang}, and the
-chemical-abundance space might be low in dimensionality.
-On the other hand, precise studies of stellar twins \citep{melendez, jofre}
-and open clusters \citep{bovy} indicate that chemical abundances ought to be
-very similar for co-eval stars.
-
-The second issue for chemical tagging---and the one we address
-here---is measurement precision (and accuracy).
-The current precision on abundance measurements in the published
-survey catalogs is not high enough (for example, \citealt{martell, ting}).
-However, our recent work \citep{thecannon} suggests that the \emph{data}
-are precise enough: There is enough $S/N$ at the relevant
-locations in spectrum space to deliver high-precision tags.
-The existence of very large data sets, homogeneous in spectrum
-space, suggests that data-driven approaches to the determination
-of stellar abundances might considerably outperform traditional
-methods.
-These traditional approaches are based on \foreign{ab-initio} physical models
-that have shortcomings that become apparent in this age of
-high-quality spectra, and the data-analysis methods do not make use of
-all of the information in the data sets.
-Improving the models and exploiting the entire information
-content in the data is critical if we are going to deliver useful
-chemical tags.
-
-Our specific contribution in this space has been to develop
-\thecannon\ \citep{thecannon, ages}, which is a data-driven model for
-stellar spectra.
-This model can deliver stellar parameters and chemical abundances for
-stars, making use of every pixel of every stellar spectrum (that is,
-all the information in the data) but making no use at all of physical
-models of stars.
-It relies only on there being training data---\emph{some} reference stars for
-which parameters and abundances are known and believed.
-In companion papers \citep{casey16, ness16} we show that
-\thecannon\ can deliver 15 to 19 abundances for stars in the
-\apogee\ survey, at precisions higher than even the training (reference) data on
-which the model is trained.
-We say more about this below.
-We will show here that \thecannon\ improves chemical abundance measurements
-to the point that \emph{chemical tagging is now possible.}
-
-One note on \emph{accuracy} and \emph{precision:}
-In principle, the problem of chemical tagging does not require
-absolute accuracy for chemical-abundance measurements.
-It only requires that we can precisely see that two stars are similar
-in their abundances, even if we happen to be wrong about the absolute values of
-those abundances.
-This point might make it seem like we don't care that our models are
-wrong, so long as they are \emph{consistent.}
-However, this is a bit misleading:
-For chemical tagging to succeed, we need stars with different
-atmospheric parameters $(\teff, \logg)$ but the same chemical
-abundances to be assigned the same position in chemical-abundance
-space.
-That doesn't require \emph{overall} accuracy, but it requires that the models
-have the right dependencies on atmospheric parameters such that the
-wrongness in abundance space is consistent across the Hertzsprung-Russell diagram.
-That is, we need a substantial amount of accuracy to achieve our goals.
-
-It is important to realize that analysis techniques based on \foreign{ab-initio}
-physical models are inaccurate, yet they too strive to improve precision
-in the face of knowingly inaccurate models.  Incomplete atomic data, 
-simplifications of photospheric structure, assumptions about convective
-motion, and inconsistencies resulting from positing local thermal
-equilibrium all contribute to produce inaccurate abundances.
-\thecannon\ stands out because it demonstrably improves the precision on
-chemical abundances, whereas the accuracy of those labels is limited only
-by the training (reference) set employed.
-
-In what follows, we are going to use \apogee\ \acronym{DR12} data \citep{dr12}, in
-which we can re-derive 15 element abundances (C, N, O, Na, Mg, Al, Si,
-S, K, Ca, Ti, V, Mn, Fe, Ni), using \thecannon\ \citep{thecannon}.
-The \apogee\ data set covers a huge radial extent and---because the
-data are taken in the infrared---explores the thin disk.
-However, it has the disadvantage that its spatial coverage is incomplete
-(that is, limited pointings produce a fractured sky map),
-which makes it hard to see, within the data set, linear or
-extended stellar structures.
-In many ways, \galah\ will deliver improvements, both because it will
-have more abundances (possibly 29) and contiguous sky coverage.
-
-One final note:
-we think of this \documentname\ as performing a proof of concept.
-We know that the stellar members of open and globular clusters---stars
-that are identified by being close in phase space---contain highly
-informative abundance information that identifies them also in
-chemical-abundance space.  Does this work the other way around? 
-Can chemical tagging identify small subsets of stars, among a vastly
-greater background sample, that have a common birth origin? If we 
-find stars purely by their clustering in abundance space and subsequently
-show that they are part of a still spatially coherent cluster, then
-we will have resolved all practical outstanding issues plaguing 
-chemical tagging, thereby bringing us much closer to
-unravelling the formation of the Milky Way.
-
-\section{Data: \apogee\ giants with abundances from \thecannon}\label{sec:data}
-
-Our analysis draws on the spectra of \totalnumber\ giant stars ($\logg
-< 3.9$) from \apogee\ \acronym{DR12} \citep{dr12}, with no warning flags
-set in the \apogee\ \aspcap\ \citep{aspcap} pipeline
-reductions.
-We re-analyze these \emph{spectra} using \thecannon, because it can
-deliver stellar abundance labels of higher precision, especially for
-stars of $S/N \le 150$.
-The details about how we select, reduce, and analyze the \apogee\ data
-are given in full detail in the companion papers \citep{casey16,
-  ness16}, and we only summarize briefly here:
-We re-derived 17 labels ($\teff$, $\logg$, and 15 abundances
-referenced to H) from the \apogee\ \acronym{DR12} spectra. 
-For the training step of \thecannon, 
-12,681 red-giant stars with spectral $S/N \ge 200$ were used.
-For this small fraction of the sample with the highest $S/N$, the
-labels provided by \aspcap\ provide consistent, relatively low-scatter, sensible
-abundance-space measurements.
-The training step fixes a spectral model
-that predicts the normalized spectrum as a quadratic function of all
-labels ($\sim 1.5\times10^6$ model coefficients). In the second stage---the test step---each unlabelled star is
-used to establish a single-star likelihood function for its labels,
-holding the spectral model coefficients (the parameters of the model) fixed.
-\thecannon\ finds the labels that minimize the single-star likelihood
-function for each test-set star.
-This optimization is not convex, but it is embarrassingly parallel,
-and therefore fast. The test
-step for 150,000 spectra takes less than 30~minutes on a small 
-research cluster in Cambridge.
-
-Six two-dimensional projections of the abundance data, plus some other
-data quantities, are shown in \figurename~\ref{fig:all}.  Throughout
-this work we show two-dimensional abundance projections for only a few
-elements. The entire sequence of 15-choose-2 combinations is too
-immense to visualize. The projections shown are ones which
-have been extensively used in abundance studies of globular clusters
-and satellite systems. These are typically light element abundances,
-and our projections include C--N, Na--O, and Mg--Al. Globular clusters
-famously demonstrate correlations in these elements, thereby making
-them suitable projections for us to show in verifying and examining
-any substructure identified by k-means (or any other clustering algorithm).
-
-\begin{figure}[!p]
-\insanefigure{all}
-\caption{The full sample of \totalnumber\ stars used in this study.
-  The top six panels show six two-dimensional projections of the
-  empirical abundance space distribution.
-  The clustering algorithm (described in
-  \sectionname~\ref{sec:method}) performs clustering in a full
-  15-dimensional abundance space of elements referenced to H (not Fe),
-  but plots are shown here referenced to Fe for familiarity reasons.
-  The bottom three panels show stellar meta-data not used in the
-  clustering described below.\label{fig:all}}
-\end{figure}
-
-\clearpage
-\section{Identifying abundance-space clusters}\label{sec:method}
-
-Although only hints of small-scale clustering in the abundance space
-are visible in \figurename~\ref{fig:all}, exploration of the data by hand indicates that
-known clusters do appear in the high-dimensional abundance space as over-densities.
-In general, the collapse of 15 dimensions down to 2-d projections
-will hide, smooth, or dilute any structure; there is no guarantee that
-a highly featured distribution in 15-d will show features in
-\emph{any} 2-d projection, let alone axis-aligned, human-selected projections.
-This encourages us to look for over-densities automatically in the
-abundance space and see if anything found that way would be over-dense
-in phase space.
-The simplest method for clustering points in $D$-dimensional space is
-the \emph{k-means} algorithm (see \citealt{bishop} for a pedagogical
-introduction and references to the original literature).
-
-Briefly, the k-means algorithm is the following:
-$\bullet$~Start at some initial guess for the locations of $K$ $D$-space cluster centers ($K$
-locations in the $D$-dimensional space).
-$\bullet$~Assign each point in the
-space (each star in our case) to the closest of the $K$ centers.
-$\bullet$~Given this assignment of stars to centers, update each center (each of
-the $K$ $D$-space positions) by taking the mean of the locations of
-the stars assigned to that center.
-$\bullet$~Iterate these steps (assignment of points, followed by taking of means) to convergence.
-The output of this algorithm is the converged locations of the $K$ centers and the assignments of
-all points to those centers.
-This algorithm is fast and performs well in practice in problems of
-this nature; also we are not the first to use the k-means algorithm in
-abundance space \citep{gratton}.
-
-The k-means algorithm has a number of limitations:
-One is that $K$ must be chosen by hand (or heuristically at best).
-Here we are only demonstrating a concept; we don't need to have the
-best possible clustering.
-For this reason we simply choose $K=128$, $K=256$, and $K=512$ and look at all the
-results.
-Also, the k-means algorithm only performs local optimization.
-At each $K$ we perform 32 restarts with different initializations, and
-preserve the best clustering (best according to the k-means score).
-Each of the 32 initializations was performed with the
-\project{scikit-learn} standard k-means initialization procedure
-\citep{sklearn}.
-
-Another issue with k-means is that it effectively uses metric
-distances in the $D$-space; it presumes Euclidean isotropy.
-We choose here to work in the hydrogen-normalized abundance space, the
-space of [C/H], [N/H], [O/H], [Na/H], [Mg/H], [Al/H], [Si/H], [S/H],
-[K/H], [Ca/H], [Ti/H], [V/H], [Mn/H], [Fe/H], [Ni/H].
-But in addition to this, we re-scale these by approximate typical measurement
-precisions obtained by \thecannon\ before running k-means.
-These scalings were
-  [C/H]/0.041,
-  [N/H]/0.044,
-  [O/H]/0.037,
-  [Na/H]/0.111,
-  [Mg/H]/0.032,
-  [Al/H]/0.055,
-  [Si/H]/0.041,
-  [S/H]/0.054,
-  [K/H]/0.069,
-  [Ca/H]/0.043,
-  [Ti/H]/0.072,
-  [V/H]/0.146,
-  [Mn/H]/0.041,
-  [Fe/H]/0.019, and
-  [Ni/H]/0.034.
-This scaling makes the space close to isotropic in measurement uncertainty or
-observational precision.
-Importantly we use \emph{only} abundance-space information, and no
-positional or velocity information (nor $\teff$ nor $\logg$ nor any
-targeting or observational meta-data) as input to the clustering
-algorithm.
-
-Nothing about this algorithm or our choices are particularly
-tuned or optimized; this is in no sense the
-algorithm or the method that reveals the best structures.
-Indeed, it is a feature of this work that even the simplest, most
-generic clustering algorithm returns interesting
-structures (as shown below).
-The prospects for Milky Way science only improve as the clustering
-algorithm improves.
-
-When the k-means results are returned, the $D\times D$ empirical
-variance tensor for the members of each cluster can be constructed.
-From this, an effective density in the abundance space can be computed
-as the number of points in the cluster divided by the square-root of
-the determinant of the tensor.
-This density was used to rank abundance-space overdensities (from highest
-to lowest) for visual inspection.
-In \figurename~\ref{fig:densities} we show the distribution of cluster membership
-and density for the $K=256$ run of k-means.
-The densest clusters tend to have fewer members, presumably because
-the bulk of the population is not extremely strongly clustered, or not
-in ways to which k-means is sensitive.
-
-Importantly, the k-means algorithm assigns \emph{every} star to at
-least one cluster.
-For this reason, there is no sense in which \emph{every} ``cluster'' returned
-by k-means is a distinct overdensity in abundance space.
-In what follows, we only consider high-density clusters---clusters
-that are much more dense than average; these ought to represent true
-over-densities in abundance space.
-
-We chose a few interesting cases from the high-density clusters
-in the $K=256$ run and show them in
-\figurename~\ref{fig:M13}, \ref{fig:Pal5}, \ref{fig:Sgr},
-\ref{fig:halo}, and \ref{fig:disk}.
-The first three of these are dominated by stars in the known clusters
-M13, Pal5, and Sagittarius.
-The fourth is a halo structure with high velocity dispersion and
-possibly accreted.
-The fifth is a thin-disk star-formation feature.
-We will discuss the astrophysical implications of these results in
-\sectionname~\ref{sec:discussion}.
-
-The abundance-space clusters shown in the \figurename s
-might not look dense in 2-d projections of the abundance space, but they
-are very dense in the 15-dimensional space.  Indeed,
-\figurename~\ref{fig:M13} is the densest cluster found in the 15-d
-space \emph{by far}.
-The challenge of this work is to find structure in the high
-dimensionality space that is not obviously visible in any 2-d projection.
-Although we have by no means any model of the 15-d space, we do know
-that the structures shown in the figures are very high in 15-d
-density.
-
-We have only shown results from the $K=256$ run of k-means.
-This run was chosen because its densest abundance-space clusters map
-well onto known stellar clusters.
-At $K=128$ the densest abundance-space clusters tended to combine
-multiple known stellar clusters into single, large abundance-space
-groupings.
-At $K=512$ k-means tended to split even mono-abundance groups into
-smaller sub-groups.
-This is a reminder that k-means has been chosen just for simplicity
-here; it is by no means matched to the discovery of
-stellar structures.
-An important follow-up project is to build a model of abundance space
-that captures the features expected from stellar populations (and observational uncertainties).
-Even if we were able to tune k-means in some sense ``perfectly'', it
-would still break up many stellar clusters, since they often show
-multiple stellar populations and non-linear correlations between element abundances.
-It is also the case that each of the abundance-space clusters we
-\emph{do} show in the \figurename s includes both star-cluster members
-and some background contamination, and also is missing some true
-star-cluster members.
-
-Although we show only three stellar clusters in
-\figurename~\ref{fig:M13}, \ref{fig:Pal5}, and \ref{fig:Sgr}, many
-other clusters are visible among our densest abundance-space k-means
-clusters.
-These include M15, M92, and M107, among others.
-We haven't asked yet whether we detect---as abundance-space
-overdensities---all of the stellar clusters we expect to find. (Many
-clusters within the \apogee\ data set only contain a few plausible 
-members, thereby  complicating any inferences we wish to make about 
-detection completeness using simple k-means.) 
-This \documentname\ is simply a demonstration of the chemical-tagging
-concept; a full investigation of whether we can construct complete
-catalogs of stellar clusters is beyond our present scope.
-
-\begin{figure}[!bp]
-\includegraphics[width=1.5\figwidth]{./figs/clusters_0256.png}
-\caption{The distribution of membership and density for the 256
-  abundance-space clusters returned by the k-means algorithm at
-  $K=256$.  The densest cluster (which is displayed in more detail in 
-  \figurename~\ref{fig:M13}) is far more dense than the second
-  densest.\label{fig:densities}}
-\end{figure}
-\begin{figure}[!p]
-\insanefigure{cluster_0256_0253}
-\caption{Same as \figurename~\ref{fig:all} except that the entire
-  sample has been made gray and the members of the most dense
-  $K=256$ abundance-space cluster have been rendered as unique, prominent triangles
-  (color is velocity rank, orientation is $\logg$ rank).
-  The lower-right plot shows this cluster (circle) in context of the other clusters (dots).
-  This cluster, which was identified only in abundance space (six
-  projections of which are the top six panels in this \figurename),
-  turns out to be dominated by the halo globular cluster
-  M13.\label{fig:M13}}
-\end{figure}
-\begin{figure}[!p]
-\insanefigure{cluster_0256_0034}
-\caption{Same as \figurename~\ref{fig:M13} but for another dense
-  abundance-space cluster.
-  This one turns out to be dominated by the disrupting globular
-  cluster Palomar 5.\label{fig:Pal5}}
-\end{figure}
-\begin{figure}[!p]
-\insanefigure{cluster_0256_0177}
-\caption{Same as \figurename~\ref{fig:M13} but for another dense
-  abundance-space cluster.
-  This one turns out to be dominated by the Sagittarius dwarf spheroidal galaxy.\label{fig:Sgr}}
-\end{figure}
-\begin{figure}[!p]
-\insanefigure{cluster_0256_0010}
-\caption{Same as \figurename~\ref{fig:M13} but for another dense
-  abundance-space cluster.
-  This one turns out to be dominated by a hitherto unrecognized high-velocity-dispersion
-  structure in the Galaxy halo.\label{fig:halo}}
-\end{figure}
-\begin{figure}[!p]
-\insanefigure{cluster_0256_0141}
-\caption{Same as \figurename~\ref{fig:M13} but for another dense
-  abundance-space cluster.
-  This one turns out to be dominated by a very thin young stellar
-  structure in the Galaxy disk, probably one that has been discovered previously \citep{wegg}.\label{fig:disk}}
-\end{figure}
-
-\clearpage
-\section{Discussion}\label{sec:discussion}
-
-We have demonstrated for the first time that, in a large-scale
-spectroscopic survey, stars that appear similar in
-chemical-abundance properties are often associated physically in
-clusters or other structures.
-The reverse is well established---that is, that stars that are
-associated physically in clusters are similar in chemical-abundance
-properties.
-That is, although they often show chemical diversity, that chemical
-diversity is small, and follows well-defined trends or a subspace in
-abundance space (for example, \citealt{gratton, meszaros, bovy},
-and references within those).
-However, it was not known---prior to this work---whether abundance-space information would be
-informative enough, or \emph{measurable} precisely enough to find this
-structure in surveys dominated by a large mix of stars from different
-origins and ages (and there was good reason to be pessimistic;
-\citealt{ting}).
-We have demonstrated with a straightforward (and in no way optimal)
-clustering in chemical-abundance space that it is easy to find co-eval
-stellar structures.
-Although this is very strong evidence that chemical tagging will work,
-we have not performed, in any sense, the canonical procedure of
-looking for features that are consistent with being delta functions in
-chemical-abundance space (convolved with a measurement noise
-distribution); we have looked only for over-densities, not delta
-functions.
-
-What made our success here possible is substantial improvements in precision
-delivered by \thecannon.
-We don't have an absolute measure of the precision of the measurements,
-but from looking at open and globular clusters, it appears to be on the
-order of or better than 0.04~dex for the median element in the list of 15.
-These improvements are discussed in detail in the companion papers
-\citep{casey16, ness16} about the abundance measurements.
-We believe these improvements come from a combination of factors, not
-limited to (a)~improvements in the determination of pseudo-continuum,
-(b)~the use of more spectral range than just unblended element windows
-\citep{aspcap}, and (c)~accurate spectral predictions
-(\thecannon\ delivers accurate predictions because it is fit to
-observed data). 
-Furthermore, because abundance space is high in dimensionality, even
-small improvements in abundance measurement get taken to a significant
-power when thinking about information gains.
-
-There are a few points to highlight about the abundance anti-correlations
-found in globular clusters. First, the detection of these signatures gives
-confirmation to our analysis: The training (reference) set was not purposefully selected
-to contain members of any globular or open cluster. However amongst the test
-set, we recover globular cluster members and their anti-correlations 
-in chemical abundances. That is to
-say we recover peculiar chemical abundances that are not dominant in our
-training set, but are expected from previous studies. Secondly, the very
-presence of complex abundance correlations makes the detection of clusters
-(by k-means) substantially more difficult. Briefly, the k-means algorithm
-is most effective for near-normally distributed, isotropic clusters in high-dimensional space.
-While globular clusters often show distributions of this kind in a few dimensions,
-they also demonstrate non-linear correlations which would be
-sub-optimally selected by k-means. For these reasons the fact that (a)~\thecannon\
-reports these abundance correlations and (b)~k-means (as a simple algorithm
-without information about expectations in abundance space) does not appear to be
-severely impacted by these correlations, gives us great hope for more sophisticated approaches.
-
-The abundances of known clusters that we identify by k-means are in
-excellent agreement with the literature. For the group shown in \figurename~\ref{fig:M13},
-which we attribute to M13, we find a mean of [Fe/H]$ = -1.55$.
-The mean and spread agree well with other work on this cluster
-\citep{Kraft_1992,Cohen_Melendez_2005,Johnson_Pilachowski_2012}, and compiled catalogs of
-globular cluster properties \citep[][accessed 2016]{Harris_1996}. The detailed abundances are also
-consistent: we find a correlation in [C/Fe] and [N/Fe] abundances \citep{Smith_2005}, and their
-projection in \figurename~\ref{fig:M13} show a spread that reflects deep mixing
-along the red giant branch \citep{Briley_2004}. We find that the light element
-abundances (most notably Mg--Al) for this cluster are anti-correlated, as
-expected from other studies \citep[for example,][]{gratton}. However we find a
-smaller spread in [O/Fe] than reported by others who have looked exclusively
-at M13 \citep{Johnson_Pilachowski_2012}.  As we discuss above, this is expected:
-k-means is by no means optimal for identifying arbitrary-shaped structures
-in high-dimensional space, and the first stars that would be assigned to another k-means
-cluster would be those with the most extreme abundances: low [O/Fe] and
-high [Na/Fe].
-
-There are scant abundance measurements reported for Palomar 5, for which
-most stars in the cluster in \figurename~\ref{fig:Pal5} belong. We find a mean
-metallicity of [Fe/H]$ = -1.3$, which lies between the only
-photometric \citep{Geisler_1997} and spectroscopic \citep{Smith_1985,Smith_2002} abundance
-measurements for this cluster.  The anti-correlation in C--N and Na--O abundance
-have been seen elsewhere \citep[][respectively]{Smith_1985,Smith_2002}, although
-to our knowledge the Mg--Al anti-correlation has not previously been shown for this cluster.
-Indeed, this work contains the largest number of abundances for Palomar 5,
-permitting a thorough search of tidally disrupted members \citep[for example,][]{Kuzma_2015} 
-that have retained the unique cluster abundance signature.
-
-The high velocity-dispersion structure seen in \figurename~\ref{fig:halo} stands
-in stark contrast to the rest of the clusters we have identified. In
-addition to being highly clustered in 15-dimensions, it is visibly
-clustered in our 2-d projections. Indeed, it is more clustered in our
-abundance projections than the young stellar structure in the disk,
-and comparably so to the other known globular clusters that we have
-shown here. The disparity between this structure and others presented
-here is the lack of co-spatial stars. 12 stars within this cluster are
-marked as `M53' candidates by \apogee, but the remaining 145 stars
-are spread amongst fields throughout the halo.
-
-The k-means algorithm employed here is by no means optimal, and other clusters
-we identify are accompanied by a few stars that are not currently co-spatial.
-Those stars may be unassociated interlopers that have been misclassified,
-or they may be true cluster members that are now unbound. However in the
-case of the high-velocity-dispersion structure in \figurename~\ref{fig:halo},
-the situation is far more extreme. We identify a group of stars with
-very similar abundances in 15 dimensions that are now spread throughout
-the Galaxy halo.
-Because some stars ($<8\%$) are candidate members of the massive globular
-cluster M53, it provides tantalizing possibility that these stars may
-indeed be all members of the same co-natal gas cloud, implying that 
-these stars have been accreted onto the halo by a single globular
-cluster early in the Milky Way's formation. While speculative,
-this idea demonstrates the promise of chemical tagging.
-
-The abundance-space group of stars in the disk structure shown in
-\figurename~\ref{fig:disk} may be associated with a very thin
-Milky-Way bar component reported previously \citep{wegg}, which is
-found to exist predominantly toward the end of the bar (at $l$
-$\approx 30$~deg).
-The stars in this structure are metal-rich and have a low
-[C/N] ratio and are therefore likely young (following the [C/N]--mass
-relationship; \citealt{martig}).
-It is possible that this structure has very low scale height because
-it is very young and hasn't experienced dynamical heating or radial
-migration.
-
-It is worth reiterating here what is written above:
-There is no sense at all in which \figurename~\ref{fig:M13},
-\ref{fig:Pal5}, \ref{fig:Sgr}, \ref{fig:halo}, and \ref{fig:disk} show
-a representative or complete set of features found in abundance space.
-These features were hand-chosen to be obviously interesting and
-interpetable.
-There are many other things to be found in this data set, and many
-more effective models to be built of abundance-space structure.
-The only strong conclusion of this investigation is that chemical
-tagging \emph{can} work; the abundance measurements with \thecannon\ are
-precise enough and the chemical-abundance space is informative enough.
-
-The \apogee\ project shows great promise for these studies, and with
-the appearance of the companion papers, we will release the
-chemical-abundance measurements we used here for further study.
-The future is even brighter, however: \apogee\ is expanding to 19
-elements, more stars, and all-sky angular coverage with \apogee2, and
-\galah\ is nearly ready to release chemical abundances on a larger set
-of 29 elements.
-The \apogee\ elements employed in this project include alpha
-(Mg, O, Si, Ca, Ti), light proton-capture (C, N, S), odd-Z (Na, Al, K)
-and iron peak (Fe, Ni, V, Mn) elements.
-\galah\ will deliver element abundances from all the major
-nucleosynthetic processes, including light proton capture elements,
-alpha, odd-Z, iron-peak, light and heavy slow neutron capture, and
-rapid neutron capture.
-Chemical tagging capabilities are expected to grow as the
-\emph{product} of the measured nucleosynthetic pathways.
+bar.
 

### 2016-02-21 papers/gradients/gradients.tex
moving furniture in prep for ref response

--- 2016-02-16 papers/gradients.tex
@@
 
-\title{The fifteen-dimensional chemical-abundance gradient in the Milky Way disk}
+\title{The fifteen-dimensional chemical-abundance gradient \\ in the Milky Way disk}
 \author{David~W.~Hogg\altaffilmark{1,2,3,4},
@@
 \altaffiltext{5}{Institute of Astronomy, University of Cambridge, Madingley Road, Cambridge CB3~0HA, UK}
-\email{david.hogg@nyu.edu}
 
@@
   ---
+  Galaxy: disk
+  ---
+  Galaxy: kinematics and dynamics
+  ---
   Galaxy: stellar content
@@
   stars: kinematics and dynamics
-  ---
-  Galaxy: kinematics and dynamics
 }
@@
 
+\section{Data}
+
+\apogee.
+
+\thecannon\ and abundances.
+
+distances.
+
+\section{Analyses and results}
+
 bar.
+
+\section{Discussion}
+
+whatever.
 
@@
 It is a pleasure to thank
+  Dan~Foreman-Mackey (UW),
   Anna~Y.~Q.~Ho (Caltech),
-  Jeremy~Magland (\acronym{SCDA}), and
-  the Blanton--Hogg group meeting
+  and the Blanton--Hogg group meeting
 for valuable comments and discussions.
@@
   European Union's Seventh Framework Programme (FP~7)
-  \acronym{ERC} Grant Agreement n.~\acronym{[320360, 321035]}).
+  \acronym{ERC} Grant Agreement nos.~\acronym{[320360, 321035]}.
 This research made use of the \acronym{NASA} \project{Astrophysics Data System}.

### 2016-02-21 papers/kmeans/response_to_referee.txt
placeholder for referee response

Dear Editor:

Please find below the response to our (very helpful) referee report on
manuscript ApJ/ApJ101852.  We have taken all of the referees
suggestions, and made some other changes from comments we have
received from colleagues.

In what follows, the Referee comments are indented and our responses
are not.  In the revised manuscript, we have made the changes in bold
text so they are apparent to the editor and referee.

     This paper presents new element abundance results using The
     Cannon, a novel, powerful method for spectroscopic abundance
     measurements. The results presented in this paper are new and
     very much relevant to the many large scale observing projects
     that are currently in operation.

     I highlight some comments for the authors to address:

     Abstract: The last sentence is not true, as the authors note
     later in the text the first blind search purely in abundance
     space was carried out by Mitschang et al 2014.

HOGG

     Introduction: The first paragraph is missing a reference or two,
     to support claim that most stars end up in different parts of the
     Galaxy. Or this is something that is assumed and yet to be
     tested?

HOGG

     There are some missing references to the original demonstrations
     of chemical tagging of unbound groups, namely the Wolf 630 group
     and HR 1614 group. Those were key to building up the case for
     this "tagging" research.

HOGG

     The text needs to highlight that the training set is critical -
     Cannon doesn't know physics. It needs to be told what is "true"
     and that the subsequent results will be limited to the believed
     level of truth. While it should be specified that tagging per se
     doesn't require absolute abundances, but the subsequent
     interpretation of the chemical accretion history will require
     "true" abundances - not just precision. Also some clarification
     about the Cannon's performance at the boundaries of the training
     set parameters would have been good to see.

HOGG

     Section 2:

     A reference for the famously demonstrated correlations in
     globular clusters is needed.

HOGG

     Section 3:

     The text doesn't clarify why only k-means was chosen to do the
     'tagging' given the issues discussed, was it simply easy and
     fast? Fine as the focus of the paper is the precision of the
     abundance results with Cannon, not the fidelity of group finding.

HOGG

     I find the group in Fig 6 dubious. Its at the very edge of the
     metallicity scale, which raises again how the Cannon performs at
     the boundaries of the training parameters. Could this not simply
     be metal-poor Halo stars, tagged as metal-poor Halo stars,
     nothing more?

HOGG

-----
Other changes:

In addition to all the above, we made the following additional
changes:

HOGG:  We added authors!

HOGG:  We added citations!

HOGG:  Adjusted what we say about Ting et al?

HOGG:  Go through github issues and emails.

### 2016-02-24 code/gradients.py
read data

@@ -1,0 +1,51 @@
+"""
+This project is part of the Platypus project.
+Copyright 2016 David W. Hogg (NYU).
+"""
+import os
+import pickle as cp
+import numpy as np 
+import sklearn.cluster as cl
+from astropy.io import fits
+import pylab as plt
+
+def pickle_to_file(fn, stuff):
+    fd = open(fn, "wb")
+    cp.dump(stuff, fd)
+    print("writing", fn)
+    fd.close()
+
+def read_pickle_file(fn):
+    fd = open(fn, "rb")
+    stuff = cp.load(fd)
+    fd.close()
+    return stuff
+
+def get_data(fn, labels, metadata_labels):
+    print("get_data(): reading table from", fn)
+    hdulist = fits.open(fn)
+    hdu = hdulist[1]
+    cols = hdu.columns
+    table = hdu.data
+    mask = ((table.field("TEFF_ASPCAP") > 3500.) *
+            (table.field("TEFF_ASPCAP") < 5500.) *
+            (table.field("LOGG_ASPCAP") > 0.) *
+            (table.field("LOGG_ASPCAP") < 3.9)) # MKN recommendation
+    metadata = np.vstack((table.field(label) for label in metadata_labels)).T
+    data = np.vstack((table.field(label) for label in labels)).T
+    mask *= np.all(np.isfinite(data), axis=1)
+    data = data[mask]
+    print("get_data()", fn, data.shape, metadata.shape, mask.shape)
+    return data, metadata, mask
+
+if __name__ == "__main__":
+    print("Hello World!")
+
+    data_labels = ["FE_H",
+                   "AL_H", "CA_H", "C_H", "K_H",  "MG_H", "MN_H", "NA_H",
+                   "NI_H", "N_H",  "O_H", "SI_H", "S_H",  "TI_H", "V_H"]
+    metadata_labels = ["APOGEE_ID", "RA", "DEC", "GLON", "GLAT", "VHELIO_AVG", "DM_Padova",
+                       "TEFF_ASPCAP", "LOGG_ASPCAP"]
+    dfn = "./data/cannon-distances.fits"
+    data, metadata, mask = get_data(dfn, data_labels, metadata_labels)
+    print(data.shape, metadata.shape, mask.shape)

### 2016-02-24 code/gradients.py
bug

@@ -35,6 +35,7 @@ def get_data(fn, labels, metadata_labels):
     data = np.vstack((table.field(label) for label in labels)).T
     mask *= np.all(np.isfinite(data), axis=1)
     data = data[mask]
+    metadata = metadata[mask]
     print("get_data()", fn, data.shape, metadata.shape, mask.shape)
     return data, metadata, mask
 

### 2016-02-24 code/gradients.py
compute x, y, z

@@ -30,7 +30,8 @@ def get_data(fn, labels, metadata_labels):
     mask = ((table.field("TEFF_ASPCAP") > 3500.) *
             (table.field("TEFF_ASPCAP") < 5500.) *
             (table.field("LOGG_ASPCAP") > 0.) *
-            (table.field("LOGG_ASPCAP") < 3.9)) # MKN recommendation
+            (table.field("LOGG_ASPCAP") < 3.9) *
+            (table.field("DIST_Padova") > 0.))
     metadata = np.vstack((table.field(label) for label in metadata_labels)).T
     data = np.vstack((table.field(label) for label in labels)).T
     mask *= np.all(np.isfinite(data), axis=1)
@@ -42,11 +43,28 @@ def get_data(fn, labels, metadata_labels):
 if __name__ == "__main__":
     print("Hello World!")
 
-    data_labels = ["FE_H",
-                   "AL_H", "CA_H", "C_H", "K_H",  "MG_H", "MN_H", "NA_H",
-                   "NI_H", "N_H",  "O_H", "SI_H", "S_H",  "TI_H", "V_H"]
-    metadata_labels = ["APOGEE_ID", "RA", "DEC", "GLON", "GLAT", "VHELIO_AVG", "DM_Padova",
-                       "TEFF_ASPCAP", "LOGG_ASPCAP"]
+    data_labels = np.array(["FE_H",
+                            "AL_H", "CA_H", "C_H", "K_H",  "MG_H", "MN_H", "NA_H",
+                            "NI_H", "N_H",  "O_H", "SI_H", "S_H",  "TI_H", "V_H"])
+    metadata_labels = np.array(["APOGEE_ID", "RA", "DEC", "GLON", "GLAT", "VHELIO_AVG", "DIST_Padova",
+                                "TEFF_ASPCAP", "LOGG_ASPCAP"])
     dfn = "./data/cannon-distances.fits"
     data, metadata, mask = get_data(dfn, data_labels, metadata_labels)
     print(data.shape, metadata.shape, mask.shape)
+
+    # TO-DO: MOVE THIS LOGIC INTO get_data
+    distances = (metadata[:, metadata_labels == "DIST_Padova"]).flatten().astype(float)
+    ls = (np.pi / 180.) * (metadata[:, metadata_labels == "GLON"]).flatten().astype(float)
[…]

### 2016-02-24 code/gradients.py
median abundances vs radius

@@ -21,7 +21,14 @@ def read_pickle_file(fn):
     fd.close()
     return stuff
 
-def get_data(fn, labels, metadata_labels):
+def get_data(fn):
+    data_labels = np.array(["FE_H",
+                            "AL_H", "CA_H", "C_H", "K_H",  "MG_H", "MN_H", "NA_H",
+                            "NI_H", "N_H",  "O_H", "SI_H", "S_H",  "TI_H", "V_H"])
+    metadata_labels = np.array(["APOGEE_ID", "RA", "DEC", "GLON", "GLAT", "VHELIO_AVG", "DIST_Padova",
+                                "TEFF_ASPCAP", "LOGG_ASPCAP"])
+
+    # read FITS file
     print("get_data(): reading table from", fn)
     hdulist = fits.open(fn)
     hdu = hdulist[1]
@@ -32,27 +39,13 @@ def get_data(fn, labels, metadata_labels):
             (table.field("LOGG_ASPCAP") > 0.) *
             (table.field("LOGG_ASPCAP") < 3.9) *
             (table.field("DIST_Padova") > 0.))
+    data = np.vstack((table.field(label) for label in data_labels)).T
     metadata = np.vstack((table.field(label) for label in metadata_labels)).T
-    data = np.vstack((table.field(label) for label in labels)).T
     mask *= np.all(np.isfinite(data), axis=1)
     data = data[mask]
     metadata = metadata[mask]
-    print("get_data()", fn, data.shape, metadata.shape, mask.shape)
-    return data, metadata, mask
 
-if __name__ == "__main__":
-    print("Hello World!")
-
-    data_labels = np.array(["FE_H",
-                            "AL_H", "CA_H", "C_H", "K_H",  "MG_H", "MN_H", "NA_H",
-                            "NI_H", "N_H",  "O_H", "SI_H", "S_H",  "TI_H", "V_H"])
-    metadata_labels = np.array(["APOGEE_ID", "RA", "DEC", "GLON", "GLAT", "VHELIO_AVG", "DIST_Padova",
-                                "TEFF_ASPCAP", "LOGG_ASPCAP"])
-    dfn = "./data/cannon-distances.fits"
-    data, metadata, mask = get_data(dfn, data_labels, metadata_labels)
-    print(data.shape, metadata.shape, mask.shape)
-
-    # TO-DO: MOVE THIS LOGIC INTO get_data
[…]

### 2016-02-25 code/gradients.py
gradients

@@ -74,30 +74,47 @@ def stats_in_slices(data, xs):
         mask = (ranks > rmin) * (ranks < rmax)
         xmedians.append(np.median(xs[mask]))
         medians.append(np.median(data[mask], axis=0))
-    return xmedians, medians
+    return np.array(xmedians), np.array(medians)
+
+def hogg_savefig(fn):
+    print("hogg_savefig():", fn)
+    return plt.savefig(fn)
 
 if __name__ == "__main__":
     print("Hello World!")
 
     # read data
     dfn = "./data/cannon-distances.fits"
-    data, data_labels, metadata, metadata_labels = get_data(dfn)
-    print(data.shape, metadata.shape)
-    print(metadata_labels)
+    pfn = "./data/gradients.pkl"
+    try:
+        print("attempting to read pickle", pfn)
+        data, data_labels, metadata, metadata_labels = read_pickle_file(pfn)
+        print(pfn, data.shape, metadata.shape)
+    except:
+        print("failed to read pickle", pfn)
+        data, data_labels, metadata, metadata_labels = get_data(dfn)
+        pickle_to_file(pfn, (data, data_labels, metadata, metadata_labels))
+        print(pfn, data.shape, metadata.shape)
 
     # compute shit
     Rs = (np.sqrt(metadata[:, metadata_labels == "GX"].astype(float) ** 2 +
                   metadata[:, metadata_labels == "GY"].astype(float) ** 2 +
                   metadata[:, metadata_labels == "GZ"].astype(float) ** 2)).flatten()
-    xmedians, datamedians = stats_in_slices(data, Rs)
-    print(xmedians, datamedians)
+    Rmedians, datamedians = stats_in_slices(data, Rs)
 
-    # TO-DO: MAKE PRETTY PLOTS of slices
+    # plot whole sample
     plt.clf()
     plt.plot(metadata[:, metadata_labels == "GX"],
              metadata[:, metadata_labels == "GY"], "k.", alpha=0.25)
-    plt.xlabel("X (kpc)")
-    plt.ylabel("Y (kpc)")
-    plt.savefig("foo.png")
+    plt.xlabel("Galactic X (kpc)")
+    plt.ylabel("Galactic Y (kpc)")
+    hogg_savefig("GX_GY.png")
 
-    # TO-DO: COMPUTE MEDIANS AND ROOT MEDIAN SQUARES OF ABUNDANCES in slices
+    # plot slices
[…]

### 2016-02-25 code/kmeans.py

@@ -128,7 +128,7 @@ def plot_cluster_context(sizes, densities, dir, name=None, k=None, suffix="png")
         fn = "{}/clusters_{:04d}.{}".format(dir, K, suffix)
     else:
         fn = "{}/{}_context.{}".format(dir, name, suffix)
-    if os.path.exists(fn):
+    if os.th.exists(fn):
         print("plot_cluster_context(): {} exists already".format(fn))
         return
     if k is None:

### 2016-02-25 code/gradients.py
works

@@ -22,9 +22,9 @@ def read_pickle_file(fn):
     return stuff
 
 def get_data(fn):
-    data_labels = np.array(["FE_H",
-                            "AL_H", "CA_H", "C_H", "K_H",  "MG_H", "MN_H", "NA_H",
-                            "NI_H", "N_H",  "O_H", "SI_H", "S_H",  "TI_H", "V_H"])
+    data_labels = ["FE_H",
+                   "AL_H", "CA_H", "C_H", "K_H",  "MG_H", "MN_H", "NA_H",
+                   "NI_H", "N_H",  "O_H", "SI_H", "S_H",  "TI_H", "V_H"]
     metadata_labels = np.array(["APOGEE_ID", "RA", "DEC", "GLON", "GLAT", "VHELIO_AVG", "DIST_Padova",
                                 "TEFF_ASPCAP", "LOGG_ASPCAP"])
 
@@ -63,18 +63,26 @@ def get_data(fn):
 def stats_in_slices(data, xs):
     M = 32
     N = len(xs)
-    assert len(data) == N
+    NN, D = data.shape
+    assert NN == N
     ranks = np.zeros(N)
     ranks[np.argsort(xs)] = np.arange(N)
     ranklims = np.arange(N / M, N, N / M)
-    xmedians = []
-    medians = []
-    rmses = []
-    for rmin, rmax in zip(ranklims[:-2], ranklims[2:]):
+    M = len(ranklims)
+    xmedians = np.zeros(M-2)
+    medians = np.zeros((M-2, D))
+    rmses = np.zeros((M-2, D))
+    for i in range(M-2):
+        rmin, rmax = ranklims[i], ranklims[i+2]
         mask = (ranks > rmin) * (ranks < rmax)
-        xmedians.append(np.median(xs[mask]))
-        medians.append(np.median(data[mask], axis=0))
-    return np.array(xmedians), np.array(medians)
+        xmedians[i] = np.median(xs[mask])
+        medians[i] = np.median(data[mask], axis=0)
+        for d in range(D):
+            diffs = data[mask, d] - medians[i, d]
+            mad = np.median(np.abs(diffs))
+            dmask = (np.abs(diffs) < (5. * mad)) # WRONG
+            rmses[i, d] = np.sqrt(np.mean(diffs[dmask] ** 2))
+    return xmedians, medians, rmses
 
 def hogg_savefig(fn):
     print("hogg_savefig():", fn)
[…]

### 2016-02-25 code/gradients.py
tweak

@@ -80,8 +80,9 @@ def stats_in_slices(data, xs):
         for d in range(D):
             diffs = data[mask, d] - medians[i, d]
             mad = np.median(np.abs(diffs))
-            dmask = (np.abs(diffs) < (5. * mad)) # WRONG
+            dmask = (np.abs(diffs) < (5. * mad))
             rmses[i, d] = np.sqrt(np.mean(diffs[dmask] ** 2))
+            print(i, d, np.sum(mask), np.sum(dmask))
     return xmedians, medians, rmses
 
 def hogg_savefig(fn):

### 2016-02-25 code/gradients.py
bugfix and various adjustments

@@ -50,7 +50,7 @@ def get_data(fn):
     ls = (np.pi / 180.) * (metadata[:, metadata_labels == "GLON"]).flatten().astype(float)
     bs = (np.pi / 180.) * (metadata[:, metadata_labels == "GLAT"]).flatten().astype(float)
     print(np.cos(1.2))
-    GXs = distances * np.cos(ls) * np.cos(bs) + 8.0 # kpc MAGIC
+    GXs = distances * np.cos(ls) * np.cos(bs) - 8.0 # kpc MAGIC
     GYs = distances * np.sin(ls) * np.cos(bs)
     GZs = distances * np.sin(bs)
     metadata = np.vstack((metadata.T, GXs, GYs, GZs)).T # T craziness
@@ -71,6 +71,7 @@ def stats_in_slices(data, xs):
     M = len(ranklims)
     xmedians = np.zeros(M-2)
     medians = np.zeros((M-2, D))
+    sigmas = np.zeros((M-2, D))
     rmses = np.zeros((M-2, D))
     for i in range(M-2):
         rmin, rmax = ranklims[i], ranklims[i+2]
@@ -83,7 +84,8 @@ def stats_in_slices(data, xs):
             dmask = (np.abs(diffs) < (5. * mad))
             rmses[i, d] = np.sqrt(np.mean(diffs[dmask] ** 2))
             print(i, d, np.sum(mask), np.sum(dmask))
-    return xmedians, medians, rmses
+        sigmas[i] = rmses[i] / np.sqrt(np.sum(mask))
+    return xmedians, medians, sigmas, rmses
 
 def hogg_savefig(fn):
     print("hogg_savefig():", fn)
@@ -123,7 +125,7 @@ if __name__ == "__main__":
     plotmetadata = metadata.copy()
 
     # cut in vertical direction
-    zcut = np.abs(metadata[:, metadata_labels == "GZ"].astype(float).flatten()) < 0.1 # kpc
+    zcut = np.abs(metadata[:, metadata_labels == "GZ"].astype(float).flatten()) < 0.2 # kpc
     plotdata = plotdata[zcut]
     plotmetadata = plotmetadata[zcut]
 
@@ -131,15 +133,16 @@ if __name__ == "__main__":
     Rs = (np.sqrt(plotmetadata[:, metadata_labels == "GX"].astype(float) ** 2 +
                   plotmetadata[:, metadata_labels == "GY"].astype(float) ** 2 +
[…]

### 2016-02-26 code/gradients.py
some refactoring; vertical gradients too

@@ -9,6 +9,8 @@ import sklearn.cluster as cl
 from astropy.io import fits
 import pylab as plt
 
+Xsun = -8.0 # kpc MAGIC
+
 def pickle_to_file(fn, stuff):
     fd = open(fn, "wb")
     cp.dump(stuff, fd)
@@ -50,7 +52,7 @@ def get_data(fn):
     ls = (np.pi / 180.) * (metadata[:, metadata_labels == "GLON"]).flatten().astype(float)
     bs = (np.pi / 180.) * (metadata[:, metadata_labels == "GLAT"]).flatten().astype(float)
     print(np.cos(1.2))
-    GXs = distances * np.cos(ls) * np.cos(bs) - 8.0 # kpc MAGIC
+    GXs = distances * np.cos(ls) * np.cos(bs) + Xsun
     GYs = distances * np.sin(ls) * np.cos(bs)
     GZs = distances * np.sin(bs)
     metadata = np.vstack((metadata.T, GXs, GYs, GZs)).T # T craziness
@@ -61,13 +63,12 @@ def get_data(fn):
     return data, data_labels, metadata, metadata_labels
 
 def stats_in_slices(data, xs):
-    M = 32
     N = len(xs)
     NN, D = data.shape
     assert NN == N
     ranks = np.zeros(N)
     ranks[np.argsort(xs)] = np.arange(N)
-    ranklims = np.arange(N / M, N, N / M)
+    ranklims = np.arange(0.5, N, 512)
     M = len(ranklims)
     xmedians = np.zeros(M-2)
     medians = np.zeros((M-2, D))
@@ -124,25 +125,23 @@ if __name__ == "__main__":
     print(plotdata_labels)
     plotmetadata = metadata.copy()
 
-    # cut in vertical direction
-    zcut = np.abs(metadata[:, metadata_labels == "GZ"].astype(float).flatten()) < 0.2 # kpc
-    plotdata = plotdata[zcut]
-    plotmetadata = plotmetadata[zcut]
-
     # compute shit
     Rs = (np.sqrt(plotmetadata[:, metadata_labels == "GX"].astype(float) ** 2 +
                   plotmetadata[:, metadata_labels == "GY"].astype(float) ** 2 +
                   plotmetadata[:, metadata_labels == "GZ"].astype(float) ** 2)).flatten()
-    Rmedians, datamedians, sigmas, rmses = stats_in_slices(plotdata, Rs)
[…]

### 2016-02-27 papers/kmeans/response_to_referee.txt
citations, authors

@@
 
-HOGG:  We added authors!
+Because of a large number of valuable contributions both prior to and
+subsequent to our submission of this paper, we expanded the author
+list.  There are now several additional new authors, most of whom
+contributed by building the APOGEE project (and hence the input data).
 
-HOGG:  We added citations!
+We added a few citations: Eisenstein 2011 (SDSS-III), Holtzman 2015
+(APOGEE DR12) and ...HOGG
 
@@
 HOGG:  Go through github issues and emails.
+
+We also added a few names to the acknowledgements for providing minor
+comments.

### 2016-02-27 papers/kmeans/kmeans.tex
citations, authors

--- 2016-01-20 papers/clusters.tex
@@
         Hans-Walter~Rix\altaffilmark{4},
-    and~Daniel~Foreman-Mackey\altaffilmark{6,7}}
+        Daniel~Foreman-Mackey\altaffilmark{6,7},
+        Anna~Y.~Q.~Ho\altaffilmark{8},
+        Jon~A.~Holtzman\altaffilmark{9},
+        David~L.~Nidever\altaffilmark{10}}
 \altaffiltext{1}{Simons Center for Data Analysis, 160 Fifth Avenue, 7th floor, New York, NY 10010, USA}
@@
 \altaffiltext{7}{Department of Astronomy, University of Washington, Box 351580, Seattle, WA 98195, USA}
+\altaffiltext{8}{Astronomy Department, California Institute of Technology, MC 249-17, 1200 East California Blvd, Pasadena, CA 91125, USA}
+\altaffiltext{9}{New Mexico State University, Las Cruces, NM 88003, USA}
+\altaffiltext{10}{Steward Observatory, 933 North Cherry Ave, Tuscon, AZ 85719, USA}
 \email{david.hogg@nyu.edu}
@@
 
-In what follows, we are going to use \apogee\ \acronym{DR12} data \citep{dr12}, in
+In what follows, we are going to use \apogee\ \acronym{DR12}
+data \citep{apogeedr12, dr12}
+from \sdssiii\ \citep{sdssiii}, in
 which we can re-derive 15 element abundances (C, N, O, Na, Mg, Al, Si,
@@
 Our analysis draws on the spectra of \totalnumber\ giant stars ($\logg
-< 3.9$) from \apogee\ \acronym{DR12} \citep{dr12}, with no warning flags
+< 3.9$) from \apogee\ \acronym{DR12} \citep{apogeedr12}, with no warning flags
 set in the \apogee\ \aspcap\ \citep{aspcap} pipeline
@@
 function for each test-set star.
-This optimization is not convex, but it is embarrassingly parallel,
+This optimization is not convex, but it is trivially parallel,
 and therefore fast. The test
@@
 It is a pleasure to thank
-  Anna~Y.~Q.~Ho (Caltech),
-  Jeremy~Magland (\acronym{SCDA}), and
+  Jeremy~Magland (\acronym{SCDA}),
+  Don Schneider (PSU), and
   the Blanton--Hogg group meeting
@@
 De~Silva,~G.~M., Freeman,~K.~C., Bland-Hawthorn,~J., \etal, 2015, \mnras, 449, 2604
+\bibitem[Eisenstein \etal(2011)]{sdssiii}
+Eisenstein, D.~J., Weinberg, D.~H., Agol, E., \etal, 2011, \aj, 142, 72 
 \bibitem[Font \etal(2006)]{font}
@@
 Harris,~W.~E., 1996, \aj, 112, 1487 
+\bibitem[Holtzman \etal(2015)]{apogeedr12}
+Holtzman, J.~A., Shetrone, M., Johnson, J.~A., \etal, 2015, \aj, 150, 148
 \bibitem[Hunter(2007)]{matplotlib}

### 2016-03-19 code/alpha.py
all works; don't understand results

@@ -162,12 +162,7 @@ class abundance_model:
         thislog10vecs = np.zeros(self.K)
         offset1 = np.max(self.log10amps)
         amps = tento(self.log10amps - offset1)
-        if np.any(self.log10amps > 10.):
-            for n,l in enumerate(self.log10amps):
-                if np.max(l) > 10.:
-                    print("fucked", n, l, self.data[n], self.predicted_data()[n])
-                    assert False
-        for d in range(1, self.D): # don't mess with the zero component [Fe/H]
+        for d in range(self.D):
             thislog10vecs[:] = self.log10vecs[:,d]
             data = self.data[:,d]
             ivars = self.ivars[:,d]
@@ -175,11 +170,6 @@ class abundance_model:
                 offset2 = np.max(log10vecs)
                 foo = tento(log10vecs - offset2)
                 denominator = np.dot(amps, foo)
-                if np.any(denominator <= 0.):
-                    for n,den in enumerate(denominator):
-                        if den <= 0.:
-                            print("denominator fucked", amps[n], log10vecs, offset1, offset2, foo, den, data[n], self.predicted_data()[n,d])
-                            assert False
                 derivs = (amps * foo[None,:]) / denominator[:,None]
                 resids = data - (np.log10(denominator) + offset1 + offset2)
                 return np.sum(resids * ivars * resids), -2. * np.sum(resids[:,None] * ivars[:,None] * derivs, axis=0)
@@ -194,6 +184,24 @@ class abundance_model:
             """
             thisresult = op.minimize(obj, thislog10vecs, method="BFGS", jac="True")
             self.log10vecs[:,d] = thisresult["x"]
+
+    def optimize(self, tol=0.1):
+        prevchisq = np.Inf
+        for t in range(512):
+            print(t)
+            self.set_amps()
+            print(t, self.chisq())
+            self.set_vecs()
+            chisq = self.chisq()
+            print(t, chisq)
+            if chisq > prevchisq:
+                print(t, "optimize(): The sky is falling", prevchisq, chisq)
+                assert False
+            if chisq > (prevchisq - tol):
[…]

### 2016-03-22 papers/nucleosynthesis/nuke.tex
placeholder

\documentclass[12pt, preprint]{aastex}

\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\apogee}{\project{\acronym{APOGEE}}}

\newcommand{\solar}{\odot}

\begin{document}

\title{The minimal nucleosynthesis project with \apogee}
\author{DWH, HWR, JR, MKN, ARC, others}

\begin{abstract}
We describe the minimal thing you could do with the \apogee\ data to
improve our understanding of nucleosynthesis.
\end{abstract}

\keywords{
  ---
  World: Hello
  ---
}

\section{Introduction}

Nucleosynthesis is important.  Models are wrong.  Data are now abundant.

\section{Model}

\begin{eqnarray}
  [X/Y] &\equiv& \log_{10}\left(\frac{X}{Y}\right) -
                 \log_{10}\left(\frac{X}{Y}\right)_\solar
\end{eqnarray}

\end{document}

### 2016-03-24 papers/nucleosynthesis/nuke.tex
model in equations
augment equations with words; equation tweaks

@@
 
+% text commands
+\newcommand{\equationname}{equation}
 \newcommand{\acronym}[1]{{\small{#1}}}
@@
 
+% math commands
 \newcommand{\solar}{\odot}
+\newcommand{\abundance}[1]{\lbrack{#1}\rbrack}
+\newcommand{\given}{\,|\,}
+\newcommand{\normal}{\mathcal{N}}
 
@@
 
-Nucleosynthesis is important.  Models are wrong.  Data are now abundant.
+Nucleosynthesis is important.  Models are (at least slightly) wrong.
+Data are now abundant.  We want data-driven models that are
+nonetheless interpretable in terms of nucleosynthetic pathways.
 
@@
 
+We adopt the standard terminology, that an abundance in square brackets
+is taken to be log (base 10) and relative to Solar:
 \begin{eqnarray}
-  [X/Y] &\equiv& \log_{10}\left(\frac{X}{Y}\right) -
-                 \log_{10}\left(\frac{X}{Y}\right)_\solar
+  \abundance{X/Y}_n &\equiv& \log_{10}\left(\left.\frac{X}{Y}\right|_n\right) -
+                             \log_{10}\left(\left.\frac{X}{Y}\right|_\solar\right)
+\quad ,
 \end{eqnarray}
+where the index $n$ indicates the identity (\apogee\ label) of the star.
+Our model obtains some complexity, because we assume (unrealistically)
+that the uncertainties are Gaussian in the log (base 10) but our
+nucleosynthetic model will be linear in the linear abundance ratios.
+In particular, we assume that each logarithmic abundance $\abundance{X/Y}$
+can be written as the log (base 10) of a linear abundance ratio $y$ with
+an error applied (outside the log)
+\begin{eqnarray}
+  \abundance{X/Y}_n &=& \log_{10}(y_{nd}) + e_{nd} \label{eq:model1}
+\\
+  p(e_{nd}) &=& \normal(e_{nd}\given 0,\sigma^2_{nd}) \label{eq:model2}
+\quad ,
+\end{eqnarray}
+where the index $n$ indicates the star and the index $d$ indicates the
+particular abundance ratio (for example, $\abundance{Fe/H}$ might get
+$d=0$, and $\abundance{C/Fe}$ might get $d=1$, and so on), $y_{nd}$ is
+the linear abundance ratio as given by the nucleosynthetic mean model
+prediction (see below), and $e_{nd}$ is an error drawn from a Gaussian
+frequency distribution with known variance $\sigma^2_{nd}$.
+
+The model is that the linear abundances are generated by a linear model
+that is a mixture of $K$ supernova and AGB yield contributions.
+\begin{eqnarray}
+  y_{nd} &=& \sum_{k=1}^K a_{nk}\,g_{kd}
+\quad ,
+\end{eqnarray}
+where the $a_{nk}$ are the $K$ supernova and AGB mixture coefficients
+for star $n$, and the $g_{kd}$ are the yield values for each of the
+elements $d$ for each of the supernova (and AGB---hereafter we will
+just call all nucleosynthetic pathways ``supernovae'') types $k$.
+We can (with no trouble) let all the $a_{nk}$ and $g_{kd}$ go free and
+optimize the likelihood function (implied by
+\equationname s~\ref{eq:model1} and \ref{eq:model2}) with respect to them all.
+This would make the model nearly equivalent to the HMF we have published
+previously (CITE Tsalmantza).
+This would be similar to (but more responsible than) performing a kind
+of principal component analysis, which is principal but not
+principled---and, importantly, not interpretable physically.
+
+However, we wish to be data-driven, but also produce results with some
+hope of being interpretable in terms of supernova and AGB yields.
+For this reason, we require that the yield vector components $g_{kd}$
+be nearby to some theoretical predictions $G_{kd}$.
+To be specfic, we require
+\begin{eqnarray}
+  g_{kd} &=& G_{kd} + \epsilon_{kd}
+\\
+  p(\epsilon_{kd}) &\propto& \left\{\begin{array}{rr}
+  \normal(\epsilon_{kd}\given 0,\Sigma^2_{kd}) & \mbox{for}~(G_{kd}+\epsilon_{kd})>0 \\
+                                             0 & \mbox{for}~(G_{kd}+\epsilon_{kd})<0 \end{array}\right.
+\quad ,
+\end{eqnarray}
+where the $\epsilon_{kd}$ are departures of the data-driven model from
+the theoretical expectation, and our prior pdfs on these departures
+are zero-mean and Gaussian (but censored at unphysical negative values).
+The $\Sigma^2_{kd}$ are hyper-parameters parameterizing our credulity
+with respect to the theory!
+
+To kick this model-fest off, Jan will supply two effective Type II
+supernova yield curves, one Type Ia curve, and one AGB curve, to make
+$K=4$.  We will use $D=15$ (with $1<d<D$) \apogee\ abundance ratios
+for $N=95,000$ red-giant stars\ldots
 

### 2016-04-01 papers/nucleosynthesis/nuke.tex
reduce authors
committing uncommitted changes (is this a useful comment?)

@@
+% This file is part of the Platypus project.
+% Copyright 2016 the authors.
+
+% # to-do list
+% - fix prior on the \epsilon_{kd} to make everything normal in the log!
+
 \documentclass[12pt, preprint]{aastex}
@@
 \newcommand{\abundance}[1]{\lbrack{#1}\rbrack}
+\newcommand{\element}[1]{\mathrm{#1}}
 \newcommand{\given}{\,|\,}
@@
 \title{The minimal nucleosynthesis project with \apogee}
-\author{DWH, HWR, JR, MKN, ARC, others}
+\author{DWH, HWR, JR, others}
 
@@
 where the index $n$ indicates the star and the index $d$ indicates the
-particular abundance ratio (for example, $\abundance{Fe/H}$ might get
-$d=0$, and $\abundance{C/Fe}$ might get $d=1$, and so on), $y_{nd}$ is
+particular abundance ratio (for example, $\abundance{\element{Fe}/\element{H}}$ might get
+$d=1$, and $\abundance{\element{C}/\element{Fe}}$ might get $d=2$, and so on), $y_{nd}$ is
 the linear abundance ratio as given by the nucleosynthetic mean model
 prediction (see below), and $e_{nd}$ is an error drawn from a Gaussian
-frequency distribution with known variance $\sigma^2_{nd}$.
+frequency distribution with zero mean and known variance $\sigma^2_{nd}$.
 
@@
 be nearby to some theoretical predictions $G_{kd}$.
-To be specfic, we require
+To be specfic, we set very informative priors on the $g_{kd}$ that they
+be close to the $G_{kd}$, with (for example)
 \begin{eqnarray}
-  g_{kd} &=& G_{kd} + \epsilon_{kd}
-\\
-  p(\epsilon_{kd}) &\propto& \left\{\begin{array}{rr}
-  \normal(\epsilon_{kd}\given 0,\Sigma^2_{kd}) & \mbox{for}~(G_{kd}+\epsilon_{kd})>0 \\
-                                             0 & \mbox{for}~(G_{kd}+\epsilon_{kd})<0 \end{array}\right.
+  p(\log_{10} g_{kd}) &=& \normal(\log_{10} g_{kd}\given \log_{10} G_{kd},\Sigma^2_{kd})
 \quad ,
 \end{eqnarray}
-where the $\epsilon_{kd}$ are departures of the data-driven model from
-the theoretical expectation, and our prior pdfs on these departures
-are zero-mean and Gaussian (but censored at unphysical negative values).
-The $\Sigma^2_{kd}$ are hyper-parameters parameterizing our credulity
+where the the $\Sigma^2_{kd}$ are hyper-parameters parameterizing our credulity
 with respect to the theory!

### 2016-04-01 papers/nucleosynthesis/nuke.tex
abstract and dex

@@
 % # to-do list
-% - fix prior on the \epsilon_{kd} to make everything normal in the log!
+% - put in figures showing the G vectors.
 
@@
 We describe the minimal thing you could do with the \apogee\ data to
-improve our understanding of nucleosynthesis.
+improve our understanding of nucleosynthesis:
+Model the detailed distribution of stars in 15-dimensional abundance
+space as a non-negative mixture of supernova and AGB yield vectors,
+but permitting those yield vectors to be distorted where the data
+require it.
 \end{abstract}
@@
 the linear abundance ratio as given by the nucleosynthetic mean model
-prediction (see below), and $e_{nd}$ is an error drawn from a Gaussian
+prediction (see below), and $e_{nd}$ is an error (in dex) drawn from a Gaussian
 frequency distribution with zero mean and known variance $\sigma^2_{nd}$.

### 2016-04-13 papers/kmeans/kmeans.tex
Addressed issues raised by the referee

@@
 This confirms the chemical-tagging hypothesis and verifies the precision of the abundance measurements delivered by \thecannon.
-This is the first-ever project to identify phase-space structures by blind search purely in abundance space;
+This is the first-ever project to identify phase-space structures at survey-scale by blind search purely in abundance space;
 the prospects for future data sets are very good.
@@
 (as things that fall in are expected to be chemically distinct from
-those that form in the parent body; for example, \citealt{font}) and
+those that form in the parent body; for example, 
+\citealt{font,De_Silva_2007,eggen_1970,bubar_king_2010}) and
 stellar-orbit diffusion process like radial mixing and radial
@@
 chemical abundances, whereas the accuracy of those labels is limited only
-by the training (reference) set employed.
+by the training (reference) set employed.  At the same time, it is crucial
+to be cognisant of the constraints in the training set: the
+results of \thecannon\ will be limited by the quality of the training 
+set labels. While chemical tagging does not necessarily require
+accurate abundance labels (precision is paramount), comparing abundance
+labels to models of Galactic chemical enrichment requires a firm level
+of belief in the label accuracy. Particularly for the abundance labels
+of stars in the training set.
 
@@
 and our projections include C--N, Na--O, and Mg--Al. Globular clusters
-famously demonstrate correlations in these elements, thereby making
+famously demonstrate correlations in these elements \citep[e.g., see][and references therein]{Norris_1995,Carretta_2009}, thereby making
 them suitable projections for us to show in verifying and examining
@@
 algorithm or the method that reveals the best structures.
+We chose k-means as a simple and tractable approach to identifying
+clusters in high dimensional label space. In practice more complex
+clustering algorithms may perform better than k-means, or even
+tuning k-means by heuristically setting $K$ is likely to improve
+upon the results here. This \documentname\ serves as a proof of concept.
 Indeed, it is a feature of this work that even the simplest, most
@@
 Briley, M.~M., Cohen, J.~G., Stetson, P.~B., 2004, \aj, 127, 1579 
+\bibitem[Bubar \& King(2010)]{bubar_king_2010} Bubar, E.~J., \& King, J.~R.\ 2010, \aj, 140, 293 
+\bibitem[Carretta et al.(2009)]{Carretta_2009} Carretta, E., Bragaglia, A., Gratton, R.~G., et al.\ 2009, \aap, 505, 117 
 \bibitem[Casey \etal(2016)]{casey16}
@@
 Cohen, J.~G.,  Mel{\'e}ndez, J., 2005, \aj, 129, 303 
+\bibitem[De Silva et al.(2007)]{De_Silva_2007} De Silva, G.~M., Freeman, K.~C., Bland-Hawthorn, J., Asplund, M., \& Bessell, M.~S.\ 2007, \aj, 133, 694
 \bibitem[De~Silva \etal(2015)]{galah}
@@
 Eisenstein, D.~J., Weinberg, D.~H., Agol, E., \etal, 2011, \aj, 142, 72 
+\bibitem[Eggen(1970)]{eggen_1970} Eggen, O.~J.\ 1970, \pasp, 82, 99 
 \bibitem[Font \etal(2006)]{font}
@@
 Ness,~M., Hogg,~D.~W., Casey,~A.~R., Rix,~H.-W., 2016b, in preparation
+\bibitem[Norris \& Da Costa(1995)]{Norris_1995} Norris, J.~E., \& Da Costa, G.~S.\ 1995, \apjl, 441, L81 
 \bibitem[Pedregosa \etal(2011)]{sklearn}

### 2016-04-15 papers/kmeans/response_to_referee.txt

@@
 
-HOGG
+        We have referenced Mitschang et al. (2014) for their work in
+        chemical tagging, but we believe the last sentence of the
+        abstract is correct. To clarify: Mitschang et al. (2014) did
+        perform a blind search in chemical abundances to identify
+        substructure, but one of the key conclusions of their work
+        was that ".. these groups may not represent distinct 
+        formation sites, e.g. as in dissolved open clusters, as 
+        previously thought". While the Mitschang et al. (2014) work
+        showed their chemically-identified structures had similar 
+        (U, V, W) space velocities (e.g., their Figure 2), those 
+        velocities are not dissimilar from the background velocity
+        distribution. In contrast, our work is the first blind
+        search in chemical abundances that has identified structures
+        that are *also* structures in phase-space: specifically
+        on-sky position and radial velocities.
+
+        Nevertheless, to avoid unwarranted claims, we have qualified
+        this statement from:
+
+        "This is the first-ever project to identify phase-space 
+        structures by blind search purely in abundance space;
+        the prospects for future data sets are very good."
+
+        to:
+
+        "This is the first-ever project to identify phase-space 
+        structures at survey-scale by blind search purely in abundance space;
+        the prospects for future data sets are very good."
+
+        since the Mitschang et al. (2014) work had a sample size of
+        714 stars.
 
@@
 
-HOGG
+        The text claims that stars end up in different parts of
+        phase-space in the Galaxy: that is to say that any common
+        kinematic (or dynamic: actions, angles) signature between
+        stars born in the same molecular gas cloud at the same time
+        will eventually disperse, presumably throughout the disk.
+        This line of reasoning follows from expectations that most
+        stars form in clusters, but the disk is full of stars that
+        do not appear to be kinematically bound to a cluster. For
+        these reasons we would claim that the opening paragraph is
+        more of an introduction to the expectations from star
+        formation, or rather introducing the line of reasoning for
+        why elemental abundances can be instructive for reconstructing
+        the evolution of the Galaxy.
 
@@
 
-HOGG
+        We have added the following references: De Silva et al. 
+        (2007), Eggen (1970), and Bubar & King (2010).
 
@@
 
-HOGG
+        We have added references to: Norris & Da Costa (1995), which
+        (to our knowledge) was the first to unambiguously show anti-
+        correlations in Na, Al, and O, and to Carretta et al. (2009),
+        which provides a more recent and comprehensive view of our
+        current knowledge of elemental abundance correlations in
+        gloublar clusters.
 
@@
 
-HOGG
+        We have added text to clarify why k-means was chosen.
+        Specifically we have updated the paragraph:
+
+        "Nothing about this algorithm or our choices are particularly
+        tuned or optimized; this is in no sense the
+        algorithm or the method that reveals the best structures.
+        Indeed, it is a feature of this work that even the simplest, most
+        generic clustering algorithm returns interesting
+        structures (as shown below). The prospects for Milky Way science only 
+        improve as the clustering algorithm improves."
+
+        To now read as:
+
+        "Nothing about this algorithm or our choices are particularly
+        tuned or optimized; this is in no sense the
+        algorithm or the method that reveals the best structures.
+        We chose k-means as a simple and tractable approach to identifying
+        clusters in high dimensional label space. In practice more complex
+        clustering algorithms may perform better than k-means, or even
+        tuning k-means by heuristically setting $K$ is likely to improve
+        upon the results here. This \documentname\ serves as a proof of concept.
+        Indeed, it is a feature of this work that even the simplest, most
+        generic clustering algorithm returns interesting
+        structures (as shown below). The prospects for Milky Way science only 
+        improve as the clustering algorithm improves."
 

### 2016-04-16 papers/kmeans/response_to_referee.txt
tweaks to ARC's work

@@
 
-        We have referenced Mitschang et al. (2014) for their work in
-        chemical tagging, but we believe the last sentence of the
-        abstract is correct. To clarify: Mitschang et al. (2014) did
-        perform a blind search in chemical abundances to identify
-        substructure, but one of the key conclusions of their work
-        was that ".. these groups may not represent distinct 
-        formation sites, e.g. as in dissolved open clusters, as 
-        previously thought". While the Mitschang et al. (2014) work
-        showed their chemically-identified structures had similar 
-        (U, V, W) space velocities (e.g., their Figure 2), those 
-        velocities are not dissimilar from the background velocity
-        distribution. In contrast, our work is the first blind
-        search in chemical abundances that has identified structures
-        that are *also* structures in phase-space: specifically
-        on-sky position and radial velocities.
+We have referenced Mitschang et al. (2014) for their work in chemical
+tagging, but we believe the last sentence of the abstract is
+correct. To clarify: Mitschang et al. (2014) did perform a blind
+search in chemical abundances to identify substructure, but one of the
+key conclusions of their work was that ".. these groups may not
+represent distinct formation sites, e.g. as in dissolved open
+clusters, as previously thought". While the Mitschang et al. (2014)
+work showed their chemically-identified structures had similar (U, V,
+W) space velocities (e.g., their Figure 2), those velocities are not
+dissimilar from the background velocity distribution. In contrast, our
+work is the first blind search in chemical abundances that has
+identified structures that are *also* structures in phase-space:
+specifically on-sky position and radial velocities.
 
-        Nevertheless, to avoid unwarranted claims, we have qualified
-        this statement from:
-
-        "This is the first-ever project to identify phase-space 
-        structures by blind search purely in abundance space;
-        the prospects for future data sets are very good."
-
-        to:
-
-        "This is the first-ever project to identify phase-space 
-        structures at survey-scale by blind search purely in abundance space;
-        the prospects for future data sets are very good."
-
-        since the Mitschang et al. (2014) work had a sample size of
-        714 stars.
+Nevertheless, to avoid unwarranted claims, we have qualified this
+statement in the abstract.
 
@@
 
-        The text claims that stars end up in different parts of
-        phase-space in the Galaxy: that is to say that any common
-        kinematic (or dynamic: actions, angles) signature between
-        stars born in the same molecular gas cloud at the same time
-        will eventually disperse, presumably throughout the disk.
-        This line of reasoning follows from expectations that most
-        stars form in clusters, but the disk is full of stars that
-        do not appear to be kinematically bound to a cluster. For
-        these reasons we would claim that the opening paragraph is
-        more of an introduction to the expectations from star
-        formation, or rather introducing the line of reasoning for
-        why elemental abundances can be instructive for reconstructing
-        the evolution of the Galaxy.
+The text claims that stars end up in different parts of phase-space in
+the Galaxy: that is to say that any common kinematic (or dynamic:
+actions, angles) signature between stars born in the same molecular
+gas cloud at the same time will eventually disperse, presumably
+throughout the disk.  This line of reasoning follows from expectations
+that most stars form in clusters, but the disk is full of stars that
+do not appear to be kinematically bound to a cluster. For these
+reasons we would claim that the opening paragraph is more of an
+introduction to the expectations from star formation, or rather
+introducing the line of reasoning for why elemental abundances can be
+instructive for reconstructing the evolution of the Galaxy.
 
@@
 
-        We have added the following references: De Silva et al. 
-        (2007), Eggen (1970), and Bubar & King (2010).
+We have added the following references: De Silva et al (2007), Eggen
+(1970), and Bubar & King (2010).
 
@@
 
-        We have added references to: Norris & Da Costa (1995), which
-        (to our knowledge) was the first to unambiguously show anti-
-        correlations in Na, Al, and O, and to Carretta et al. (2009),
-        which provides a more recent and comprehensive view of our
-        current knowledge of elemental abundance correlations in
-        gloublar clusters.
+We have added references to: Norris & Da Costa (1995), which (to our
+knowledge) was the first to unambiguously show anti- correlations in
+Na, Al, and O, and to Carretta et al. (2009), which provides a more
+recent and comprehensive view of our current knowledge of elemental
+abundance correlations in gloublar clusters.
 
@@
 We have added text to clarify why k-means was chosen.
-        Specifically we have updated the paragraph:
-
-        "Nothing about this algorithm or our choices are particularly
-        tuned or optimized; this is in no sense the
-        algorithm or the method that reveals the best structures.
-        Indeed, it is a feature of this work that even the simplest, most
-        generic clustering algorithm returns interesting
-        structures (as shown below). The prospects for Milky Way science only 
-        improve as the clustering algorithm improves."
-
-        To now read as:
-
-        "Nothing about this algorithm or our choices are particularly
-        tuned or optimized; this is in no sense the
-        algorithm or the method that reveals the best structures.
-        We chose k-means as a simple and tractable approach to identifying
-        clusters in high dimensional label space. In practice more complex
-        clustering algorithms may perform better than k-means, or even
-        tuning k-means by heuristically setting $K$ is likely to improve
-        upon the results here. This \documentname\ serves as a proof of concept.
-        Indeed, it is a feature of this work that even the simplest, most
-        generic clustering algorithm returns interesting
-        structures (as shown below). The prospects for Milky Way science only 
-        improve as the clustering algorithm improves."
 

### 2016-04-16 papers/kmeans/kmeans.tex
Put in 12pt manuscript format
tweaks to ARC's work

@@
 
-\documentclass[12pt, letterpaper, preprint]{aastex}
+\documentclass[12pt, letterpaper, preprint]{aastex6}
 
@@
 those that form in the parent body; for example, 
-\citealt{font,De_Silva_2007,eggen_1970,bubar_king_2010}) and
+\citealt{eggen_1970,font,De_Silva_2007,bubar_king_2010}) and
 stellar-orbit diffusion process like radial mixing and radial
@@
 and our projections include C--N, Na--O, and Mg--Al. Globular clusters
-famously demonstrate correlations in these elements \citep[e.g., see][and references therein]{Norris_1995,Carretta_2009}, thereby making
+famously demonstrate correlations in these elements \citep[for example,][and references therein]{Norris_1995,Carretta_2009}, thereby making
 them suitable projections for us to show in verifying and examining
@@
 
-All of the code for the data manipulation, clustering, and
-visualization is built on \project{numpy} \citep{numpy},
-\project{scikit-learn} \citep{sklearn}, and \project{matplotlib}
-\citep{matplotlib}.
 All of the code written specifically for this project is available in
@@
 \url{https://github.com/davidwhogg/Platypus}.
+
+\software{
+	\project{numpy} \citep{numpy},
+	\project{scikit-learn}, \citep{sklearn},
+	\project{matplotlib} \citep{matplotlib}}.
 

### 2016-04-16 code/alpha.py

@@ -10,6 +10,8 @@ Copyright 2016 David W. Hogg (NYU).
 - Or show that there must be stochastic nucleosynthetic processes.
 
 ## bugs / notes
+- Need to add command-line options.
+- Need to make a "refine" option that takes the current state and further refines it.
 - Need to get a fake data set from Jan.
 - Too much content below the "__main__" conditional.
 - Ought to restrict to RC and nearby (Jan arguments).

### 2016-04-16 papers/kmeans/kmeans.tex
faster make, reformatting aastex badness

@@
 
-\documentclass[12pt, letterpaper, preprint]{aastex6}
+\documentclass[manuscript, letterpaper]{aastex6}
+
+% omg AASTEX 6.0
+\setlength{\voffset}{0in}
+\setlength{\hoffset}{0in}
+\setlength{\textwidth}{6.5in}
+\setlength{\textheight}{9in}
+\setlength{\headheight}{0ex}
+\setlength{\headsep}{\baselineskip} % this is 2 lines in ``manuscript''
+\setlength{\topmargin}{-1\headsep}
+\linespread{0.54} % close to 10/13 spacing in ``manuscript''
+\setlength{\parindent}{0.54\baselineskip}
+\hypersetup{
+  colorlinks = false,
+}
+\makeatletter % you know you are living your life wrong when you need to do this
+\long\def\frontmatter@title@above{
+\vspace*{-\headsep}\vspace*{\headheight}
+\noindent\footnotesize
+{\noindent\footnotesize\textsc{\@journalinfo}}\par
+{\noindent\scriptsize Preprint typeset using \LaTeX\ style AASTeX6 v.\ 1.0 but heavily modified by David W. Hogg (NYU).
+}\par\vspace*{-\baselineskip}\vspace*{0.625in}
+}%
+\makeatother
 
@@
 \newlength{\figwidth}\setlength{\figwidth}{0.3\textwidth}
-\newcommand{\insanefigure}[1]{%
+\newcommand{\insanefigure}[1]{\rule{0ex}{\headsep}\\% This shouldn't be necessary, but something is fucked with AASTEX 6.0
 \includegraphics[width=\figwidth]{./figs/#1_MgFe.png}%
@@
 \newcommand{\totalnumber}{98,462}
+
+\shorttitle{Chemical tagging can work}
+\shortauthors{Hogg, Casey, Ness, \etal}
+\submitted{Submitted to \textit{The Astrophysical Journal}}
 
@@
 \altaffiltext{10}{Steward Observatory, 933 North Cherry Ave, Tuscon, AZ 85719, USA}
-\email{david.hogg@nyu.edu}
 
@@
 Although undeniably promising---and motivating the 
-launch of costly large-scale spectroscopic surveys---
+launch of costly large-scale spectroscopic surveys---%
 chemical-tagging as a search technique has yet to be proven in
@@
 \bibitem[Bubar \& King(2010)]{bubar_king_2010} Bubar, E.~J., \& King, J.~R.\ 2010, \aj, 140, 293 
-\bibitem[Carretta et al.(2009)]{Carretta_2009} Carretta, E., Bragaglia, A., Gratton, R.~G., et al.\ 2009, \aap, 505, 117 
+\bibitem[Carretta \etal(2009)]{Carretta_2009} Carretta, E., Bragaglia, A., Gratton, R.~G., \etal, 2009, \aap, 505, 117 
 \bibitem[Casey \etal(2016)]{casey16}
@@
 Cohen, J.~G.,  Mel{\'e}ndez, J., 2005, \aj, 129, 303 
-\bibitem[De Silva et al.(2007)]{De_Silva_2007} De Silva, G.~M., Freeman, K.~C., Bland-Hawthorn, J., Asplund, M., \& Bessell, M.~S.\ 2007, \aj, 133, 694
+\bibitem[De Silva \etal(2007)]{De_Silva_2007} De Silva, G.~M., Freeman, K.~C., Bland-Hawthorn, J., Asplund, M., \& Bessell, M.~S.\ 2007, \aj, 133, 694
 \bibitem[De~Silva \etal(2015)]{galah}

### 2016-06-22 papers/nucleosynthesis/diversity.tex
start paper for Jan

% This file is part of the Platypus project.
% Copyright 2016 the authors.

% # to-do list
% - put in figures showing the G vectors.

\documentclass[12pt, preprint]{aastex}

% text commands
\newcommand{\equationname}{equation}
\newcommand{\acronym}[1]{{\small{#1}}}
\newcommand{\project}[1]{\textsl{#1}}
\newcommand{\apogee}{\project{\acronym{APOGEE}}}

% math commands
\newcommand{\solar}{\odot}
\newcommand{\abundance}[1]{\lbrack{#1}\rbrack}
\newcommand{\element}[1]{\mathrm{#1}}
\newcommand{\given}{\,|\,}
\newcommand{\normal}{\mathcal{N}}

\begin{document}

\title{No straightforward nucleosynthesis scenario can explain the
  observed diversity of stellar abundances}
\author{DWH, JR, HWR, others}

\begin{abstract}
The elements heavier than Li are believed to be produced in stars and
supernovae, and delivered to the interstellar medium by supernova
explosions and stellar winds.
There are finite supernova and wind channels: Chemical enrichment is
expected to be dominated by core-collapse supernovae, type Ia
supernovae, and AGB stars.
If this is correct, and even if there are substantial mass- and
metallicity-dependent yields, it is impossible for even very
permissive nucleosynthesis models to produce a high dimensional family
of chemical abundance patterns in stars.
Here we compare the output of a huge diversity of nucleosynthetic
scenarios (varying star-formation histories, infall and outflow
parameters, and finite-numbers effects) to chemical abundances
observed in the \apogee\ data.
Although we are liberal with models, we restrict our consideration of
the data to red-clump stars, so we aren't comparing chemical abundance
measurements in very different kinds of stars.
We show that there are necessarily directions in chemical-abundance
space that can have no diversity, but that these same dimensions
\emph{do} show significant diversity in the data.
No reasonable adjustment to yield tables can resolve this discrepancy;
there is significant variance in the data along too many axes.
There either must be other important processes, such as spallation, or
diversity in depletion of elements onto dust, or stochasticity in
supernova and wind yields, or chemical separation in ejecta and winds.
These results provide support and hope for the idea of \emph{chemical
tagging} as a tool to understand common stellar origins.
\end{abstract}

\keywords{
  ---
  World: Hello
  ---
}

\section{Introduction}

Nucleosynthesis is important.  Models are (at least slightly) wrong.
Data are now abundant. People have tried to measure dimensionality; that
isn't a well-posed question.

\acknowledgements
We thank Yuan-Sen Ting (Harvard) for valuable discussions.

\end{document}

### 2016-06-22 papers/nucleosynthesis/diversity.tex
tweak

@@
 measurements in very different kinds of stars.
-We show that there are necessarily directions in chemical-abundance
-space that can have no diversity, but that these same dimensions
-\emph{do} show significant diversity in the data.
-No reasonable adjustment to yield tables can resolve this discrepancy;
-there is significant variance in the data along too many axes.
-There either must be other important processes, such as spallation, or
+We show that---from a theoretical perspective---there are directions
+in chemical-abundance space that are forced to have vanishing
+diversity, but that these same dimensions---in the data---\emph{do}
+show significant diversity.
+We also show that no reasonable adjustment to yield tables can resolve
+this discrepancy; there is significant variance in the data along too
+many axes.
+There either must be other important processes (such as spallation), or
 diversity in depletion of elements onto dust, or stochasticity in

### 2016-07-04 code/kinematics/disperse.py
start integrating orbits

@@ -1,0 +1,31 @@
+# This file is part of the Platypus project.
+# Copyright 2016 David W. Hogg.
+
+import numpy as np
+
+CIRCULAR_VELOCITY = 220 * 1000. # m / s (Julianne Dalcanton)
+PARSEC = 3.086e16 # m (Google)
+GYR = 3.1556926e13 # s (Google)
+
+def flat_rotation_curve(x):
+    return -1. * CIRCULAR_VELOCITY * CIRCULAR_VELOCITY * x / np.dot(x, x)
+
+def leapfrog_step(delta_t, x_zero, v_half, accelerationlaw):
+    x_one = x_zero + v_half * delta_t
+    v_one_and_a_half = v_half + accelerationlaw(x_one) * delta_t
+    return x_one, v_one_and_a_half
+
+if __name__ == "__main__":
+    import pylab as plt
+    x_00 = np.array((8.0 * PARSEC, 0., 0.))
+    v_05 = np.array((0., CIRCULAR_VELOCITY, 0.))
+    nsteps = 1000
+    xs = np.zeros((nsteps, 3))
+    vs = np.zeros_like(xs)
+    dt = 0.01 * GYR
+    for n in range(nsteps):
+        x_00, v_05 = leapfrog_step(dt, x_00, v_05, flat_rotation_curve)
+        xs[n], vs[n] = x_00, v_05
+    plt.clf()
+    plt.plot(xs[:,0], xs[:,1], "k-", alpha=0.8)
+    plt.savefig("foo.png")

### 2016-07-05 code/kinematics/disperse.py
works

@@ -4,8 +4,8 @@
 import numpy as np
 
 CIRCULAR_VELOCITY = 220 * 1000. # m / s (Julianne Dalcanton)
-PARSEC = 3.086e16 # m (Google)
-GYR = 3.1556926e13 # s (Google)
+KPC = 3.086e19 # m (Google)
+GYR = 3.1556926e16 # s (Google)
 
 def flat_rotation_curve(x):
     return -1. * CIRCULAR_VELOCITY * CIRCULAR_VELOCITY * x / np.dot(x, x)
@@ -17,15 +17,42 @@ def leapfrog_step(delta_t, x_zero, v_half, accelerationlaw):
 
 if __name__ == "__main__":
     import pylab as plt
-    x_00 = np.array((8.0 * PARSEC, 0., 0.))
-    v_05 = np.array((0., CIRCULAR_VELOCITY, 0.))
-    nsteps = 1000
-    xs = np.zeros((nsteps, 3))
-    vs = np.zeros_like(xs)
-    dt = 0.01 * GYR
+    x0 = np.array((8.0 * KPC, 0., 0.))
+    v0 = np.array((0., CIRCULAR_VELOCITY, 0.))
+    x1 = 1. * x0
+    v1 = 1. * v0 + 1000. * np.random.normal(3) # add a 1 km/s dispersion
+    v0 = 1. * v0 + 1000. * np.random.normal(3) # add a 1 km/s dispersion
+    dv = np.sqrt(np.dot(v0 - v1, v0 - v1))
+    nsteps = 10000
+    x0s = np.zeros((nsteps, 3))
+    v0s = np.zeros_like(x0s)
+    x1s = np.zeros_like(x0s)
+    v1s = np.zeros_like(x0s)
+    dt = 0.0003 * GYR
     for n in range(nsteps):
-        x_00, v_05 = leapfrog_step(dt, x_00, v_05, flat_rotation_curve)
-        xs[n], vs[n] = x_00, v_05
+        x0, v0 = leapfrog_step(dt, x0, v0, flat_rotation_curve)
+        x1, v1 = leapfrog_step(dt, x1, v1, flat_rotation_curve)
+        x0s[n], v0s[n] = x0, v0
+        x1s[n], v1s[n] = x1, v1
+    ts = dt * np.arange(nsteps)
     plt.clf()
[…]

### 2016-08-06 papers/kmeans/response_to_referee.txt
authors, affils, response text

@@
 Please find below the response to our (very helpful) referee report on
-manuscript ApJ/ApJ101852.  We have taken all of the referees
+manuscript ApJ/ApJ101852.  We have taken all of the referee's
 suggestions, and made some other changes from comments we have
@@
 
-HOGG
+This statement and point is true and wise. We have modified the text
+in [HOGG]
 
@@
 
-We have added text to clarify why k-means was chosen.
+We have added text to clarify why k-means was chosen. It is
+fundamentally because k-means is the clustering algorithm of "first
+resort", which contains very few assumptions and only one (integer)
+adjustable parameter (not counting the metric on the vector space).
 
@@
 
-Because of a large number of valuable contributions both prior to and
-subsequent to our submission of this paper, we expanded the author
-list.  There are now several additional new authors, most of whom
-contributed by building the APOGEE project (and hence the input data).
+Because of discussions and checking of results that happened
+subsequent to our submission of this paper with the APOGEE team, some
+of which made use of proprietary data and information about the
+survey, we have expanded the author list in accordance with the
+APOGEE/SDSS-III authorship policies.  The consequence is that there
+are now several additional new authors, most of whom contributed to
+this project most directly by building the APOGEE project (and hence
+the input data and tests thereof).
 
-We added a few citations: Eisenstein 2011 (SDSS-III), Holtzman 2015
-(APOGEE DR12) and ...HOGG
+We added the two crucial APOGEE references: Eisenstein 2011 (SDSS-III)
+Holtzman 2015 (APOGEE DR12).
 
@@
 
-HOGG:  Go through github issues and emails.
+HOGG:  Commented on, eg, the Na abundances?
+
+HOGG:  Go through github issues and emails?
 

### 2016-08-06 papers/kmeans/kmeans.tex
omg figurename too?
frenchspacing
nothing
authors, affils, response text

@@
 \makeatother
+\frenchspacing
+\sloppy\sloppypar
 
@@
 \newcommand{\sectionname}{Section}
+\renewcommand{\figurename}{Figure} % omg AASTEX 6
 
@@
         Daniel~Foreman-Mackey\altaffilmark{6,7},
-        Anna~Y.~Q.~Ho\altaffilmark{8},
-        Jon~A.~Holtzman\altaffilmark{9},
-        David~L.~Nidever\altaffilmark{10}}
+        Sten~Hasselquist\altaffilmark{8,9},
+        Anna~Y.~Q.~Ho\altaffilmark{10},
+        Jon~A.~Holtzman\altaffilmark{8},
+        Steven~R.~Majewski\altaffilmark{9},
+        Sarah~L.~Martell\altaffilmark{11},
+        Szabolcs~M\'esz\'aros\altaffilmark{12},
+        David~L.~Nidever\altaffilmark{13},
+        Matthew~Shetrone\altaffilmark{14}}
 \altaffiltext{1}{Simons Center for Data Analysis, 160 Fifth Avenue, 7th floor, New York, NY 10010, USA}
@@
 \altaffiltext{7}{Department of Astronomy, University of Washington, Box 351580, Seattle, WA 98195, USA}
-\altaffiltext{8}{Astronomy Department, California Institute of Technology, MC 249-17, 1200 East California Blvd, Pasadena, CA 91125, USA}
-\altaffiltext{9}{New Mexico State University, Las Cruces, NM 88003, USA}
-\altaffiltext{10}{Steward Observatory, 933 North Cherry Ave, Tuscon, AZ 85719, USA}
+\altaffiltext{8}{New Mexico State University, Las Cruces, NM 88003, USA}
+\altaffiltext{9}{Department of Astronomy, University of Virginia, Charlottesville, VA 22904-4325, USA}
+\altaffiltext{10}{Astronomy Department, California Institute of Technology, MC 249-17, 1200 East California Blvd, Pasadena, CA 91125, USA}
+\altaffiltext{11}{School of Physics, University of South Wales, Sydney 2052, Australia}
+\altaffiltext{12}{ELTE Gothard Astrophysical Observatory, H-9704 Szombathely, Szent Imre Herceg st. 112, Hungary}
+\altaffiltext{13}{Steward Observatory, 933 North Cherry Ave, Tuscon, AZ 85719, USA}
+\altaffiltext{14}{University of Texas at Austin, McDonald Observatory, USA}
 
@@
 However, we have succeeded in substantially improving spectroscopic
-measurements with \thecannon, which has delivered $15$ individual
+measurements with \thecannon, which has now delivered $15$ individual
 abundances for $\sim 10^5$ stars observed as part of the
@@
 Although we have by no means any model of the 15-d space, we do know
-that the structures shown in the figures are very high in 15-d
+that the structures shown in the \figurename s are very high in 15-d
 density.
@@
 It is a pleasure to thank
-  Jeremy~Magland (\acronym{SCDA}),
-  Don Schneider (PSU), and
-  the Blanton--Hogg group meeting
+  Katia Cunha (NOAO),
+  Jeremy Magland (\acronym{SCDA}),
+  Don Schneider (PSU),
+  Keivan Stassun (Vanderbilt),
+  and the Blanton--Hogg group meeting
 for valuable comments and discussions.
 This project was funded in part by
-  the \acronym{NSF} (grants \acronym{IIS-1124794}, \acronym{AST-1517237}),
+  the \acronym{NSF} (grants \acronym{IIS-1124794}, \acronym{AST-1312863}, \acronym{AST-1517237}),
   \acronym{NASA} (grant \acronym{NNX12AI50G}),

### 2016-08-06 papers/disruption/birthdays.py
new project: birthday paradox

@@ -1,0 +1,74 @@
+"""
+This file is part of the Platypus project.
+Copyright 2016 David W. Hogg (NYU).
+
+# Purpose
+- Develop the analog of the birthday paradox for constraining the
+  formation of the Milky Way halo.
+- Rule out stupid models of the Milky Way halo.
+
+## Issues
+- Not yet written
+"""
+
+import numpy as np
+import pylab as plt
+
+def _make_one_tag(fsmooth, Neff):
+    if np.random.uniform() < fsmooth:
+        return 0
+    return np.random.randint(1, Neff + 1)
+
+
+def make_tags(N, fsmooth, Neff, verbose=False):
+    """
+    Tag stars with tags 0...`Neff`.
+
+    # inputs:
+    - `N`: number of stars to tag.
+    - `fsmooth`: fraction of stars to tag with a zero.
+    - `Neff`: number of non-zero tags.
+
+    # outputs:
+    - `np.ndarray` of tags
+    """
+    assert fsmooth >= 0.
+    assert fsmooth <= 1.
+    if verbose:
+        print("Making N={} tags for fsmooth={}, Neff={}".format(N, fsmooth, Neff))
+    return np.array([_make_one_tag(fsmooth, Neff) for n in range(N)])
+
+def coincidence(tags, verbose=False):
+    nztags = np.sort(tags[tags > 0])
+    bools = nztags[1:] == nztags[:-1]
+    if verbose:
+        if np.any(bools):
+            print("YES: shared non-zero tags: ", (nztags[1:])[bools])
+        else:
+            print("NO")
+    return np.sum(bools)
+
+def plot_one_case(Nstr, fsmoothstr, Neffstr):
+    fsmooth = float(fsmoothstr)
+    N = int(Nstr)
+    Neff = int(Neffstr)
+    ns = np.zeros(10000).astype("int")
+    for i in range(len(ns)):
+        tags = make_tags(N, fsmooth, Neff)
+        ns[i] = coincidence(tags)
+    print(ns)
+    plt.clf()
+    plt.hist(ns, bins=np.arange(np.max(ns))-0.5, facecolor="none")
+    plt.title("N = {} / fsmooth = {} / Neff = {}".format(Nstr, fsmoothstr, Neffstr))
+    foo, xt = plt.xlim()
+    foo, yt = plt.ylim()
[…]

### 2016-08-06 papers/kmeans/response_to_referee.txt
majewski comments

@@
 
+We softened very slightly the "no previous successes" sentence in the
+abstract and also cited Majewski et al 2012. We also added a caveat
+that globular-cluster members are over-represented in APOGEE because
+of targeting decisions.
+
 HOGG:  Adjusted what we say about Ting et al?
@@
 We also added a few names to the acknowledgements for providing minor
-comments.
+comments, and made other small, non-content-changing corrections to
+wording.

### 2016-08-06 papers/kmeans/kmeans.tex
majewski comments

@@
        Identification of stellar phase-space structures \\
-       purely by chemical-abundance similarity}
+       purely by chemical-abundance similarity\vspace{-7ex}}
 \author{David~W.~Hogg\altaffilmark{1,2,3,4},
@@
 (in the same molecular cloud), long ago.
-This idea has not previously yielded scientific successes, probably because of the
+This idea has not yielded much success, probably because of the
 noise and incompleteness in chemical-abundance measurements.
@@
 \citealt{eggen_1970,font,De_Silva_2007,bubar_king_2010}) and
-stellar-orbit diffusion process like radial mixing and radial
+stellar-orbit diffusion processes like radial mixing and radial
 migration (for example, \citealt{roskar, quillen}).  After
@@
 disrupted---associations or groups will disperse, through
-two-body processes, interactions with resonances, or tides from the whole Galaxy.
+two-body mechanisms, interactions with resonances, or tides from the whole Galaxy.
 
@@
 and open clusters \citep{bovy} indicate that chemical abundances ought to be
-very similar for co-eval stars.
+very similar for co-eval stars, and peculiar abundance ratios
+have been successfully used to identify disrupted cluster members (for example, \citealt{ocen}).
 
@@
 The \apogee\ data set covers a huge radial extent and---because the
-data are taken in the infrared---explores the thin disk.
+data are taken in the infrared---is capable of exploring all parts of the disk,
+including the thin parts.
 However, it has the disadvantage that its spatial coverage is incomplete
@@
 In many ways, \galah\ will deliver improvements, both because it will
-have more abundances (possibly 29) and contiguous sky coverage.
+have more abundances (possibly 29) and contiguous sky coverage; that
+said, it will not observe much of the thin disk.
 
@@
 find stars purely by their clustering in abundance space and subsequently
-show that they are part of a still spatially coherent cluster, then
+show that they are part of a still spatially coherent cluster, group, or stream, then
 we will have resolved all practical outstanding issues plaguing 
@@
 \caption{Same as \figurename~\ref{fig:all} except that the entire
-  sample has been made gray and the members of the most dense
-  $K=256$ abundance-space cluster have been rendered as unique, prominent triangles
-  (color is velocity rank, orientation is $\logg$ rank).
-  The lower-right plot shows this cluster (circle) in context of the other clusters (dots).
-  This cluster, which was identified only in abundance space (six
-  projections of which are the top six panels in this \figurename),
-  turns out to be dominated by the halo globular cluster
-  M13.\label{fig:M13}}
+  sample has been made gray and the members of the most dense $K=256$
+  abundance-space cluster have been rendered as unique, prominent
+  triangles (color is velocity rank, orientation is $\logg$ rank).
+  The lower-right plot shows this cluster (circle) in context of the
+  other clusters (dots).  This cluster, which was identified only in
+  abundance space (six projections of which are the top six panels in
+  this \figurename), turns out to be dominated by the halo globular
+  cluster M13. The symbol orientations and colors have no meaning;
+  they are randomly generated---one unique color and orientation for
+  each highlighted star---to make the points cross-identifiable across
+  panels.\label{fig:M13}}
 \end{figure}
@@
 severely impacted by these correlations, gives us great hope for more sophisticated approaches.
+That said, globular cluster members are over-represented in teh \apogee\ data set,
+because many of them were specifically targeted for observations for being cluster
+members \citep{apogee}; that is, globular-cluster structure is over-represented relative to the
+field in this data set.
 
@@
 shown here. The disparity between this structure and others presented
-here is the lack of co-spatial stars. 12 stars within this cluster are
+here is the lack of co-spatial stars. Twelve stars within this cluster are
 marked as `M53' candidates by \apogee, but the remaining 145 stars
@@
 indeed be all members of the same co-natal gas cloud, implying that 
-these stars have been accreted onto the halo by a single globular
+these stars have been accreted onto the halo from a single globular
 cluster early in the Milky Way's formation. While speculative,
@@
 chemical-abundance measurements we used here for further study.
-The future is even brighter, however: \apogee\ is expanding to 19
+The future is even brighter, however: \apogee\ is expanding to more measured
 elements, more stars, and all-sky angular coverage with \apogee2, and
-\galah\ is nearly ready to release chemical abundances on a larger set
+\galah\ is working towards releasing chemical abundances on a larger set
 of 29 elements.
@@
 Kuzma, P.~B., Da Costa, G.~S., Keller, S.~C., Maunder, E., 2015, \mnras, 446, 3297 
+\bibitem[Majewski \etal(2012)]{ocen}
+Majewski,~S.~R., Nidever,~D.~L., Smith,~V.~V., Damke,~G.~J., Kunkel,~W.~E.,
+Patterson,~R.~J., Bizyaev,~D., \& Garca~P ́erez~ A~ E,. 2012,\apjl, 747, L37
 \bibitem[Majewski \etal(2015)]{apogee}

### 2016-08-06 papers/kmeans/kmeans.tex
more corrections

@@
 The first three of these are dominated by stars in the known clusters
-M13, Pal5, and Sagittarius.
+M13 and Pal5, and the Sagittarius stream (we identify overlap with these objects
+by looking at stellar position and velocity, and, in some cases,
+\apogee\ targeting flags).
 The fourth is a halo structure with high velocity dispersion and
@@
 It is a pleasure to thank
-  Katia Cunha (NOAO),
+  Joss Bland-Hawthorn (Sydney),
+  Jo Bovy (Toronto),
+  Charlie Conroy (Harvard),
+  Katia Cunha (\acronym{NOAO}),
+  Amina Helmi (Kapteyn),
   Jeremy Magland (\acronym{SCDA}),
-  Don Schneider (PSU),
+  Don Schneider (\acronym{PSU}),
   Keivan Stassun (Vanderbilt),

### 2016-08-06 papers/kmeans/response_to_referee.txt
more corrections
requested citations

@@
 
+We fixed a small bug with the cluster density measure and updated the
+figure. This changes no conclusions.
+
+We fixed the (embarrassing) bug that everywhere we say "Pal 5" we mean
+"M5". This was a straight-up mistake in the paper, and leads to some
+changes in the discussion, including some reference changes. No
+fundamental conclusions change because of this.
+
 HOGG:  Adjusted what we say about Ting et al?

### 2016-08-06 papers/kmeans/kmeans.tex
requested citations

--- 2016-08-06 papers/kmeans/kmeans.tex
@@
 This idea---dubbed ``chemical tagging''
-\citep{freeman}---is one of the principal motivations for a number
+\citep{freeman, jbh}---is one of the principal motivations for a number
 of surveys, including \apogee\ \citep{apogee},
@@
 On the other hand, precise studies of stellar twins \citep{melendez, jofre}
-and open clusters \citep{bovy} indicate that chemical abundances ought to be
-very similar for co-eval stars, and peculiar abundance ratios
+indicate that pairs of stars can be found with unusually similar abundances, 
+open clusters show remarkably uniform chemical abundances \citep{bovy},
+and peculiar abundance ratios
 have been successfully used to identify disrupted cluster members (for example, \citealt{ocen}).
+There have also been hints seen of relationships between chemistry and kinematics
+(for example, \citealt{helmi, helmi2}).
 
@@
 The first three of these are dominated by stars in the known clusters
-M13, Pal5, and Sagittarius.
+M13 and M5, and the Sagittarius stream (we identify overlap with these objects
+by looking at stellar position and velocity, and, in some cases,
+\apogee\ targeting flags).
 The fourth is a halo structure with high velocity dispersion and
@@
   abundance-space cluster.
-  This one turns out to be dominated by the disrupting globular
-  cluster Palomar 5.\label{fig:Pal5}}
+  This one turns out to be dominated by globular
+  cluster M5.\label{fig:Pal5}}
 \end{figure}
@@
 
+%HOGG: THIS PARAGRAPH IS ALL WRONG!
 There are scant abundance measurements reported for Palomar 5, for which
@@
 The \apogee\ elements employed in this project include alpha
-(Mg, O, Si, Ca, Ti), light proton-capture (C, N, S), odd-Z (Na, Al, K)
+(Mg, O, Si, S, Ca, Ti), light proton-capture (C, N, S), odd-Z (Na, Al, K)
 and iron peak (Fe, Ni, V, Mn) elements.
@@
 It is a pleasure to thank
-  Katia Cunha (NOAO),
+  Joss Bland-Hawthorn (Sydney),
+  Jo Bovy (Toronto),
+  Charlie Conroy (Harvard),
+  Katia Cunha (\acronym{NOAO}),
+  Amina Helmi (Kapteyn),
   Jeremy Magland (\acronym{SCDA}),
-  Don Schneider (PSU),
+  Don Schneider (\acronym{PSU}),
   Keivan Stassun (Vanderbilt),
+  Angus Williams (Cambridge),
   and the Blanton--Hogg group meeting
@@
 Blanco-Cuaresma,~S., Soubiran,~C., Heiter,~U., \etal, 2015, \aap, 577, A47
+\bibitem[Bland-Hawthorn \etal(2010)]{jbh}
+Bland-Hawthorn,~J., Krumholz,~M.~R., \& Freeman,~K., 2010, \apj, 713, 166
 \bibitem[Bovy(2015)]{bovy}
@@
 Harris,~W.~E., 1996, \aj, 112, 1487
+\bibitem[Helmi \etal(2006)]{helmi}
+Helmi,~A., Navarro,~J.~F., Nordstr{\"o}m,~B., Holmberg,~J., Abadi,~M.~G., \& Steinmetz,~M., 2006, \mnras, 365, 1309
+\bibitem[Helmi \etal(2014)]{helmi2}
+Helmi,~A., Williams,~M., Freeman,~K.~C., Bland-Hawthorn,~J., \& De~Silva,~G., 2014, \apj, 791, 135
 \bibitem[Holtzman \etal(2015)]{apogeedr12}

### 2016-08-06 papers/kmeans/kmeans.tex
ting references; holtzmann comment

@@
 structure in surveys dominated by a large mix of stars from different
-origins and ages (and there was good reason to be pessimistic;
-\citealt{ting}).
+origins and ages.
+There was good reason to be pessimistic, but plausibly the increased
+precision obtained here by \thecannon\ made this possible; even tiny improvements
+in individual abundances can enormously increase contrast in the high-dimensional
+abundance space \citep{ting}.
 We have demonstrated with a straightforward (and in no way optimal)
@@
 functions.
+
+One clear result here is that it is easier for the clustering algorithm
+to find structures at low metallicity than at high metallicity:
+Some of our low-metallicity features are very compact in phase space,
+while none of our higher metallicity features are.
+This probably relates to the much higher background
+(in abundance space) of unclustered stars at higher metallicities than
+at low metallicities; in general it is easier to find abundance-space features
+at the edges of the distribution than in the center \citep{ting}.
 
@@
 
-%HOGG: THIS PARAGRAPH IS ALL WRONG!
-There are scant abundance measurements reported for Palomar 5, for which
-most stars in the cluster in \figurename~\ref{fig:Pal5} belong. We find a mean
-metallicity of [Fe/H]$ = -1.3$, which lies between the only
-photometric \citep{Geisler_1997} and spectroscopic \citep{Smith_1985,Smith_2002} abundance
-measurements for this cluster.  The anti-correlation in C--N and Na--O abundance
-have been seen elsewhere \citep[][respectively]{Smith_1985,Smith_2002}, although
-to our knowledge the Mg--Al anti-correlation has not previously been shown for this cluster.
-Indeed, this work contains the largest number of abundances for Palomar 5,
-permitting a thorough search of tidally disrupted members \citep[for example,][]{Kuzma_2015} 
-that have retained the unique cluster abundance signature.
+% HOGG: THIS PARAGRAPH IS ALL WRONG!
+% There are scant abundance measurements reported for Palomar 5, for which
+% most stars in the cluster in \figurename~\ref{fig:Pal5} belong. We find a mean
+% metallicity of [Fe/H]$ = -1.3$, which lies between the only
+% photometric \citep{Geisler_1997} and spectroscopic \citep{Smith_1985,Smith_2002} abundance
+% measurements for this cluster.  The anti-correlation in C--N and Na--O abundance
+% have been seen elsewhere \citep[][respectively]{Smith_1985,Smith_2002}, although
+% to our knowledge the Mg--Al anti-correlation has not previously been shown for this cluster.
+% Indeed, this work contains the largest number of abundances for Palomar 5,
+% permitting a thorough search of tidally disrupted members \citep[for example,][]{Kuzma_2015} 
+% that have retained the unique cluster abundance signature.
+
+[HOGG, CASEY: INSERT PARAGRAPH ABOUT M5 HERE]
 

### 2016-08-06 papers/kmeans/response_to_referee.txt
tweaking

@@
 
-We have referenced Mitschang et al. (2014) for their work in chemical
+We have referenced Mitschang et al (2014) for their work in chemical
 tagging, but we believe the last sentence of the abstract is
-correct. To clarify: Mitschang et al. (2014) did perform a blind
+correct. To clarify: Mitschang et al (2014) did perform a blind
 search in chemical abundances to identify substructure, but one of the
@@
 represent distinct formation sites, e.g. as in dissolved open
-clusters, as previously thought". While the Mitschang et al. (2014)
+clusters, as previously thought". While the Mitschang et al (2014)
 work showed their chemically-identified structures had similar (U, V,
@@
 We have added the following references: De Silva et al (2007), Eggen
-(1970), and Bubar & King (2010).
+(1970), and Bubar & King (2010). Also Helmi et al (2006) and (2014).
 
@@
 knowledge) was the first to unambiguously show anti- correlations in
-Na, Al, and O, and to Carretta et al. (2009), which provides a more
+Na, Al, and O, and to Carretta et al (2009), which provides a more
 recent and comprehensive view of our current knowledge of elemental
@@
 
-We added the two crucial APOGEE references: Eisenstein 2011 (SDSS-III)
-Holtzman 2015 (APOGEE DR12).
+We added the two crucial APOGEE references: Eisenstein et al (2011;
+SDSS-III) Holtzman et al (2015; APOGEE DR12).
 
 We softened very slightly the "no previous successes" sentence in the
-abstract and also cited Majewski et al 2012. We also added a caveat
+abstract and also cited Majewski et al (2012). We also added a caveat
 that globular-cluster members are over-represented in APOGEE because
@@
 
-HOGG:  Adjusted what we say about Ting et al?
+We noted the point (related to the referee's comments) that everything
+we find that is dense in phase space is at low metallicity; this is
+probably because of our sensitivity to clusters above background (in
+abundance space).
 
 HOGG:  Commented on, eg, the Na abundances?
-
-HOGG:  Go through github issues and emails?
 

### 2016-08-07 papers/kmeans/response_to_referee.txt
referee comments

@@
 
-This statement and point is true and wise. We have modified the text
-in [HOGG]
+This statement and point is true and wise. We have added a paragraph
+in Section 2 describing and emphasizing this point, and also
+addressing the point that The Cannon can use element-element
+correlations to improve its precision. All these things reduce the
+interpretability of the absolute abundances shown in the paper, as the
+referee notes. We refer to the Casey et al paper to see the
+performance of the method at the edges.
 
@@
 resort", which contains very few assumptions and only one (integer)
-adjustable parameter (not counting the metric on the vector space).
+adjustable parameter (not counting the metric on the vector space);
+it is also well tested and understood.
 
@@
 
-HOGG:  Commented on, eg, the Na abundances?
-
 We also added a few names to the acknowledgements for providing minor

### 2016-08-07 papers/kmeans/kmeans.tex
referee comments

@@
 
+HOGG: Paragraph here about limitations of \thecannon: Dependence on
+training set, makes use of element-element correlations (especially, say, Na),
+and delivers precision but no guarantees of accuracy beyond that in the training set.
+Comment on its performance at the edges of the abundance distribution...? As per referee issue.
+
 Six two-dimensional projections of the abundance data, plus some other
@@
 algorithm or the method that reveals the best structures.
-We chose k-means as a simple and tractable approach to identifying
-clusters in high dimensional label space. In practice more complex
+We chose k-means as a simple and straightforward approach to identifying
+clusters in high dimensional label space, with very few control parameters
+or decisions required.
+It is also an algorithm that is well studied in the machine-learning literature,
+so it has well understood properties.
+In practice more complex
 clustering algorithms may perform better than k-means, or even

### 2016-08-07 papers/kmeans/response_to_referee.txt
Szabolcs comments

@@
 
-This statement and point is true and wise. We have added a paragraph
-in Section 2 describing and emphasizing this point, and also
-addressing the point that The Cannon can use element-element
-correlations to improve its precision. All these things reduce the
-interpretability of the absolute abundances shown in the paper, as the
-referee notes. We refer to the Casey et al paper to see the
-performance of the method at the edges.
+This statement and point is true and wise. We have added a long and
+substantial paragraph in Section 2 describing and emphasizing this
+point, and also addressing the point that The Cannon can use
+element-element correlations to improve its precision. All these
+things reduce the interpretability of the absolute abundances shown in
+the paper, as the referee notes. We also discuss the concern that
+things could diverge at the boundaries of the training set.
+
+We also add a paragraph in the discussion bringing up issues with
+Na/Fe abundances that relate to this referee comment.
 
@@
 
-HOGG
+The lower-right panel of Fig 6 shows that the group a strong
+over-density in abundance space. The referee makes a good point: It is
+indeed conceivable that this is created somehow by issues with The
+Cannon at low metallicity. We now include a couple of sentences of
+discussion of this point in the discussion section.  We make the
+prediction for this structure that if it is followed up with data that
+are sensitive to more elements, it could (in principle) be shown to
+have abundances that are consistent with cluster or dwarf-galaxy
+formation; if this succeeded, it might represent a discovery related
+to the accretion formation of the halo.
 

### 2016-08-07 papers/kmeans/kmeans.tex
Szabolcs comments

@@
 
-HOGG: Paragraph here about limitations of \thecannon: Dependence on
-training set, makes use of element-element correlations (especially, say, Na),
-and delivers precision but no guarantees of accuracy beyond that in the training set.
-Comment on its performance at the edges of the abundance distribution...? As per referee issue.
+It is important here to note some of the limitations of stellar labels
+delivered by \thecannon.
+The first is that \thecannon\ is only as good as its training set!
+All of the biases and calibration issues in the input training set will be
+delivered to the output labels.
+This means that there is no sense in which \thecannon\ delivers
+absolute abundances any better than \aspcap\ does.
+The second is that \thecannon\ (in the form used here) is aggressively
+data-driven.
+It will use anything it can to measure, say, the Na abundance, not just
+actual Na lines.
+This means that population-level correlations in  element abundances are
+being used to deliver information about individual elements.
+We bring up Na as an example here for the important reason that at low
+metallicity, there are no significant Na lines in the spectra; we are
+measuring Na only indirectly at low metallicity.
+A third limitation of \thecannon\ is that it is operating a regression
+in a very high dimensional label space.
+This regression is hard to test and validate near the edges of the range
+of applicability.
+One way to say this is that the convex hull of the training set of
+points in 17-dimensional space is potentially very small, and it is
+hard to visualize what is going on outside that hull.
+For these reasons, there might be label biases that grow with displacement
+from the bulk of the training data.
 
@@
 
+All that said, we reiterate a caveat here that we also mentioned in
+\sectionname~\ref{sec:data}, which is that at very low metallicity, Na
+does not show strong lines in any \apogee\ spectrum.  For this reason,
+the [Na/Fe] shown for the low-metallicity structures are obtained not
+by measuring Na lines but rather the lines of elements that correlate
+strongly with Na at the population level.
+The true [Na/Fe] may show correlations, variations, or anomalies in
+the clusters that are not captured by \thecannon\ working at low
+metallicity.
+
 The high velocity-dispersion structure seen in \figurename~\ref{fig:halo} stands
@@
 are spread amongst fields throughout the halo.
+
+It is conceivable that this structure, being near the outskirts of the
+abundance distribution, is somehow concentrated in abundance space by
+some kind of shrinkage (in the statistical sense) induced by
+\thecannon's regression.
+However, it is a very dense feature in abundance space, and is worth
+following up.
+If follow-up observations show that the stars have abundances and that
+are consistent with having origin in a single stellar population or
+low-metallicty dwarf galaxy, it could represent an accretion event in
+the Milky Way halo.
 
@@
 \acknowledgements
-It is a pleasure to thank
+It is a pleasure to thank an anonymous referee for useful comments that have led to improvements in the manuscript.
+We also thank
   Joss Bland-Hawthorn (Sydney),
@@
   and the Blanton--Hogg group meeting
-for valuable comments and discussions.
+for valuable discussions and comments.
 This project was funded in part by

### 2016-08-07 papers/kmeans/kmeans.tex
Fix teh typo; added a 'the'

@@
 severely impacted by these correlations, gives us great hope for more sophisticated approaches.
-That said, globular cluster members are over-represented in teh \apogee\ data set,
+That said, globular cluster members are over-represented in the \apogee\ data set,
 because many of them were specifically targeted for observations for being cluster
@@
 Because some stars ($<8\%$) are candidate members of the massive globular
-cluster M53, it provides tantalizing possibility that these stars may
+cluster M53, it provides the tantalizing possibility that these stars may
 indeed be all members of the same co-natal gas cloud, implying that 

### 2016-08-07 papers/kmeans/kmeans.tex
Clarified heavy/slow neutron-capture, and shut down argument about 'what is an alpha element'

@@
 of 29 elements.
-The \apogee\ elements employed in this project include alpha
-(Mg, O, Si, S, Ca, Ti), light proton-capture (C, N, S), odd-Z (Na, Al, K)
-and iron peak (Fe, Ni, V, Mn) elements.
+The \apogee\ elements employed in this project include alpha, light proton-capture, odd-Z, and iron peak elements.
 \galah\ will deliver element abundances from all the major
 nucleosynthetic processes, including light proton capture elements,
-alpha, odd-Z, iron-peak, light and heavy slow neutron capture, and
-rapid neutron capture.
+alpha, odd-Z, iron-peak, as well as neutron-capture elements.
 Chemical tagging capabilities are expected to grow as the

### 2016-08-07 papers/kmeans/kmeans.tex
Removed wrong paragraph on Pal5 and associated references. Added better paragraph for M5

@@
 in the $K=256$ run and show them in
-\figurename~\ref{fig:M13}, \ref{fig:Pal5}, \ref{fig:Sgr},
+\figurename~\ref{fig:M13}, \ref{fig:M5}, \ref{fig:Sgr},
 \ref{fig:halo}, and \ref{fig:disk}.
@@
 Although we show only three stellar clusters in
-\figurename~\ref{fig:M13}, \ref{fig:Pal5}, and \ref{fig:Sgr}, many
+\figurename~\ref{fig:M13}, \ref{fig:M5}, and \ref{fig:Sgr}, many
 other clusters are visible among our densest abundance-space k-means
@@
   This one turns out to be dominated by globular
-  cluster M5.\label{fig:Pal5}}
+  cluster M5.\label{fig:M5}}
 \end{figure}
@@
 
-% HOGG: THIS PARAGRAPH IS ALL WRONG!
-% There are scant abundance measurements reported for Palomar 5, for which
-% most stars in the cluster in \figurename~\ref{fig:Pal5} belong. We find a mean
-% metallicity of [Fe/H]$ = -1.3$, which lies between the only
-% photometric \citep{Geisler_1997} and spectroscopic \citep{Smith_1985,Smith_2002} abundance
-% measurements for this cluster.  The anti-correlation in C--N and Na--O abundance
-% have been seen elsewhere \citep[][respectively]{Smith_1985,Smith_2002}, although
-% to our knowledge the Mg--Al anti-correlation has not previously been shown for this cluster.
-% Indeed, this work contains the largest number of abundances for Palomar 5,
-% permitting a thorough search of tidally disrupted members \citep[for example,][]{Kuzma_2015} 
-% that have retained the unique cluster abundance signature.
-
-[HOGG, CASEY: INSERT PARAGRAPH ABOUT M5 HERE]
+We find a mean metallicity of [Fe/H]$ = -1.3$ for the cluster we associate as M~5 (\figurename~\ref{fig:M5}), which agrees well with the [Fe/H]$ = -1.33 \pm 0.03$ measurement reported by \citet{Koch_2010}, and the $-1.29 \pm 0.02$ value listed in the compilation of \citet[][accessed 2016]{Harris_1996}.
+The correlations in light elements---specifically Mg--Al and C--N---also agree well with other studies \citep{Ivans_2001,meszaros}.
+In particular we find only a weak correlation in the [Na/Fe]--[O/Fe] abundance ratios \citep{Lai_2011}.
+However, when we consider the extent of the literature on M~5, it would suggest that we do not recover the full extent of these correlations: the stars with the highest [Na/Fe], [Mg/Al], and lowest [O/Fe] abundance ratios are not represented in the cluster that we associate as M~5.
+This is likely a consequence of our (poor) choice of the k-means algorithm, which is most effective for near-normal distributions, and clearly less powerful for clusters (in a statistical sense) that have polynomial relationships in dimensional space (e.g., [Na/Fe]--[O/Fe]).
+Indeed, \citet{meszaros} report abundances for 122 members of M~5 using \apogee\ \acronym{DR12} data, whereas the cluster we associate as M~5 only contains $\approx60$ members, and only covers about half of the extent of [Na/Fe]--[O/Fe] relationship.
 
@@
 There is no sense at all in which \figurename~\ref{fig:M13},
-\ref{fig:Pal5}, \ref{fig:Sgr}, \ref{fig:halo}, and \ref{fig:disk} show
+\ref{fig:M5}, \ref{fig:Sgr}, \ref{fig:halo}, and \ref{fig:disk} show
 a representative or complete set of features found in abundance space.
@@
 Garc{\'{\i}}a~P{\'e}rez,~A.~E., Allende~Prieto,~C., Holtzman,~J.~A., \etal, 2015, arXiv:1510.07635
-\bibitem[Geisler \etal(1997)]{Geisler_1997}
-Geisler, D., Claria, J.~J., Minniti, D., 1997, \pasp, 109, 799 
 \bibitem[Gilmore \etal(2012)]{gaiaeso}
@@
 Hunter,~J.~D., 2007, Computing in Science \& Engineering, 9, 90
+\bibitem[Ivans et al.(2001)]{Ivans_2001} Ivans, I.~I., Kraft, R.~P., Sneden, C., et al.\ 2001, \aj, 122, 1438 
 \bibitem[Jofr{\'e} \etal(2015)]{jofre}
@@
 Johnson, C.~I.,  Pilachowski, C.~A., 2012, \apjl, 754, L38 
+\bibitem[Koch \& McWilliam(2010)]{Koch_2010} Koch, A., \& McWilliam, A.\ 2010, \aj, 139, 2289 
 \bibitem[Kraft \etal(1992)]{Kraft_1992}
 Kraft, R.~P., Sneden, C., Langer, G.~E.,  Prosser, C.~F., 1992, \aj, 104, 645 
-\bibitem[Kuzma \etal(2015)]{Kuzma_2015}
-Kuzma, P.~B., Da Costa, G.~S., Keller, S.~C., Maunder, E., 2015, \mnras, 446, 3297 
+\bibitem[Lai et al.(2011)]{Lai_2011} Lai, D.~K., Smith, G.~H., Bolte, M., et al.\ 2011, \aj, 141, 62 
 \bibitem[Majewski \etal(2012)]{ocen}
@@
 Ro{\v s}kar,~R., Debattista,~V.~P., Quinn,~T.~R., Stinson,~G.~S., Wadsley,~J., 2008, \apjl, 684, L79
-\bibitem[Smith(1985)]{Smith_1985}
-Smith, G.~H., 1985, \apj, 298, 249 
-\bibitem[Smith \etal(2002)]{Smith_2002}
-Smith, G.~H., Sneden, C., Kraft, R.~P., 2002, \aj, 123, 1502 
 \bibitem[Smith \etal(2005)]{Smith_2005}

### 2016-08-09 code/kmeans.py
make pickle for MKN and fix .th. bug

@@ -128,7 +128,7 @@ def plot_cluster_context(sizes, densities, dir, name=None, k=None, suffix="png")
         fn = "{}/clusters_{:04d}.{}".format(dir, K, suffix)
     else:
         fn = "{}/{}_context.{}".format(dir, name, suffix)
-    if os.th.exists(fn):
+    if os.path.exists(fn):
         print("plot_cluster_context(): {} exists already".format(fn))
         return
     if k is None:
@@ -330,6 +330,12 @@ if __name__ == "__main__":
                 plotdata_labels[d] = data_labels[d] + " - " + data_labels[0]
             metadata_labels = ["RA", "DEC", "GLON", "GLAT", "VHELIO_AVG", "TEFF_ASPCAP", "LOGG_ASPCAP", "FIELD", "APOGEE_ID"]
             metadata = get_metadata(dfn, metadata_labels, mask)
+
+            mknpfn = "data/mkn_{:04d}.pkl".format(K)
+            print("writing pickle")
+            pickle_to_file(mknpfn, (data, metadata))
+            print(mknpfn)
+
             fields =   metadata[:, -2]
             names  =   metadata[:, -1]
             metadata = metadata[:, 0:-2].astype(float)

### 2016-08-12 papers/kmeans/kmeans.tex
ready to resubmit

@@
 (in the same molecular cloud), long ago.
-This idea has not yielded much success, probably because of the
+This idea has not yielded much practical success, presumably because of the
 noise and incompleteness in chemical-abundance measurements.
-However, we have succeeded in substantially improving spectroscopic
+We have succeeded in substantially improving spectroscopic
 measurements with \thecannon, which has now delivered $15$ individual
@@
 We recover some known phase-space clusters and find other interesting structures.
-This confirms the chemical-tagging hypothesis and verifies the precision of the abundance measurements delivered by \thecannon.
-This is the first-ever project to identify phase-space structures at survey-scale by blind search purely in abundance space;
-the prospects for future data sets are very good.
+This is the first-ever project to identify phase-space structures
+at survey-scale by blind search purely in abundance space;
+it verifies the precision of the abundance measurements delivered by \thecannon;
+the prospects for future data sets appear very good.
 \end{abstract}
@@
 In many ways, \galah\ will deliver improvements, both because it will
-have more abundances (possibly 29) and contiguous sky coverage; that
+have more abundances (possibly 29) and contiguous sky coverage \citep{galah}; that
 said, it will not observe much of the thin disk.
@@
 observational precision.
+Finally and related, because it looks for clusters compact in
+metric distance, k-means is more sensitive to clusters that are spherical
+in the scaled abundance space than clusters of the same density
+that are elongated in any sense.
 Importantly we use \emph{only} abundance-space information, and no
@@
 
-We find a mean metallicity of [Fe/H]$ = -1.3$ for the cluster we associate as M~5 (\figurename~\ref{fig:M5}), which agrees well with the [Fe/H]$ = -1.33 \pm 0.03$ measurement reported by \citet{Koch_2010}, and the $-1.29 \pm 0.02$ value listed in the compilation of \citet[][accessed 2016]{Harris_1996}.
-The correlations in light elements---specifically Mg--Al and C--N---also agree well with other studies \citep{Ivans_2001,meszaros}.
-In particular we find only a weak correlation in the [Na/Fe]--[O/Fe] abundance ratios \citep{Lai_2011}.
-However, when we consider the extent of the literature on M~5, it would suggest that we do not recover the full extent of these correlations: the stars with the highest [Na/Fe], [Mg/Al], and lowest [O/Fe] abundance ratios are not represented in the cluster that we associate as M~5.
-This is likely a consequence of our (poor) choice of the k-means algorithm, which is most effective for near-normal distributions, and clearly less powerful for clusters (in a statistical sense) that have polynomial relationships in dimensional space (e.g., [Na/Fe]--[O/Fe]).
-Indeed, \citet{meszaros} report abundances for 122 members of M~5 using \apogee\ \acronym{DR12} data, whereas the cluster we associate as M~5 only contains $\approx60$ members, and only covers about half of the extent of [Na/Fe]--[O/Fe] relationship.
+We find a mean metallicity of [Fe/H]$ = -1.3$ for the cluster we
+associate as M~5 (\figurename~\ref{fig:M5}), which agrees well with
+the [Fe/H]$ = -1.33 \pm 0.03$ measurement 
+\citep{Koch_2010}, and the $-1.29 \pm 0.02$ value listed in the
+standard compilation (\citealt{Harris_1996}; accessed 2016).  The correlations
+in light elements---specifically Mg--Al and C--N---also agree well
+with other studies \citep{Ivans_2001,meszaros}.  In particular we find
+only a weak correlation in the [Na/Fe]--[O/Fe] abundance ratios
+\citep{Lai_2011}.  However, when we consider the extent of the
+literature on M~5, it would suggest that we do not recover the full
+extent of these correlations: the stars with the highest [Na/Fe],
+[Mg/Al], and lowest [O/Fe] abundance ratios are not represented in the
+cluster that we associate as M~5.  This is likely a consequence of our
+(poor) choice of the k-means algorithm, which is most effective for
+near-circular distributions in the scaled abundance space, and reduced in power for clusters
+ that have polynomial relationships in dimensional
+space (for example, [Na/Fe]--[O/Fe]).  Indeed, 
+abundances for 122 members of M~5 have been reported in \apogee\ \acronym{DR12} data (\citealt{meszaros}),
+whereas the cluster we associate as M~5 only contains $\approx60$
+members, and only covers about half of the extent of [Na/Fe]--[O/Fe]
+relationship.
 
@@
 are consistent with having origin in a single stellar population or
-low-metallicty dwarf galaxy, it could represent an accretion event in
+low-metallicity dwarf galaxy, it could represent an accretion event in
 the Milky Way halo.
@@
 the Galaxy halo.
-Because some stars ($<8\%$) are candidate members of the massive globular
+Because some stars ($<8$\,percent) are candidate members of the massive globular
 cluster M53, it provides the tantalizing possibility that these stars may
@@
 these stars have been accreted onto the halo from a single globular
-cluster early in the Milky Way's formation. While speculative,
+cluster early in the Milky Way's formation (as
+has been hypothesized elsewhere; for example in \citealt{grebel}). While speculative,
 this idea demonstrates the promise of chemical tagging.
@@
   \acronym{NASA} (grant \acronym{NNX12AI50G}),
-  the Moore-Sloan Data Science Environment at \acronym{NYU}, and
+  the Moore-Sloan Data Science Environment at \acronym{NYU},
+  the Australian Research Council (DECRA Fellowship DE140100598),
+  and
   the European Research Council under the
@@
 Martell,~S.~L., 2015, arXiv:1507.00079
+\bibitem[Martell \& Grebel(2010)]{grebel}
+Martell, S.~L., \& Grebel, E.~K., 2010, \aap, 519, A14 
 \bibitem[Martig \etal(2016)]{martig}

### 2016-08-13 papers/kmeans/kmeans.tex
final tweaks

@@
         Daniel~Foreman-Mackey\altaffilmark{6,7},
-        Sten~Hasselquist\altaffilmark{8,9},
-        Anna~Y.~Q.~Ho\altaffilmark{10},
+        Sten~Hasselquist\altaffilmark{8},
+        Anna~Y.~Q.~Ho\altaffilmark{9},
         Jon~A.~Holtzman\altaffilmark{8},
-        Steven~R.~Majewski\altaffilmark{9},
+        Steven~R.~Majewski\altaffilmark{10},
         Sarah~L.~Martell\altaffilmark{11},
@@
 \altaffiltext{8}{New Mexico State University, Las Cruces, NM 88003, USA}
-\altaffiltext{9}{Department of Astronomy, University of Virginia, Charlottesville, VA 22904-4325, USA}
-\altaffiltext{10}{Astronomy Department, California Institute of Technology, MC 249-17, 1200 East California Blvd, Pasadena, CA 91125, USA}
-\altaffiltext{11}{School of Physics, University of South Wales, Sydney 2052, Australia}
+\altaffiltext{9}{Astronomy Department, California Institute of Technology, MC 249-17, 1200 East California Blvd, Pasadena, CA 91125, USA}
+\altaffiltext{10}{Department of Astronomy, University of Virginia, Charlottesville, VA 22904-4325, USA}
+\altaffiltext{11}{School of Physics, University of New South Wales, Sydney 2052, Australia}
 \altaffiltext{12}{ELTE Gothard Astrophysical Observatory, H-9704 Szombathely, Szent Imre Herceg st. 112, Hungary}

### 2016-08-24 code/kmeans.py
rasterize

@@ -51,7 +51,7 @@ def get_metadata(fn, labels, mask):
     print("get_metadata()", dfn, metadata.shape)
     return metadata
 
-def plot_one_cluster(data, labels, mask, name, dir, suffix="png"):
+def plot_one_cluster(data, labels, mask, name, dir, suffix):
     print("plot_one_cluster(): plotting", name)
     for ly, lx, plotname in [
         ("MG_H - FE_H", "FE_H", "MgFe"),
@@ -84,9 +84,9 @@ def plot_one_cluster(data, labels, mask, name, dir, suffix="png"):
             plt.clf()
             kwargs = {"ls": "none", "marker": "."}
             if mask is None:
-                plt.plot(data[:,x], data[:,y], c="k",    mec="none", ms=0.5, alpha=0.20, **kwargs)
+                plt.plot(data[:,x], data[:,y], c="k",    mec="none", ms=0.5, alpha=0.20, **kwargs, rasterized=True)
             else:
-                plt.plot(data[:,x], data[:,y], c="0.75", mec="none", ms=0.5, alpha=0.20, **kwargs)
+                plt.plot(data[:,x], data[:,y], c="0.75", mec="none", ms=0.5, alpha=0.20, **kwargs, rasterized=True)
                 cc = plt.get_cmap('viridis')
                 M = len(data[mask])
                 angles = (120. / M) * (data[mask, logg].argsort().argsort())
@@ -118,7 +118,7 @@ def _clusterplot(xs, ys, k, ms=5.0):
     if k is not None:
         plt.plot(xs[k], thisys[k], marker="o", c="k", mfc="none", ms=1.5*ms, mew=2., alpha=1., ls="none")
 
-def plot_cluster_context(sizes, densities, dir, name=None, k=None, suffix="png"):
+def plot_cluster_context(sizes, densities, dir, suffix, name=None, k=None):
     """
     so many conditionals!
     """
@@ -157,7 +157,7 @@ def plot_cluster_context(sizes, densities, dir, name=None, k=None, suffix="png")
 if __name__ == "__main__":
     dfn = "./data/results-unregularized-matched.fits.gz"
     dir = "./kmeans_figs"
-    suffix = "png"
+    suffix = "pdf"
     data = None
     plotdata = None
     Ks = 2 ** np.arange(8, 12) # do everything
@@ -319,7 +319,7 @@ if __name__ == "__main__":
[…]

### 2016-08-24 code/kmeans.py
update to pdf figures

@@ -84,9 +84,9 @@ def plot_one_cluster(data, labels, mask, name, dir, suffix):
             plt.clf()
             kwargs = {"ls": "none", "marker": "."}
             if mask is None:
-                plt.plot(data[:,x], data[:,y], c="k",    mec="none", ms=0.5, alpha=0.20, **kwargs, rasterized=True)
+                plt.plot(data[:,x], data[:,y], c="k",   mec="none", ms=0.5, alpha=0.50, **kwargs, rasterized=True)
             else:
-                plt.plot(data[:,x], data[:,y], c="0.75", mec="none", ms=0.5, alpha=0.20, **kwargs, rasterized=True)
+                plt.plot(data[:,x], data[:,y], c="0.5", mec="none", ms=0.5, alpha=0.50, **kwargs, rasterized=True)
                 cc = plt.get_cmap('viridis')
                 M = len(data[mask])
                 angles = (120. / M) * (data[mask, logg].argsort().argsort())
@@ -163,8 +163,8 @@ if __name__ == "__main__":
     Ks = 2 ** np.arange(8, 12) # do everything
     Ks = [256, ] # just do the winner
     # Ks = [2048, ] # just do the craziest
-    plot_everything = False # set to true only for MKN atlas
-    only_high_Z = True # set to true to only plot things in the metallicity bulk
+    plot_everything = True # set to true only for MKN atlas
+    only_high_Z = False # set to true to only plot things in the metallicity bulk
 
     scale_dict = {"FE_H": 0.0191707168068, # all from AC
                   "AL_H": 0.0549037045265,
@@ -351,7 +351,7 @@ if __name__ == "__main__":
         plot_one_cluster(plotdata, plotdata_labels, NGC6819_indices, "NGC_6819", dir, suffix)
 
         # plot clusters from this K
-        plotcount = 0
+        plotcount = 0 # fix!
         for k in (np.argsort(densities))[::-1]:
             clustername = "cluster_{:04d}_{:04d}".format(K, k)
             if (not only_high_Z) or (fehs[k] > -0.5 and fehs[k] < 0.3):

### 2016-08-24 papers/kmeans/kmeans.tex
update to pdf figures
back off slightly on density wording

@@
 \newcommand{\insanefigure}[1]{\rule{0ex}{\headsep}\\% This shouldn't be necessary, but something is fucked with AASTEX 6.0
-\includegraphics[width=\figwidth]{./figs/#1_MgFe.png}%
-\includegraphics[width=\figwidth]{./figs/#1_MgK.png}%
-\includegraphics[width=\figwidth]{./figs/#1_AlMg.png}\\
-\includegraphics[width=\figwidth]{./figs/#1_NaO.png}%
-\includegraphics[width=\figwidth]{./figs/#1_CN.png}%
-\includegraphics[width=\figwidth]{./figs/#1_SAl.png}\\
-\includegraphics[width=2\figwidth]{./figs/#1_gal.png}%
-\includegraphics[width=\figwidth]{./figs/#1_HR.png}\\
-\includegraphics[width=2\figwidth]{./figs/#1_vlon.png}%
-\includegraphics[width=\figwidth]{./figs/#1_context.png}}
+\includegraphics[width=\figwidth]{./figs/#1_MgFe.pdf}%
+\includegraphics[width=\figwidth]{./figs/#1_MgK.pdf}%
+\includegraphics[width=\figwidth]{./figs/#1_AlMg.pdf}\\
+\includegraphics[width=\figwidth]{./figs/#1_NaO.pdf}%
+\includegraphics[width=\figwidth]{./figs/#1_CN.pdf}%
+\includegraphics[width=\figwidth]{./figs/#1_SAl.pdf}\\
+\includegraphics[width=2\figwidth]{./figs/#1_gal.pdf}%
+\includegraphics[width=\figwidth]{./figs/#1_HR.pdf}\\
+\includegraphics[width=2\figwidth]{./figs/#1_vlon.pdf}%
+\includegraphics[width=\figwidth]{./figs/#1_context.pdf}}
 
@@
 and density for the $K=256$ run of k-means.
-The densest clusters tend to have fewer members, presumably because
-the bulk of the population is not extremely strongly clustered, or not
-in ways to which k-means is sensitive.
 
@@
 In what follows, we only consider high-density clusters---clusters
-that are much more dense than average; these ought to represent true
+that are more dense than average for their occupation number (total membership); these ought to represent true
 over-densities in abundance space.
@@
 Although we have by no means any model of the 15-d space, we do know
-that the structures shown in the \figurename s are very high in 15-d
-density.
+that the structures shown in the \figurename s are high in 15-d
+density, or at least relative to other clusters with the same occupation number.
 
@@
 \begin{figure}[!bp]
-\includegraphics[width=1.5\figwidth]{./figs/clusters_0256.png}
+\includegraphics[width=1.5\figwidth]{./figs/clusters_0256.pdf}
 \caption{The distribution of membership and density for the 256

### 2016-08-25 papers/kmeans/kmeans.tex
pull back on density

@@
 the determinant of the tensor.
-This density was used to rank abundance-space overdensities (from highest
-to lowest) for visual inspection.
+This density was used to rank abundance-space overdensities
+for visual inspection.
 In \figurename~\ref{fig:densities} we show the distribution of cluster membership
 and density for the $K=256$ run of k-means.
+There is a bulk trend of larger clusters (clusters with more members or higher occupation number)
+being more dense, but the most interesting
+k-means densities are those that are more dense than this trend.
 
@@
   abundance-space clusters returned by the k-means algorithm at
-  $K=256$.  The densest cluster (which is displayed in more detail in 
-  \figurename~\ref{fig:M13}) is far more dense than the second
-  densest.\label{fig:densities}}
+  $K=256$.  There is a bulk trend and then clusters that are much more
+  dense than the trend. The densest cluster is displayed in more detail in 
+  \figurename~\ref{fig:M13}.\label{fig:densities}}
 \end{figure}
@@
 \thecannon's regression.
-However, it is a very dense feature in abundance space, and is worth
+However, it is a dense feature in abundance space, and is worth
 following up.
